### 2022-07-18 analysis/paper/paper.Rmd
Use Rmd and trackdown for paper

---
title: "Assessment of forestry road network projection methods using the “roads” package in R"
author:
  - Josie Hughes:
      email: josie.hughes@ec.gc.ca
      institute: [ECCC]
      correspondence: true
  - Sarah Endicott:
      email: sarah.endicott@ec.gc.ca
      institute: [ECCC]
      correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]
Global biodiversity is being lost at an unprecedented rate as a result of human activities. One of the major world-wide threats to biodiversity is increasing infrastructural development (Benítez-López et al., 2010). Land transformation has resulted in the loss and fragmentation of habitats in many different ecosystem types worldwide (Benítez-López et al., 2010; Sanderson et al., 2002). Roads and other linear infrastructure, in particular, are key components of landscapes as they have become a major cause of habitat loss, fragmentation, and degradation and are ubiquitous in most landscapes around the world (van der Ree et al., 2011). Although linear infrastructure is important for society because it provides connectivity for people, it also exerts significant negative effects on adjacent habitats, wildlife populations, communities, and ecosystems (van der Ree et al., 2011).

# Methods
## Study Area
The Revelstoke timber supply area (TSA 27), in British Columbia, Canada, was selected as the main study site for this paper. With timber harvesting being the dominant resource extraction activity in this region, much of Revelstoke TSA’s total road network is composed of forestry resource roads which were built solely to gain access to forest harvest cutblocks; this makes for a suitable location to test our forest road projections which build networks using the same goal of connecting cutblocks. With approximately 55% of Revelstoke TSA’s area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The projections were tested in various TSAs across the province with similar patterns being found in comparable mountainous landscapes where forestry is the dominant resource extraction activity (such as Arrow and Cranbrook TSAs). Revelstoke TSA contains wildlife species that are at risk due to their declining populations, such as members of the southern mountain woodland caribou herds (Brown & Hachey, 2010), and therefore is an example of a landscape where these projections could be used to manage critical wildlife habitats. The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).

# Results

<!-- Here's some example analysis code: -->

```{r get-data}
##### Metric table bar plots
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)

data_path_drvd <- "analysis/data/derived_data/"

fig_path <- "analysis/figures/"

fig_dev <- "jpeg"

pal_nm <- "Egypt"

fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)

# Figure 3 #====================================================================

# data (from "meanTable") in the runProjections script
meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))

```

```{r fig3, fig.height=7, fig.width=fig_widths["two"], fig.cap="Revelstoke TSA mean road metrics using different projection methods. Observed road network metric values are represented by the horizontal dotted lines for each metric. The “overall” column represents mean metric values from across the entire area of the TSA,  whereas the right “cutover” column represents mean values from within cutblock areas only."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleDens = factor(sampleDens,
                             levels = c("centroid",
                                        "low sample density",
                                        "high sample density",
                                        "klementQGIS"),
                             labels = c("Centroid", "Low\ndensity", "High\ndensity",
                                        "Klement\nQGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road density", "Road presence",
                                    "Distance to road",
                                    "Forestry disturbance footprint",
                                    "Road disturbance footprint")))

observed_values <- filter(meanTable_long, sampleType == "observed")
projected_values <- filter(meanTable_long, !is.na(sampleDens))
# Create bar plot
fig3 <- projected_values %>% group_by(areaMean) %>% nest() %>% ungroup() %>%
  mutate(obs = observed_values %>% group_by(areaMean) %>% nest() %>% pull(data)) %>%
  group_by(areaMean) %>%
  mutate(data = map(data, ~mutate(.x, area = areaMean)),
         plot = map2(data, obs,
                    ~ ggplot(.x, aes(x = sampleDens, response, fill = sampleType))+
                      geom_col(position = position_dodge2(preserve = "single",
                                                         width = 0.75))+
                      geom_hline(aes(yintercept = response),
                                 .y, color = "black", linetype = "dashed")+
                      facet_wrap(~metric, scales = "free_y", ncol = 1)+
                      theme_classic()+
                      scale_fill_manual(values = met.brewer(pal_nm, 4))+
                      theme(text = element_text(size = 15), axis.title.x = element_blank(),
                            axis.title.y = element_blank(),
                            legend.position = "none")+
                      ggtitle(str_to_title(areaMean)))) %>%
  pull(plot) %>%
  {ggarrange(plotlist = .)}

fig3
```

Figure \@ref(fig:fig3) shows how we can have a caption and cross-reference for a plot

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2022-07-18 analysis/paper/paper.Rmd
bring in all of David's text

@@
 
-# Introduction
+# Introduction[a]
 
-Here is a citation [@Marwick2017]
 Global biodiversity is being lost at an unprecedented rate as a result of human activities.
@@
 Although linear infrastructure is important for society because it provides connectivity for people, it also exerts significant negative effects on adjacent habitats, wildlife populations, communities, and ecosystems (van der Ree et al., 2011).
+Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada’s forested area.
+In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province’s total road network (83%)  (Environmental Reporting BC, 2018; Forest Practices Board, 2015).
+Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015).
+Resource extraction activities provide jobs and renewable raw materials (Canada, 2013b), but they also can create threats for wildlife.
+For example, the main threats to woodland caribou, a species at risk in Canada, are human disturbances such as forest harvesting, oil and gas exploration/extraction, and road networks.
+These particular disturbances fragment their habitat, and create open areas and extensive young forests that attract other species such as moose and deer which in turn, attracts an increased number of predators, such as wolves and even poachers (Canada, 2013a).
+Woodland caribou are often considered to be an umbrella species, which means that protecting and restoring their habitat would benefit other species also dependent on old forests, such as fishers, pine martens, and many species of birds (Mulvihill, 2018).
+With certain wildlife, such as woodland caribou, being highly sensitive to road development and use, there is a need to spatially simulate future roads in order to test the implications of alternative recovery policies and resource development scenarios.
+Landscape models are mathematical simulations designed to represent at least one landscape pattern-process relationship of interest.
+Landscape models can be great ways to make predictions about specific scenarios that allow ecologists and land managers to assess alternative management scenarios or to assess the potential impact of anthropogenic change scenarios (climate or land use change) (McGarigal, n.d.).
+Simulating the effects of disturbances, such as forestry road development, helps to broaden the scope of models, especially for species with complex ecological requirements who reside in disturbance-prone landscapes that undergo dynamic changes which may substantially alter habitat quality (Whitman et al., 2017).
+Land managers are often responsible for multiple species and populations distributed over large regions and often rely on models to determine habitat suitability (Whitman et al., 2017).
+Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011).
+Stochastic simulation approaches use probability distributions in combination with random number generators to determine events (He & Mladenoff, 1999).
+Despite the importance of roads on the environment, previous spatial stochastic harvest simulation models rarely include forest road network development in their projections.
+The “roads” package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface.
+This paper uses the “roads” package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.
+
+(Introduction round 2: road context)
+Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use.
+These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network.
+These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes.
+On the other hand, many spatial stochastic forest management simulations haven’t included forest road development at all (citations).
+One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
+There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes.
+As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).
+The goal of our model is to provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations, and can also be applied as an element of more complex models.
+Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes.
+Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network.
+In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially.
+In the first section of the paper (methods), we introduce the study area used to assess the accuracy of the projection, we describe the forest road projection methods, the data required, and the various metrics used in order to analyze the similarities and differences between the projected and observed forest road networks across the landscape.
+Next, we present and discuss the results of our projections across the landscape and compare the metrics spatially and aspatially.
 
@@
 ## Study Area
-The Revelstoke timber supply area (TSA 27), in British Columbia, Canada, was selected as the main study site for this paper.
-With timber harvesting being the dominant resource extraction activity in this region, much of Revelstoke TSA’s total road network is composed of forestry resource roads which were built solely to gain access to forest harvest cutblocks; this makes for a suitable location to test our forest road projections which build networks using the same goal of connecting cutblocks.
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA’s total road network is composed of forestry resource roads built to gain access to forest harvest cutblocks.
+This allows us to project road building based on forest harvest data.
 With approximately 55% of Revelstoke TSA’s area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.
-The projections were tested in various TSAs across the province with similar patterns being found in comparable mountainous landscapes where forestry is the dominant resource extraction activity (such as Arrow and Cranbrook TSAs).
-Revelstoke TSA contains wildlife species that are at risk due to their declining populations, such as members of the southern mountain woodland caribou herds (Brown & Hachey, 2010), and therefore is an example of a landscape where these projections could be used to manage critical wildlife habitats.
+The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%).
+The TSA contains wildlife species that are at risk due to their declining populations and are believed to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010).
+The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats.
 The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).
 
+## Road network projection methods
+Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes.
+Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network.
+We used three methods for solving the MTAP in the roads package.
+These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST).
+In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
+The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
+For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959).
+Dijkstra’s algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section “weighted graph”).
+Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction.
+This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP.
+They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
+The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
+Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
+Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph.
+The minimum spanning tree was solved using Kruskal’s algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop.
+This method was most similar to Clark et al.
+(2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling.
+An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.
+        A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods.
+All the default settings such as skidding distance, etc.
+were left as they were within the QGIS plugin.
+
+## Weighted graph
+A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal’s or Dijkstra’s algorithms.
+When solving least cost paths using Dijkstra’s algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution).
+Whereas, in the graph used for solving the minimum spanning tree via Kruskal’s algorithm, each node represented either a landing or a point on the existing road network.
+Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point.
+The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up.
+The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies.
+The edges connecting the nodes follow a “queens” case or allow a maximum of eight possible edges for a given node.
+Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only.
+Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs.
+When solving least cost paths, a cost surface raster was used.
+The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels.
+When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.
+
+## Data
+To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster.
+We used several spatial raster layers describing the roading cost structure, barriers to road development and road impedance.
+For intermediate nodes, the roading cost structure followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope.
+Separate costs for river and pipeline crossings were then added to the roading cost.
+Barriers for road development masked the cost surface raster and these included the major water bodies of BC.
+Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned the smallest cost (Table 1).
+The cost surface for Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.
+
+Table 1. Definitions of road building costs for building the cost surface. 
+Cost Type
+	Road Attribute
+	Value ($ per km)
+	Source
+	Barriers of road development
+	Major lakes
+	infinity
+	Major water bodies of BC
+	Impedance
+	Existing roads
+	0.01
+	Assumed
+	Road Construction Cost Structure
+
+	Pipe crossing
+	1911
+	Interior Appraisal Manual
+	River crossing
+	2130
+	Interior Appraisal Manual
+	Base cost
+	16178 + 504*slope
+	Interior Appraisal Manual
+
+The road network used was a combination of the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) datasets from the Government of British Columbia’s Data Catalogue.
+The existing road network inputted into the model was created by filtering out the forest tenure roads with an award date greater than 1990.
+Because the objective of the roads package is to project forest resource roads, the DRA roads were left untouched for the existing 1990 road network.
+The projection results would be compared with the observed modern road network which was simply a complete and unfiltered combination of the DRA and Forest Tenure road datasets, representing the most complete and up-to-date road network for the area.
+Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) Government of British Columbia’s Data Catalogue (Government of British Columbia, 2021b).
+Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
+The dataset does not exclude non-harvested ‘reserved’ areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b).
+Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.
+Analysis
+Projections were run in R (R Core Team, 2020) within the Revelstoke TSA.
+The forest road network projections were run using two different resolutions of the cost surface in order to assess how sensitive the results were to pixel size; the finer resolution used cost surface pixels of 100 meters by 100 meters and was the focus of this study, whereas the coarse resolution used an aggregated cost surface with pixels of 1000 meters by 1000 meters.
+The least cost path method yields projections that are obviously less realistic than the minimum spanning tree method (Figure 2), so we focussed on assessing the performance of the minimum spanning tree method and investigated sensitivity to sampling types (regular and random) and sampling densities (0.000001 for low and 0.0001 for high) (Figure 2).
+        In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network.
+Road presence is a simple measure of whether or not roads are present in each pixel using the cost surface as a template.
+Road disturbance footprint was created as done by Ibisch et al.
+(2016) where a 1 km buffer was applied to all roads.
+Forestry disturbance footprint was created by buffering 500 metres around all roads and cutblocks.
+This metric relates to the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and can be calculated using the caribouMetrics R package (Endicott et al.,in prep).
+The 100m x 100m resolution cost surface was used as the template to derive the forestry disturbance footprint metric values.
+Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004).
+Road density is represented in meters per hectare (m/ha) and was calculated using the rasterizeLineDensity function within the caribouMetrics package (Endicott et al., in prep).
+This was done by intersecting the roads with the cost surface raster grid, calculating the length of road in each grid cell and dividing by the grid cell resolution.
+Distance to nearest road assigns each cell within the TSA a distance (meters) to the nearest road segment.
+The metric can be calculated using the “dist_to_nearest” function in the “starsExtra” package (Dorman, 2021), but this implementation can be prohibitively slow.
+We have implemented a faster moving window method in the XX function of the roads package.
+        Using the metrics described above, we assessed the aspatial accuracy of the projections by calculating the mean values of the projection metrics rasters and then compared them with the means of the observed metric rasters.
+This was done at an overall TSA level (where means were taken from across the entire study area) and at a cutblock level (where metric rasters were masked by the cutblock polygons and the means were taken only from within these areas).
+        The projections were assessed for their spatial accuracy in two different ways.
+First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to further highlight discrepancies between the networks.
+The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in GIS (QGIS) to gain a better understanding of what was happening on the ground.
+Secondly, a roaded/roadless matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed.
+This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA which matched and didn’t match.
+All pixels in the Revelstoke TSA for the projected networks         fell into 5 categories when compared with the observed raster: Roaded prior to 1990 is the roaded area of the existing 1990 road network that was inputted into the models to project future forestry roads; agree roadless is the roadless area where pixels between projected and observed match; agree roaded after 1990 is the roaded areas where projected forestry roads match with observed forestry roads; project roaded & observe roadless is areas where the projection was roaded but where the observed network was roadless; and finally project roadless & observe roaded represents areas where the projection was roadless but the observed network was roaded.
+Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized (figure 3).
+
 # Results
+## Aspatial accuracy
+Various projection methods provided more accurate aspatial results than others when compared with the observed metrics depending on the metric used.
+The forestry disturbance footprint metric produced means across all projections that were closest with observed values (Figure 3).
+Projections of road disturbance footprint and distance to road were also reasonably accurate across the whole landscape, with distance to road being less accurate within cutblocks.
+Road presence and road density were less accurate and much more sensitive to assumptions about the distribution of roads within cutblocks (i.e.
+sample density).
+The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm.
+Road presence and road density were also the most sensitive to pixel size (reference appendix table for coarse resolution projections).
+The projection using the Clement’s (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.
+
+## Spatial accuracy
+        Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks.
+In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a).
+Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data.
+There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b).
+There are also instances where areas were likely misidentified as cutblocks (Figure 4c).
+This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
+Figure 5 shows the percentage of matching pixels by category between observed and projected rasters.
+The percentage of roaded prior to 1990 was the same across all projection methods for the respective metrics because the same 1990 road network was used for all projection as the input for the existing road network.
+The agree roadless category represented most of the landscape for all projection methods.
+The subsequent 3 categories (Agree roaded after 1990, project roaded & observe roadless, project roadless & observe roaded) represent the areas where change occurred following 1990.
+With the road disturbance footprint metric, the agree roaded after 1990 represented the largest percentage of the landscape which underwent change after 1990.
+The project roaded & observe roadless category represented higher percentages of the landscape than the project roadless & observe roaded, especially when using a higher sampling density, which shows that there were more areas where the model over projected roads than under projected roads.
+The small percentage of project roadless & observe roaded across all projection methods shows that there were very few observed roaded areas that the projections did not address.
+Using the forestry disturbance footprint metric, the post-1990 roaded pixels lined up very well.
+Nearly 70% of all post-1990 projected roaded pixels matched with the observed network when using this metric.
+Project roadless & observe roaded very slightly made up a higher percentage of the landscape when compared to project roaded & observe roadless (by less than half a percent across all projection methods).
+When using the fine resolution road presence raster to compare with the observed road presence raster, very few of the post-1990 pixels align.
+Nearly all post-1990 roaded pixels did not match between observed and projected.
+The sampling density used determined whether there were more project roaded & observe roadless pixels (higher sampling density) or project roadless & observe roaded pixels (centroid and lower sampling density) amongst the post-1990 roaded pixels.
 
@@
 # Discussion
+Through assessing the performance of our model by comparing its spatial and aspatial
+accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters.
+However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model.
+The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc.
+Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.
+        Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data; these discrepancies should be treated with caution as they are not necessarily the fault of the projections themselves and rather with issues in available data.
+These potential issues were identified through the use of the difference maps paired with recent satellite imagery that allowed us to better understand what was happening on the ground.
+Often major discrepancies were related to cutblocks that were likely falsely labelled (most often through LANDSAT change detection) in places where there weren’t actually cutblocks.
+These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections.
+In order to get more accurate road projections, one could simply remove the cutblocks derived from LANDSAT as these were often the cutblocks which caused issues, however, these cutblocks were left in as a major goal of this assessment was to create road projections while adjusting input parameters as little as possible.
+Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue.
+Another potential issue was with roads not being included in the data set that were seen in the satellite imagery.
+This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets.
+Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford’s doc).
+This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery.
+These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn’t have and should be treated with caution.
+It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets.
+Again, this would not present an issue when predicting future impacts on landscapes, only when assessing the model accuracy in this context.
+Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.
+Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines).
+Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc.
+Because of this, forestry roads make up a much smaller percentage of the overall road network.
+In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors.
+The model was attempted in the Fort Nelson TSA however the resulting road network (Figure A-1) did not accurately portray the observed network like it did in the Revelstoke TSA.
+However, other resource extraction activities could likely be incorporated into this model relatively easily.
+The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas.
+ It is also worth noting that although the model can project road networks fairly accurately across a large landscape, it may differ from the way in which roads are planned in real life.
+The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries.
+In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).
+Certain projection methods were more accurate at projected some metrics compared to others.
+Similarly, certain metrics were sensitive to pixel size (road density, road presence) and others were not.
+Depending on what the goal of running the projections is, then the projection method and resolution may or may not play an important role.
+If you would like to predict road density as accurately as possible, then running the projections using as fine a resolution of cost surface as possible is the best option.
+If you’re wanting to only predict roughly the location of the roads/save computing time and are more interested in a larger area, then a coarse resolution may suit these requirements perfectly fine.
+The tradeoffs of using a coarser resolution of cost surface may be less extreme when projecting roads across a flatter landscape compared to a mountainous landscape like was used for this paper since there would likely be less variability in cost across larger areas.
+     With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which “roads” can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios.
+Because caribou are also considered an umbrella species within forests (Drever et al., 2019), projecting forestry impacts on their habitats may also provide valuable insight towards other key species as well.
+With the input of future cutblocks and the projections of forest road networks to these cutblocks, the results from the projection can provide a great outlook at the impacts of forestry scenarios on woodland caribou habitats using the forestry disturbance footprint metric.
+The distance to nearest road metric appears to be a very interesting metric to consider as well when looking specifically at larger areas.
+Using this metric to make projections at larger scales, such as at a TSA level, may provide very beneficial information about forestry impacts especially in a wildlife context when assessing the influence of forestry on various species with different tolerances to disturbance.
+Further improvements could be made that would increase the accuracy of the model.
+The cost surface used for this paper was developed at a provincial scale, however, a higher resolution cost surface with more detail could further improve the accuracy of the projection results.
+A finer resolution cost surface with more detail will ultimately create a more realistic road network and would possibly allow for more accuracy within cutblocks (where fine micro-topography, water tables, etc.
+will play a role in road development).
+The downside being that it would also increase computing time and may not always be necessary depending on the application of the model, especially when using the model for a large-scale landscape.
+Adding too much complexity to the cost surface may also take away from the goal of this model to be simple and easy to use, however, the algorithm itself would work on a very fine and detailed cost surface.
+Along with increased computing time, a key limitation of using this model on a finer cost surface would be that eventually cost pixels would need to have directionality associated with them, which is something this model does not take into account.
+The accuracy of comparing the projections to the observed network could also simply be improved by using higher quality/more up to date road and cutblock data.
+Assessing the performance of this model also presented some unique challenges with comparing two networks with each other across a landscape.
+We needed to carefully think about what we were trying to achieve.
+The goal wasn’t to present a model that matched the observed road network pixel by pixel, but instead was to simulate disturbance across a large-scale landscape to an acceptable level while still being a relatively simple and straightforward tool.
+It’s for this reason why metrics were chosen that are commonly used in road impact assessments on wildlife; such as the road density and forestry disturbance footprint metrics.
+Once the vector networks were converted into these raster layers, it also facilitated comparing observed versus projected as we could compare the means across the TSA and study where pixels matched and didn’t match.
+Incorporating these metrics into landscape level simulations is also a realistic goal for this tool (ie projecting forest road density across a landscape) so a goal was to therefore see how well the model did at simulating these metrics when compared to the observed landscape.
+Many spatial stochastic models don’t include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010).
+Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these.
+For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.
 
 # Conclusion
+The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort.
+This model, however, is not suitable for strategic planning.
+The results from our projections in our study area overall matched up adequately with the observed present day forest road networks, with major discrepancies being attributed to issues in the available road and cutblock data.
+These discrepancies would not be issues in future landscape simulations, only when assessing accuracy.
+In areas where the quality of the data was high, projections matched well with the observed network.
+Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations.
+        This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape.
+With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale.
+With land use activities transforming a large proportion of the planet’s land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats.
+Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of ~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005).
+With only 7% of the planet’s roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016).
+The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.
 
@@
 ```
+
+[a]A comment

### 2022-07-18 analysis/paper/paper.Rmd
formatting

@@
 Although linear infrastructure is important for society because it provides connectivity for people, it also exerts significant negative effects on adjacent habitats, wildlife populations, communities, and ecosystems (van der Ree et al., 2011).
+
 Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada’s forested area.
@@
 Woodland caribou are often considered to be an umbrella species, which means that protecting and restoring their habitat would benefit other species also dependent on old forests, such as fishers, pine martens, and many species of birds (Mulvihill, 2018).
+
 With certain wildlife, such as woodland caribou, being highly sensitive to road development and use, there is a need to spatially simulate future roads in order to test the implications of alternative recovery policies and resource development scenarios.
@@
 Land managers are often responsible for multiple species and populations distributed over large regions and often rely on models to determine habitat suitability (Whitman et al., 2017).
+
 Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011).
@@
 As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).
+
 The goal of our model is to provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations, and can also be applied as an element of more complex models.
@@
 In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially.
+
 In the first section of the paper (methods), we introduce the study area used to assess the accuracy of the projection, we describe the forest road projection methods, the data required, and the various metrics used in order to analyze the similarities and differences between the projected and observed forest road networks across the landscape.
@@
 In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
-The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
+
+LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
 For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959).
@@
 They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
+
 The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
@@
 An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.
+
 A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods.
@@
 Table 1. Definitions of road building costs for building the cost surface. 
-Cost Type
-	Road Attribute
-	Value ($ per km)
-	Source
-	Barriers of road development
-	Major lakes
-	infinity
-	Major water bodies of BC
-	Impedance
-	Existing roads
-	0.01
-	Assumed
-	Road Construction Cost Structure
-
-	Pipe crossing
-	1911
-	Interior Appraisal Manual
-	River crossing
-	2130
-	Interior Appraisal Manual
-	Base cost
-	16178 + 504*slope
-	Interior Appraisal Manual
+```{r}
+tbl <- "Cost Type|Road Attribute|Value ($ per km)|Source
+Barriers of road development|Major lakes|infinity|Major water bodies of BC
+Impedance|Existing roads|0.01|Assumed
+Road Construction Cost Structure|Pipe crossing|1911|Interior Appraisal Manual
+|River crossing|2130|Interior Appraisal Manual
+|Base cost|16178 + 504*slope|Interior Appraisal Manual
+"
+df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df) <- unname(as.list(df[1,]))
+df <- df[-1,] # remove first row
+row.names(df)<-NULL
+knitr::kable(df)
+```
 
@@
 Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.
-Analysis
+
+## Analysis
+
 Projections were run in R (R Core Team, 2020) within the Revelstoke TSA.
@@
 The least cost path method yields projections that are obviously less realistic than the minimum spanning tree method (Figure 2), so we focussed on assessing the performance of the minimum spanning tree method and investigated sensitivity to sampling types (regular and random) and sampling densities (0.000001 for low and 0.0001 for high) (Figure 2).
+
 In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network.
@@
 We have implemented a faster moving window method in the XX function of the roads package.
+
 Using the metrics described above, we assessed the aspatial accuracy of the projections by calculating the mean values of the projection metrics rasters and then compared them with the means of the observed metric rasters.
 This was done at an overall TSA level (where means were taken from across the entire study area) and at a cutblock level (where metric rasters were masked by the cutblock polygons and the means were taken only from within these areas).
+
 The projections were assessed for their spatial accuracy in two different ways.
@@
 This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
+
 Figure 5 shows the percentage of matching pixels by category between observed and projected rasters.
@@
 The subsequent 3 categories (Agree roaded after 1990, project roaded & observe roadless, project roadless & observe roaded) represent the areas where change occurred following 1990.
+
 With the road disturbance footprint metric, the agree roaded after 1990 represented the largest percentage of the landscape which underwent change after 1990.
@@
 The small percentage of project roadless & observe roaded across all projection methods shows that there were very few observed roaded areas that the projections did not address.
+
 Using the forestry disturbance footprint metric, the post-1990 roaded pixels lined up very well.
@@
 Project roadless & observe roaded very slightly made up a higher percentage of the landscape when compared to project roaded & observe roadless (by less than half a percent across all projection methods).
+
 When using the fine resolution road presence raster to compare with the observed road presence raster, very few of the post-1990 pixels align.
@@
 The sampling density used determined whether there were more project roaded & observe roadless pixels (higher sampling density) or project roadless & observe roaded pixels (centroid and lower sampling density) amongst the post-1990 roaded pixels.
-
-<!-- Here's some example analysis code: -->
 
@@
 Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.
+
 Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data; these discrepancies should be treated with caution as they are not necessarily the fault of the projections themselves and rather with issues in available data.
@@
 Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.
+
 Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines).
@@
 The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas.
+
 It is also worth noting that although the model can project road networks fairly accurately across a large landscape, it may differ from the way in which roads are planned in real life.
@@
 In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).
+
 Certain projection methods were more accurate at projected some metrics compared to others.
@@
 The tradeoffs of using a coarser resolution of cost surface may be less extreme when projecting roads across a flatter landscape compared to a mountainous landscape like was used for this paper since there would likely be less variability in cost across larger areas.
+
 With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which “roads” can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios.
@@
 Using this metric to make projections at larger scales, such as at a TSA level, may provide very beneficial information about forestry impacts especially in a wildlife context when assessing the influence of forestry on various species with different tolerances to disturbance.
+
 Further improvements could be made that would increase the accuracy of the model.
@@
 The accuracy of comparing the projections to the observed network could also simply be improved by using higher quality/more up to date road and cutblock data.
+
 Assessing the performance of this model also presented some unique challenges with comparing two networks with each other across a landscape.
@@
 Incorporating these metrics into landscape level simulations is also a realistic goal for this tool (ie projecting forest road density across a landscape) so a goal was to therefore see how well the model did at simulating these metrics when compared to the observed landscape.
+
 Many spatial stochastic models don’t include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010).
@@
 Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations.
+
 This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape.

### 2022-07-19 analysis/paper/paper.Rmd
working on paper text and figures. Added new map

@@
   fig.path = "../figures/",
-  dpi = 300
+  dpi = 300,
+  cache = TRUE
 )
@@
 
-# Introduction[a]
+# Introduction
 
@@
 
-Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada’s forested area.
-In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province’s total road network (83%)  (Environmental Reporting BC, 2018; Forest Practices Board, 2015).
+Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area.
+In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015).
 Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015).
@@
 Despite the importance of roads on the environment, previous spatial stochastic harvest simulation models rarely include forest road network development in their projections.
-The “roads” package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface.
-This paper uses the “roads” package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.
+The "roads" package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface.
+This paper uses the "roads" package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.
 
-(Introduction round 2: road context)
-Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use.
+(Introduction round 2: road context) Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use.
 These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network.
 These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes.
-On the other hand, many spatial stochastic forest management simulations haven’t included forest road development at all (citations).
+On the other hand, many spatial stochastic forest management simulations haven't included forest road development at all (citations).
 One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
 There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes.
-As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).
+As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) ("Bridging the divide between ecological forecasts and environmental decision-making").
 
@@
 # Methods
+
 ## Study Area
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA’s total road network is composed of forestry resource roads built to gain access to forest harvest cutblocks.
+
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1).
 This allows us to project road building based on forest harvest data.
-With approximately 55% of Revelstoke TSA’s area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.
-The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%).
-The TSA contains wildlife species that are at risk due to their declining populations and are believed to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010).
+With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.
+The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) (Ministry of Forests, n.d.-b).
+The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010).
 The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats.
@@
 
-## Road network projection methods
-Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes.
-Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network.
-We used three methods for solving the MTAP in the roads package.
-These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST).
-In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
+```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel."}
+##### Metric table bar plots
+library(ggplot2)
+library(dplyr)
+library(stringr)
+library(tidyr)
+library(purrr)
+library(ggpubr)
+library(MetBrewer)
+library(here)
+library(tmap)
+library(sf)
 
-LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
-For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959).
-Dijkstra’s algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section “weighted graph”).
-Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction.
-This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP.
-They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
+#set paths
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
 
-The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
-Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
-Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph.
-The minimum spanning tree was solved using Kruskal’s algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop.
-This method was most similar to Clark et al.
-(2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling.
-An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.
+pal_nm <- "Egypt"
 
-A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods.
-All the default settings such as skidding distance, etc.
-were left as they were within the QGIS plugin.
+fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
 
-## Weighted graph
-A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal’s or Dijkstra’s algorithms.
-When solving least cost paths using Dijkstra’s algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution).
-Whereas, in the graph used for solving the minimum spanning tree via Kruskal’s algorithm, each node represented either a landing or a point on the existing road network.
-Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point.
-The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up.
-The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies.
-The edges connecting the nodes follow a “queens” case or allow a maximum of eight possible edges for a given node.
-Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only.
-Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs.
-When solving least cost paths, a cost surface raster was used.
-The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels.
-When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.
+tsaBoundary <- read_sf(here(data_path_raw, "new_tsa27_boundaries.shp"))
+#cost surface raster layer
+bc_cost_surface <- terra::rast(here(data_path_raw, "cost_surface_bc_ha.tif"))
+lakeValue <- 65000
+
+tsaCost <- terra::crop(bc_cost_surface, tsaBoundary)
+tsaCost <- terra::subst(tsaCost, from = NA, to = lakeValue)
+tsaCost <- terra::mask(tsaCost, terra::vect(tsaBoundary))
+
+canada <- geodata::gadm("CAN", level = 1, path = here(data_path_raw))
+
+canada <- canada %>% 
+  st_as_sf() %>% 
+  filter(NAME_0 == "Canada") %>% 
+  st_transform(st_crs(tsaBoundary)) %>% 
+  st_simplify(dTolerance = 10000)
+
+canada_overlay <- tm_shape(canada, projection = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")+
+  tm_borders(col="black")+
+  tm_shape(st_buffer(tsaBoundary, 10000))+
+  tm_polygons(border.col = met.brewer(pal_nm, 1), col = met.brewer(pal_nm, 1))+
+  tm_layout(frame=TRUE)
+
+tsa <- tm_shape(tsaCost, name = "Cost")+
+  tm_raster(style = "cont", title = "Cost")+
+  tm_shape(st_union(tsaBoundary))+
+  tm_borders(col = "black")+
+  tm_compass(position = c(0.3, 0.11), size = 1)+
+  tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
+  tm_layout(legend.position = c("left", "bottom"),
+            legend.bg.color = "white")
+
+vp <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
+               just=c("right", "top"))
+
+tmap_save(tsa, filename = here("analysis/figures/Figure1_StudyArea.png"),
+  dpi = 600, insets_tm = canada_overlay, insets_vp = vp,
+  height = 6, width = fig_widths["single"], units = "in")
+
+knitr::include_graphics(here("analysis/figures/Figure1_StudyArea.png"))
+
+```
 
 ## Data
+
 To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster.
-We used several spatial raster layers describing the roading cost structure, barriers to road development and road impedance.
-For intermediate nodes, the roading cost structure followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope.
-Separate costs for river and pipeline crossings were then added to the roading cost.
-Barriers for road development masked the cost surface raster and these included the major water bodies of BC.
-Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned the smallest cost (Table 1).
-The cost surface for Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.
+We used several spatial raster layers to construct our cost surface.
+For all cells the base cost followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope.
+Separate costs for river and pipeline crossings were then added to the cost.
+Major water bodies were considered barriers and assigned a very high cost.
+Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned a cost of 0 (Table 1).
+The cost surface for the Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.
 
 Table 1. Definitions of road building costs for building the cost surface.
+
 ```{r}
 tbl <- "Cost Type|Road Attribute|Value ($ per km)|Source
-Barriers of road development|Major lakes|infinity|Major water bodies of BC
-Impedance|Existing roads|0.01|Assumed
+Barriers of road development|Major lakes|65000|Major water bodies of BC
+Impedance|Existing roads|0|Assumed
 Road Construction Cost Structure|Pipe crossing|1911|Interior Appraisal Manual
@@
 
-The road network used was a combination of the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) datasets from the Government of British Columbia’s Data Catalogue.
-The existing road network inputted into the model was created by filtering out the forest tenure roads with an award date greater than 1990.
-Because the objective of the roads package is to project forest resource roads, the DRA roads were left untouched for the existing 1990 road network.
-The projection results would be compared with the observed modern road network which was simply a complete and unfiltered combination of the DRA and Forest Tenure road datasets, representing the most complete and up-to-date road network for the area.
-Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) Government of British Columbia’s Data Catalogue (Government of British Columbia, 2021b).
+Our road network data was obtained by combining the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) data sets for BC.
+The existing road network, used as the starting point for projections, was created by filtering out the forest tenure roads with an award date greater than 1990.
+Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset (Government of British Columbia, 2021b).
 Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-The dataset does not exclude non-harvested ‘reserved’ areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b).
-Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.
+The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b).
+Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
+
+## Road Projection Methods
+
+We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e.
+harvest cutblocks) and the landscape connecting the two.
+There are several different methods available for solving this problem, including snapping, least cost paths, and minimum spanning trees.
+Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape.
+The least cost paths (LCP) method uses Dijkstra's shortest path algorithm (1959) to determine the optimal path between a cutblock and existing road based on a the cost of building a road at any point between the two.
+The cost is represented by a cost raster described below.
+Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock.
+Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm (1956).
+(See Appendix X for a more detailed explanation of these methods).
+In the remainder of this paper, we focus on the LCP and MST approaches given their assumptions are better aligned with how road networks are designed.
+
+In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>).
+This plugin uses a similar least cost path algorithm to the `roads` package but instead of building roads along the least cost path to each cutblock independently the Forest Road Network Plugin uses a heuristic to determine which cutblocks to connect first and then builds the roads sequentially.
+The default heuristic is to connect the closest cutblocks first.
+We supplied our cost surface, cutblocks and road network data and kept all the default settings within the QGIS plugin.
 
@@
 
-Projections were run in R (R Core Team, 2020) within the Revelstoke TSA.
-The forest road network projections were run using two different resolutions of the cost surface in order to assess how sensitive the results were to pixel size; the finer resolution used cost surface pixels of 100 meters by 100 meters and was the focus of this study, whereas the coarse resolution used an aggregated cost surface with pixels of 1000 meters by 1000 meters.
-The least cost path method yields projections that are obviously less realistic than the minimum spanning tree method (Figure 2), so we focussed on assessing the performance of the minimum spanning tree method and investigated sensitivity to sampling types (regular and random) and sampling densities (0.000001 for low and 0.0001 for high) (Figure 2).
+We projected road development in the Revelstoke TSA from 1990 to the present using several different inputs and compared the projections to the observed present day road network in order to evaluate the accuracy of the projections and their sensitivity to different inputs.
+We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution.
+We also compared methods for locating landing points with in cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 100 pts/km^2^ (Figure \@ref(fig:fig2)).
 
-In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network.
-Road presence is a simple measure of whether or not roads are present in each pixel using the cost surface as a template.
-Road disturbance footprint was created as done by Ibisch et al.
-(2016) where a 1 km buffer was applied to all roads.
-Forestry disturbance footprint was created by buffering 500 metres around all roads and cutblocks.
-This metric relates to the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and can be calculated using the caribouMetrics R package (Endicott et al.,in prep).
-The 100m x 100m resolution cost surface was used as the template to derive the forestry disturbance footprint metric values.
+```{r fig2, fig.height=5, fig.width=fig_widths["two"], fig.cap="Example of the different projection methods and sampling strategies."}
+knitr::include_graphics(here("analysis/figures/projection methods figure.jpg"))
+```
+
+In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network: 1) Road presence, a simple measure of whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer applied to observed and projected roads (Ibisch et al., 2016); 3) Forestry disturbance footprint, a 500 m buffer applied to observed and projected roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road.
+The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and was calculated using the `caribouMetrics` R package (Endicott et al.,in prep).
 Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004).
-Road density is represented in meters per hectare (m/ha) and was calculated using the rasterizeLineDensity function within the caribouMetrics package (Endicott et al., in prep).
-This was done by intersecting the roads with the cost surface raster grid, calculating the length of road in each grid cell and dividing by the grid cell resolution.
-Distance to nearest road assigns each cell within the TSA a distance (meters) to the nearest road segment.
-The metric can be calculated using the “dist_to_nearest” function in the “starsExtra” package (Dorman, 2021), but this implementation can be prohibitively slow.
-We have implemented a faster moving window method in the XX function of the roads package.
+Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package (Endicott et al., in prep).
+Distance to nearest road was calculated using a fast moving window method in the `getDistFromSource` function of the roads package.
+Each metric was applied to the results from each projection method and the observed roads with the results were stored as rasters.
 
-Using the metrics described above, we assessed the aspatial accuracy of the projections by calculating the mean values of the projection metrics rasters and then compared them with the means of the observed metric rasters.
-This was done at an overall TSA level (where means were taken from across the entire study area) and at a cutblock level (where metric rasters were masked by the cutblock polygons and the means were taken only from within these areas).
+The aspatial accuracy of the projections was assessed by calculating the mean values of the projection metric rasters and then comparing them with the means of the observed metric rasters.
+This was done across the TSA and at the cutblock level, where metric rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.
 
 The projections were assessed for their spatial accuracy in two different ways.
-First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to further highlight discrepancies between the networks.
-The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in GIS (QGIS) to gain a better understanding of what was happening on the ground.
-Secondly, a roaded/roadless matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed.
-This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA which matched and didn’t match.
-All pixels in the Revelstoke TSA for the projected networks fell into 5 categories when compared with the observed raster: Roaded prior to 1990 is the roaded area of the existing 1990 road network that was inputted into the models to project future forestry roads; agree roadless is the roadless area where pixels between projected and observed match; agree roaded after 1990 is the roaded areas where projected forestry roads match with observed forestry roads; project roaded & observe roadless is areas where the projection was roaded but where the observed network was roadless; and finally project roadless & observe roaded represents areas where the projection was roadless but the observed network was roaded.
-Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized (figure 3).
+First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to highlight discrepancies between the networks.
+The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS to investigate potential causes of the discrepancy.
+Secondly, a matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed.
+This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA that matched and didn't match.
+All pixels in the Revelstoke TSA for the projected networks fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed.
+Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized.
 
 # Results
+
 ## Aspatial accuracy
+
 Various projection methods provided more accurate aspatial results than others when compared with the observed metrics depending on the metric used.
@@
 Road presence and road density were also the most sensitive to pixel size (reference appendix table for coarse resolution projections).
-The projection using the Clement’s (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.
+The projection using the Clement's (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.
 
 ## Spatial accuracy
+
 Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks.
@@
 ```{r get-data}
-##### Metric table bar plots
-library(ggplot2)
-library(dplyr)
-library(stringr)
-library(tidyr)
-library(purrr)
-library(ggpubr)
-library(MetBrewer)
-library(here)
+# load csv
+meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
 
-data_path_drvd <- "analysis/data/derived_data/"
-
-fig_path <- "analysis/figures/"
-
-fig_dev <- "jpeg"
-
-pal_nm <- "Egypt"
-
-fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
-
-# Figure 3 #====================================================================
-
-# data (from "meanTable") in the runProjections script
-meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
 
@@
                              labels = c("Centroid", "Low\ndensity", "High\ndensity",
-                                        "Klement\nQGIS")),
+                                        "Hardy\nQGIS")),
          metric = factor(metric,
@@
 
-Figure \@ref(fig:fig3) shows how we can have a caption and cross-reference for a plot
+```{r fig4, fig.height=5, fig.width=fig_widths["two"], fig.cap="Spatial accuracy"}
+
+stackedBar <- matchData %>% ungroup() %>%
+  mutate(agreement = factor(agreement,
+                            levels = c("False negative", "False positive",
+                                       "Agree roaded","Agree roadless",
+                                       "Pre-existing roads")),
+         sampleDens = paste(sampleType, sampleDens) %>%
+           str_replace("1e-06", "low density") %>%
+           str_replace("1e-04", "high density") %>%
+           str_replace("centroid high density", "centroid") %>%
+           str_replace("klementQGIS NA", "klementQGIS") %>%
+           factor(levels = c("centroid",
+                             "random low density",
+                             "random high density",
+                             "regular low density",
+                             "regular high density",
+                             "klementQGIS"),
+                  labels = c("Centroid", "Random low density", "Random high density",
+                             "Regular low density",
+                             "Regular high density",
+                             "Hardy QGIS")),
+         metric = factor(metric,
+                         levels = c("roadPresence",
+                                    "forestryDisturbance",
+                                    "roadDisturbance"),
+                         labels = c("Road presence",
+                                    "Forestry disturbance footprint",
+                                    "Road disturbance footprint"))) %>%
+  group_by(metric, sampleDens, agreement) %>%
+  summarise(count = sum(count), .groups = "drop_last") %>%
+  mutate(perc = count/sum(count)*100) %>% 
+  ggplot(aes(x = sampleDens, y =perc, fill = agreement))+
+  geom_col(position = "stack", width = 0.75)+
+  theme_classic()+
+  theme(axis.text.x=element_text(color = "black", size=11, angle=50, vjust=1, hjust=1))+
+  scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
+                               "#999999"))+
+  theme(text = element_text(size = 15), axis.title.x = element_blank())+
+  theme(legend.position="bottom", legend.text = element_text(size = 10), 
+        legend.title = element_blank(),
+        strip.background = element_blank())+
+  labs(y = "Percent of Landscape")+
+  facet_wrap(~metric)
+
+stackedBar
+
+```
 
 # Discussion
-Through assessing the performance of our model by comparing its spatial and aspatial
-accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters.
+
+Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters.
 However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model.
@@
 These potential issues were identified through the use of the difference maps paired with recent satellite imagery that allowed us to better understand what was happening on the ground.
-Often major discrepancies were related to cutblocks that were likely falsely labelled (most often through LANDSAT change detection) in places where there weren’t actually cutblocks.
+Often major discrepancies were related to cutblocks that were likely falsely labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks.
 These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections.
@@
 This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets.
-Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford’s doc).
+Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc).
 This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery.
-These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn’t have and should be treated with caution.
+These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn't have and should be treated with caution.
 It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets.
@@
 If you would like to predict road density as accurately as possible, then running the projections using as fine a resolution of cost surface as possible is the best option.
-If you’re wanting to only predict roughly the location of the roads/save computing time and are more interested in a larger area, then a coarse resolution may suit these requirements perfectly fine.
+If you're wanting to only predict roughly the location of the roads/save computing time and are more interested in a larger area, then a coarse resolution may suit these requirements perfectly fine.
 The tradeoffs of using a coarser resolution of cost surface may be less extreme when projecting roads across a flatter landscape compared to a mountainous landscape like was used for this paper since there would likely be less variability in cost across larger areas.
 
-With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which “roads” can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios.
+With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which "roads" can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios.
 Because caribou are also considered an umbrella species within forests (Drever et al., 2019), projecting forestry impacts on their habitats may also provide valuable insight towards other key species as well.
@@
 We needed to carefully think about what we were trying to achieve.
-The goal wasn’t to present a model that matched the observed road network pixel by pixel, but instead was to simulate disturbance across a large-scale landscape to an acceptable level while still being a relatively simple and straightforward tool.
-It’s for this reason why metrics were chosen that are commonly used in road impact assessments on wildlife; such as the road density and forestry disturbance footprint metrics.
-Once the vector networks were converted into these raster layers, it also facilitated comparing observed versus projected as we could compare the means across the TSA and study where pixels matched and didn’t match.
+The goal wasn't to present a model that matched the observed road network pixel by pixel, but instead was to simulate disturbance across a large-scale landscape to an acceptable level while still being a relatively simple and straightforward tool.
+It's for this reason why metrics were chosen that are commonly used in road impact assessments on wildlife; such as the road density and forestry disturbance footprint metrics.
+Once the vector networks were converted into these raster layers, it also facilitated comparing observed versus projected as we could compare the means across the TSA and study where pixels matched and didn't match.
 Incorporating these metrics into landscape level simulations is also a realistic goal for this tool (ie projecting forest road density across a landscape) so a goal was to therefore see how well the model did at simulating these metrics when compared to the observed landscape.
 
-Many spatial stochastic models don’t include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010).
+Many spatial stochastic models don't include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010).
 Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these.
@@
 # Conclusion
+
 The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort.
@@
 With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale.
-With land use activities transforming a large proportion of the planet’s land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats.
-Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of ~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005).
-With only 7% of the planet’s roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016).
+With land use activities transforming a large proportion of the planet's land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats.
+Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of \~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005).
+With only 7% of the planet's roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016).
 The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.
@@
 
-<div id="refs"></div>
+::: {#refs}
+:::
 
@@
 
-[a]A comment
+# Appendix
+
+## Road network projection methods
+
+Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes.
+Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network.
+We used three methods for solving the MTAP in the roads package.
+These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST).
+In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
+
+LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
+For each STAP, a least cost path from a single source to sink was solved using Dijkstra's shortest path algorithm (1959).
+Dijkstra's algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section "weighted graph").
+Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction.
+This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP.
+They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
+
+The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
+Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
+Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph.
+The minimum spanning tree was solved using Kruskal's algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop.
+This method was most similar to Clark et al.
+(2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling.
+An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.
+
+A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods.
+All the default settings such as skidding distance, etc.
+were left as they were within the QGIS plugin.
+
+## Weighted graph
+
+A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms.
+When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution).
+Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network.
+Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point.
+The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up.
+The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies.
+[The edges connecting the nodes follow a "queens" case or allow a maximum of eight possible edges for a given node.]{.ul} Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only.
+Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs.
+When solving least cost paths, a cost surface raster was used.
+The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels.
+When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

### 2022-07-21 analysis/paper/paper.Rmd
finished updating methods results and figures.

@@
   These are the highlights. 
+editor_options: 
+  chunk_output_type: console
 ---
@@
 This allows us to project road building based on forest harvest data.
-With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.
-The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) (Ministry of Forests, n.d.-b).
-The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010).
+With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) [@brown2010], the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.
+The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests].
+The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population [@brown2010].
 The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats.
-The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).
+The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue [@governmentofbritishcolumbia2021].
 
@@
 library(sf)
+library(waffle) # Used github version 1.0.1
 
@@
 We used several spatial raster layers to construct our cost surface.
-For all cells the base cost followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope.
+For all cells the base cost followed the Interior Appraisal Manual of BC where the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
 Separate costs for river and pipeline crossings were then added to the cost.
@@
 
-Our road network data was obtained by combining the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) data sets for BC.
+Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
 The existing road network, used as the starting point for projections, was created by filtering out the forest tenure roads with an award date greater than 1990.
-Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset (Government of British Columbia, 2021b).
+Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
 Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b).
+The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b].
 Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
@@
 Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape.
-The least cost paths (LCP) method uses Dijkstra's shortest path algorithm (1959) to determine the optimal path between a cutblock and existing road based on a the cost of building a road at any point between the two.
+The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two.
+The path to each cutblock is determined independently of the other cutblocks.
 The cost is represented by a cost raster described below.
 Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock.
-Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm (1956).
+Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956].
 (See Appendix X for a more detailed explanation of these methods).
@@
 We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points with in cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 100 pts/km^2^ (Figure \@ref(fig:fig2)).
+We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 100 pts/km^2^ (Figure \@ref(fig:fig2)).
+Note that the default skidding distance used by the Hardy QGIS plugin is 100 m which corresponds to 25 pts/km^2^ .
 
@@
 
-In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network: 1) Road presence, a simple measure of whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer applied to observed and projected roads (Ibisch et al., 2016); 3) Forestry disturbance footprint, a 500 m buffer applied to observed and projected roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road.
-The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and was calculated using the `caribouMetrics` R package (Endicott et al.,in prep).
+In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network: 1) Road presence, a simple measure of whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer applied to observed and projected roads [@ibisch2016a]; 3) Forestry disturbance footprint, a 500 m buffer applied to observed and projected roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road.
+The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
 Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004).
-Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package (Endicott et al., in prep).
+Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package.
 Distance to nearest road was calculated using a fast moving window method in the `getDistFromSource` function of the roads package.
@@
 Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized.
+We also calculated the sensitivity, precision and F-measure where $sensitivity = \frac{\text{agree roaded}}{(\text{agree roaded} + \text{false negative)}}$, $precision = \frac{\text{agree roaded}}{(\text{agree roaded} + \text{false positive)}}$ and $F\text{-}measure = \frac{(2 * precision * sensitivity)}{(precision + sensitivity)}$.
 
@@
 
-Various projection methods provided more accurate aspatial results than others when compared with the observed metrics depending on the metric used.
-The forestry disturbance footprint metric produced means across all projections that were closest with observed values (Figure 3).
-Projections of road disturbance footprint and distance to road were also reasonably accurate across the whole landscape, with distance to road being less accurate within cutblocks.
-Road presence and road density were less accurate and much more sensitive to assumptions about the distribution of roads within cutblocks (i.e.
-sample density).
-The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm.
-Road presence and road density were also the most sensitive to pixel size (reference appendix table for coarse resolution projections).
-The projection using the Clement's (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.
+Projection accuracy varied widely across the 5 metrics and two scales compared (Figure \@ref(fig:fig3)).
+Across all projection methods road projections were less accurate within cutblocks than in the whole TSA.
+Metrics that used a buffer around the roads, forestry and road disturbance footprint, showed closer agreement between the observed and projected roads than the road density and road presence metrics at both scales.
+For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA.
+The road presence and road density metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (reference appendix table for coarse resolution projections).
 
-## Spatial accuracy
-
-Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks.
-In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a).
-Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data.
-There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b).
-There are also instances where areas were likely misidentified as cutblocks (Figure 4c).
-This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
-
-Figure 5 shows the percentage of matching pixels by category between observed and projected rasters.
-The percentage of roaded prior to 1990 was the same across all projection methods for the respective metrics because the same 1990 road network was used for all projection as the input for the existing road network.
-The agree roadless category represented most of the landscape for all projection methods.
-The subsequent 3 categories (Agree roaded after 1990, project roaded & observe roadless, project roadless & observe roaded) represent the areas where change occurred following 1990.
-
-With the road disturbance footprint metric, the agree roaded after 1990 represented the largest percentage of the landscape which underwent change after 1990.
-The project roaded & observe roadless category represented higher percentages of the landscape than the project roadless & observe roaded, especially when using a higher sampling density, which shows that there were more areas where the model over projected roads than under projected roads.
-The small percentage of project roadless & observe roaded across all projection methods shows that there were very few observed roaded areas that the projections did not address.
-
-Using the forestry disturbance footprint metric, the post-1990 roaded pixels lined up very well.
-Nearly 70% of all post-1990 projected roaded pixels matched with the observed network when using this metric.
-Project roadless & observe roaded very slightly made up a higher percentage of the landscape when compared to project roaded & observe roadless (by less than half a percent across all projection methods).
-
-When using the fine resolution road presence raster to compare with the observed road presence raster, very few of the post-1990 pixels align.
-Nearly all post-1990 roaded pixels did not match between observed and projected.
-The sampling density used determined whether there were more project roaded & observe roadless pixels (higher sampling density) or project roadless & observe roaded pixels (centroid and lower sampling density) amongst the post-1990 roaded pixels.
+When comparing between projection methods we found that the density of landing points chosen within cutblocks made a large difference to the results with the high density method overestimating the road density and road presence metrics while the low density and centroid methods underestimated them.
+The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results for the low sample density but regular sampling was less accurate for all metrics with the high sample density.
+Hardy's QGIS plugin performed well at the overall scale but tended to overestimate the amount of roads within cutblocks.
 
@@
 
-```{r fig3, fig.height=7, fig.width=fig_widths["two"], fig.cap="Revelstoke TSA mean road metrics using different projection methods.
-Observed road network metric values are represented by the horizontal dotted lines for each metric.
-The “overall” column represents mean metric values from across the entire area of the TSA,  whereas the right “cutover” column represents mean values from within cutblock areas only."}
+```{r fig3, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
 # organize data for ggplot
@@
                                         "klementQGIS"),
-                             labels = c("Centroid", "Low\ndensity", "High\ndensity",
-                                        "Hardy\nQGIS")),
+                             labels = c("Centroid", "Low density", "High density",
+                                        "Hardy QGIS")),
          metric = factor(metric,
@@
                                     "roadDisturbanceMean"),
-                         labels = c("Road density", "Road presence",
-                                    "Distance to road",
-                                    "Forestry disturbance footprint",
-                                    "Road disturbance footprint")))
+                         labels = c("Road\ndensity", "Road\npresence",
+                                    "Distance\nto road",
+                                    "Forestry\ndisturbance\nfootprint",
+                                    "Road\ndisturbance\nfootprint")),
+         sampleType = factor(sampleType,
+                             levels = c("centroid",
+                                        "random",
+                                        "regular",
+                                        "klementQGIS", "observed"),
+                             labels = c("Centroid", "Random", "Regular",
+                                        "Hardy QGIS", "Observed")))
 
-observed_values <- filter(meanTable_long, sampleType == "observed")
+observed_values <- filter(meanTable_long, sampleType == "Observed")
 projected_values <- filter(meanTable_long, !is.na(sampleDens))
-# Create bar plot
+
+# get proportional difference from observed
+prop_dif <- projected_values %>% 
+  left_join(observed_values %>% select(areaMean, metric, response),
+            by = c("areaMean", "metric"), 
+            suffix = c("_proj", "_obs")) %>% 
+  mutate(prop_dif = (response_proj-response_obs)/response_obs,
+         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall")))
+
+prop_dif %>% 
+  ggplot(aes(x = sampleDens, prop_dif, fill = sampleType))+
+  geom_hline(aes(yintercept = 0), color = "grey")+
+  geom_col(position = position_dodge2(preserve = "single",
+                                      width = 0.75))+
+  facet_grid(areaMean~ metric)+
+  theme_classic()+
+  scale_fill_manual(values = met.brewer(pal_nm, 4), name = "Sampling\nMethod")+
+  theme(text = element_text(size = 10), axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        legend.position = "right",
+        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))
+
+# Create bar plot This is the old version keep in case
 fig3 <- projected_values %>% group_by(areaMean) %>% nest() %>% ungroup() %>%
@@
 
-fig3
 ```
 
-```{r fig4, fig.height=5, fig.width=fig_widths["two"], fig.cap="Spatial accuracy"}
+## Spatial accuracy
 
-stackedBar <- matchData %>% ungroup() %>%
+Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks.
+In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a).
+Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data.
+There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b).
+There are also instances where areas were likely misidentified as cutblocks (Figure 4c).
+This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
+
+```{r fig4, fig.height=9, fig.width=fig_widths["two"], fig.cap="Difference maps used to investigate the performance of the projections.
+Image “a” shows an area of low discrepancy where the projections matched well.
+When zooming in on the discrepancy shown in image “b”, one can see from the satellite imagery that there are forest roads in the imagery that are missing from the observed data.
+Image “c” reveals one example of a cutblock that has most likely been mislabelled by remote sensing as it is located on the summit of a mountain."}
+
+knitr::include_graphics(here("analysis/figures/Figure4_diff_map.png"))
+
+```
+
+```{r fig5, fig.height=9, fig.width=fig_widths["two"], fig.cap="Spatial accuracy"}
+
+matchData_sum <- matchData %>% ungroup() %>%
   mutate(agreement = factor(agreement,
@@
                              "klementQGIS"),
-                  labels = c("Centroid", "Random low density", "Random high density",
-                             "Regular low density",
-                             "Regular high density",
+                  labels = c("Centroid", "Random\nlow density", "Random\nhigh density",
+                             "Regular\nlow density",
+                             "Regular\nhigh density",
                              "Hardy QGIS")),
@@
   summarise(count = sum(count), .groups = "drop_last") %>%
-  mutate(perc = count/sum(count)*100) %>% 
+  mutate(perc = count/sum(count)*100)
+
+stackedBar <- matchData_sum %>%
   ggplot(aes(x = sampleDens, y =perc, fill = agreement))+
@@
 
-stackedBar
+matchData_sum %>%
+  #filter(metric == "Road presence") %>%
+  ggplot(aes(values = perc*3,
+             fill = factor(agreement, levels = levels(agreement))))+
+  geom_waffle(size = 0.25, colour = "white", flip = TRUE, n_rows = 9)+
+  facet_grid(metric~ sampleDens)+
+  scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
+                               "#999999"), name = "Agreement")+
+  scale_x_discrete(name = NULL) +
+  scale_y_continuous(labels = function(x) x * 9/3,
+                     expand = c(0,0), name =  "Percent of Landscape") +
+  coord_equal()+
+  theme_minimal() +
+  theme(text = element_text(size = 10),
+        panel.grid = element_blank(),
+        axis.ticks.y = element_line(),
+        legend.position = "bottom") +
+  guides(fill=guide_legend(nrow=2))
+
+# performance table
+perf_tbl <- matchData_sum %>% 
+  pivot_wider(c(metric, sampleDens), names_from = agreement, 
+              values_from = count) %>% 
+  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
+         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
+         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
+  select(metric, sampleDens, sensitivity, precision, F_measure)
+
+perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
+  summarise(rng_F = range(F_measure) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
+
+perf_tbl %>% 
+  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
+               values_to = "value") %>% 
+  ggplot(aes(sampleDens, value))+
+  geom_col()+
+  facet_grid(perf_meas~metric)+
+  coord_flip()+
+  ylim(0, 1)+
+  theme_classic()
 
 ```
+
+As expected all the projection methods preformed worst with with road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance footprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
+For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives.
+For the forestry and road disturbance footprint metrics, the lower density projection methods had slightly higher F-measure scores due to fewer false positives and only slightly more false negatives.
 
@@
 
-Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data; these discrepancies should be treated with caution as they are not necessarily the fault of the projections themselves and rather with issues in available data.
-These potential issues were identified through the use of the difference maps paired with recent satellite imagery that allowed us to better understand what was happening on the ground.
-Often major discrepancies were related to cutblocks that were likely falsely labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks.
+Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data.
+Often major discrepancies were related to cutblocks that were likely incorrectly labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks.
 These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections.
@@
 For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.
+
+-The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm.
 

### 2022-07-21 analysis/scripts/makeFigures.R

@@ -1,127 +1,0 @@
-##### Metric table bar plots
-library(ggplot2)
-library(dplyr)
-library(stringr)
-library(tidyr)
-library(purrr)
-library(ggpubr)
-library(MetBrewer)
-
-data_path_drvd <- "analysis/data/derived_data/"
-
-fig_path <- "analysis/figures/"
-
-fig_dev <- "jpeg"
-
-pal_nm <- "Egypt"
-
-
-# Figure 3 #====================================================================
-
-# data (from "meanTable") in the runProjections script
-meanTable <- read.csv(paste0(data_path_drvd, "mean_table.csv"))
-
-# organize data for ggplot
-meanTable_long <- meanTable %>%
-  pivot_longer(-c(sampleType, sampleDens, areaMean),
-               names_to = "metric", values_to = "response")%>%
-  mutate(sampleDens = factor(sampleDens,
-                             levels = c("centroid",
-                                        "low sample density",
-                                        "high sample density",
-                                        "klementQGIS"),
-                             labels = c("Centroid", "Low\ndensity", "High\ndensity",
-                                        "Klement\nQGIS")),
-         metric = factor(metric,
-                         levels = c("roadDensityMean", "roadPresenceMean",
-                                    "distanceToRoadMean",
-                                    "forestryDisturbanceMean",
-                                    "roadDisturbanceMean"),
-                         labels = c("Road density", "Road presence",
-                                    "Distance to road",
-                                    "Forestry disturbance footprint",
-                                    "Road disturbance footprint")))
-
-
-
-observed_values <- filter(meanTable_long, sampleType == "observed")
-projected_values <- filter(meanTable_long, !is.na(sampleDens))
-# Create bar plot
-fig3 <- projected_values %>% group_by(areaMean) %>% nest() %>% ungroup() %>%
-  mutate(obs = observed_values %>% group_by(areaMean) %>% nest() %>% pull(data)) %>%
-  group_by(areaMean) %>%
-  mutate(data = map(data, ~mutate(.x, area = areaMean)),
-         plot = map2(data, obs,
-                    ~ ggplot(.x, aes(x = sampleDens, response, fill = sampleType))+
-                      geom_col(position = position_dodge2(preserve = "single",
-                                                         width = 0.75))+
-                      geom_hline(aes(yintercept = response),
-                                 .y, color = "black", linetype = "dashed")+
-                      facet_wrap(~metric, scales = "free_y", ncol = 1)+
[…]

### 2022-08-23 analysis/scripts/runProjections.R
Use new versions of data

@@ -20,16 +20,16 @@ library(pfocal)
 devtools::load_all()
 
 # set rasterOptions to allow higher memory usage
-prevOpts <- terraOptions(memfrac = 0.9)
+#prevOpts <- terraOptions(memfrac = 0.9)
 
 ######### load in data for projections ########################################
 
 #forest harvest cutblocks
-cutblocks <- st_make_valid(st_read(paste0(data_path_raw, "cutblocks_revelstoke.shp")))
+cutblocks <- st_make_valid(st_read(paste0(data_path_raw, "cutblocks_revelstoke.gpkg")))
 #modern observed roads
-roads <- st_read(paste0(data_path_raw, "roads_revelstoke.shp"))
+roads <- st_read(paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"))
 #boundary for running projection
-tsaBoundary <- st_read(paste0(data_path_raw, "new_tsa27_boundaries.shp"))
+tsaBoundary <- st_read(paste0(data_path_raw, "tsa27_boundaries.gpkg"))
 #cost surface raster layer
 bc_cost_surface <- terra::rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
 
@@ -42,7 +42,7 @@ if(use_sub){
                          full.names = TRUE), st_read)
   # tsbs <- tsbs[1:3] #for testing on smaller area
 } else {
-  tsbs <- list(tsaBoundary)
+  tsbs <- list(st_union(tsaBoundary))
 }
 
 #Klement QGIS projection results shapefile
@@ -52,12 +52,16 @@ klementProj <- st_read(paste0(data_path_drvd,
 ###### Parameters #############################################################
 
 # set years for projection start (the projection is set to go from 1990 onwards)
-roadsYear <- 19900000
+roadsYear <- as.Date("1990-01-01")
+# need to limit roads included to only those from max harvest year or earlier
+maxHarvYear <- cutblocks$HARVEST_YEAR %>% max %>% as.character() %>%
+  paste0("-12-31") %>%  as.Date(format = "%Y-%m-%d")
+
 # set cutblock year
 cutblocksYear <- 1990
 #set high and low sampling densities for projections
-low <- 0.000001
-high <- 0.0001
+low  <- 0.000001
+high <- 0.00001
 #if lakes block path then set high value for lakes (NA) on cost surface.
 lakeValue <- 65000 #often this isn't needed but Revelstoke TSA requires this.
[…]

### 2022-08-23 R/functionsSource.R
add runtime to results

@@ -241,6 +241,7 @@ projectAll<-function(tsbs,paramTable, costSurface,
   for(i in 1:nrow(paramTable)) {
     # i = 1
     cRow = paramTable[i,]
+    start <- Sys.time()
 
     projectionsList <- map(tsbs,
                            fineResProject,
@@ -253,9 +254,12 @@ projectAll<-function(tsbs,paramTable, costSurface,
 
     projections <- do.call(rbind, projectionsList)
 
+    end <- Sys.time()
+
 
     paramTable$output[[i]] <- paste0(fileLocation, "_", cRow$sampleType, "_",
                                      cRow$sampleDens, ".shp")
+    paramTable$runTime[[i]] <- as.numeric(end - start)
 
     # save the projected roads to a file
     sf::write_sf(projections, paramTable$output[[i]])

### 2022-08-23 analysis/paper/paper.Rmd
minor changes to figures

@@
 author:
-  - Josie Hughes:
+  - name: Josie Hughes
     email: josie.hughes@ec.gc.ca
@@
     correspondence: true
-  - Sarah Endicott:
+  - name: Sarah Endicott
     email: sarah.endicott@ec.gc.ca
@@
 
-tsaBoundary <- read_sf(here(data_path_raw, "new_tsa27_boundaries.shp"))
+tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
 #cost surface raster layer
@@
 Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
-The existing road network, used as the starting point for projections, was created by filtering out the forest tenure roads with an award date greater than 1990.
+To avoid situations where the same road was included in both data sets with slightly different coordinates we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set.
+The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990.
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
@@
 
-```{r fig3, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+```{r fig3, fig.height=3.3, fig.width=fig_widths["single"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
 Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
@@
                names_to = "metric", values_to = "response")%>%
-  mutate(sampleDens = factor(sampleDens,
-                             levels = c("centroid",
-                                        "low sample density",
-                                        "high sample density",
-                                        "klementQGIS"),
-                             labels = c("Centroid", "Low density", "High density",
+  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
+                             sampleType),
+         sampleDens = paste(sampleType, sampleDens) %>% 
+           factor(levels = c("NA centroid",
+                             "NA observed",
+                             "random low sample density",
+                             "regular low sample density",
+                             "random high sample density",
+                             "regular high sample density",
+                             "NA klementQGIS"),
+                  labels = c("Centroid", 
+                             "Observed",
+                             "Random\nlow density",
+                             "Regular\nlow density",
+                             "Random\nhigh density",
+                             "Regular\nhigh density",
                              "Hardy QGIS")),
@@
                                     "Forestry\ndisturbance\nfootprint",
-                                    "Road\ndisturbance\nfootprint")),
-         sampleType = factor(sampleType,
-                             levels = c("centroid",
-                                        "random",
-                                        "regular",
-                                        "klementQGIS", "observed"),
-                             labels = c("Centroid", "Random", "Regular",
-                                        "Hardy QGIS", "Observed")))
+                                    "Road\ndisturbance\nfootprint")))
 
-observed_values <- filter(meanTable_long, sampleType == "Observed")
-projected_values <- filter(meanTable_long, !is.na(sampleDens))
+observed_values <- filter(meanTable_long, sampleDens == "Observed")
+projected_values <- filter(meanTable_long, sampleDens != "Observed")
 
@@
 prop_dif %>% 
-  ggplot(aes(x = sampleDens, prop_dif, fill = sampleType))+
+  ggplot(aes(x = sampleDens, prop_dif, fill = sampleDens))+
   geom_hline(aes(yintercept = 0), color = "grey")+
@@
   theme_classic()+
-  scale_fill_manual(values = met.brewer(pal_nm, 4), name = "Sampling\nMethod")+
+  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:4, 6)],
+                    name = "Sampling\nMethod", guide = "none")+
   theme(text = element_text(size = 10), axis.title.x = element_blank(),
@@
         legend.position = "right",
-        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))
+        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
+  coord_flip()
 
@@
                          labels = c("Road presence",
-                                    "Forestry disturbance footprint",
+                                    "Forestry disturbance\nfootprint",
                                     "Road disturbance footprint"))) %>%
@@
 
-matchData_sum %>%
-  #filter(metric == "Road presence") %>%
-  ggplot(aes(values = perc*3,
-             fill = factor(agreement, levels = levels(agreement))))+
-  geom_waffle(size = 0.25, colour = "white", flip = TRUE, n_rows = 9)+
-  facet_grid(metric~ sampleDens)+
-  scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
-                               "#999999"), name = "Agreement")+
-  scale_x_discrete(name = NULL) +
-  scale_y_continuous(labels = function(x) x * 9/3,
-                     expand = c(0,0), name =  "Percent of Landscape") +
-  coord_equal()+
-  theme_minimal() +
-  theme(text = element_text(size = 10),
-        panel.grid = element_blank(),
-        axis.ticks.y = element_line(),
-        legend.position = "bottom") +
-  guides(fill=guide_legend(nrow=2))
+# waffle stacked bar
+# matchData_sum %>%
+#   #filter(metric == "Road presence") %>%
+#   ggplot(aes(values = perc*3,
+#              fill = factor(agreement, levels = levels(agreement))))+
+#   geom_waffle(size = 0.25, colour = "white", flip = TRUE, n_rows = 9)+
+#   facet_grid(metric~ sampleDens)+
+#   scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
+#                                "#999999"), name = "Agreement")+
+#   scale_x_discrete(name = NULL) +
+#   scale_y_continuous(labels = function(x) x * 9/3,
+#                      expand = c(0,0), name =  "Percent of Landscape") +
+#   coord_equal()+
+#   theme_minimal() +
+#   theme(text = element_text(size = 10),
+#         panel.grid = element_blank(),
+#         axis.ticks.y = element_line(),
+#         legend.position = "bottom") +
+#   guides(fill=guide_legend(nrow=2))
 
@@
                values_to = "value") %>% 
-  ggplot(aes(sampleDens, value))+
+  mutate(perf_meas = factor(perf_meas,
+                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
+  ggplot(aes(sampleDens, value, fill = sampleDens))+
   geom_col()+
-  facet_grid(perf_meas~metric)+
+  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:4, 6)],
+                    name = "Sampling\nMethod", guide = "none")+
+  facet_grid(metric ~ perf_meas)+
   coord_flip()+
   ylim(0, 1)+
-  theme_classic()
+  theme_classic()+
+  labs(y = "Performance", x = NULL)
 
@@
 
-As expected all the projection methods preformed worst with with road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance footprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
+As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance footprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
 For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives.

### 2022-08-23 analysis/paper/paper.Rmd
update description

@@
 library(sf)
-library(waffle) # Used github version 1.0.1
 

### 2022-08-23 analysis/paper/paper.Rmd
add ref for Fmeasure etc #5

@@
 Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized.
-We also calculated the sensitivity, precision and F-measure where $sensitivity = \frac{\text{agree roaded}}{(\text{agree roaded} + \text{false negative)}}$, $precision = \frac{\text{agree roaded}}{(\text{agree roaded} + \text{false positive)}}$ and $F\text{-}measure = \frac{(2 * precision * sensitivity)}{(precision + sensitivity)}$.
+We used the performance metrics sensitivity, precision and F-measure to compare the results for each projection method [@lee2021].
 
@@
 The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies.
-[The edges connecting the nodes follow a "queens" case or allow a maximum of eight possible edges for a given node.]{.ul} Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only.
+[The edges connecting the nodes follow a "queens" case or allow a maximum of eight possible edges for a given node.]{.underline} Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only.
 Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs.

### 2022-08-23 analysis/paper/paper.Rmd
Author data not working so commented out

@@
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
+      # pandoc_args:
+      # - --lua-filter=../templates/scholarly-metadata.lua
+      # - --lua-filter=../templates/author-info-blocks.lua
+      # - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib

### 2022-08-23 analysis/paper/paper.Rmd
fix colour scale and ordering in figs

@@
 
-```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel."}
+```{r setup2}
 ##### Metric table bar plots
@@
 
-pal_nm <- "Egypt"
+pal_nm <- "Redon"
 
 fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
+```
 
+```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel."}
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
@@
 
-```{r fig3, fig.height=3.3, fig.width=fig_widths["single"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+```{r fig3, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
 Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
@@
 
-```{r fig5, fig.height=9, fig.width=fig_widths["two"], fig.cap="Spatial accuracy"}
+```{r fig5, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method.
+F measure is the harmonic mean of precision and sensitivity, precision is the proportion of  predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
 
@@
                              "random low density",
+                             "regular low density",
                              "random high density",
-                             "regular low density",
                              "regular high density",
                              "klementQGIS"),
-                  labels = c("Centroid", "Random\nlow density", "Random\nhigh density",
+                  labels = c("Centroid", 
+                             "Random\nlow density",
                              "Regular\nlow density",
+                             "Random\nhigh density",
                              "Regular\nhigh density",

### 2022-08-26 analysis/scripts/prepareData.R
Move maxHarvYear filter to here and add Ft Nelson

@@ -103,6 +103,15 @@ dra_rds_dif <- st_difference(dra_rds, for_rds_buf)
 comb_rds <- bind_rows(for_rds, dra_rds_dif) %>%
   filter(is.na(ROAD_CLASS) | ROAD_CLASS != "trail")
 
+# need to limit roads included to only those from max harvest year or earlier
+maxHarvYear <- cutblocks$HARVEST_YEAR %>% max %>% as.character() %>%
+  paste0("-12-31") %>%  as.Date(format = "%Y-%m-%d")
+
+# set roads with no year as existing in 1990
+roadsYear <- as.Date("1990-01-01")
+comb_rds$AWARD_DATE[is.na(comb_rds$AWARD_DATE)] <- roadsYear
+comb_rds <- filter(comb_rds, AWARD_DATE <= maxHarvYear)
+
 write_sf(comb_rds, here(data_path_drvd, "combined_revelstoke_roads.gpkg"))
 
 # Another option might be this "integrated roads dataset". Their method is
@@ -120,3 +129,53 @@ bcdc_get_record("9568a219-819d-417a-be68-8431b6fb5de0")
 
 
 
+# Repeat for Fort Nelson --------------------------------------------------
+
+tsa_8 <- bcdc_query_geodata("8daa29da-d7f4-401c-83ae-d962e3a28980") %>%
+  filter(TSA_NUMBER == "08") %>%
+  collect()
+
+# using .gpkg because writing to .shp abbreviates the column names
+write_sf(tsa_8, here(data_path_raw, "tsa8_boundaries.gpkg"))
+
+dra_rds <- bcdc_query_geodata("bb060417-b6e6-4548-b837-f9060d94743e") %>%
+  filter(INTERSECTS(tsa_8)) %>%
+  collect() %>%
+  st_filter(tsa_8)
+
+write_sf(dra_rds, here(data_path_raw, "dra_roads_ft_nelson.gpkg"))
+
+for_rds <- bcdc_query_geodata("243c94a1-f275-41dc-bc37-91d8a2b26e10") %>%
+  filter(INTERSECTS(tsa_8)) %>%
+  collect() %>%
+  st_filter(tsa_8)
+
+write_sf(for_rds, here(data_path_raw, "forest_tenure_roads_ft_nelson.gpkg"))
+
+cutblocks <- bcdc_query_geodata("b1b647a6-f271-42e0-9cd0-89ec24bce9f7") %>%
[…]

### 2022-08-26 analysis/scripts/runProjections.R
Convert to function and add coarse and Ft Nelson

@@ -22,156 +22,66 @@ devtools::load_all()
 # set rasterOptions to allow higher memory usage
 #prevOpts <- terraOptions(memfrac = 0.9)
 
-######### load in data for projections ########################################
-
-#forest harvest cutblocks
-cutblocks <- st_make_valid(st_read(paste0(data_path_raw, "cutblocks_revelstoke.gpkg")))
-#modern observed roads
-roads <- st_read(paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"))
-#boundary for running projection
-tsaBoundary <- st_read(paste0(data_path_raw, "tsa27_boundaries.gpkg"))
-#cost surface raster layer
-bc_cost_surface <- terra::rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
-
-
-#Use subsets of TSA (in order to avoid memory issues)?
-use_sub <- FALSE
-
-if(use_sub){
-  tsbs <- map(list.files(paste0(data_path_raw, "subs/"), pattern = ".shp",
-                         full.names = TRUE), st_read)
-  # tsbs <- tsbs[1:3] #for testing on smaller area
-} else {
-  tsbs <- list(st_union(tsaBoundary))
-}
-
-#Klement QGIS projection results shapefile
-klementProj <- st_read(paste0(data_path_drvd,
-                              "klementProjection.shp"))
-
-###### Parameters #############################################################
-
-# set years for projection start (the projection is set to go from 1990 onwards)
-roadsYear <- as.Date("1990-01-01")
-# need to limit roads included to only those from max harvest year or earlier
-maxHarvYear <- cutblocks$HARVEST_YEAR %>% max %>% as.character() %>%
-  paste0("-12-31") %>%  as.Date(format = "%Y-%m-%d")
-
-# set cutblock year
-cutblocksYear <- 1990
-#set high and low sampling densities for projections
-low  <- 0.000001
+#densities to test
+low <- 0.000001
 high <- 0.00001
-#if lakes block path then set high value for lakes (NA) on cost surface.
-lakeValue <- 65000 #often this isn't needed but Revelstoke TSA requires this.
-aggFact <- 1 #factor of aggregation of cost surface. 1 = no aggregation.
 
-###############################################################################
-
-# parameter table creation for running projections
-sampleDens <- c(low,high,low,high,low)
-sampleType <- c("regular","regular","random","random","centroid")
[…]

### 2022-08-26 analysis/paper/paper.Rmd
change high density #2 and #4

@@
 data_path_raw <- "analysis/data/raw_data/"
-data_path_drvd <- "analysis/data/derived_data/"
+data_path_drvd <- "analysis/data/derived_data/TSA27"
 
@@
 We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 100 pts/km^2^ (Figure \@ref(fig:fig2)).
+We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
 Note that the default skidding distance used by the Hardy QGIS plugin is 100 m which corresponds to 25 pts/km^2^ .
@@
 
-When comparing between projection methods we found that the density of landing points chosen within cutblocks made a large difference to the results with the high density method overestimating the road density and road presence metrics while the low density and centroid methods underestimated them.
-The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results for the low sample density but regular sampling was less accurate for all metrics with the high sample density.
+When comparing between projection methods we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods.
+The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results.
 Hardy's QGIS plugin performed well at the overall scale but tended to overestimate the amount of roads within cutblocks.
@@
 
-```{r fig5, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method.
+```{r fig5, fig.height=6, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method.
 F measure is the harmonic mean of precision and sensitivity, precision is the proportion of  predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
@@
            str_replace("1e-06", "low density") %>%
-           str_replace("1e-04", "high density") %>%
-           str_replace("centroid high density", "centroid") %>%
+           str_replace("1e-05", "high density") %>%
+           str_replace("centroid low density", "centroid") %>%
            str_replace("klementQGIS NA", "klementQGIS") %>%

### 2022-08-26 R/functionsSource.R
fix issue in calcAgree for coarse resolution

@@ -433,6 +433,9 @@ getMetricMeans <- function(paramTable, cutblocks){
 #' @return a table
 #' @noRd
 calcAgree <- function(obs_rast, proj_rast, prex_rast){
+  if(res(proj_rast)[1]!= res(obs_rast)){
+    proj_rast <- aggregate(proj_rast, aggFact)
+  }
   proj_rast <- terra::crop(proj_rast, obs_rast)
   res <- obs_rast + prex_rast + proj_rast
 
@@ -458,7 +461,7 @@ calcAgree <- function(obs_rast, proj_rast, prex_rast){
 #' @param boundary polygon boundary to mask rasters
 
 
-agreeMetricsAll <- function(paramTable, prex_rast, prex_vect, boundary, cutblocks){
+agreeMetricsAll <- function(paramTable, prex_rast, prex_vect, boundary, cutblocks, nonAggregatedCostSurface){
   obs_tbl <- paramTable %>% filter(sampleType == "observed") %>%
     dplyr::select(roadDisturbance, roadPresence, forestryDisturbance) %>%
     tidyr::pivot_longer(everything(), names_to = "metric", values_to = "obs_rast") %>%
@@ -472,9 +475,9 @@ agreeMetricsAll <- function(paramTable, prex_rast, prex_vect, boundary, cutblock
     prex_rast = list(roadDisturbanceFootprint(prex_vect, !is.na(prex_rast), boundary),
                      prex_rast,
                      disturbanceMetrics(linFeat = prex_vect,
-                                        landCover = as(!is.na(prex_rast), "Raster"),
+                                        landCover = as(nonAggregatedCostSurface, "Raster"),
                                         projectPoly = boundary,
-                                        anthroDist = terra::rasterize(terra::vect(cutblocks), prex_rast) %>%
+                                        anthroDist = terra::rasterize(terra::vect(cutblocks), nonAggregatedCostSurface) %>%
                                           as("Raster"))@processedData$Anthro %>% terra::rast())
   ) %>%
     mutate(prex_rast = prex_rast %>%
@@ -576,9 +579,9 @@ run_projections <- function(cutblocksPth, roadsPth, tsaBoundaryPth, costPth,
                            method = method)
 
   # recreate allResults after a restart using saved files
-  # allResults <- paramTable[c(1,3,5),] %>%
-  #   mutate(output = paste0(data_path_drvd, "TSA27", "_", sampleType, "_",
-  #                          sampleDens, ".shp"))
[…]

### 2022-08-26 analysis/scripts/method_example_fig.R
Redo figure showing different methods and parameters

@@ -1,0 +1,108 @@
+# Example area for demonstrating different methods
+
+library(raster)
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(tmap)
+library(purrr)
+library(caribouMetrics)
+library(pfocal)
+
+devtools::load_all(".")
+
+#set path for data
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
+
+cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg")
+roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg")
+tsaBoundaryPth = paste0(data_path_raw, "tsa27_58_testing.gpkg")
+costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif")
+
+# densities
+high <- 0.00001
+low <-  0.000005
+
+
+cutblocks <- read_sf(cutblocksPth)
+roads <- read_sf(roadsPth)
+tsaCost <- rast(costPth)
+
+projRoads <- read_sf(paste0(data_path_drvd, "TSA27/", "TSA27_random_1e-05.shp"))
+
+exRoads <- roads %>% filter(AWARD_DATE <= as.Date("1990-01-01"))
+
+# find a good location
+qtm(st_as_sf(cutblocks), fill = NULL, borders.lwd = 2)+
+  qtm(st_as_sf(roads), lines.col = "red", lines.lwd = 2)+
+  qtm(st_as_sf(projRoads), lines.col = "green", lines.lwd = 2)+
+  qtm(st_as_sf(exRoads), lines.lwd = 2)+
+    tm_layout(legend.show = FALSE)
+
+# # get extent
+# plot(roads)
+#
+# bound <- draw()
+#SpatExtent : 1537685.65481532, 1546695.23062125, 741830.821988938, 750131.44756752 (xmin, xmax, ymin, ymax)
+bound <- ext(c(1537685.65481532, 1546695.23062125, 741830.821988938, 750131.44756752))
+
+roads <- vect(roads) %>% crop(bound) %>% st_as_sf()
+cutblocks <- vect(cutblocks) %>% crop(bound) %>% st_as_sf()
+projRoads <- vect(projRoads) %>% crop(bound) %>% st_as_sf()
+exRoads <- vect(exRoads) %>% crop(bound) %>% st_as_sf()
[…]

### 2022-08-26 R/functionsSource.R
fix calcAgree for coarse res

@@ -433,10 +433,8 @@ getMetricMeans <- function(paramTable, cutblocks){
 #' @return a table
 #' @noRd
 calcAgree <- function(obs_rast, proj_rast, prex_rast){
-  if(res(proj_rast)[1]!= res(obs_rast)){
-    proj_rast <- aggregate(proj_rast, aggFact)
-  }
   proj_rast <- terra::crop(proj_rast, obs_rast)
+  prex_rast <- terra::crop(prex_rast, obs_rast)
   res <- obs_rast + prex_rast + proj_rast
 
   lu_tbl <- tibble::tribble(~value, ~agreement,

### 2022-08-26 analysis/paper/paper.Rmd
new projection methods figure and add appendix with coarse results

@@
 ```{r fig2, fig.height=5, fig.width=fig_widths["two"], fig.cap="Example of the different projection methods and sampling strategies."}
-knitr::include_graphics(here("analysis/figures/projection methods figure.jpg"))
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.jpg"))
 ```
@@
 For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA.
-The road presence and road density metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (reference appendix table for coarse resolution projections).
+The road density and road distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).
 
@@
 For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives.
-For the forestry and road disturbance footprint metrics, the lower density projection methods had slightly higher F-measure scores due to fewer false positives and only slightly more false negatives.
+For the forestry and road disturbance footprint metrics, the F-measure was similar across projection methods.
 
@@
 
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```
-
 # Appendix
@@
 When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.
+
+## Coarse resolution results
+
+```{r get-data2}
+data_path_drvd <- "analysis/data/derived_data/TSA27_1000"
+
+# load csv
+meanTable2 <- read.csv(here(data_path_drvd, "mean_table.csv")) %>% 
+  mutate(resolution = "1000")
+
+matchData2 <- read.csv(here(data_path_drvd, "agree_table.csv")) %>% 
+  mutate(resolution = "1000")
+
+meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2)
+
+matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2)
+
+```
+
+```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters.
+Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
+# organize data for ggplot
+meanTable_long <- meanTable %>%
+  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution),
+               names_to = "metric", values_to = "response")%>%
+  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
+                             sampleType),
+         sampleDens = paste(sampleType, sampleDens) %>% 
+           factor(levels = c("NA centroid",
+                             "NA observed",
+                             "random low sample density",
+                             "regular low sample density",
+                             "random high sample density",
+                             "regular high sample density",
+                             "NA klementQGIS"),
+                  labels = c("Centroid", 
+                             "Observed",
+                             "Random\nlow density",
+                             "Regular\nlow density",
+                             "Random\nhigh density",
+                             "Regular\nhigh density",
+                             "Hardy QGIS")),
+         metric = factor(metric,
+                         levels = c("roadDensityMean", "roadPresenceMean",
+                                    "distanceToRoadMean",
+                                    "forestryDisturbanceMean",
+                                    "roadDisturbanceMean"),
+                         labels = c("Road\ndensity", "Road\npresence",
+                                    "Distance\nto road",
+                                    "Forestry\ndisturbance\nfootprint",
+                                    "Road\ndisturbance\nfootprint")))
+
+observed_values <- filter(meanTable_long, sampleDens == "Observed")
+projected_values <- filter(meanTable_long, sampleDens != "Observed")
+
+# get proportional difference from observed
+prop_dif <- projected_values %>% 
+  left_join(observed_values %>% select(areaMean, metric, resolution, response),
+            by = c("areaMean", "metric", "resolution"), 
+            suffix = c("_proj", "_obs")) %>% 
+  mutate(prop_dif = (response_proj-response_obs)/response_obs,
+         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall")))
+
+prop_dif %>% 
+  ggplot(aes(x = sampleDens, prop_dif, fill = resolution))+
+  geom_hline(aes(yintercept = 0), color = "grey")+
+  geom_col(position = position_dodge2(preserve = "single",
+                                      width = 0.75))+
+  facet_grid(areaMean~ metric)+
+  theme_classic()+
+  scale_fill_manual(values = met.brewer(pal_nm))+
+  theme(text = element_text(size = 10), axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        legend.position = "right",
+        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
+  coord_flip()
+
+```
+
+```{r figA2, fig.height=6, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method.
+F measure is the harmonic mean of precision and sensitivity, precision is the proportion of  predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
+
+matchData_sum <- matchData %>% ungroup() %>%
+  mutate(agreement = factor(agreement,
+                            levels = c("False negative", "False positive",
+                                       "Agree roaded","Agree roadless",
+                                       "Pre-existing roads")),
+         sampleDens = paste(sampleType, sampleDens) %>%
+           str_replace("1e-06", "low density") %>%
+           str_replace("1e-05", "high density") %>%
+           str_replace("centroid low density", "centroid") %>%
+           str_replace("klementQGIS NA", "klementQGIS") %>%
+           factor(levels = c("centroid",
+                             "random low density",
+                             "regular low density",
+                             "random high density",
+                             "regular high density",
+                             "klementQGIS"),
+                  labels = c("Centroid", 
+                             "Random\nlow density",
+                             "Regular\nlow density",
+                             "Random\nhigh density",
+                             "Regular\nhigh density",
+                             "Hardy QGIS")),
+         metric = factor(metric,
+                         levels = c("roadPresence",
+                                    "forestryDisturbance",
+                                    "roadDisturbance"),
+                         labels = c("Road presence",
+                                    "Forestry disturbance\nfootprint",
+                                    "Road disturbance footprint"))) %>%
+  group_by(metric, sampleDens, agreement, resolution) %>%
+  summarise(count = sum(count), .groups = "drop_last") %>%
+  mutate(perc = count/sum(count)*100)
+
+# performance table
+perf_tbl <- matchData_sum %>% 
+  pivot_wider(c(metric, sampleDens, resolution), names_from = agreement, 
+              values_from = count) %>% 
+  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
+         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
+         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
+  select(resolution, metric, sampleDens, sensitivity, precision, F_measure)
+
+perf_tbl_rng <- perf_tbl %>% group_by(metric, resolution) %>% 
+  summarise(rng_F = range(F_measure) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
+
+perf_tbl %>% 
+  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
+               values_to = "value") %>% 
+  mutate(perf_meas = factor(perf_meas,
+                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
+  ggplot(aes(sampleDens, value, fill = resolution))+
+  geom_col(position = position_dodge2(preserve = "single",
+                                      width = 0.75))+
+  scale_fill_manual(values = met.brewer(pal_nm))+
+  facet_grid(metric ~ perf_meas)+
+  coord_flip()+
+  ylim(0, 1)+
+  theme_classic()+
+  labs(y = "Performance", x = NULL)
+
+```
+
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2022-08-26 analysis/scripts/method_example_fig.R

@@ -55,6 +55,7 @@ exRoads <- vect(exRoads) %>% crop(bound) %>% st_as_sf()
 tsaCost <- crop(tsaCost, bound)
 
 # add a large cutblock
+plot(roads %>% st_geometry())
 cut_add <- draw("polygon")
 cut_add <- set.crs(cut_add, crs(roads))
 
@@ -62,6 +63,14 @@ cutblocks <- bind_rows(cutblocks, st_as_sf(cut_add))
 
 #
 tsb <- st_bbox(roads) %>% st_as_sfc() %>% list()
+
+
+# save example area for other testing
+write_sf(exRoads, paste0(data_path_drvd, "testing_ex_roads.gpkg"))
+write_sf(tsb[[1]], paste0(data_path_drvd, "testing_tsb.gpkg"))
+write_sf(cutblocks, paste0(data_path_drvd, "testing_cutblocks.gpkg"))
+writeRaster(tsaCost, paste0(data_path_drvd, "testing_cost.tif"))
+
 
 sampleDens <- c(low,high,low,high,low)
 sampleType <- c("regular","regular","random","random","centroid")
@@ -105,4 +114,5 @@ allMaps <- purrr::map(1:nrow(allProj),
                                                 ifelse(allProj$sampleType[.x] == "centroid", "",
                                                        allProj$sampleDens[.x]))))
 
-tmap_arrange(allMaps, ncol = 2)
+tmap_save(tmap_arrange(allMaps, ncol = 2), "analysis/figures/projection_methods_figure.jpg",
+          dpi = 300, height = 8, width = 7.48)

### 2022-08-26 analysis/paper/paper.Rmd
add ft nelson results

@@
 
+## Fort Nelson map
+
+```{r ftNelson, fig.height=6, fig.width=fig_widths["two"], fig.cap="Road projections for the Fort Nelson Timber Supply Area.
+Black lines show projected and pre-existing roads.
+Red lines show observed roads that were not captured by the projections."}
+ftN_roads_proj <- read_sf(here("analysis/data/derived_data/TSA8_1000/centroid_1e-05.shp"))
+
+ftN_roads_obs <- read_sf(here("analysis/data/derived_data/combined_ft_nelson_roads.gpkg"))
+
+qtm(ftN_roads_obs, lines.col = "red")+
+  qtm(ftN_roads_proj)
+```
+
 \newpage

### 2022-08-29 analysis/paper/paper.Rmd
fix scholarly metadata issue

@@
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      # pandoc_args:
-      # - --lua-filter=../templates/scholarly-metadata.lua
-      # - --lua-filter=../templates/author-info-blocks.lua
-      # - --lua-filter=../templates/pagebreak.lua
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib

### 2022-10-04 analysis/scripts/prepareData.R
update raw data because some roads were missing on the edge
Also updated R packages and fixed some things

@@ -141,24 +141,27 @@ tsa_8 <- bcdc_query_geodata("8daa29da-d7f4-401c-83ae-d962e3a28980") %>%
 # using .gpkg because writing to .shp abbreviates the column names
 write_sf(tsa_8, here(data_path_raw, "tsa8_boundaries.gpkg"))
 
+# use a buffered version of the tsa boundary to include some nearby roads
+tsa_8_buf <- st_buffer(tsa_8, 2000)
+
 dra_rds <- bcdc_query_geodata("bb060417-b6e6-4548-b837-f9060d94743e") %>%
   filter(INTERSECTS(tsa_8)) %>%
   collect() %>%
-  st_filter(tsa_8)
+  st_filter(tsa_8_buf)
 
 write_sf(dra_rds, here(data_path_raw, "dra_roads_ft_nelson.gpkg"))
 
 for_rds <- bcdc_query_geodata("243c94a1-f275-41dc-bc37-91d8a2b26e10") %>%
   filter(INTERSECTS(tsa_8)) %>%
   collect() %>%
-  st_filter(tsa_8)
+  st_filter(tsa_8_buf)
 
 write_sf(for_rds, here(data_path_raw, "forest_tenure_roads_ft_nelson.gpkg"))
 
 cutblocks <- bcdc_query_geodata("b1b647a6-f271-42e0-9cd0-89ec24bce9f7") %>%
   filter(INTERSECTS(tsa_8)) %>%
   collect() %>%
-  st_filter(tsa_8)
+  st_filter(tsa_8, .predicate = st_covered_by)
 
 write_sf(cutblocks, here(data_path_raw, "cutblocks_ft_nelson.gpkg"))
 

### 2022-10-04 analysis/scripts/runProjections.R

@@ -1,4 +1,4 @@
-## Running projections on full TSA using subsets (to do high resolution)
+## Running projections
 
 #set path for data
 data_path_raw <- "analysis/data/raw_data/"
@@ -26,7 +26,33 @@ devtools::load_all()
 low <- 0.000001
 high <- 0.00001
 
-# # For Revelstoke at fine resolution
+# For Revelstoke at fine resolution
+
+# First run prepInputs to create filtered inputs for QGIS plugin
+prepInputs(
+  cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
+  roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
+  tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
+  costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif"),
+  outPth = paste0(data_path_drvd, "TSA27/"),
+  aggFact = 1, #factor of aggregation of cost surface. 1 = no aggregation.
+  saveInputs = TRUE
+)
+
+# Code copied from QIS for running the plugin
+# processing.run("FRNC:Forest Road Network Creation",
+# {'INPUT_COST_RASTER':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_cost.tif',
+# 'INPUT_RASTER_BAND':1,
+# 'INPUT_POLYGONS_TO_ACCESS':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_cutblocks.gpkg',
+# 'INPUT_ROADS_TO_CONNECT_TO':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_roads.gpkg',
+# 'SKIDDING_DISTANCE':100,'METHOD_OF_GENERATION':1,'HEURISTIC_IN_POLYGONS':'',
+# 'ANGLES_CONSIDERED':1,'PUNISHER_45DEGREES':1.25,'PUNISHER_90DEGREES':2,
+# 'PUNISHER_135DEGREES':5,
+# 'OUTPUT':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/klementProjection.shp'})
+
+
+
+#
 run_projections(
[…]

### 2022-10-04 R/functionsSource.R
add a prepInputs function so it can be done separately as well. remove old getDistFromSource, use package remove assumption that tsb is a list

@@ -69,9 +69,7 @@ fineResProject <- function(TSAsubset, cutblockPoylgons, sampleDensity,
                            sampleType, costRaster, existingRoads,
                            projectionMethod) {
 
-  subCutblocks <- st_filter(cutblockPoylgons, TSAsubset)
-  existingRoads <- st_filter(existingRoads, TSAsubset)
-  subLandings <- getLandingsFromTarget(subCutblocks,
+  subLandings <- getLandingsFromTarget(cutblockPoylgons,
                                        landingDens = sampleDensity,
                                        sampleType = sampleType)
   if(nrow(subCutblocks) == 0){
@@ -89,87 +87,56 @@ fineResProject <- function(TSAsubset, cutblockPoylgons, sampleDensity,
 }
 
 
-
-# this function is to project using subset list and table of all different parameters
-projectParamTable<-function(tsbs,paramTable) {
-
-  for(i in 1:nrow(paramTable)) {
-
-    cRow = paramTable[i,]
-
-    projectionsList <- map(tsbs,
-                           fineResProject,
-                           cutblockPoylgons = cutblocks,
-                           sampleDensity = cRow$sampleDens,
-                           sampleType = cRow$sampleType,
-                           costRaster = tsaCost_st,
-                           existingRoads = roadsExist,
-                           projectionMethod = "mst")
-
-    projections <- do.call(rbind, projectionsList)
-
-
-    paramTable$output[[i]] <- projections
-
-  }
-
-  return(paramTable)
-
-}
-
-
-
-
 #Josie's distance function
-
-getDistKernelFromMax <- function(kdim) {
-  # kdim=2
-  kdim <- ceiling(kdim)
-  wDim <- kdim * 2 + 1
-  locSeq <- seq(-kdim, kdim)
-  y <- matrix(locSeq, wDim, wDim)
-  xx <- t(y)
-  d <- (xx^2 + y^2)^0.5
-  return(d)
-}
-
-uniformKernel <- function(dmax, cellDim = 1, useAveDist = F) {
-  if (useAveDist) {
-    # https://math.stackexchange.com/questions/3019165/average-distance-from-center-of-circle
-    dmax <- 3 * dmax / 2
-  }
-  hdim <- ceiling(dmax / cellDim)
-  weights <- getDistKernelFromMax(hdim)
-  weights <- weights <= dmax / cellDim
-  weights <- weights / sum(weights)
-  return(weights)
-}
-
[…]

### 2022-10-04 R/functionsSource.R
remove not in use funs

@@ -19,27 +19,6 @@ rasterizeLineDensity <- function(x, r) {
   return(terra::rast(spst_rast))
 }
 
-
-
-# Function to run projections at multiple landing densities
-
-
-testLandingDens_mst <- function(cutblockPolygons, landingDens, sampleType) {
-
-  variousLandingNum <- getLandingsFromTarget(cutblockPolygons, landingDens, sampleType)
-
-  resultProj_mst <- projectRoads(landings = variousLandingNum,
-                                 cost = tsaCost_st,
-                                 roads = roadsExist,
-                                 roadMethod = "mst")
-
-  message(landingDens)
-
-  return(resultProj_mst)
-
-}
-
-
 roadDisturbanceFootprint <- function(x, r, b) {
  # x = cRow$output[[1]]; r = tsaCost_st ; b = tsaBoundary
 
@@ -85,120 +64,6 @@ fineResProject <- function(TSAsubset, cutblockPoylgons, sampleDensity,
   return(subProjectionResults$roads)
 
 }
-
-
-#Josie's distance function
-# use the version in the roads package
-# getDistKernelFromMax <- function(kdim) {
-#   # kdim=2
-#   kdim <- ceiling(kdim)
-#   wDim <- kdim * 2 + 1
-#   locSeq <- seq(-kdim, kdim)
-#   y <- matrix(locSeq, wDim, wDim)
-#   xx <- t(y)
-#   d <- (xx^2 + y^2)^0.5
-#   return(d)
-# }
-#
-# uniformKernel <- function(dmax, cellDim = 1, useAveDist = F) {
-#   if (useAveDist) {
-#     # https://math.stackexchange.com/questions/3019165/average-distance-from-center-of-circle
-#     dmax <- 3 * dmax / 2
-#   }
-#   hdim <- ceiling(dmax / cellDim)
-#   weights <- getDistKernelFromMax(hdim)
-#   weights <- weights <= dmax / cellDim
-#   weights <- weights / sum(weights)
-#   return(weights)
-# }
-#
-# getDistFromSource <- function(src,maxDist,kwidth=3,dissag=F){
-#   if(dissag){
-#     mwidth=res(src)[1]
-#     src=disaggregate(src,fact=kwidth)
-#   }else{
-#     mwidth=res(src)[1]*kwidth
-#   }
-#   src[src>0]=1
[…]

### 2022-10-04 R/functionsSource.R
try to use existing sim object so don't have to recalulate graph

@@ -46,7 +46,7 @@ roadDisturbanceFootprint <- function(x, r, b) {
 
 fineResProject <- function(TSAsubset, cutblockPoylgons, sampleDensity,
                            sampleType, costRaster, existingRoads,
-                           projectionMethod) {
+                           projectionMethod, sim = NULL) {
 
   subLandings <- getLandingsFromTarget(cutblockPoylgons,
                                        landingDens = sampleDensity,
@@ -55,13 +55,27 @@ fineResProject <- function(TSAsubset, cutblockPoylgons, sampleDensity,
     return(NULL)
   }
 
-  subProjectionResults <- projectRoads(subLandings,
-                                       cost = costRaster,
-                                       roads = existingRoads,
-                                       roadMethod = projectionMethod)
+  if(!is.null(sim)){
+    # if sim list supplied use it so don't redo getGraph but change all the
+    # other parts to the current inputs
+    sim$costSurface <- costRaster
+
+    sim$roads <- existingRoads
+
+    sim$roadMethod <- projectionMethod
+
+    subProjectionResults <- projectRoads(subLandings, sim = sim)
+  } else{
+    subProjectionResults <- projectRoads(subLandings,
+                                         cost = costRaster,
+                                         roads = existingRoads,
+                                         roadMethod = projectionMethod)
+  }
+
+
   gc(verbose = TRUE)
 
-  return(subProjectionResults$roads)
+  return(subProjectionResults)
 
 }
 
@@ -75,15 +89,20 @@ projectAll <- function(tsbs,paramTable, costSurface,
     cRow = paramTable[i,]
     start <- Sys.time()
 
-    projections <- fineResProject(tsbs,
-                                  cutblockPoylgons = cutblocks,
-                                  sampleDensity = cRow$sampleDens,
-                                  sampleType = cRow$sampleType,
-                                  costRaster = costSurface,
-                                  existingRoads = existingRoads,
-                                  projectionMethod = method)
+    projectionsList <- vector("list", nrow(paramTable))
 
-    projections <- do.call(rbind, projectionsList)
+    # use sim object so graph not re-calced, will be NULL for first iter
+    projectionsList[[i]] <- fineResProject(tsbs,
+                                           cutblockPoylgons = cutblocks,
+                                           sampleDensity = cRow$sampleDens,
+                                           sampleType = cRow$sampleType,
+                                           costRaster = costSurface,
+                                           existingRoads = existingRoads,
[…]

### 2022-10-04 R/functionsSource.R
record runTime in metric mean table

@@ -198,6 +198,7 @@ getMetricMeans <- function(paramTable, cutblocks){
 
   metricsTable <- tibble(sampleType = paramTable$sampleType,
                          sampleDens = paramTable$sampleDens,
+                         runTime = paramTable$runTime,
                          areaMean = "overall",
                          roadDisturbanceMean = vector("numeric", nrow(paramTable)),
                          roadDensityMean = vector("numeric", nrow(paramTable)),
@@ -228,6 +229,7 @@ getMetricMeans <- function(paramTable, cutblocks){
 
   metricsTable1 <- tibble(sampleType = paramTable$sampleType,
                           sampleDens = paramTable$sampleDens,
+                          runTime = paramTable$runTime,
                           areaMean = "cutover",
                           roadDisturbanceMean = vector("numeric", nrow(paramTable)),
                           roadDensityMean = vector("numeric", nrow(paramTable)),

### 2022-10-04 analysis/paper/paper.Rmd
update raw data because some roads were missing on the edge
Also updated R packages and fixed some things
update projection method example figure

@@
 editor_options: 
-  chunk_output_type: console
+  chunk_output_type: inline
 ---
@@
   fig.path = "../figures/",
-  dpi = 300,
-  cache = TRUE
+  dpi = 300
 )
@@
 
-```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel."}
+```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel.", cache=TRUE}
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
@@
   st_as_sf() %>% 
-  filter(NAME_0 == "Canada") %>% 
+  filter(COUNTRY == "Canada") %>% 
   st_transform(st_crs(tsaBoundary)) %>% 
@@
 
-```{r fig2, fig.height=5, fig.width=fig_widths["two"], fig.cap="Example of the different projection methods and sampling strategies."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.jpg"))
+```{r, cache=TRUE}
+# Example area for demonstrating different methods
+
+library(raster)
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(tmap)
+library(purrr)
+library(caribouMetrics)
+library(pfocal)
+
+pal <- met.brewer(pal_nm, 8) %>% .[c(1,3,5,7)]
+
+# read in example area data
+exRoads <- read_sf(here(data_path_drvd, "testing_ex_roads.gpkg"))
+tsb <- read_sf(here(data_path_drvd, "testing_tsb.gpkg"))
+cutblocks <- read_sf(here(data_path_drvd, "testing_cutblocks.gpkg"))
+tsaCost <- raster(here(data_path_drvd, "testing_cost.tif"))
+roads <- read_sf(here(data_path_drvd, "testing_obs_roads.gpkg"))
+
+sampleDens <- c(low,high,low,high,low)
+sampleType <- c("regular","regular","random","random","centroid")
+paramTable <- tibble(sampleType, sampleDens,
+                     runTime = vector("list", length(sampleDens)),
+                     output = vector("list", length(sampleDens)),
+                     roadDisturbance = vector("list", length(sampleDens)),
+                     roadDensity = vector("list", length(sampleDens)),
+                     roadPresence = vector("list", length(sampleDens)),
+                     distanceToRoad = vector("list", length(sampleDens)),
+                     forestryDisturbance = vector("list", length(sampleDens))) %>%
+  distinct()
+
+# re-run projections for just the small area with different parameters
+mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
+                      costSurface = tsaCost,
+                      cutblocks = cutblocks,
+                      existingRoads = exRoads,
+                      fileLocation = paste0(data_path_drvd, "for_fig/mst"),
+                      method = "mst")
+
+dlcpProj <- projectAll(tsbs = tsb,
+                      paramTable = paramTable %>% filter(sampleType == "regular"),
+                      costSurface = tsaCost,
+                      cutblocks = cutblocks,
+                      existingRoads = exRoads,
+                      fileLocation = paste0(data_path_drvd, "for_fig/dlcp"),
+                      method = "dlcp")
+
+lcpProj <- projectAll(tsbs = tsb,
+                       paramTable = paramTable %>% filter(sampleType == "centroid"),
+                       costSurface = tsaCost,
+                       cutblocks = cutblocks,
+                       existingRoads = exRoads,
+                       fileLocation = paste0(data_path_drvd, "for_fig/lcp"),
+                       method = "lcp")
+
+allProj <- bind_rows(mstProj, dlcpProj, lcpProj, .id = "method") %>%
+  arrange(desc(sampleType), sampleDens)
+
+allMaps <- purrr::map(
+  1:nrow(allProj),
+  ~ qtm(tsaCost, raster.style = "cont", raster.palette = "Greys",
+        raster.alpha = 0.5)+
+    qtm(cutblocks, fill = NULL, borders = pal[1], borders.lwd = 1.5)+
+    qtm(roads, lines.col = pal[3], lines.lwd = 2)+
+    qtm(read_sf(here(allProj$output[[.x]])), lines.col = pal[2], lines.lwd = 2)+
+    qtm(exRoads, lines.col = "black", lines.lwd = 2)+
+    tm_layout(legend.show = FALSE,
+              main.title = paste(ifelse(allProj$method[.x]== 1, "MST",
+                                        ifelse(allProj$method[.x]== 2, "DLCP", "LCP")),
+                                 allProj$sampleType[.x],
+                                 ifelse(allProj$sampleType[.x] == "centroid", "",
+                                        allProj$sampleDens[.x])),
+              main.title.size = 0.75))
+
+tmap_save(tmap_arrange(allMaps, ncol = 2),
+          here("analysis/figures/projection_methods_figure.png"),
+          dpi = 300, height = 8, width = 3.543)
+```
+
+```{r fig2, fig.cap="Example of the different projection methods and sampling strategies.
+Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by green lines, observed roads by red lines, and the cost of road building by the grey shading."}
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
 ```

### 2022-10-04 R/functionsSource.R

@@ -44,14 +44,14 @@ roadDisturbanceFootprint <- function(x, r, b) {
 }
 
 
-fineResProject <- function(TSAsubset, cutblockPoylgons, sampleDensity,
+fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
                            sampleType, costRaster, existingRoads,
                            projectionMethod, sim = NULL) {
 
-  subLandings <- getLandingsFromTarget(cutblockPoylgons,
+  subLandings <- getLandingsFromTarget(cutblockPolygons,
                                        landingDens = sampleDensity,
                                        sampleType = sampleType)
-  if(nrow(subCutblocks) == 0){
+  if(nrow(cutblockPolygons) == 0){
     return(NULL)
   }
 
@@ -93,7 +93,7 @@ projectAll <- function(tsbs,paramTable, costSurface,
 
     # use sim object so graph not re-calced, will be NULL for first iter
     projectionsList[[i]] <- fineResProject(tsbs,
-                                           cutblockPoylgons = cutblocks,
+                                           cutblockPolygons = cutblocks,
                                            sampleDensity = cRow$sampleDens,
                                            sampleType = cRow$sampleType,
                                            costRaster = costSurface,
@@ -107,7 +107,7 @@ projectAll <- function(tsbs,paramTable, costSurface,
     end <- Sys.time()
 
     paramTable$output[[i]] <- paste0(fileLocation, cRow$sampleType, "_",
-                                     cRow$sampleDens, ".shp")
+                                     cRow$sampleDens, ".gpkg")
     paramTable$runTime[[i]] <- as.numeric(end - start)
 
     # save the projected roads to a file
@@ -455,7 +455,7 @@ run_projections <- function(cutblocksPth, roadsPth, tsaBoundaryPth, costPth,
   # recreate allResults after a restart using saved files
   # allResults <- paramTable %>%
   #   mutate(output = paste0(outPth, sampleType, "_",
-  #                          sampleDens, ".shp"))
+  #                          sampleDens, ".gpkg"))
 
   # Using David's saved results
   # allResults <- paramTable %>%

### 2022-10-04 analysis/scripts/method_example_fig.R

@@ -23,23 +23,23 @@ costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif")
 
 # densities
 high <- 0.00001
-low <-  0.000005
+low <-  0.000001
 
 
 cutblocks <- read_sf(cutblocksPth)
 roads <- read_sf(roadsPth)
 tsaCost <- rast(costPth)
 
-projRoads <- read_sf(paste0(data_path_drvd, "TSA27/", "TSA27_random_1e-05.shp"))
+projRoads <- read_sf(paste0(data_path_drvd, "TSA27/", "random_1e-05.shp"))
 
 exRoads <- roads %>% filter(AWARD_DATE <= as.Date("1990-01-01"))
 
 # find a good location
-qtm(st_as_sf(cutblocks), fill = NULL, borders.lwd = 2)+
-  qtm(st_as_sf(roads), lines.col = "red", lines.lwd = 2)+
-  qtm(st_as_sf(projRoads), lines.col = "green", lines.lwd = 2)+
-  qtm(st_as_sf(exRoads), lines.lwd = 2)+
-    tm_layout(legend.show = FALSE)
+# qtm(st_as_sf(cutblocks), fill = NULL, borders.lwd = 2)+
+#   qtm(st_as_sf(roads), lines.col = "red", lines.lwd = 2)+
+#   qtm(st_as_sf(projRoads), lines.col = "green", lines.lwd = 2)+
+#   qtm(st_as_sf(exRoads), lines.lwd = 2)+
+#     tm_layout(legend.show = FALSE)
 
 # # get extent
 # plot(roads)
@@ -70,49 +70,4 @@ write_sf(exRoads, paste0(data_path_drvd, "testing_ex_roads.gpkg"))
 write_sf(tsb[[1]], paste0(data_path_drvd, "testing_tsb.gpkg"))
 write_sf(cutblocks, paste0(data_path_drvd, "testing_cutblocks.gpkg"))
 writeRaster(tsaCost, paste0(data_path_drvd, "testing_cost.tif"))
-
-
-sampleDens <- c(low,high,low,high,low)
-sampleType <- c("regular","regular","random","random","centroid")
-paramTable <- tibble(sampleType, sampleDens,
-                     runTime = vector("list", length(sampleDens)),
-                     output = vector("list", length(sampleDens)),
-                     roadDisturbance = vector("list", length(sampleDens)),
-                     roadDensity = vector("list", length(sampleDens)),
-                     roadPresence = vector("list", length(sampleDens)),
[…]

### 2022-10-04 analysis/scripts/testRunProjections.R

@@ -1,0 +1,43 @@
+## Running projections on test area
+
+#set path for data
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
+
+###### Source scripts ######
+# load in required libraries
+library(raster)
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(tmap)
+library(purrr)
+library(caribouMetrics)
+library(pfocal)
+
+# load functions used in script
+devtools::load_all()
+
+#densities to test
+low <-  0.000001
+high <- 0.00001
+
+
+run_projections(
+  cutblocksPth = paste0(data_path_drvd, "testing_cutblocks.gpkg"),
+  roadsPth = paste0(data_path_drvd, "testing_ex_roads.gpkg"),
+  tsaBoundaryPth = paste0(data_path_drvd, "testing_tsb.gpkg"),
+  costPth = paste0(data_path_drvd, "testing_cost.tif"),
+  outPth = paste0(data_path_drvd, "test_tsb/"),
+
+  #Klement QGIS projection results shapefile
+  klementProj = NULL,
+
+  #set high and low sampling densities for projections
+  low  = low,
+  high = high,
+
+  aggFact = 1, #factor of aggregation of cost surface. 1 = no aggregation.
+  saveInputs = FALSE
+)

### 2022-10-05 R/functionsSource.R
remove redundant map/list in calcMetrics
keep method in output

@@ -145,20 +145,17 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
     print(i)
     cRow = paramTable[i,]
 
-    out <- list(sf::read_sf(cRow$output))
+    out <- sf::read_sf(cRow$output)
 
-    roadDisturbanceResults <- map(out,
-                                  roadDisturbanceFootprint,
-                                  r = costSurface,
-                                  b = boundary)
+    roadDisturbanceResults <- roadDisturbanceFootprint(out, r = costSurface,
+                                                       b = boundary)
 
-    paramTable$roadDisturbance[[i]] <- roadDisturbanceResults[[1]]
+    paramTable$roadDisturbance[[i]] <- roadDisturbanceResults
 
-    roadDensityResults <- map(out,
-                              rasterizeLineDensity,
-                              r = as(costSurface, "Raster"))
+    roadDensityResults <- rasterizeLineDensity(out,
+                                               r = as(costSurface, "Raster"))
 
-    paramTable$roadDensity[[i]] <- roadDensityResults[[1]]
+    paramTable$roadDensity[[i]] <- roadDensityResults
 
     roadPresenceResults <- paramTable$roadDensity[[i]]
 
@@ -199,6 +196,7 @@ getMetricMeans <- function(paramTable, cutblocks){
   metricsTable <- tibble(sampleType = paramTable$sampleType,
                          sampleDens = paramTable$sampleDens,
                          runTime = paramTable$runTime,
+                         method = paramTable$method,
                          areaMean = "overall",
                          roadDisturbanceMean = vector("numeric", nrow(paramTable)),
                          roadDensityMean = vector("numeric", nrow(paramTable)),
@@ -230,6 +228,7 @@ getMetricMeans <- function(paramTable, cutblocks){
   metricsTable1 <- tibble(sampleType = paramTable$sampleType,
                           sampleDens = paramTable$sampleDens,
                           runTime = paramTable$runTime,
+                          method = paramTable$method,
                           areaMean = "cutover",
                           roadDisturbanceMean = vector("numeric", nrow(paramTable)),
                           roadDensityMean = vector("numeric", nrow(paramTable)),
[…]

### 2022-10-05 analysis/paper/paper.Rmd
make figs show dlcp results too

@@
 # read in example area data
-exRoads <- read_sf(here(data_path_drvd, "testing_ex_roads.gpkg"))
-tsb <- read_sf(here(data_path_drvd, "testing_tsb.gpkg"))
-cutblocks <- read_sf(here(data_path_drvd, "testing_cutblocks.gpkg"))
-tsaCost <- raster(here(data_path_drvd, "testing_cost.tif"))
-roads <- read_sf(here(data_path_drvd, "testing_obs_roads.gpkg"))
+exRoads <- read_sf(here(data_path_drvd, "..", "testing_ex_roads.gpkg"))
+tsb <- read_sf(here(data_path_drvd, "..", "testing_tsb.gpkg"))
+cutblocks <- read_sf(here(data_path_drvd, "..", "testing_cutblocks.gpkg"))
+tsaCost <- raster(here(data_path_drvd, "..", "testing_cost.tif"))
+roads <- read_sf(here(data_path_drvd, "..", "testing_obs_roads.gpkg"))
 
@@
 paramTable <- tibble(sampleType, sampleDens,
+                     method = "mst",
                      runTime = vector("list", length(sampleDens)),
@@
                       existingRoads = exRoads,
-                      fileLocation = paste0(data_path_drvd, "for_fig/mst"),
-                      method = "mst")
+                      fileLocation = here(data_path_drvd, "..", "for_fig"))
 
 dlcpProj <- projectAll(tsbs = tsb,
-                      paramTable = paramTable %>% filter(sampleType == "regular"),
+                      paramTable = paramTable %>% 
+                        filter(sampleType == "regular") %>% 
+                        mutate(method = "dlcp"),
                       costSurface = tsaCost,
@@
                       existingRoads = exRoads,
-                      fileLocation = paste0(data_path_drvd, "for_fig/dlcp"),
-                      method = "dlcp")
+                      fileLocation = here(data_path_drvd, "..", "for_fig"))
 
 lcpProj <- projectAll(tsbs = tsb,
-                       paramTable = paramTable %>% filter(sampleType == "centroid"),
+                       paramTable = paramTable %>% 
+                        filter(sampleType == "centroid") %>% 
+                        mutate(method = "lcp"),
                        costSurface = tsaCost,
@@
                        existingRoads = exRoads,
-                       fileLocation = paste0(data_path_drvd, "for_fig/lcp"),
-                       method = "lcp")
+                       fileLocation = here(data_path_drvd, "..", "for_fig"))
 
@@
 For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA.
-The road density and road distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).
+The road density and distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).
 
@@
 # organize data for ggplot
-meanTable_long <- meanTable %>%
-  pivot_longer(-c(sampleType, sampleDens, areaMean),
+meanTable_long <- meanTable %>% select(-runTime) %>% 
+  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
                names_to = "metric", values_to = "response")%>%
@@
                              sampleType),
-         sampleDens = paste(sampleType, sampleDens) %>% 
-           factor(levels = c("NA centroid",
-                             "NA observed",
-                             "random low sample density",
-                             "regular low sample density",
-                             "random high sample density",
-                             "regular high sample density",
+         sampleDens = paste(method, sampleType, sampleDens) %>% 
+           factor(levels = c("mst NA centroid",
+                             "NA NA observed",
+                             "mst random low sample density",
+                             "mst regular low sample density",
+                             "mst random high sample density",
+                             "mst regular high sample density",
+                             "dlcp regular high sample density",
                              "NA klementQGIS"),
@@
                              "Regular\nhigh density",
+                             "DLCP Regular\nhigh density",
                              "Hardy QGIS")),
@@
   theme_classic()+
-  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:4, 6)],
+  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:6)],
                     name = "Sampling\nMethod", guide = "none")+
@@
   coord_flip()
-
-# Create bar plot This is the old version keep in case
-fig3 <- projected_values %>% group_by(areaMean) %>% nest() %>% ungroup() %>%
-  mutate(obs = observed_values %>% group_by(areaMean) %>% nest() %>% pull(data)) %>%
-  group_by(areaMean) %>%
-  mutate(data = map(data, ~mutate(.x, area = areaMean)),
-         plot = map2(data, obs,
-                    ~ ggplot(.x, aes(x = sampleDens, response, fill = sampleType))+
-                      geom_col(position = position_dodge2(preserve = "single",
-                                                         width = 0.75))+
-                      geom_hline(aes(yintercept = response),
-                                 .y, color = "black", linetype = "dashed")+
-                      facet_wrap(~metric, scales = "free_y", ncol = 1)+
-                      theme_classic()+
-                      scale_fill_manual(values = met.brewer(pal_nm, 4))+
-                      theme(text = element_text(size = 15), axis.title.x = element_blank(),
-                            axis.title.y = element_blank(),
-                            legend.position = "none")+
-                      ggtitle(str_to_title(areaMean)))) %>%
-  pull(plot) %>%
-  {ggarrange(plotlist = .)}
 
@@
                                        "Pre-existing roads")),
-         sampleDens = paste(sampleType, sampleDens) %>%
+         sampleDens = paste(method, sampleType, sampleDens) %>%
            str_replace("1e-06", "low density") %>%
@@
            str_replace("klementQGIS NA", "klementQGIS") %>%
-           factor(levels = c("centroid",
-                             "random low density",
-                             "regular low density",
-                             "random high density",
-                             "regular high density",
+           factor(levels = c("mst centroid",
+                             "mst random low density",
+                             "mst regular low density",
+                             "mst random high density",
+                             "mst regular high density",
+                             "dlcp regular high density",
                              "klementQGIS"),
@@
   mutate(perc = count/sum(count)*100)
-
-stackedBar <- matchData_sum %>%
-  ggplot(aes(x = sampleDens, y =perc, fill = agreement))+
-  geom_col(position = "stack", width = 0.75)+
-  theme_classic()+
-  theme(axis.text.x=element_text(color = "black", size=11, angle=50, vjust=1, hjust=1))+
-  scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
-                               "#999999"))+
-  theme(text = element_text(size = 15), axis.title.x = element_blank())+
-  theme(legend.position="bottom", legend.text = element_text(size = 10),
-        legend.title = element_blank(),
-        strip.background = element_blank())+
-  labs(y = "Percent of Landscape")+
-  facet_wrap(~metric)
-
-# waffle stacked bar
-# matchData_sum %>%
-#   #filter(metric == "Road presence") %>%
-#   ggplot(aes(values = perc*3,
-#              fill = factor(agreement, levels = levels(agreement))))+
-#   geom_waffle(size = 0.25, colour = "white", flip = TRUE, n_rows = 9)+
-#   facet_grid(metric~ sampleDens)+
-#   scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
-#                                "#999999"), name = "Agreement")+
-#   scale_x_discrete(name = NULL) +
-#   scale_y_continuous(labels = function(x) x * 9/3,
-#                      expand = c(0,0), name =  "Percent of Landscape") +
-#   coord_equal()+
-#   theme_minimal() +
-#   theme(text = element_text(size = 10),
-#         panel.grid = element_blank(),
-#         axis.ticks.y = element_line(),
-#         legend.position = "bottom") +
-#   guides(fill=guide_legend(nrow=2))
 
@@
   geom_col()+
-  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:4, 6)],
+  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:6)],
                     name = "Sampling\nMethod", guide = "none")+

### 2022-10-05 analysis/scripts/testRunProjections.R

@@ -45,7 +45,7 @@ paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>%
 run_projections(
   paramTable,
   cutblocksPth = here(data_path_drvd, "testing_cutblocks.gpkg"),
-  roadsPth = here(data_path_drvd, "testing_ex_roads.gpkg"),
+  roadsPth = here(data_path_drvd, "testing_obs_roads.gpkg"),
   tsaBoundaryPth = here(data_path_drvd, "testing_tsb.gpkg"),
   costPth = here(data_path_drvd, "testing_cost.tif"),
   outPth = here(data_path_drvd, "test_tsb/"),

### 2022-10-06 analysis/paper/paper.Rmd
re-run projections and updates to the paper. WIP waiting on Hardy results at 200m

@@
 
+devtools::load_all(here())
+
 #set paths
@@
 
-To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster.
+To spatially define the cost of constructing and traveling on a forestry road we developed a cost raster.
 We used several spatial raster layers to construct our cost surface.
-For all cells the base cost followed the Interior Appraisal Manual of BC where the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
-Separate costs for river and pipeline crossings were then added to the cost.
-Major water bodies were considered barriers and assigned a very high cost.
-Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned a cost of 0 (Table 1).
-The cost surface for the Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.
+Following the Interior Appraisal Manual of BC, the base the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
+Separate costs for river and pipeline crossings were then added and major water bodies were considered barriers and assigned a very high cost.
+Lastly, existing roads were assigned a cost of 0 (Table 1).
+The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.
 
@@
 Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
-To avoid situations where the same road was included in both data sets with slightly different coordinates we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set.
-The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990.
+To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set.
+The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990.
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
@@
 
+All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher2021].
+
 ## Road Projection Methods
@@
 harvest cutblocks) and the landscape connecting the two.
-There are several different methods available for solving this problem, including snapping, least cost paths, and minimum spanning trees.
+There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees.
 Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape.
@@
 The path to each cutblock is determined independently of the other cutblocks.
-The cost is represented by a cost raster described below.
+The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
+The sequence is detemined by a heuristic such as the distance to the existing road.
 Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock.
@@
 (See Appendix X for a more detailed explanation of these methods).
-In the remainder of this paper, we focus on the LCP and MST approaches given their assumptions are better aligned with how road networks are designed.
+In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed.
 
 In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>).
-This plugin uses a similar least cost path algorithm to the `roads` package but instead of building roads along the least cost path to each cutblock independently the Forest Road Network Plugin uses a heuristic to determine which cutblocks to connect first and then builds the roads sequentially.
-The default heuristic is to connect the closest cutblocks first.
-We supplied our cost surface, cutblocks and road network data and kept all the default settings within the QGIS plugin.
+This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks.
 
@@
 We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
-Note that the default skidding distance used by the Hardy QGIS plugin is 100 m which corresponds to 25 pts/km^2^ .
+We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid and regularly spaced and random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
+For the Hardy QGIS plugind supplied our cost surface, cutblocks, road network data and kept all the default settings except for the skidding distance which was set to 200m.
+Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^ .
 
@@
 
-library(raster)
 library(dplyr)
@@
 cutblocks <- read_sf(here(data_path_drvd, "..", "testing_cutblocks.gpkg"))
-tsaCost <- raster(here(data_path_drvd, "..", "testing_cost.tif"))
+tsaCost <- raster::raster(here(data_path_drvd, "..", "testing_cost.tif"))
 roads <- read_sf(here(data_path_drvd, "..", "testing_obs_roads.gpkg"))
+
+# densities
+high <- 0.00001
+low <-  0.000001
 
@@
 
-```{r fig3, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
-Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
+```{r fig3, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
 # organize data for ggplot
-meanTable_long <- meanTable %>% select(-runTime) %>% 
+meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
   pivot_longer(-c(sampleType, sampleDens, areaMean, method),
@@
                              "dlcp regular high sample density",
-                             "NA klementQGIS"),
+                             "NA NA klementQGIS"),
                   labels = c("Centroid", 
@@
                              "Regular\nhigh density",
-                             "DLCP Regular\nhigh density",
+                             "DLCP regular\nhigh density",
                              "Hardy QGIS")),
@@
                              "dlcp regular high density",
-                             "klementQGIS"),
+                             "NA klementQGIS"),
                   labels = c("Centroid", 
@@
 # organize data for ggplot
-meanTable_long <- meanTable %>%
-  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution),
+meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
+  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
                names_to = "metric", values_to = "response")%>%

### 2022-10-06 R/functionsSource.R

@@ -379,7 +379,7 @@ prepInputs <- function(cutblocksPth, roadsPth, tsaBoundaryPth, costPth,
   cutblocks <- filter(cutblocks, HARVEST_YEAR > cutblocksYear)
 
   ### prepare cost surface layer
-  tsaCost <- crop(bc_cost_surface, tsaBoundary)
+  tsaCost <- crop(bc_cost_surface, st_buffer(tsaBoundary, 2000))
 
   if(aggFact > 1){
     tsaCost <- terra::aggregate(tsaCost, fact = aggFact, fun = terra::mean)

### 2022-10-06 analysis/scripts/runProjections.R

@@ -39,16 +39,32 @@ prepInputs(
   saveInputs = TRUE
 )
 
-# Code copied from QIS for running the plugin
-# processing.run("FRNC:Forest Road Network Creation",
-# {'INPUT_COST_RASTER':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_cost.tif',
-# 'INPUT_RASTER_BAND':1,
-# 'INPUT_POLYGONS_TO_ACCESS':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_cutblocks.gpkg',
-# 'INPUT_ROADS_TO_CONNECT_TO':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_roads.gpkg',
-# 'SKIDDING_DISTANCE':100,'METHOD_OF_GENERATION':1,'HEURISTIC_IN_POLYGONS':'',
-# 'ANGLES_CONSIDERED':1,'PUNISHER_45DEGREES':1.25,'PUNISHER_90DEGREES':2,
-# 'PUNISHER_135DEGREES':5,
-# 'OUTPUT':'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/klementProjection.shp'})
+# Logs copied from QIS for running the plugin and unioning the results to the input roads
+# QGIS version: 3.12.0-București
+# QGIS code revision: cd141490ec
+# Qt version: 5.11.2
+# GDAL version: 3.0.4
+# GEOS version: 3.8.0-CAPI-1.13.1
+# PROJ version: Rel. 6.3.1, February 10th, 2020
+# Processing algorithm…
+# Algorithm 'Forest Road Network Creation' starting…
+# Input parameters:
+#   { 'ANGLES_CONSIDERED' : 1, 'HEURISTIC_IN_POLYGONS' : '',
+#     'INPUT_COST_RASTER' : 'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_cost.tif',
+#     'INPUT_POLYGONS_TO_ACCESS' : 'C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/derived_data/TSA27/input_cutblocks.gpkg|layername=input_cutblocks',
[…]

### 2022-10-13 analysis/paper/paper.Rmd
add cutblocks only and save metrics
Also update figs and paper

@@
 ```{r setup2}
-##### Metric table bar plots
 library(ggplot2)
@@
 
-In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>).
+In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method.
 This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks.
+
+```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks.
+The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP).
+Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density.
+Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the cost of road building by the grey shading."}
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
+```
 
@@
 
-We projected road development in the Revelstoke TSA from 1990 to the present using several different inputs and compared the projections to the observed present day road network in order to evaluate the accuracy of the projections and their sensitivity to different inputs.
-We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid and regularly spaced and random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
-For the Hardy QGIS plugind supplied our cost surface, cutblocks, road network data and kept all the default settings except for the skidding distance which was set to 200m.
-Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^ .
+We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods.
+We also explored the effects of the cost surface resolution and different methods of selecting landing points within cutblocks by repeating the MST projections with several variations of the input parameters.
+We compared two resolutions for the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
+We also compared methods for locating landing points within cutblocks by using the polygon centroid and regularly spaced and random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
+The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings.
+To allow comparisons, we also included a cutblocks only projection which did not project roads but assumed all disturbances were with in cutblocks.
+For the Hardy QGIS plugin we input the 100 m resolution cost raster, cutblocks, and road network data.
+We kept all the default settings except for the skidding distance which was set to 200m.
+Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
+This resulted in a total of 15 road projection variants to be compared (Table 2).
+
+Table 2: Road projection method variations 
+```{r}
+tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
+MST|Regular|1|100 m and 1000 m
+MST|Regular|10|100 m and 1000 m
+MST|Random|1|100 m and 1000 m
+MST|Random|10|100 m and 1000 m
+MST|Centroid|N/A|100 m and 1000 m
+DLCP|Regular|10|100 m 
+Hardy QGIS|N/A|~9|100 m
+Cutblocks only|N/A|N/A|100 m
+
+"
+df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df2) <- unname(as.list(df2[1,]))
+df2 <- df2[-1,] # remove first row
+row.names(df2)<-NULL
+knitr::kable(df2)
+```
 
@@
 library(pfocal)
-
-pal <- met.brewer(pal_nm, 8) %>% .[c(1,3,5,7)]
 
@@
         raster.alpha = 0.5)+
-    qtm(cutblocks, fill = NULL, borders = pal[1], borders.lwd = 1.5)+
-    qtm(roads, lines.col = pal[3], lines.lwd = 2)+
-    qtm(read_sf(here(allProj$output[[.x]])), lines.col = pal[2], lines.lwd = 2)+
+    qtm(cutblocks, fill = NULL, borders = "grey40", borders.lwd = 1.5)+
+    qtm(roads, lines.col = "red", lines.lwd = 2)+
+    qtm(read_sf(here(allProj$output[[.x]])), lines.col = "yellow", lines.lwd = 2)+
     qtm(exRoads, lines.col = "black", lines.lwd = 2)+
@@
                                  ifelse(allProj$sampleType[.x] == "centroid", "",
-                                        allProj$sampleDens[.x])),
+                                        ifelse(allProj$sampleDens[.x] == 1e-06, "low density", "high density"))),
               main.title.size = 0.75))
@@
 
-```{r fig2, fig.cap="Example of the different projection methods and sampling strategies.
-Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by green lines, observed roads by red lines, and the cost of road building by the grey shading."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
-```
+In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 3) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road.
+The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
+Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004].
+Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package.
+Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package.
+All road characteristics were stored in raster format.
 
-In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network: 1) Road presence, a simple measure of whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer applied to observed and projected roads [@ibisch2016a]; 3) Forestry disturbance footprint, a 500 m buffer applied to observed and projected roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road.
-The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
-Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004).
-Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package.
-Distance to nearest road was calculated using a fast moving window method in the `getDistFromSource` function of the roads package.
-Each metric was applied to the results from each projection method and the observed roads with the results were stored as rasters.
-
-The aspatial accuracy of the projections was assessed by calculating the mean values of the projection metric rasters and then comparing them with the means of the observed metric rasters.
+The aspatial accuracy of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values.
 This was done across the TSA and at the cutblock level, where metric rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.
@@
 First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to highlight discrepancies between the networks.
-The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS to investigate potential causes of the discrepancy.
-Secondly, a matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed.
-This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA that matched and didn't match.
-All pixels in the Revelstoke TSA for the projected networks fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed.
-Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized.
-We used the performance metrics sensitivity, precision and F-measure to compare the results for each projection method [@lee2021].
+The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS to investigate potential causes of the discrepancy.
+Secondly, the performance of each projection for the three binary characteristics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed.
+All pixels in the projected networks fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed.
+These categories were summarized for each projection and characteristic and used to calculate the performance metrics sensitivity, precision and F-measure [@lee2021].
+Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true.
+Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed.
+F-measure attempts to balance sensitivity and precision to give an overall score.
 
@@
 
-## Aspatial accuracy
+## Aspatial performance
 
-Projection accuracy varied widely across the 5 metrics and two scales compared (Figure \@ref(fig:fig3)).
-Across all projection methods road projections were less accurate within cutblocks than in the whole TSA.
-Metrics that used a buffer around the roads, forestry and road disturbance footprint, showed closer agreement between the observed and projected roads than the road density and road presence metrics at both scales.
-For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA.
+For each projection method the aspatial projection performance varied widely across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)).
+The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks while the other characteristics, presence, density and distance to nearest road, performed better across the whole landscape.
+Across the whole landscape the the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated.
 The road density and distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).
@@
 
-```{r fig3, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
-Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
+```{r fig3, fig.height=5, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+Values greater than 0 indicate the projection overestimated the characteristic and negative values indicate an underestimation.
+The cublocks only method is not defined for the road density, road presence or distance to road characteristics."}
 # organize data for ggplot
@@
                              "dlcp regular high sample density",
-                             "NA NA klementQGIS"),
+                             "NA NA klementQGIS",
+                             "NA NA NA"),
                   labels = c("Centroid", 
@@
                              "DLCP regular\nhigh density",
-                             "Hardy QGIS")),
+                             "Hardy QGIS",
+                             "Cutblocks only")),
          metric = factor(metric,
@@
   mutate(prop_dif = (response_proj-response_obs)/response_obs,
-         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall")))
+         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall"))) %>% 
+  mutate(prop_dif = ifelse(sampleDens == "Cutblocks only" & 
+                             metric %in% c("Road\ndensity", "Road\npresence",
+                                           "Distance\nto road"),
+                           NA_real_, prop_dif))
 
@@
   theme_classic()+
-  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:6)],
+  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
                     name = "Sampling\nMethod", guide = "none")+
@@
 
-## Spatial accuracy
+## Spatial performance
 
-Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks.
-In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a).
-Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data.
-There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b).
-There are also instances where areas were likely misidentified as cutblocks (Figure 4c).
-This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
+Using difference maps to highlight areas of poor projections we found several potential causes of discrepancy.
+In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)b).
+However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections often cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley.
+In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)b).
+There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)c).
 
-```{r fig4, fig.height=9, fig.width=fig_widths["two"], fig.cap="Difference maps used to investigate the performance of the projections.
-Image “a” shows an area of low discrepancy where the projections matched well.
-When zooming in on the discrepancy shown in image “b”, one can see from the satellite imagery that there are forest roads in the imagery that are missing from the observed data.
-Image “c” reveals one example of a cutblock that has most likely been mislabelled by remote sensing as it is located on the summit of a mountain."}
+```{r fig4, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network.
+The primary map shows the observed distance to the nearest road minus the projected distance and the insets show close-ups of the road networks with satellite imagery to give more context.
+a) Projected and observed road networks follow similar patterns although projected roads cross valley bottoms more frequently.
+b) Projected roads do not match observed switchback roads.
+c) An example of a misidentified cutblock.
+The satellite images are from Google (2022)."}
 
-knitr::include_graphics(here("analysis/figures/Figure4_diff_map.png"))
+knitr::include_graphics(here("analysis/figures/Figure4_diff_map_update.png"))
 
@@
 
-```{r fig5, fig.height=6, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method.
-F measure is the harmonic mean of precision and sensitivity, precision is the proportion of  predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
+```{r fig5, fig.height=7, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method.
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
 
@@
                              "dlcp regular high density",
-                             "NA klementQGIS"),
+                             "NA klementQGIS", 
+                             "NA cutOnly NA"),
                   labels = c("Centroid", 
@@
                              "DLCP regular\nhigh density",
-                             "Hardy QGIS")),
+                             "Hardy QGIS", 
+                             "Cutblocks only")),
          metric = factor(metric,
@@
   mutate(perf_meas = factor(perf_meas,
-                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
+                            labels = c("F-measure", "Precision", "Sensitivity"))) %>% 
   ggplot(aes(sampleDens, value, fill = sampleDens))+
   geom_col()+
-  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:6)],
+  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
                     name = "Sampling\nMethod", guide = "none")+
@@
 
-As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance footprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
+As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
 For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives.
@@
 
-```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters.
+```{r figA1, fig.height=10, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters.
 Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
@@
                              "dlcp regular high sample density",
-                             "NA klementQGIS"),
+                             "NA klementQGIS", 
+                             "NA NA NA"),
                   labels = c("Centroid", 
@@
                                       width = 0.75))+
-  facet_grid(areaMean~ metric)+
+  facet_grid(areaMean + metric ~1)+
   theme_classic()+
@@
 
-```{r ftNelson, fig.height=6, fig.width=fig_widths["two"], fig.cap="Road projections for the Fort Nelson Timber Supply Area.
-Black lines show projected and pre-existing roads.
-Red lines show observed roads that were not captured by the projections."}
-ftN_roads_proj <- read_sf(here("analysis/data/derived_data/TSA8_1000/centroid_1e-05.shp"))
-
+```{r ftNelson, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson.
+Cost based road projection methods are not expected to be useful in areas where roads are built for purposes other than accessing a target location at minimal cost.
+For example, in the Northeast corner of the Fort Nelson Timber Supply Area seismic lines show a different pattern than foresty roads."}
 ftN_roads_obs <- read_sf(here("analysis/data/derived_data/combined_ft_nelson_roads.gpkg"))
 
-qtm(ftN_roads_obs, lines.col = "red")+
-  qtm(ftN_roads_proj)
+# only show top right corner
+bb <- st_bbox(ftN_roads_obs)
+
+bb[["xmin"]] <- bb[["xmax"]] - (bb[["xmax"]] - bb[["xmin"]])/2
+bb[["ymin"]] <- bb[["ymax"]] - (bb[["ymax"]] - bb[["ymin"]])/2
+
+ftN_roads_obs_NE <- st_crop(ftN_roads_obs, bb)
+
+tm_shape(ftN_roads_obs_NE, is.master = TRUE)+
+  tm_lines()
 ```

### 2022-10-13 R/functionsSource.R

@@ -121,22 +121,28 @@ projectAll <- function(tsbs,paramTable, costSurface,
 }
 
 calcMetrics <- function(paramTable, klementProj, cutblocks,
-                        nonAggregatedCostSurface, observedRoads, boundary,
+                        nonAggregatedCostSurface, observedRoads, boundary, cutblocksPth,
                         costSurface){
-
-  observedRow <- c("observed", NA, NA, NA, NA, NA, NA, NA)
+  # row for observed
+  observedRow <- c("observed", NA, NA, NA, NA, NA, NA, NA, NA, NA)
 
   paramTable <- rbind(paramTable, observedRow)
 
   paramTable$output[[nrow(paramTable)]] <- observedRoads
 
+  # row for QGIS Plugin
   if(!is.null(klementProj)){
-    klementRow <- c("klementQGIS", NA, NA, NA, NA, NA, NA, NA)
+    klementRow <- c("klementQGIS", NA, NA, NA, NA, NA, NA, NA, NA, NA)
 
     paramTable <- rbind(paramTable, klementRow)
 
     paramTable$output[[nrow(paramTable)]] <- klementProj
   }
+
+  # row for if only consider roads in disturbance
+  paramTable <- rbind(paramTable, c("cutOnly", NA, NA, NA, NA, NA, NA, NA, NA, NA))
+
+  paramTable$output[[nrow(paramTable)]] <- cutblocksPth
 
   cutblocksRaster <- terra::rasterize(terra::vect(cutblocks), nonAggregatedCostSurface)
 
@@ -461,7 +467,19 @@ run_projections <- function(paramTable,cutblocksPth, roadsPth, tsaBoundaryPth, c
                             observedRoads = roadsPth,
                             klementProj = klementProj,
                             cutblocks = cutblocks,
+                            cutblocksPth = cutblocksPth,
                             costSurface = tsaCost_st)
+
+  # save the metrics for further study
+  select(allMetrics, -c(sampleType, sampleDens, method, runTime, output)) %>%
+    colnames() %>%
+    purrr::cross2(1:nrow(allMetrics)) %>%
+    purrr::walk(~terra::writeRaster(allMetrics[[.x[[1]]]][[.x[[2]]]],
+                                    file.path(outPth, paste0(.x[[1]], allMetrics$sampleType[.x[[2]]], "_",
+                                                             allMetrics$sampleDens[.x[[2]]], "_",
+                                                             allMetrics$method[.x[[2]]], ".tif")),
[…]

### 2022-10-13 analysis/scripts/runProjections.R

@@ -114,23 +114,6 @@ run_projections(
   aggFact = 10 #factor of aggregation of cost surface. 1 = no aggregation.
 )
 
-# For Fort Nelson at coarse resolution
-run_projections(
-  paramTable %>% filter(sampleType == "regular", sampleDens == "low",
-                        method = "mst"),
-  cutblocksPth = paste0(data_path_raw, "cutblocks_ft_nelson.gpkg"),
-  roadsPth = paste0(data_path_drvd, "combined_ft_nelson_roads.gpkg"),
-  tsaBoundaryPth = paste0(data_path_raw, "tsa8_boundaries.gpkg"),
-  costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif"),
-  outPth = paste0(data_path_drvd, "TSA8_1000/"),
-
-  #Klement QGIS projection results shapefile
-  klementProj = NULL,
-
-  aggFact = 10 #factor of aggregation of cost surface. 1 = no aggregation.
-)
-beepr::beep()
-
 # set rasterOptions back to previous value
 # terraOptions(memfrac = prevOpts$memfrac)
 print(Sys.time())

### 2022-10-13 analysis/scripts/testRunProjections.R

@@ -76,3 +76,4 @@ allMaps <- purrr::map(
 
 tmap_arrange(allMaps)
 
+meanTable <- read.csv(here(data_path_drvd, "test_tsb", "mean_table.csv"))

### 2022-10-13 analysis/scripts/reconstructRoadYear.R
add reconstruct road year
not using in the end

@@ -1,0 +1,71 @@
+# Reconstruct road building year from cublock harvest date by iterating over
+# cutblocks for each year
+
+#set path for data
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
+
+###### Source scripts ######
+# load in required libraries
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(tmap)
+library(here)
+
+# Use sequence of landings to assign sequence of roads built
+
+realRoads <- read_sf(here(data_path_drvd, "combined_revelstoke_roads.gpkg"))
+cost <- rast(here(data_path_raw, "cost_surface_bc_ha_proj.tif"))
+cutblocks <- read_sf(here(data_path_raw, "cutblocks_revelstoke.gpkg"))
+
+cost <- crop(cost, vect(realRoads))
+
+cost <- subst(cost, from = NA, to = 65000)
+
+# burn in realRoads to have cost of 0 so that projected roads will follow them.
+roadsRast <- rasterize(vect(realRoads), cost,
+                              background = 0) == 0
+
+# doesn't work if cost is 0 because areas with 0 cost are assumed to already be
+# roads so no new ones will be built
+cost <- cost * (roadsRast + 0.00001)
+
+# Use highways, arterial, collector, local etc as non-forestry existing roads
+roadsExist <- realRoads %>%
+  filter(ROAD_CLASS %in% c("local", "arterial", "collector", "highway", "ramp",
+                           "pedestrian", "runway", "yield", "driveway", "strata",
+                           "lane", "service", "ferry"))
+
+# set pre-existing roads to year 1989
+roadsExist$Year <- 1950
+
+landings <- getLandingsFromTarget(cutblocks, sampleType = "centroid")
+
+# initialize sim list with first landings set
+oldSim <- projectRoads(landings[landings$HARVEST_YEAR == min(landings$HARVEST_YEAR),],
+                                   cost,
+                                   roadsExist)
+
+oldSim$roads <- oldSim$roads %>%
+  mutate(Year = ifelse(is.na(Year), min(landings$HARVEST_YEAR), Year))
+
+harv_yrs <- sort(unique(landings$HARVEST_YEAR))
+
+# iterate over landings sets using the sim list from the previous run as input
+for (i in 2:length(harv_yrs)) {
+  print(i)
[…]

### 2022-10-14 analysis/paper/paper.Rmd
paper edits

@@
       fig_caption: yes
+      number_sections: no
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
@@
 The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population [@brown2010].
-The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats.
-The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue [@governmentofbritishcolumbia2021].
+The Revelstoke TSA is an example of a landscape where road projections could be used to inform management of critical wildlife habitats.
 
@@
 We used several spatial raster layers to construct our cost surface.
-Following the Interior Appraisal Manual of BC, the base the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
+Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
 Separate costs for river and pipeline crossings were then added and major water bodies were considered barriers and assigned a very high cost.
@@
 
-Table 1. Definitions of road building costs for building the cost surface.
+Table 1. Definitions of road building costs used to build the cost raster.
 
@@
 harvest cutblocks) and the landscape connecting the two.
-There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees.
+There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)).
 Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape.
@@
 The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
-The sequence is detemined by a heuristic such as the distance to the existing road.
+The sequence is determined by a heuristic such as the distance to the existing road.
 Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock.
@@
 In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed.
-
-In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method.
+In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method.
 This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks.
-
-```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks.
-The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP).
-Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density.
-Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the cost of road building by the grey shading."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
-```
-
-## Analysis
-
-We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods.
-We also explored the effects of the cost surface resolution and different methods of selecting landing points within cutblocks by repeating the MST projections with several variations of the input parameters.
-We compared two resolutions for the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points within cutblocks by using the polygon centroid and regularly spaced and random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
-The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings.
-To allow comparisons, we also included a cutblocks only projection which did not project roads but assumed all disturbances were with in cutblocks.
-For the Hardy QGIS plugin we input the 100 m resolution cost raster, cutblocks, and road network data.
-We kept all the default settings except for the skidding distance which was set to 200m.
-Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
-This resulted in a total of 15 road projection variants to be compared (Table 2).
-
-Table 2: Road projection method variations 
-```{r}
-tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
-MST|Regular|1|100 m and 1000 m
-MST|Regular|10|100 m and 1000 m
-MST|Random|1|100 m and 1000 m
-MST|Random|10|100 m and 1000 m
-MST|Centroid|N/A|100 m and 1000 m
-DLCP|Regular|10|100 m 
-Hardy QGIS|N/A|~9|100 m
-Cutblocks only|N/A|N/A|100 m
-
-"
-df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
-names(df2) <- unname(as.list(df2[1,]))
-df2 <- df2[-1,] # remove first row
-row.names(df2)<-NULL
-knitr::kable(df2)
-```
+Finally, in some cases the impact of forestry on wildlife is projected without considering the road development that will accompany it, we refer to this as the cutblocks only projection method and compare it to projections that do consider roads.
 
@@
 
-In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 3) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road.
-The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
+```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks.
+The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP).
+Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density.
+Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the cost of road building by the grey shading."}
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
+```
+
+## Analysis
+
+We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods.
+We also explored the effects of several variations of the input parameters by repeating the MST projections for each variant.
+We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
+We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced pointd or random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
+The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings.
+For the Hardy QGIS plugin we used the 100 m resolution cost raster.
+We kept all the default settings except for the skidding distance which was set to 200m.
+Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
+This resulted in a total of 15 road projection variants to be compared (Table 2).
+
+Table 2: Road projection method variations 
+```{r}
+tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
+MST|Regular|1|100 m and 1000 m
+MST|Regular|10|100 m and 1000 m
+MST|Random|1|100 m and 1000 m
+MST|Random|10|100 m and 1000 m
+MST|Centroid|N/A|100 m and 1000 m
+DLCP|Regular|10|100 m 
+Hardy QGIS|N/A|~9|100 m
+Cutblocks only|N/A|N/A|100 m
+
+"
+df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df2) <- unname(as.list(df2[1,]))
+df2 <- df2[-1,] # remove first row
+row.names(df2)<-NULL
+knitr::kable(df2)
+```
+
+In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road.
+4) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
 Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004].
@@
 Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package.
+Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable with out roads.
 All road characteristics were stored in raster format.
@@
 
-For each projection method the aspatial projection performance varied widely across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)).
-The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks while the other characteristics, presence, density and distance to nearest road, performed better across the whole landscape.
-Across the whole landscape the the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated.
-The road density and distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).
+For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)).
+The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarily or better across the whole landscape.
+Across the whole landscape the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated.
+With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).
 
@@
 The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results.
-Hardy's QGIS plugin performed well at the overall scale but tended to overestimate the amount of roads within cutblocks.
+Hardy's QGIS plugin was closest to the observed mean for the road presence and density characteristics but was among the furthest from the observed mean for the distance to the nearest road and road disturbance footprint characteristic.
+The cutblocks only method showed that not including road development underestimates anthropogenic disturbance.
 
@@
 
-```{r fig3, fig.height=5, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection method variations with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
 Values greater than 0 indicate the projection overestimated the characteristic and negative values indicate an underestimation.
@@
                              "NA NA NA"),
-                  labels = c("Centroid", 
+                  labels = c("MST centroid", 
                              "Observed",
-                             "Random\nlow density",
-                             "Regular\nlow density",
-                             "Random\nhigh density",
-                             "Regular\nhigh density",
-                             "DLCP regular\nhigh density",
+                             "MST random low density",
+                             "MST regular low density",
+                             "MST random high density",
+                             "MST regular high density",
+                             "DLCP regular high density",
                              "Hardy QGIS",
@@
 
+# save runTimes for later comparison
+dlcp_time <- meanTable %>% filter(method == "dlcp", areaMean == "overall") %>% 
+  pull(runTime) %>% round(2)
+
+mst_time <- meanTable %>% filter(method == "mst", areaMean == "overall",
+                                 sampleType == "regular", sampleDens == "high sample density") %>% 
+  pull(runTime) %>% round(2) %>% `*`(60)
+
 prop_dif %>% 
@@
 Using difference maps to highlight areas of poor projections we found several potential causes of discrepancy.
-In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)b).
-However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections often cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley.
-In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)b).
-There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)c).
+In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)a).
+However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley.
+There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)b).
+In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)c).
 
@@
 a) Projected and observed road networks follow similar patterns although projected roads cross valley bottoms more frequently.
-b) Projected roads do not match observed switchback roads.
-c) An example of a misidentified cutblock.
+b) An example of a misidentified cutblock.
+c) Projected roads do not match observed switchback roads.
 The satellite images are from Google (2022)."}
@@
 
-```{r fig5, fig.height=7, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method.
-F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
+```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 100 m resolution cost raster.
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
+Values closer to one indicate better performance."}
 
@@
                              "NA cutOnly NA"),
-                  labels = c("Centroid", 
-                             "Random\nlow density",
-                             "Regular\nlow density",
-                             "Random\nhigh density",
-                             "Regular\nhigh density",
-                             "DLCP regular\nhigh density",
+                  labels = c("MST centroid", 
+                             "MST random low density",
+                             "MST regular low density",
+                             "MST random high density",
+                             "MST regular high density",
+                             "DLCP regular high density",
                              "Hardy QGIS", 
@@
                                     "Forestry disturbance\nfootprint",
-                                    "Road disturbance footprint"))) %>%
+                                    "Road disturbance\nfootprint"))) %>%
   group_by(metric, sampleDens, agreement) %>%
@@
 
-As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
-For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives.
-For the forestry and road disturbance footprint metrics, the F-measure was similar across projection methods.
+As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
+The cutblocks only method performed worst for all the road network characteristics.
+For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives.
+For the forestry and road disturbance footprint metrics, the F-measure was similar across projection method variants, while sensitivity was higher at high sampling densities and precision was higher at low sampling densities.
+The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
+The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` for the MST method.
 
@@
 
-meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2)
+meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
+  filter((method != "dlcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))
 
-matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2)
+matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
+  filter(method != "dlcp", sampleType != "cutOnly")
 
@@
                              "NA klementQGIS"),
-                  labels = c("Centroid", 
-                             "Random\nlow density",
-                             "Regular\nlow density",
-                             "Random\nhigh density",
-                             "Regular\nhigh density",
-                             "DLCP regular\nhigh density",
+                  labels = c("MST centroid", 
+                             "Random low density",
+                             "Regular low density",
+                             "Random high density",
+                             "Regular high density",
+                             "DLCP regular high density",
                              "Hardy QGIS")),
@@
 
-```{r ftNelson, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson.
+```{r ftNelson, cache=TRUE, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson.
 Cost based road projection methods are not expected to be useful in areas where roads are built for purposes other than accessing a target location at minimal cost.

### 2022-10-14 analysis/paper/paper.Rmd
remove part about rivers since I don't think there cost was actually included.
Use cost with existing roads burned in for fig 1

@@
 #cost surface raster layer
-bc_cost_surface <- terra::rast(here(data_path_raw, "cost_surface_bc_ha.tif"))
-lakeValue <- 65000
+tsaCost <- terra::rast(here(data_path_drvd, "input_cost.tif"))
 
-tsaCost <- terra::crop(bc_cost_surface, tsaBoundary)
-tsaCost <- terra::subst(tsaCost, from = NA, to = lakeValue)
 tsaCost <- terra::mask(tsaCost, terra::vect(tsaBoundary))
@@
 
-To spatially define the cost of constructing and traveling on a forestry road we developed a cost raster.
-We used several spatial raster layers to construct our cost surface.
+Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
+To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set.
+The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990.
+Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
+Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
+The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b].
+Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
+
+To spatially define the cost of building and traveling on a forestry road we developed a simple cost raster based on the slope and major barriers.
 Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
-Separate costs for river and pipeline crossings were then added and major water bodies were considered barriers and assigned a very high cost.
+Major water bodies were considered barriers and assigned a very high cost.
 Lastly, existing roads were assigned a cost of 0 (Table 1).
@@
 ```{r}
-tbl <- "Cost Type|Road Attribute|Value ($ per km)|Source
-Barriers of road development|Major lakes|65000|Major water bodies of BC
-Impedance|Existing roads|0|Assumed
-Road Construction Cost Structure|Pipe crossing|1911|Interior Appraisal Manual
-|River crossing|2130|Interior Appraisal Manual
-|Base cost|16178 + 504*slope|Interior Appraisal Manual
+tbl <- "Cost Type|Value ($ per km)|Source
+Slope|16178 + 504*slope|Interior appraisal manual
+Major lakes|65000|Major water bodies of BC
+Existing roads|0|Digital road atlas and forest tenure roads for BC
 "
@@
 ```
-
-Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
-To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set.
-The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990.
-Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
-Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b].
-Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
 

### 2022-10-14 analysis/paper/paper.Rmd
paper edits

@@
 Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
-To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set.
+To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set.
 The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990.
@@
 To spatially define the cost of building and traveling on a forestry road we developed a simple cost raster based on the slope and major barriers.
-Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a].
+Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope (Table 1) [@governmentofbritishcolumbia2021a].
 Major water bodies were considered barriers and assigned a very high cost.
-Lastly, existing roads were assigned a cost of 0 (Table 1).
+Lastly, existing roads were assigned a cost of 0.
 The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.
 
-Table 1. Definitions of road building costs used to build the cost raster.
+Table 1. Definitions of road construction costs used to build the cost raster.
 
@@
 We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e.
-harvest cutblocks) and the landscape connecting the two.
+harvest cutblocks) and the cost of traversing the landscape connecting the two.
 There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)).
 Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape.
-The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two.
-The path to each cutblock is determined independently of the other cutblocks.
+The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the road building cost between the two.
+The least cost path to each cutblock is determined independently of the other cutblocks.
 The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
@@
 (See Appendix X for a more detailed explanation of these methods).
-In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed.
+In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed (Figure \@ref(fig:fig2)).
 In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method.
@@
 Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density.
-Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the cost of road building by the grey shading."}
+Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the 100 m resolution cost raster by the grey shading."}
 knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
@@
 We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced pointd or random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)).
+We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced points or random points.
+Regular and random  point densities of 1 and 10 pts/km^2^ were tested (Figure \@ref(fig:fig2)).
 The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings.
@@
 Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
-This resulted in a total of 15 road projection variants to be compared (Table 2).
+This resulted in a total of 13 road projection variants to be compared (Table 2).
 
@@
 In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road.
-4) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
+4) Road disturbance footprint, a 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks.
+The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
 Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004].
@@
 
-The aspatial accuracy of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values.
-This was done across the TSA and at the cutblock level, where metric rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.
+The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values.
+This was done across the TSA and at the cutblock level, where road network characteristic rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.
 
-The projections were assessed for their spatial accuracy in two different ways.
-First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to highlight discrepancies between the networks.
-The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS to investigate potential causes of the discrepancy.
+The projections were assessed for their spatial performance in two different ways.
+First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks.
+The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
 Secondly, the performance of each projection for the three binary characteristics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed.
-All pixels in the projected networks fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed.
+All pixels in the projected rasters fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed.
 These categories were summarized for each projection and characteristic and used to calculate the performance metrics sensitivity, precision and F-measure [@lee2021].
@@
 For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)).
-The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarily or better across the whole landscape.
-Across the whole landscape the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated.
-With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).
+The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarly or better across the whole landscape.
+Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated.
+With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).
 

### 2022-10-14 R/functionsSource.R
include existing roads in cutblocks only

@@ -121,7 +121,8 @@ projectAll <- function(tsbs,paramTable, costSurface,
 }
 
 calcMetrics <- function(paramTable, klementProj, cutblocks,
-                        nonAggregatedCostSurface, observedRoads, boundary, cutblocksPth,
+                        nonAggregatedCostSurface, observedRoads, boundary,
+                        cutblocksPth, existingRoads,
                         costSurface){
   # row for observed
   observedRow <- c("observed", NA, NA, NA, NA, NA, NA, NA, NA, NA)
@@ -142,7 +143,8 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
   # row for if only consider roads in disturbance
   paramTable <- rbind(paramTable, c("cutOnly", NA, NA, NA, NA, NA, NA, NA, NA, NA))
 
-  paramTable$output[[nrow(paramTable)]] <- cutblocksPth
+  # combine with existing roads
+  cutOnly <- st_union(cutblocks, existingRoads)
 
   cutblocksRaster <- terra::rasterize(terra::vect(cutblocks), nonAggregatedCostSurface)
 
@@ -151,26 +153,28 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
     print(i)
     cRow = paramTable[i,]
 
-    out <- sf::read_sf(cRow$output)
+    if(cRow$sampleType == "cutOnly"){
+      out <- cutOnly
+    } else {
+      out <- sf::read_sf(cRow$output)
+    }
 
-    roadDisturbanceResults <- roadDisturbanceFootprint(out, r = costSurface,
-                                                       b = boundary)
+    if(cRow$sampleType != "cutOnly"){
+      roadDensityResults <- rasterizeLineDensity(out, r = as(costSurface, "Raster"))
+      paramTable$roadDensity[[i]] <- roadDensityResults
 
-    paramTable$roadDisturbance[[i]] <- roadDisturbanceResults
+      roadPresenceResults <- paramTable$roadDensity[[i]]
+      roadPresenceResults <- terra::classify(
+        roadPresenceResults,
+        rcl = matrix(c(-1, 0, 0, 0, Inf, 1), byrow = TRUE, ncol = 3)
+      )
+      paramTable$roadPresence[[i]] <- roadPresenceResults
 
-    roadDensityResults <- rasterizeLineDensity(out,
[…]

### 2022-10-14 analysis/paper/paper.Rmd
more edits

@@
 Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated.
-With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).
+With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).
 
-When comparing between projection methods we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods.
+When comparing between projection method variants we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods.
 The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results.
@@
 
-As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`).
-The cutblocks only method performed worst for all the road network characteristics.
+As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`).
+The cutblocks only method performed worst for all the road network characteristics (Figure \@ref(fig:fig5)).
 For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives.
@@
 The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
-The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` for the MST method.
+The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
 
@@
 
+**NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion **
+
+## Sarah's thoughts
+
+- The projections work reasonably well for strategic planning purposes, especially if we are not concerned with roads inside cutblocks.
+
+- The differences between algorithms were fairly small and it would be more valuable to focus on the accuracy of input data such as costs and cutblock locations rather than fine tuning the algorithms.
+
+- This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
+The results would likely be less accurate if the landings were spread further apart.
+
+## What David had
 Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters.
@@
 Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue.
-Another potential issue was with roads not being included in the data set that were seen in the satellite imagery.
-This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets.
+*Another potential issue was with roads not being included in the data set that were seen in the satellite imagery.* (this is no longer identified in the results section since it wasn't a major discrepancy after fixing the road data source) This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets.
 Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc).
@@
 Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines).
-Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc.
+Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc (\@ref(fig:ftNelson)).
 Because of this, forestry roads make up a much smaller percentage of the overall road network.
 In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors.
-The model was attempted in the Fort Nelson TSA however the resulting road network (Figure A-1) did not accurately portray the observed network like it did in the Revelstoke TSA.
 However, other resource extraction activities could likely be incorporated into this model relatively easily.
@@
 
-## Road network projection methods
+## Road network projection method details
 
@@
 Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network.
-We used three methods for solving the MTAP in the roads package.
-These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST).
-In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
+We used four methods for solving the MTAP in the roads package.
+These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP), dynamic least cost paths (DLCP) approach and finally, the minimum spanning tree approach (MST).
+In this paper, we focus on the LCP, DLCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
 
-LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
+The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
 For each STAP, a least cost path from a single source to sink was solved using Dijkstra's shortest path algorithm (1959).
 Dijkstra's algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section "weighted graph").
-Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction.
+Witht the LCP approach each STAP is solved independently but for the DLCP approach each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction.
 This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP.
@@
 
-A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods.
-All the default settings such as skidding distance, etc.
-were left as they were within the QGIS plugin.
-
 ## Weighted graph
@@
 The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies.
-[The edges connecting the nodes follow a "queens" case or allow a maximum of eight possible edges for a given node.]{.underline} Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only.
+The edges connecting the nodes follow an "octagonal" neighbourhood or allow a maximum of eight possible edges for a given node, with diagonal edges having a higher cost.
+Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design.
+Therefore, our road projection methods are for strategic modelling purposes only.
 Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs.
@@
 
-```{r figA2, fig.height=6, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method.
+```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters.
 F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

### 2022-10-14 analysis/scripts/testRunProjections.R

@@ -77,3 +77,4 @@ allMaps <- purrr::map(
 tmap_arrange(allMaps)
 
 meanTable <- read.csv(here(data_path_drvd, "test_tsb", "mean_table.csv"))
+

### 2022-12-07 analysis/paper/paper.Rmd
work on introduction in progress

@@
 ---
-title: "Assessment of forestry road network projection methods using the “roads” package in R"
+title: "Assessment and implementation of forestry road network projection methods"
 author:
@@
     correspondence: false
+  - name: David Lapins
+    email: david.lapins@pc.gc.ca
+    institute: [ECCC]
+    correspondence: false
+  - name: Kyle Lochhead
+    email: Kyle.Lochhead@gov.bc.ca
+    institute: [FLNR]
+    correspondence: false
+  - name: Gregory Paradis
+    email: gregory.paradis@ubc.ca
+    institute: [UBC]
+    correspondence: false
+
 institute:
   - ECCC: Environment and Climate Change Canada 
+  - FLNR: British Columbia Ministry of Forests 
+  - UBC: University of British Columbia 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       - --lua-filter=../templates/pagebreak.lua
-bibliography: references.bib
+bibliography: Roads.bib
 csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
@@
 <!-- With the following code you can access and display values from the yml header above. -->
+<!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->
 
@@
 
-Global biodiversity is being lost at an unprecedented rate as a result of human activities.
-One of the major world-wide threats to biodiversity is increasing infrastructural development (Benítez-López et al., 2010).
-Land transformation has resulted in the loss and fragmentation of habitats in many different ecosystem types worldwide (Benítez-López et al., 2010; Sanderson et al., 2002).
-Roads and other linear infrastructure, in particular, are key components of landscapes as they have become a major cause of habitat loss, fragmentation, and degradation and are ubiquitous in most landscapes around the world (van der Ree et al., 2011).
-Although linear infrastructure is important for society because it provides connectivity for people, it also exerts significant negative effects on adjacent habitats, wildlife populations, communities, and ecosystems (van der Ree et al., 2011).
+The impacts of roads on wildlife are complex and varied.
+Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_a_2021;@trombulak_review_2000;@fahrig_effects_2009].
+In Canadian forests, there is ample evidence of linear features attracting predators such as black bears and wolves seeking food or easy movement, while repelling prey such as caribou and moose [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_the_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009].
+Roads can also facilitate spread of invasive species [], and enable other development activities [@johnson_growth-inducing_2020].
 
-Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area.
-In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015).
-Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015).
-Resource extraction activities provide jobs and renewable raw materials (Canada, 2013b), but they also can create threats for wildlife.
-For example, the main threats to woodland caribou, a species at risk in Canada, are human disturbances such as forest harvesting, oil and gas exploration/extraction, and road networks.
-These particular disturbances fragment their habitat, and create open areas and extensive young forests that attract other species such as moose and deer which in turn, attracts an increased number of predators, such as wolves and even poachers (Canada, 2013a).
-Woodland caribou are often considered to be an umbrella species, which means that protecting and restoring their habitat would benefit other species also dependent on old forests, such as fishers, pine martens, and many species of birds (Mulvihill, 2018).
+Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [bodner, dietze, perfict].
+Recent efforts to forecast impacts of climate change and anthropogenic disturbance on Canadian forests and forest-dwelling wildlife [@tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_;@michelletti_;recent quebec examples] build on a rich history of spatial-stochastic models of forest growth and succession following disturbance [landis;overview;stms;seles].
+Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales are not available.
 
-With certain wildlife, such as woodland caribou, being highly sensitive to road development and use, there is a need to spatially simulate future roads in order to test the implications of alternative recovery policies and resource development scenarios.
-Landscape models are mathematical simulations designed to represent at least one landscape pattern-process relationship of interest.
-Landscape models can be great ways to make predictions about specific scenarios that allow ecologists and land managers to assess alternative management scenarios or to assess the potential impact of anthropogenic change scenarios (climate or land use change) (McGarigal, n.d.).
-Simulating the effects of disturbances, such as forestry road development, helps to broaden the scope of models, especially for species with complex ecological requirements who reside in disturbance-prone landscapes that undergo dynamic changes which may substantially alter habitat quality (Whitman et al., 2017).
-Land managers are often responsible for multiple species and populations distributed over large regions and often rely on models to determine habitat suitability (Whitman et al., 2017).
+Forest roads are also essential for harvest planning.
+Some forest management decision support systems include forest road projection methods [patchworks], but methods implemented in proprietary closed-source commercial software cannot be used in other contexts.
+Published methods intended to support strategic planning [@Anderson_projecting_2004;Clark et al., 2000; Dean, 1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales.
+More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Clement [2021], and a least cost path minimum spanning tree algorithm used to project roads and reconstruct development history in British Columbia [lochhead published and not].
+The former method is available as QGIS and Landis plugins, and the latter is not easily extricated from a larger open source modelling project that is specific to British Columbia [].
+Neither method has not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].
 
-Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011).
-Stochastic simulation approaches use probability distributions in combination with random number generators to determine events (He & Mladenoff, 1999).
-Despite the importance of roads on the environment, previous spatial stochastic harvest simulation models rarely include forest road network development in their projections.
-The "roads" package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface.
-This paper uses the "roads" package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.
+Comment on metrics relevant to wildlife and what exactly accuracy means.
 
-(Introduction round 2: road context) Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use.
-These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network.
-These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes.
-On the other hand, many spatial stochastic forest management simulations haven't included forest road development at all (citations).
-One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
-There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes.
-As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) ("Bridging the divide between ecological forecasts and environmental decision-making").
+In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package.
+We compare the ability of these methods to reconstruct observed forest road network development in a mountainous region of British Columbia using performance metrics relevant to wildlife.
 
-The goal of our model is to provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations, and can also be applied as an element of more complex models.
+#One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
+
+#Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations.
 Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes.
@@
 
-In the first section of the paper (methods), we introduce the study area used to assess the accuracy of the projection, we describe the forest road projection methods, the data required, and the various metrics used in order to analyze the similarities and differences between the projected and observed forest road networks across the landscape.
-Next, we present and discuss the results of our projections across the landscape and compare the metrics spatially and aspatially.
+#value of roadless areas: [@ibisch_a_2016; @talty_conservation_2020; @psaralexi_importance_2017]
+
+#forest mgmt planning models: [@depellegrinllorente_recognizing_2017]
+
+#other of interest: caribou habitat metric [@whitman_a_2017]
+
+#Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area.
+In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015).
+Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015).
+
+#Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011).
 
@@
 
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1).
-This allows us to project road building based on forest harvest data.
+#TO DO: dynamic links to figs and tables.
+
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1).
+This allows us to project road building based on historic forest harvest data.
 With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) [@brown2010], the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.

### 2023-01-09 analysis/paper/paper.Rmd
Writing in progress...

@@
 Recent efforts to forecast impacts of climate change and anthropogenic disturbance on Canadian forests and forest-dwelling wildlife [@tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_;@michelletti_;recent quebec examples] build on a rich history of spatial-stochastic models of forest growth and succession following disturbance [landis;overview;stms;seles].
-Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales are not available.
+Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales have not been available.
 
@@
 More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Clement [2021], and a least cost path minimum spanning tree algorithm used to project roads and reconstruct development history in British Columbia [lochhead published and not].
-The former method is available as QGIS and Landis plugins, and the latter is not easily extricated from a larger open source modelling project that is specific to British Columbia [].
-Neither method has not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].
+The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [].
+Neither method has been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].
 
-Comment on metrics relevant to wildlife and what exactly accuracy means.
+<!-- Comment on metrics relevant to wildlife and what exactly accuracy means. -->
+In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package.
+We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife.
 
-In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package.
-We compare the ability of these methods to reconstruct observed forest road network development in a mountainous region of British Columbia using performance metrics relevant to wildlife.
+<!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
+-->
 
-#One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
-
-#Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations.
+<!-- Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations.
 Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes.
@@
 In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially.
+-->
 
-#value of roadless areas: [@ibisch_a_2016; @talty_conservation_2020; @psaralexi_importance_2017]
+<!-- value of roadless areas: [@ibisch_a_2016; @talty_conservation_2020; @psaralexi_importance_2017] -->
 
-#forest mgmt planning models: [@depellegrinllorente_recognizing_2017]
+<!-- forest mgmt planning models: [@depellegrinllorente_recognizing_2017] -->
 
-#other of interest: caribou habitat metric [@whitman_a_2017]
+<!-- other of interest: caribou habitat metric [@whitman_a_2017] -->
 
-#Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area.
+<!-- Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area.
 In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015).
 Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015).
+-->
 
-#Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011).
+<!-- Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011). -->
 
@@
 
-## Study Area
+## Study Area and Data
 
-#TO DO: dynamic links to figs and tables.
-
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1).
-This allows us to project road building based on historic forest harvest data.
-With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) [@brown2010], the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development.
-The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests].
-The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population [@brown2010].
-The Revelstoke TSA is an example of a landscape where road projections could be used to inform management of critical wildlife habitats.
+<!-- TO DO: dynamic links to figs and tables. -->
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1).
+This allows projection of the forest road network from historic forest harvest data.
+Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g.
+alpine, lakes, swamp, brush, rock) [@brown2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers.
+Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests].
+The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).
 
@@
 
-## Data
-
 Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
 To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set.
-The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990.
+The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990.
+All roads from the DRA data set were assumed to be built before 1990.
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
@@
 
-To spatially define the cost of building and traveling on a forestry road we developed a simple cost raster based on the slope and major barriers.
-Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope (Table 1) [@governmentofbritishcolumbia2021a].
-Major water bodies were considered barriers and assigned a very high cost.
-Lastly, existing roads were assigned a cost of 0.
-The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.
+To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1).
+Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@governmentofbritishcolumbia2021a], and that road construction across major water bodies is prohibitively expensive (Table 1).
+Existing roads are assigned a cost of 0.
+The cost raster consists of 1703936 pixels with a resolution of 100 meters by 100 meters.
 
-Table 1. Definitions of road construction costs used to build the cost raster.
-
+Table 1. A simple model of road construction costs.
 ```{r}
@@
 
-## Road Projection Methods
+<!-- TO DO: add links to repositories and explain what data is and is not there, and why -->
 
-We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e.
-harvest cutblocks) and the cost of traversing the landscape connecting the two.
-There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)).
-Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape.
-The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the road building cost between the two.
-The least cost path to each cutblock is determined independently of the other cutblocks.
-The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
-The sequence is determined by a heuristic such as the distance to the existing road.
-Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock.
-Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956].
-(See Appendix X for a more detailed explanation of these methods).
-In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed (Figure \@ref(fig:fig2)).
-In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method.
-This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks.
-Finally, in some cases the impact of forestry on wildlife is projected without considering the road development that will accompany it, we refer to this as the cutblocks only projection method and compare it to projections that do consider roads.
+## Road Projection Methods and Variants
+
+Simple methods for projecting low-cost routes from an existing road network to target locations (i.e.
+harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)).
+We use Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the least cost path between a target location (e.g.
+centroid of a cutblock) and the existing road network.
+Least cost paths to each cutblock is determined independently in simple method (LCP).
+The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
+Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples.
+The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal1956].
+See Appendix A for a more detailed explanation of these methods.
+We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)).
+We compared these methods to an alternative iterative least cost path method developed by Hardy [ref].
+
+<!-- TO DO: cite key R packages used in method implementation -->
+
+The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
+We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:fig2)).
+We used default settings in the Hardy QGIS plugin except for kidding distance which was set to 200m.
+A skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
+We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
+In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table 2).
+
+Table 2: Road projection method variants. 
+```{r}
+tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
+MST|Regular|1|100 m and 1000 m
+MST|Regular|10|100 m and 1000 m
+MST|Random|1|100 m and 1000 m
+MST|Random|10|100 m and 1000 m
+MST|Centroid|N/A|100 m and 1000 m
+DLCP|Regular|10|100 m 
+Hardy QGIS|N/A|~9|100 m
+Cutblocks only|N/A|N/A|100 m
+
+"
+df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df2) <- unname(as.list(df2[1,]))
+df2 <- df2[-1,] # remove first row
+row.names(df2)<-NULL
+knitr::kable(df2)
+```
+
+To make help make simple road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.
+
+<!-- TO DO: add links to R package documentation & resources -->
 
@@
 
-```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks.
-The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP).
-Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density.
-Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the 100 m resolution cost raster by the grey shading."}
+```{r fig2, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms.
+Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2).
+Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red.
+Cost (100 m resolution) is indicated by grey shading."}
 knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
@@
 
-## Analysis
+## Performance Metrics
 
-We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods.
-We also explored the effects of several variations of the input parameters by repeating the MST projections for each variant.
-We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
-We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced points or random points.
-Regular and random  point densities of 1 and 10 pts/km^2^ were tested (Figure \@ref(fig:fig2)).
-The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings.
-For the Hardy QGIS plugin we used the 100 m resolution cost raster.
-We kept all the default settings except for the skidding distance which was set to 200m.
-Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
-This resulted in a total of 13 road projection variants to be compared (Table 2).
-
-Table 2: Road projection method variations 
-```{r}
-tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
-MST|Regular|1|100 m and 1000 m
-MST|Regular|10|100 m and 1000 m
-MST|Random|1|100 m and 1000 m
-MST|Random|10|100 m and 1000 m
-MST|Centroid|N/A|100 m and 1000 m
-DLCP|Regular|10|100 m 
-Hardy QGIS|N/A|~9|100 m
-Cutblocks only|N/A|N/A|100 m
-
-"
-df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
-names(df2) <- unname(as.list(df2[1,]))
-df2 <- df2[-1,] # remove first row
-row.names(df2)<-NULL
-knitr::kable(df2)
-```
-
-In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road.
-4) Road disturbance footprint, a 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks.
-The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes].
-Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004].
-Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package.
-Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package.
-Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable with out roads.
+In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch2016a]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
+The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley2004; more].
+Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes].
+Distance to nearest road was calculated using a fast parallel moving window method implemented in the getDistFromSource function of the `roads` package.
+Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads.
 All road characteristics were stored in raster format.
 
+<!-- TO DO: update the details on getDistFromSource function - where is it now? -->
+<!-- TO DO: Does the cutblocks only scenario include the old roads?
+If no, it should.
+And if so, it seems like the other metrics could be calculated in the cutblocks only scenario.
+It would be interesting to have the "no projection" baseline for all the metrics -->
+
 The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values.
-This was done across the TSA and at the cutblock level, where road network characteristic rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.
-
-The projections were assessed for their spatial performance in two different ways.
+This was done across the TSA and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks.
+Spatial performance was assessed in two different ways.
 First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks.
 The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
-Secondly, the performance of each projection for the three binary characteristics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed.
-All pixels in the projected rasters fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed.
-These categories were summarized for each projection and characteristic and used to calculate the performance metrics sensitivity, precision and F-measure [@lee2021].
+Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed.
+These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee2021].
 Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true.
@@
 
-For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)).
-The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarly or better across the whole landscape.
-Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated.
+For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:fig3)).
+All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks.
+The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads.
+All methods underproject road density, in part because within 100 m pixels all the methods generate straight road segments.
+All the methods also underproject road presence, and the problem is more severe for the low density method variants.
+In general, performance differences among low and high density method variants are larger than differences among projection algorithms.
 With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).
 
-When comparing between projection method variants we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods.
-The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results.
-Hardy's QGIS plugin was closest to the observed mean for the road presence and density characteristics but was among the furthest from the observed mean for the distance to the nearest road and road disturbance footprint characteristic.
-The cutblocks only method showed that not including road development underestimates anthropogenic disturbance.
+<!-- How is the 'cutblocks only' scenario overestimating the overall road disturbance footprint? That doesn't make sense. -->
+
+<!-- Discussion pts
+- disturbance footprints and distance to road are easier to project than road density and road presence.
+- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. 
+-->
 
@@
 
-```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection method variations with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
-Values greater than 0 indicate the projection overestimated the characteristic and negative values indicate an underestimation.
-The cublocks only method is not defined for the road density, road presence or distance to road characteristics."}
+```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
+The cutblocks only method is not defined for the road density, road presence or distance to road characteristics."}
 # organize data for ggplot
@@
 
-Using difference maps to highlight areas of poor projections we found several potential causes of discrepancy.
+Difference maps highlight several potential causes of discrepancy between our projections and the observed road network.
 In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)a).
-However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley.
+However, the cost raster is not sufficiently details to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys.
 There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)b).
 In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)c).
+All these discrepancies could be addressed with more accurate harvest and cost information.
 
@@
 
-As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`).
-The cutblocks only method performed worst for all the road network characteristics (Figure \@ref(fig:fig5)).
-For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives.
-For the forestry and road disturbance footprint metrics, the F-measure was similar across projection method variants, while sensitivity was higher at high sampling densities and precision was higher at low sampling densities.
+As expected, none of the projection methods can accurately project road presence (i.e.
+the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives.
+All projection methods yielded more accurate results than the 'cutblock only' scenario, indicating that any of these projection methods are better than none (Figure \@ref(fig:fig5)).
 The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
@@
 
-**NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion **
+<!-- NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion -->
 
@@
 
-## What David had
+<!--  What David had
 Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters.
@@
 The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.
+-->
 

### 2023-01-10 analysis/paper/paper.Rmd
writing in progress

@@
 ---
-title: "Assessment and implementation of forestry road network projection methods"
+title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects on wildlife"
 author:
@@
 The impacts of roads on wildlife are complex and varied.
-Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_a_2021;@trombulak_review_2000;@fahrig_effects_2009].
+Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].
 In Canadian forests, there is ample evidence of linear features attracting predators such as black bears and wolves seeking food or easy movement, while repelling prey such as caribou and moose [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022].
-Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_the_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009].
-Roads can also facilitate spread of invasive species [], and enable other development activities [@johnson_growth-inducing_2020].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009].
+Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
 
-Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [bodner, dietze, perfict].
-Recent efforts to forecast impacts of climate change and anthropogenic disturbance on Canadian forests and forest-dwelling wildlife [@tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_;@michelletti_;recent quebec examples] build on a rich history of spatial-stochastic models of forest growth and succession following disturbance [landis;overview;stms;seles].
-Despite the importance of roads for wildlife, none of these efforts have included forecasts of the development of road networks associated with forest harvest, largely because tools to project road network development over large spatial scales have not been available.
+Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022].
+In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_climate-informed_2023;@micheletti_assessing_2021;@leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
+Despite the importance of roads for wildlife, these models generally do not include road network projections.
+For example, in a recent collection of articles focussed on “Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests” [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].
 
-Forest roads are also essential for harvest planning.
-Some forest management decision support systems include forest road projection methods [patchworks], but methods implemented in proprietary closed-source commercial software cannot be used in other contexts.
-Published methods intended to support strategic planning [@Anderson_projecting_2004;Clark et al., 2000; Dean, 1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales.
-More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Clement [2021], and a least cost path minimum spanning tree algorithm used to project roads and reconstruct development history in British Columbia [lochhead published and not].
-The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [].
-Neither method has been formally assessed, compared, published, or implemented in a way that allows for easy integration into the variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [spades/syncrosim/landis].
+Leston et al.
+(2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire.
+This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool.
+However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
+Published methods intended to support strategic planning [@anderson_projecting_2004;@clark_three-stage_2000;@dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales.
+More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
+The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
+None of these methods have been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+@barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].
 
-<!-- Comment on metrics relevant to wildlife and what exactly accuracy means. -->
 In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package.
 We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife.
+The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose.
+Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016], and density [].
+We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here.
+Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.
+
+<!--
+See notes here:
+https://docs.google.com/document/d/1UQ_iC69QX0qX86NM6zSSQ036CgslN-uCXCdBY-4nU5M/edit
+Effects of roads on wildlife include increased mortality from various types of encounters with humans, alterations in behaviour and interactions among predators and prey, and spread of invasive species (Trombulak and Frissel 2001).
+In boreal forests and coniferous forests of the Columbia and Rocky mountains, notable impacts of logging roads include facilitation of movement by wolves (Mumma et al.
+2018; Dickie et al.
+2020), attraction of bears to roadside food resources (Roever et al.
+2010; Bastille-Rousseau et al.
+2011; Dickie et al.
+2020), increased mortality caused by humans near roads and trails (Roever et al.
+2010), and avoidance by ungulates (Laurian et al.
+2015; Dickie et al.
+2020).
+Forest dwelling boreal woodland and mountain caribou avoid roads leading to a loss of available habitat (Polfus et al.
+2011; Dickie et al.
+2020), and suffer increased mortality near roads (Bastille-Rousseau et al.
+2011; Mumma et al.
+2018).
+At a range scale, survival and recruitment of boreal caribou declines with anthropogenic disturbance (Johnson et al.
+2020).
+Eight herds of Southern Mountain Caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC  and listed as threatened under the Canadian Species at Risk Act (SARA) (Palm et al.
+2019).
+Most boreal caribou populations with ranges that include intensive forest management activity are not self-sustaining (ECCC 2011; Johnson et al.
+2020), and boreal caribou are also listed as threatened under SARA.
+
+Nonlinear responses of elk to road density (Frair et al.
+2008).
+Black bear responses to roads vary with road type (Zeller et al.
+2020), and selection for roadside vegetation can increase neonatal boreal woodland caribou mortality in areas near roads .
+For more careful reading: Flyal et al.
+2019; Jaeger et al.
+2005; Laurance et al.
+2014; Psaralexi et al.
+2017; Apps et al 2014; Beyer et al.
+2014; Blagdon and Johnson 2021; Suzuki and Parker 2016; Fahrig and Rytwinski 2009; Coffin et al.
+2021; Crist et al.
+2005; Grantham et al.
+2020;Johnson et al.
+2019; Meijer et al.
+2018; Langeveld et al.
+2009
+
+Continue reviewing pdfs in downloads -
+
+-->
 
@@
 Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g.
-alpine, lakes, swamp, brush, rock) [@brown2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers.
-Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests].
+alpine, lakes, swamp, brush, rock) [@brown_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers.
+Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021].
 The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).
@@
 
-Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC.
+Our road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC.
 To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set.
@@
 All roads from the DRA data set were assumed to be built before 1990.
-Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b].
+Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021].
 Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b].
+The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021].
 Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
@@
 To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1).
-Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@governmentofbritishcolumbia2021a], and that road construction across major water bodies is prohibitively expensive (Table 1).
+Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table 1).
 Existing roads are assigned a cost of 0.
@@
 
-All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher2021].
+All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher_bcdata_2021].
 
@@
 harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)).
-We use Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the least cost path between a target location (e.g.
+We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g.
 centroid of a cutblock) and the existing road network.
@@
 Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples.
-The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal1956].
+The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956].
 See Appendix A for a more detailed explanation of these methods.
@@
 ## Performance Metrics
-
-In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch2016a]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
-The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley2004; more].
+In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
+The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; more].
 Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes].
@@
 Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed.
-These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee2021].
+These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
 Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true.
@@
 <!-- How is the 'cutblocks only' scenario overestimating the overall road disturbance footprint? That doesn't make sense. -->
-
-<!-- Discussion pts
-- disturbance footprints and distance to road are easier to project than road density and road presence.
-- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. 
--->
 
@@
 
-<!-- NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion -->
+<!-- Discussion pts
+- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
+- as expected, not great at projecting exact road locations.
+sources of discrepancy include switchback, stream crossings, inaccurate harvest locations.
+All of these could be improved with more detailed high resolution cost & harvest data.
+- density of landing pts has a larger effect than choice of road projection algorithm.
+So maybe don't get too worked up about the differences between one algorithm and another.
+if more accurate projections are required, focus first on refining cost & harvest landing locations.
+- Selected revelstoke because it is topographically complex, so a good challenge.
+However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
+The results would likely be less accurate if the landings were spread further apart.
+But also less topographically complex landscapes may be simpler to project.
+- For some purposes, projections from relatively crude cost models may be adequate.
+For this example landscape projections of disturbance footprints and distance to road are pretty good (and better than nothing).
+Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels.
+If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,).
+If interested in predictors that we did not consider, see example script.
+Can evaluate alternative metrics without rerunning the projections.
+-->
 
-## Sarah's thoughts
-
-- The projections work reasonably well for strategic planning purposes, especially if we are not concerned with roads inside cutblocks.
-
-- The differences between algorithms were fairly small and it would be more valuable to focus on the accuracy of input data such as costs and cutblock locations rather than fine tuning the algorithms.
-
-- This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
-The results would likely be less accurate if the landings were spread further apart.
+<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice.
+They don't need to run the projections to do this - just summarize the outputs using a method of their choice.
+-->
 

### 2023-01-17 R/functionsSource.R
use bind instead of union for cutblocks

@@ -144,7 +144,7 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
   paramTable <- rbind(paramTable, c("cutOnly", NA, NA, NA, NA, NA, NA, NA, NA, NA))
 
   # combine with existing roads
-  cutOnly <- st_union(cutblocks, existingRoads)
+  cutOnly <- bind_rows(cutblocks, existingRoads)
 
   cutblocksRaster <- terra::rasterize(terra::vect(cutblocks), nonAggregatedCostSurface)
 

### 2023-01-17 analysis/paper/paper.Rmd

@@
 ---
-title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects on wildlife"
+title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife"
 author:
@@
 Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].
-In Canadian forests, there is ample evidence of linear features attracting predators such as black bears and wolves seeking food or easy movement, while repelling prey such as caribou and moose [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022].
-Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter movement patterns [@tucker_moving_2018], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009].
+In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus, Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022; @proctor_effects_2020].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020].
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
@@
 The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
-None of these methods have been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+These methods have not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].
@@
 The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose.
-Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016], and density [].
+Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here.
 Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.
+
+<!-- TO DO:
+- more thorough review of SMC & grizzly papers
+- summarize road metric types used in cited papers more systematically.
+-->
 
@@
 Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g.
-alpine, lakes, swamp, brush, rock) [@brown_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers.
+alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers.
 Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021].
-The area contains wildlife species that are sensitive to roads (refs) including endangered southern mountain woodland caribou (refs).
+The area is home to wildlife species that are sensitive to roads including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010].
+Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020].
+Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022].
+Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].
 
@@
 Existing roads are assigned a cost of 0.
-The cost raster consists of 1703936 pixels with a resolution of 100 meters by 100 meters.
+The cost raster consists of 1703936 pixels with a resolution of 1 ha.
 
@@
 
-All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher_bcdata_2021].
+All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].
 
@@
 We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)).
-We compared these methods to an alternative iterative least cost path method developed by Hardy [ref].
+We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].
 
@@
 A skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
-We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution.
+We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution.
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table 2).
@@
 tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
-MST|Regular|1|100 m and 1000 m
-MST|Regular|10|100 m and 1000 m
-MST|Random|1|100 m and 1000 m
-MST|Random|10|100 m and 1000 m
-MST|Centroid|N/A|100 m and 1000 m
-DLCP|Regular|10|100 m 
-Hardy QGIS|N/A|~9|100 m
-Cutblocks only|N/A|N/A|100 m
+MST|Regular|1|1 ha and 100 ha
+MST|Regular|10|1 ha and 100 ha
+MST|Random|1|1 ha and 100 ha
+MST|Random|10|1 ha and 100 ha
+MST|Centroid|N/A|1 ha and 100 ha
+DLCP|Regular|10|1 ha 
+Hardy QGIS|N/A|~9|1 ha
+Cutblocks only|N/A|N/A|1 ha
 
@@
 
-To make help make simple road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.
+To make help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.
 
@@
 Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red.
-Cost (100 m resolution) is indicated by grey shading."}
+Cost (1 ha resolution) is indicated by grey shading."}
 knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
@@
 In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
-The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; more].
-Forestry disturbance footprint and road density were calculated using the XX and rasterizeLineDensity functions in the `caribouMetrics` R package [@hughes].
-Distance to nearest road was calculated using a fast parallel moving window method implemented in the getDistFromSource function of the `roads` package.
+The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more].
+Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
+Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
 Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads.
@@
 
-<!-- TO DO: update the details on getDistFromSource function - where is it now? -->
+<!-- TO DO: Discuss why we do not calculate the 50m buffer metric used by Kyle for SMC.
+50 m buffer is basically built in to the 100 m raster resolution.
+Road presence would be our closest equivalent measure - and we don't project that measure very well.
+-->
+
+<!-- TO DO: update the details on getDistFromSource function - where is it, or where should it be? -->
 <!-- TO DO: Does the cutblocks only scenario include the old roads?
@@
 The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values.
-This was done across the TSA and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks.
+This was done across the timber supply area and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks.
 Spatial performance was assessed in two different ways.
@@
 The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads.
-All methods underproject road density, in part because within 100 m pixels all the methods generate straight road segments.
+All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments.
 All the methods also underproject road presence, and the problem is more severe for the low density method variants.
@@
 
+<!-- TO DO: consider breaking down aspatial results into 'cutblocks', 'outside cutblocks' and 'overall': it may make the results easier to explain and interpret. -->
+
 ```{r get-data}
@@
 
-```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”).
+```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”).
 Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
@@
 
-```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 100 m resolution cost raster.
+```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
@@
 sources of discrepancy include switchback, stream crossings, inaccurate harvest locations.
-All of these could be improved with more detailed high resolution cost & harvest data.
+Missing switchbacks and too many stream crossings also reduce our total amount of roads.
+All of these could be improved with more detailed high resolution cost & harvest and landing data.
 - density of landing pts has a larger effect than choice of road projection algorithm.
@@
 However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
-The results would likely be less accurate if the landings were spread further apart.
+The results may be less accurate if the landings were spread further apart.
 But also less topographically complex landscapes may be simpler to project.
 - For some purposes, projections from relatively crude cost models may be adequate.
-For this example landscape projections of disturbance footprints and distance to road are pretty good (and better than nothing).
+For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).
 Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels.
@@
 A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms.
-When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution).
+When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (1 ha resolution).
 Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network.
@@
 
+<!-- TO DO: switch fig resolution labels from metres to ha. -->
 ```{r get-data2}
@@
 
-```{r figA1, fig.height=10, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters.
+```{r figA1, fig.height=10, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
 Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
@@
 
-```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters.
+```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
 F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

### 2023-01-17 R/functionsSource.R
change cutOnly to only use roads for road disturbance

@@ -143,9 +143,6 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
   # row for if only consider roads in disturbance
   paramTable <- rbind(paramTable, c("cutOnly", NA, NA, NA, NA, NA, NA, NA, NA, NA))
 
-  # combine with existing roads
-  cutOnly <- bind_rows(cutblocks, existingRoads)
-
   cutblocksRaster <- terra::rasterize(terra::vect(cutblocks), nonAggregatedCostSurface)
 
   for(i in 1:nrow(paramTable)) {
@@ -154,27 +151,25 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
     cRow = paramTable[i,]
 
     if(cRow$sampleType == "cutOnly"){
-      out <- cutOnly
+      out <- existingRoads
     } else {
       out <- sf::read_sf(cRow$output)
     }
 
-    if(cRow$sampleType != "cutOnly"){
-      roadDensityResults <- rasterizeLineDensity(out, r = as(costSurface, "Raster"))
-      paramTable$roadDensity[[i]] <- roadDensityResults
+    roadDensityResults <- rasterizeLineDensity(out, r = as(costSurface, "Raster"))
+    paramTable$roadDensity[[i]] <- roadDensityResults
 
-      roadPresenceResults <- paramTable$roadDensity[[i]]
-      roadPresenceResults <- terra::classify(
-        roadPresenceResults,
-        rcl = matrix(c(-1, 0, 0, 0, Inf, 1), byrow = TRUE, ncol = 3)
-      )
-      paramTable$roadPresence[[i]] <- roadPresenceResults
+    roadPresenceResults <- paramTable$roadDensity[[i]]
+    roadPresenceResults <- terra::classify(
+      roadPresenceResults,
+      rcl = matrix(c(-1, 0, 0, 0, Inf, 1), byrow = TRUE, ncol = 3)
+    )
+    paramTable$roadPresence[[i]] <- roadPresenceResults
 
-      src <- terra::trim(roadPresenceResults[[1]]) # thinks this is a list
-      maxDist=10000 #in units of res
-      distanceToRoadResults <- getDistFromSource(src, maxDist, kwidth = 3, method = "pfocal")
-      paramTable$distanceToRoad[[i]] <- distanceToRoadResults
-    }
+    src <- terra::trim(roadPresenceResults[[1]]) # thinks this is a list
+    maxDist=10000 #in units of res
[…]

### 2023-01-17 analysis/scripts/testRunProjections.R

@@ -78,3 +78,11 @@ tmap_arrange(allMaps)
 
 meanTable <- read.csv(here(data_path_drvd, "test_tsb", "mean_table.csv"))
 
+cutOnly_rdDist <- terra::rast(here(data_path_drvd, "test_tsb", "roadDisturbancecutOnly_NA_NA.tif"))
+obs_rdDist <- terra::rast(here(data_path_drvd, "test_tsb", "roadDisturbanceobserved_NA_NA.tif"))
+
+qtm(cutOnly_rdDist)+
+  qtm(read_sf(here(data_path_drvd, "testing_cutblocks.gpkg")),
+      fill = NULL, borders = pal[1], borders.lwd = 1.5)+
+  qtm(read_sf(here(data_path_drvd, "testing_ex_roads.gpkg")),
+      fill = NULL, lines.col = "black")

### 2023-01-17 analysis/paper/paper.Rmd
minor edits, meaningful fig chunk labels and dynamic links

@@
 Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].
-In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus, Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022; @proctor_effects_2020].
+In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022; @proctor_effects_2020].
 Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020].
@@
 
-<!-- TO DO: dynamic links to figs and tables. -->
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure 1).
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure \@ref(fig:study-area)).
 This allows projection of the forest road network from historic forest harvest data.
@@
 
-```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel.", cache=TRUE}
+```{r study-area, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel.", cache=TRUE}
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
@@
 To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1).
-Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table 1).
+Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)).
 Existing roads are assigned a cost of 0.
@@
 
-Table 1. A simple model of road construction costs.
-```{r}
+```{r cost-tbl}
 tbl <- "Cost Type|Value ($ per km)|Source
@@
 row.names(df)<-NULL
-knitr::kable(df)
+knitr::kable(df, caption = "A simple model of road construction costs.")
 ```
@@
 Simple methods for projecting low-cost routes from an existing road network to target locations (i.e.
-harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) \@ref(fig:fig2)).
+harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)).
 We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g.
@@
 See Appendix A for a more detailed explanation of these methods.
-We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:fig2)).
+We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)).
 We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].
@@
 The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
-We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:fig2)).
+We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)).
 We used default settings in the Hardy QGIS plugin except for kidding distance which was set to 200m.
@@
 We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution.
-In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table 2).
-
-Table 2: Road projection method variants. 
-```{r}
+In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
+
+```{r method-tbl}
 tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
@@
 row.names(df2)<-NULL
-knitr::kable(df2)
+knitr::kable(df2, caption = "Road projection method variants.")
 ```
@@
 
-```{r fig2, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms.
+```{r proj-meth-fig, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms.
 Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2).
@@
 
-For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:fig3)).
+For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:aspat-perf-fig)).
 All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks.
@@
 
-```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”).
+```{r aspat-perf-fig, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”).
 Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
@@
 Difference maps highlight several potential causes of discrepancy between our projections and the observed road network.
-In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)a).
+In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:proj-map-fig)a).
 However, the cost raster is not sufficiently details to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys.
-There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)b).
-In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)c).
+There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:proj-map-fig)b).
+In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:proj-map-fig)c).
 All these discrepancies could be addressed with more accurate harvest and cost information.
 
-```{r fig4, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network.
+```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network.
 The primary map shows the observed distance to the nearest road minus the projected distance and the insets show close-ups of the road networks with satellite imagery to give more context.
@@
 
-```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
@@
 All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives.
-All projection methods yielded more accurate results than the 'cutblock only' scenario, indicating that any of these projection methods are better than none (Figure \@ref(fig:fig5)).
+All projection methods yielded more accurate results than the 'cutblock only' scenario, indicating that any of these projection methods are better than none (Figure \@ref(fig:spat-perf-fig)).
 The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.

### 2023-01-18 analysis/paper/paper.Rmd
some more fixs for cutOnly

@@
   mutate(prop_dif = (response_proj-response_obs)/response_obs,
-         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall"))) %>% 
-  mutate(prop_dif = ifelse(sampleDens == "Cutblocks only" & 
-                             metric %in% c("Road\ndensity", "Road\npresence",
-                                           "Distance\nto road"),
-                           NA_real_, prop_dif))
+         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall")))
 
@@
   pivot_wider(c(metric, sampleDens), names_from = agreement, 
-              values_from = count) %>% 
+              values_from = count, values_fill = 0) %>% 
   mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
@@
 perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
-  summarise(rng_F = range(F_measure) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
+  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
 

### 2023-01-18 R/functionsSource.R

@@ -179,12 +179,8 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
     )
     paramTable$forestryDisturbance[[i]] <- terra::rast(forestryDisturanceResults@processedData$Anthro)
 
-
     roadDisturbanceResults <- roadDisturbanceFootprint(out, r = costSurface, b = boundary)
     paramTable$roadDisturbance[[i]] <- roadDisturbanceResults
-
-
-
   }
 
   return(paramTable)
@@ -217,17 +213,14 @@ getMetricMeans <- function(paramTable, cutblocks){
     forestryDisturbanceMean <- terra::global(cRow$forestryDisturbance[[1]], "mean", na.rm = TRUE)
     metricsTable$forestryDisturbanceMean[i] <- forestryDisturbanceMean
 
-    if(cRow$sampleType != "cutOnly"){
-      roadDensityMean <- terra::global(cRow$roadDensity[[1]], "mean", na.rm = TRUE)
-      metricsTable$roadDensityMean[i] <- roadDensityMean
+    roadDensityMean <- terra::global(cRow$roadDensity[[1]], "mean", na.rm = TRUE)
+    metricsTable$roadDensityMean[i] <- roadDensityMean
 
-      roadPresenceMean <- terra::global(cRow$roadPresence[[1]], "mean", na.rm = TRUE)
-      metricsTable$roadPresenceMean[i] <- roadPresenceMean
+    roadPresenceMean <- terra::global(cRow$roadPresence[[1]], "mean", na.rm = TRUE)
+    metricsTable$roadPresenceMean[i] <- roadPresenceMean
 
-      distanceToRoadMean <- terra::global(cRow$distanceToRoad[[1]], "mean", na.rm = TRUE)
-      metricsTable$distanceToRoadMean[i] <- distanceToRoadMean
-    }
-
+    distanceToRoadMean <- terra::global(cRow$distanceToRoad[[1]], "mean", na.rm = TRUE)
+    metricsTable$distanceToRoadMean[i] <- distanceToRoadMean
 
   }
 
@@ -256,20 +249,18 @@ getMetricMeans <- function(paramTable, cutblocks){
     croppedforestryDisturbanceMean <- terra::global(croppedforestryDisturbance, "mean", na.rm = TRUE)
     metricsTable1$forestryDisturbanceMean[i] <- croppedforestryDisturbanceMean
 
-    if(cRow$sampleType != "cutOnly"){
[…]

### 2023-01-19 R/functionsSource.R
Add metric means for outside cutblocks and simplify that code. Also add option to load saved results from file in run_projections

@@ -159,11 +159,9 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
     roadDensityResults <- rasterizeLineDensity(out, r = as(costSurface, "Raster"))
     paramTable$roadDensity[[i]] <- roadDensityResults
 
-    roadPresenceResults <- paramTable$roadDensity[[i]]
-    roadPresenceResults <- terra::classify(
-      roadPresenceResults,
-      rcl = matrix(c(-1, 0, 0, 0, Inf, 1), byrow = TRUE, ncol = 3)
-    )
+    roadPresenceResults <- terra::rasterize(terra::vect(out), costSurface,
+                                            background = 0)
+
     paramTable$roadPresence[[i]] <- roadPresenceResults
 
     src <- terra::trim(roadPresenceResults[[1]]) # thinks this is a list
@@ -192,79 +190,61 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
 
 getMetricMeans <- function(paramTable, cutblocks){
 
+  # Overall
   metricsTable <- tibble(sampleType = paramTable$sampleType,
                          sampleDens = paramTable$sampleDens,
                          runTime = paramTable$runTime,
                          method = paramTable$method,
-                         areaMean = "overall",
-                         roadDisturbanceMean = vector("numeric", nrow(paramTable)),
-                         roadDensityMean = vector("numeric", nrow(paramTable)),
-                         roadPresenceMean = vector("numeric", nrow(paramTable)),
-                         distanceToRoadMean = vector("numeric", nrow(paramTable)),
-                         forestryDisturbanceMean = vector("numeric", nrow(paramTable)))
+                         areaMean = "overall")
 
-  for(i in 1:nrow(paramTable)) {
-  # i = 1
-    cRow = paramTable[i,]
+  metricsAll <- paramTable %>% select(roadDisturbance:forestryDisturbance) %>%
+    unlist() %>%
+    map_dbl(~terra::global(.x, "mean", na.rm = TRUE)[1,1]) %>%
+    as_tibble(rownames = "name") %>%
+    tidyr::separate(name, into = c("name", "order"), sep = -1) %>%
+    mutate(name = paste0(name, "Mean")) %>%
+    tidyr::pivot_wider(names_from = "name", values_from = "value")
 
-    roadDisturbanceMean <- terra::global(cRow$roadDisturbance[[1]], "mean", na.rm = TRUE)
[…]

### 2023-01-19 analysis/paper/paper.Rmd
re-run metrics with above changes

@@
 
-```{r aspat-perf-fig, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”).
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster.
+Metrics were summarised within cutblocks, outside cutblocks, and overall.
 Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
@@
 # organize data for ggplot
-meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
+meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
   pivot_longer(-c(sampleType, sampleDens, areaMean, method),
@@
   mutate(prop_dif = (response_proj-response_obs)/response_obs,
-         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall")))
+         areaMean = factor(areaMean, labels = c("Cutblocks", "Outside\ncutblocks",
+                                                "Overall")))
 
@@
                                       width = 0.75))+
-  facet_grid(areaMean~ metric)+
+  facet_grid(areaMean~ metric, scales = "free")+
   theme_classic()+
@@
               values_from = count, values_fill = 0) %>% 
-  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
+  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`)
          precision = `Agree roaded`/(`Agree roaded` + `False positive`),
@@
   mutate(perf_meas = factor(perf_meas,
-                            labels = c("F-measure", "Precision", "Sensitivity"))) %>% 
+                            labels = c("F-measure", "Precision", "Sensitivity")),
+         value = value  +0.0001) %>%   # adding so 0 shows in figure
   ggplot(aes(sampleDens, value, fill = sampleDens))+
@@
   coord_flip()+
-  ylim(0, 1)+
+  ylim(0, 1.0001)+
   theme_classic()+
@@
 
-```{r figA1, fig.height=10, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke Timber Supply Area (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
+```{r figA1, fig.height=12, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
+Metrics were summarised within cutblocks, outside cutblocks, and overall.
 Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}

### 2023-01-19 R/functionsSource.R

@@ -260,7 +260,7 @@ getMetricMeans <- function(paramTable, cutblocks){
 #'
 #' @return a table
 #' @noRd
-calcAgree <- function(obs_rast, proj_rast, prex_rast){
+calcAgree <- function(obs_rast, proj_rast, prex_rast, return_res = FALSE){
 
   if(is(proj_rast, "character")){
     return(NULL)
@@ -269,6 +269,10 @@ calcAgree <- function(obs_rast, proj_rast, prex_rast){
   proj_rast <- terra::crop(proj_rast, obs_rast)
   prex_rast <- terra::crop(prex_rast, obs_rast)
   res <- obs_rast + prex_rast + proj_rast
+
+  if(return_res){
+    return(res)
+  }
 
   lu_tbl <- tibble::tribble(~value, ~agreement,
                             0, "Agree roadless",

### 2023-01-19 analysis/scripts/runProjections.R

@@ -96,6 +96,7 @@ run_projections(
   klementProj = paste0(data_path_drvd, "TSA27/", "klementProjection.shp"),
 
   aggFact = 1, #factor of aggregation of cost surface. 1 = no aggregation.
+  # load_file = "results",
   saveInputs = TRUE
 )
 

### 2023-01-19 analysis/scripts/testRunProjections.R

@@ -53,7 +53,8 @@ run_projections(
   #Klement QGIS projection results shapefile
   klementProj = NULL,
   aggFact = 1, #factor of aggregation of cost surface. 1 = no aggregation.
-  saveInputs = FALSE
+  saveInputs = FALSE,
+  load_file = "results"
 )
 
 resPths <- list.files(here(data_path_drvd, "test_tsb"), pattern = ".gpkg")
@@ -78,11 +79,39 @@ tmap_arrange(allMaps)
 
 meanTable <- read.csv(here(data_path_drvd, "test_tsb", "mean_table.csv"))
 
-cutOnly_rdDist <- terra::rast(here(data_path_drvd, "test_tsb", "roadDisturbancecutOnly_NA_NA.tif"))
-obs_rdDist <- terra::rast(here(data_path_drvd, "test_tsb", "roadDisturbanceobserved_NA_NA.tif"))
+meanTablenew <- read.csv(here(data_path_drvd, "test_tsb", "mean_table.csv"))
+
+# look at agree metric rasters
+
+cutOnly_rdDist <- terra::rast(here(data_path_drvd, "test_tsb", "forestryDisturbancecutOnly_NA_NA.tif"))
+obs_rdDist <- terra::rast(here(data_path_drvd, "test_tsb", "forestryDisturbanceobserved_NA_NA.tif"))
+
+cost_rast <- terra::rast(here(data_path_drvd, "testing_cost.tif"))
+
+prex_vect <- read_sf(here(data_path_drvd, "testing_ex_roads.gpkg"))
+prex_rast <- terra::rasterize(terra::vect(prex_vect), cost_rast, background = 0)
+boundary <- read_sf(here(data_path_drvd, "testing_tsb.gpkg"))
+cutblocks <-  read_sf(here(data_path_drvd, "testing_cutblocks.gpkg"))
+prex_rdDist <- roadDisturbanceFootprint(prex_vect, !is.na(prex_rast),
+                         boundary)
+prex_forDist <- disturbanceMetrics(linFeat = prex_vect,
+                                    landCover = as(cost_rast, "Raster"),
+                                    projectPoly = boundary,
+                                    anthroDist = terra::rasterize(terra::vect(cutblocks), cost_rast) %>%
+                                      as("Raster"))@processedData$Anthro %>% terra::rast()
+
+aggCutOnly <- calcAgree(terra::subst(obs_rdDist, from = 1, to = 100) %>%
+                          terra::crop(prex_forDist),
[…]

### 2023-01-19 analysis/scripts/prepareData.R
change prepare data to have option to download from OSF repo

@@ -5,73 +5,119 @@ library(dplyr)
 library(sf)
 library(here)
 library(purrr)
+library(osfr)
 
 #set paths for data
+data_path <- "analysis/data/"
 data_path_raw <- "analysis/data/raw_data/"
 data_path_drvd <- "analysis/data/derived_data/"
 
+# Download method: Raw data can be downloaded using bcdata but it may have been
+# updated since the paper was published so use OSF repository data to recreate
+# the paper results. Note that we could not get permission to store data from
+# the Digital Road Atlas so it will be downloaded with bcdata either way and the
+# cost raster was provided by Kyle Lockheed and will be downloaded from OSF
+# either way. Also note that only data for Revelstoke is included on OSF
+down_meth <- "osf"
+# down_meth <- "bcdata"
 
 # Download data -----------------------------------------------------------
 
-# Data from BC gov
-# * TSA boundaries
-# * digital road atlas and forest tenure road FLNRORD datasets.
-# * Harvest Areas of BC (Consolidated Cutblocks)
+if(down_meth == "osf"){
 
-# Do we have or can we get data and code for building the cost raster?
+  # Download the zip file of raw files from OSF
+  osf_proj <- osf_retrieve_node("https://osf.io/8vmqh/")
 
-# bcdc_search("timber supply area")
+  osf_ls_files(osf_proj) %>% filter(name == "raw_data.zip") %>%
+    osf_download(path = data_path)
 
-tsa_27 <- bcdc_query_geodata("8daa29da-d7f4-401c-83ae-d962e3a28980") %>%
-  filter(TSA_NUMBER == "27") %>%
-  collect()
+  # extract it to the data folder of the current project
+  unzip("data/raw_data.zip", exdir = data_path)
 
-# using .gpkg because writing to .shp abbreviates the column names
-write_sf(tsa_27, here(data_path_raw, "tsa27_boundaries.gpkg"))
+  # load tsa boundaries to use to get road data
+  tsa_27 <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
 
[…]

### 2023-01-20 analysis/paper/paper.Rmd
minor edits to the text and adding things JH asked for in todos

@@
 <!-- With the following code you can access and display values from the yml header above. -->
+
 <!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->
@@
 The impacts of roads on wildlife are complex and varied.
-Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021;@trombulak_review_2000;@fahrig_effects_2009].
-In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020;@polfus_identifying_2011;@mumma_predation_2018;@beauchesne_disentangling_2013;@leblond_avoidance_2013;@laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011;@whittington_towns_2022; @proctor_effects_2020].
-Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021;@apps_spatial_2013;@pinard_calving_2012;@fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020].
+Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].\
+In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020].
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
@@
 Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022].
-In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018;@snell_importance_2018;@daniel_incorporating_2017;@stewart_climate-informed_2023;@micheletti_assessing_2021;@leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
+In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
 Despite the importance of roads for wildlife, these models generally do not include road network projections.
-For example, in a recent collection of articles focussed on “Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests” [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].
+For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].
 
@@
 However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
-Published methods intended to support strategic planning [@anderson_projecting_2004;@clark_three-stage_2000;@dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales.
+Published methods intended to support strategic planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales.
 More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
@@
 The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose.
-Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020;@ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
+Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here.
@@
 
+```{=html}
 <!-- TO DO:
@@
 -->
-
+```
+```{=html}
 <!--
@@
 -->
-
+```
 <!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
@@
 
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access to forest harvest cutblocks (Figure \@ref(fig:study-area)).
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)).
 This allows projection of the forest road network from historic forest harvest data.
 Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g.
-alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observed the responses of road projection algorithms to barriers.
+alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observe the responses of road projection algorithms to barriers.
 Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021].
@@
 Existing roads are assigned a cost of 0.
-The cost raster consists of 1703936 pixels with a resolution of 1 ha.
+The cost raster consists of 1,703,936 pixels with a resolution of 1 ha.
 
@@
 All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].
-
-<!-- TO DO: add links to repositories and explain what data is and is not there, and why -->
+These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (add link).
+We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.
+
+<!-- TO DO: add link to OSF repository once public-->
 
@@
 centroid of a cutblock) and the existing road network.
-Least cost paths to each cutblock is determined independently in simple method (LCP).
+Least cost paths to each cutblock are determined independently in the simple method (LCP).
 The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
-Sequence is determined by a heuristic such as the distance to the nearest existing road in these examples.
-The minimum spanning tree (MST) method calculates the minimum cost network to connecting all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956].
+Sequence is determined by a heuristic, such as the distance to the nearest existing road in these examples.
+The minimum spanning tree (MST) method calculates the minimum cost network to connect all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956].
 See Appendix A for a more detailed explanation of these methods.
@@
 We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].
-
-<!-- TO DO: cite key R packages used in method implementation -->
 
@@
 
-To make help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package 'roads'.
-
-<!-- TO DO: add links to R package documentation & resources -->
+To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott2023].
+The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi2006a; @igraph:2022], `terra` for manipulating spatial raster data [@hijmans2023], and `sf` for manipulating spatial vector data [@pebesma2018; @pebesma2022].
+The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>).
+The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>).
 
@@
 ## Performance Metrics
+
 In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
-The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more].
+The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more].
 Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
 Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
-Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable without roads.
 All road characteristics were stored in raster format.
@@
 -->
-
-<!-- TO DO: update the details on getDistFromSource function - where is it, or where should it be? -->
-<!-- TO DO: Does the cutblocks only scenario include the old roads?
-If no, it should.
-And if so, it seems like the other metrics could be calculated in the cutblocks only scenario.
-It would be interesting to have the "no projection" baseline for all the metrics -->
 
@@
 
-For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks or overall scales (Figure \@ref(fig:aspat-perf-fig)).
+For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
+All road network projection method variants perform better for all metrics than the "cutblocks only" method where cutblocks were the only disturbance considered.
 All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks.
-The high density methods overproject the overall road disturbance footprint more than low density methods, and underproject overall distance to roads, but all method variants give reasonably good projections of overall distance to roads.
+The high density methods overproject the outside cutblock road disturbance footprint more than low density methods, and underproject the outside cutblock distance to roads, but all method variants give reasonably good projections of the outside cutblock distance to roads.
 All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments.
@@
 With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).
-
-<!-- How is the 'cutblocks only' scenario overestimating the overall road disturbance footprint? That doesn't make sense. -->
 
@@
 In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:proj-map-fig)a).
-However, the cost raster is not sufficiently details to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys.
+However, the cost raster is not sufficiently detailed to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys.
 There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:proj-map-fig)b).
@@
 All these discrepancies could be addressed with more accurate harvest and cost information.
+
+<!--# TODO: I don't know that the switchback issue could be addressed with more accurate cost information.
+Wouldn't you need to account for the direction of the slope and cost according to that in order to induce that behaviour?
+I guess that would be better cost info but is more complicated than this implies.
+-->
 
@@
               values_from = count, values_fill = 0) %>% 
-  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`)
+  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
          precision = `Agree roaded`/(`Agree roaded` + `False positive`),
@@
                             labels = c("F-measure", "Precision", "Sensitivity")),
-         value = value  +0.0001) %>%   # adding so 0 shows in figure
+         value = value  +0.01) %>%   # adding so 0 shows in figure
   ggplot(aes(sampleDens, value, fill = sampleDens))+
@@
   coord_flip()+
-  ylim(0, 1.0001)+
+  ylim(0, 1.01)+
   theme_classic()+
@@
 the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives.
-All projection methods yielded more accurate results than the 'cutblock only' scenario, indicating that any of these projection methods are better than none (Figure \@ref(fig:spat-perf-fig)).
+All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
+The performance of the "cutblocks only" method could not be assessed for road presence or disturbance footprint since no roads were projected.
+However, the "cutblocks only" method performed better than any other method for the forestry disturbance footprint.
+This reflected the fact that in the Revelstoke TSA most new roads were built inside  cutblocks, so the number of false negatives caused by not projecting roads was small.
 The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
@@
 
+```{=html}
 <!-- Discussion pts
@@
 -->
-
+```
 <!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice.
@@
 
+```{=html}
 <!--  What David had
@@
 -->
-
+```
 # Acknowledgements
@@
 <!-- TO DO: switch fig resolution labels from metres to ha. -->
+
 ```{r get-data2}

### 2023-01-20 analysis/paper/paper.Rmd
add mean_table files to git so JH can have updated results for knitting paper

@@
 # organize data for ggplot
-meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
+meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
   pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),

### 2023-01-20 analysis/scripts/runProjections.R

@@ -111,7 +111,7 @@ run_projections(
 
   #Klement QGIS projection results shapefile
   klementProj = NULL,
-
+  # load_file = "results",
   aggFact = 10 #factor of aggregation of cost surface. 1 = no aggregation.
 )
 

### 2023-04-05 analysis/paper/paper.Rmd

@@
 With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).
-
-<!-- TO DO: consider breaking down aspatial results into 'cutblocks', 'outside cutblocks' and 'overall': it may make the results easier to explain and interpret. -->
 
@@
 
-<!-- TO DO: switch fig resolution labels from metres to ha. -->
-
 ```{r get-data2}
@@
                                       width = 0.75))+
-  facet_grid(areaMean + metric ~1)+
+  facet_grid(areaMean + metric ~"")+
   theme_classic()+
-  scale_fill_manual(values = met.brewer(pal_nm))+
+  scale_fill_manual(values = met.brewer(pal_nm), labels = c("1 ha", "100 ha"))+
   theme(text = element_text(size = 10), axis.title.x = element_blank(),
         axis.title.y = element_blank(),
+        strip.background.x = element_blank(),
         legend.position = "right",
         axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
-  coord_flip()
+  coord_flip()+
+  labs(fill = "Resolution")
 
@@
                                       width = 0.75))+
-  scale_fill_manual(values = met.brewer(pal_nm))+
+  scale_fill_manual(values = met.brewer(pal_nm), labels = c("1 ha", "100 ha"))+
   facet_grid(metric ~ perf_meas)+
@@
   theme_classic()+
-  labs(y = "Performance", x = NULL)
+  labs(y = "Performance", x = NULL, fill = "Resolution")
 

### 2023-04-05 analysis/paper/paper.Rmd
end of day commit.

@@
 ## Aspatial performance
-
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
-All road network projection method variants perform better for all metrics than the "cutblocks only" method where cutblocks were the only disturbance considered.
-All road network projection method variants accurately predict the forestry disturbance footprint, and slightly overproject the road disturbance footprint within cutblocks.
-The high density methods overproject the outside cutblock road disturbance footprint more than low density methods, and underproject the outside cutblock distance to roads, but all method variants give reasonably good projections of the outside cutblock distance to roads.
+All road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered.
+Within all cutblocks, the Hardy method most accurately projects road density, and least accurately projects distance to road.
+Low density method variants most severely underproject road density within cutblocks, and most accurately project distance to road within cutblocks.
+These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
+Figure \@ref(fig:proj-map-fig)b).
+Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method more accurately projects ...
 All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments.
@@
 With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).
+
+Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)).
+Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a).
+Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy.
+Another source of discrepancy between observed and projected networks in this mountainous landscape are switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c).
+None of the simple algorithms considered here are capable of projecting switchbacks.
+Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
+
+As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
+the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
+Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
+The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
+The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
 
@@
 
-## Spatial performance
-
-Difference maps highlight several potential causes of discrepancy between our projections and the observed road network.
-In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:proj-map-fig)a).
-However, the cost raster is not sufficiently detailed to reflect costs of crossing rivers, streams or other wet areas in valley bottoms so the projections cross valleys more than the observed roads which often run parallel to one another on either side of valleys.
-There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:proj-map-fig)b).
-In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:proj-map-fig)c).
-All these discrepancies could be addressed with more accurate harvest and cost information.
-
-<!--# TODO: I don't know that the switchback issue could be addressed with more accurate cost information.
-Wouldn't you need to account for the direction of the slope and cost according to that in order to induce that behaviour?
-I guess that would be better cost info but is more complicated than this implies.
--->
-
 ```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network.
@@
 
-As expected, none of the projection methods can accurately project road presence (i.e.
-the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`), and all of the methods provide accurate projections of the forestry disturbance footprint (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-All methods also provide reasonably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
-The performance of the "cutblocks only" method could not be assessed for road presence or disturbance footprint since no roads were projected.
-However, the "cutblocks only" method performed better than any other method for the forestry disturbance footprint.
-This reflected the fact that in the Revelstoke TSA most new roads were built inside  cutblocks, so the number of false negatives caused by not projecting roads was small.
-The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
-The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
-
 # Discussion
@@
 Missing switchbacks and too many stream crossings also reduce our total amount of roads.
-All of these could be improved with more detailed high resolution cost & harvest and landing data.
+stream crossing behaviour could improved with more detailed high resolution cost.
+harvest and landing data.
 - density of landing pts has a larger effect than choice of road projection algorithm.
@@
 However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
-The results may be less accurate if the landings were spread further apart.
-But also less topographically complex landscapes may be simpler to project.
-- For some purposes, projections from relatively crude cost models may be adequate.
+Compare to Hardy results in eastern canadian shield - better road density projections in that landscape suggests better performance of simple iterative least cost path algorthms in more moderate terrain.
+- Recognize that adequacy depends on the goal.
+For some purposes, projections from relatively crude cost models and simple algorithms may be adequate.
 For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).

### 2023-04-06 analysis/paper/paper.Rmd
ongoing
to do - figure what is going wrong in closerLookAtPatterns script.

@@
 All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].
-These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (add link).
+These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (https://osf.io/8vmqh/).
 We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.
@@
 Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
-
-As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
-the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
-Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
-The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
-The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
 
@@
 
+As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
+the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
+Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
+The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
+The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
+
 # Discussion
@@
 - as expected, not great at projecting exact road locations.
-sources of discrepancy include switchback, stream crossings, inaccurate harvest locations.
-Missing switchbacks and too many stream crossings also reduce our total amount of roads.
-stream crossing behaviour could improved with more detailed high resolution cost.
-harvest and landing data.
-- density of landing pts has a larger effect than choice of road projection algorithm.
-So maybe don't get too worked up about the differences between one algorithm and another.
-if more accurate projections are required, focus first on refining cost & harvest landing locations.
-- Selected revelstoke because it is topographically complex, so a good challenge.
-However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
-Compare to Hardy results in eastern canadian shield - better road density projections in that landscape suggests better performance of simple iterative least cost path algorthms in more moderate terrain.
-- Recognize that adequacy depends on the goal.
+sources of discrepancy highlighted in this mountainous terrain include switchbacks, stream crossings, inaccurate harvest locations.
+we selected revelstoke because it is topographically complex, so a good challenge.
+Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for.
+Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface.
+But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain.
+In particular, none of these simple algorithms will project switchbacks on steep slopes.
+A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases.
+- Aspatial overall performance metrics can be difficult to interpret.
+To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies.
+Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors.
+Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
+ - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing.
+Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm, and the Hardy algorithm more accurately projects road density than either.
+Hardy algorithm is also less prone to putting roads on block edges.
+If accurate road density within blocks is primary goal, consider using Hardy algorithm.
+Thus far only implemented in Landis II and QGIS extension, but source code is open.
+If there is a need could integrate into roads R package.
+Or consider simpler adjustments to algorithms for selecting landing points such as placing landing points away from cutblock edges.
+- However, before investing in more complexity recognize that adequacy depends on the goal.
 For some purposes, projections from relatively crude cost models and simple algorithms may be adequate.
-For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).
-Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels.
-If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,).
+For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates ().
+In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).
+If our wildlife response models depend on distance to road or buffered disturbance footprint, these somewhat crude road projections are likely good enough.
+- A related point is that it may helpful to recognize that some road metrics are easier to project than others when constructing model to project wildlife responses to disturbance ...
+(Bodner,).
 If interested in predictors that we did not consider, see example script.
 Can evaluate alternative metrics without rerunning the projections.
+
+However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
 -->

### 2023-04-06 analysis/scripts/closerLookAtPatterns.R

@@ -1,0 +1,148 @@
+data_path_drvd <- "analysis/data/derived_data/"
+
+##############
+#get other little bits
+library(ggplot2)
+library(dplyr)
+library(stringr)
+library(tidyr)
+library(purrr)
+library(ggpubr)
+library(MetBrewer)
+library(here)
+library(tmap)
+library(sf)
+library(roads)
+library(terra)
+library(caribouMetrics)
+library(pfocal)
+
+devtools::load_all(here())
+klementProj = paste0(data_path_drvd, "TSA27/", "klementProjection.shp")
+out=sf::read_sf(klementProj)
+
+# set example area
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
+
+cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg")
+roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg")
+tsaBoundaryPth = paste0(data_path_raw, "tsa27_58_testing.gpkg")
+costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif")
+
+# densities
+high <- 0.00001
+low <-  0.000001
+cutblocksI <- read_sf(cutblocksPth)
+roadsI <- read_sf(roadsPth)
+tsaCostI <- rast(costPth)
+
+projRoadsI <- read_sf(paste0(data_path_drvd, "TSA27/", "random_1e-05_mst.gpkg"))
+exRoadsI <- roads %>% filter(AWARD_DATE <= as.Date("1990-01-01"))
+
+# find a good location
+ qtm(st_as_sf(cutblocksI), fill = NULL, borders.lwd = 2)+
+   qtm(st_as_sf(roadsI), lines.col = "red", lines.lwd = 2)+
+   qtm(st_as_sf(projRoadsI), lines.col = "green", lines.lwd = 2)+
+   qtm(st_as_sf(exRoadsI), lines.lwd = 2)+
+     tm_layout(legend.show = FALSE)
+
+ ext(tsaCost)
+# # get extent
+ expX = 0.02
+ expY = 0.02
+bound <- ext(c(1537685.65481532*(1-expX), 1546695.23062125*(1+expX), 741830.821988938*(1-expY), 750131.44756752*(1+expY)))
+
+roads <- vect(roadsI) %>% crop(bound) %>% st_as_sf()
[…]

### 2023-04-06 analysis/scripts/method_example_fig.R

@@ -30,7 +30,7 @@ cutblocks <- read_sf(cutblocksPth)
 roads <- read_sf(roadsPth)
 tsaCost <- rast(costPth)
 
-projRoads <- read_sf(paste0(data_path_drvd, "TSA27/", "random_1e-05.shp"))
+projRoads <- read_sf(paste0(data_path_drvd, "TSA27/", "random_1e-05_mst.gpkg"))
 
 exRoads <- roads %>% filter(AWARD_DATE <= as.Date("1990-01-01"))
 
@@ -69,5 +69,5 @@ tsb <- st_bbox(cutblocks) %>% st_as_sfc() %>% list()
 write_sf(exRoads, paste0(data_path_drvd, "testing_ex_roads.gpkg"))
 write_sf(tsb[[1]], paste0(data_path_drvd, "testing_tsb.gpkg"))
 write_sf(cutblocks, paste0(data_path_drvd, "testing_cutblocks.gpkg"))
-writeRaster(tsaCost, paste0(data_path_drvd, "testing_cost.tif"))
+writeRaster(tsaCost, paste0(data_path_drvd, "testing_cost.tif"),overwrite=T)
 write_sf(roads, paste0(data_path_drvd, "testing_obs_roads.gpkg"))

### 2023-04-11 analysis/scripts/closerLookAtPatterns.R
partially complete script - may not be needed. Will leave in for now.

@@ -49,8 +49,8 @@ exRoadsI <- roads %>% filter(AWARD_DATE <= as.Date("1990-01-01"))
 
  ext(tsaCost)
 # # get extent
- expX = 0.02
- expY = 0.02
+ expX = 0.03
+ expY = 0.03
 bound <- ext(c(1537685.65481532*(1-expX), 1546695.23062125*(1+expX), 741830.821988938*(1-expY), 750131.44756752*(1+expY)))
 
 roads <- vect(roadsI) %>% crop(bound) %>% st_as_sf()

### 2023-04-11 analysis/paper/paper.Rmd
add Hardy QGIS to proj methods fig

@@
 We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)).
-We used default settings in the Hardy QGIS plugin except for kidding distance which was set to 200m.
+We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200m.
 A skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
@@
 
-sampleDens <- c(low,high,low,high,low)
-sampleType <- c("regular","regular","random","random","centroid")
+sampleDens <- c(high, high, low)
+sampleType <- c("regular","random","centroid")
 paramTable <- tibble(sampleType, sampleDens,
@@
 
-lcpProj <- projectAll(tsbs = tsb,
-                       paramTable = paramTable %>% 
-                        filter(sampleType == "centroid") %>% 
-                        mutate(method = "lcp"),
-                       costSurface = tsaCost,
-                       cutblocks = cutblocks,
-                       existingRoads = exRoads,
-                       fileLocation = here(data_path_drvd, "..", "for_fig"))
-
-allProj <- bind_rows(mstProj, dlcpProj, lcpProj, .id = "method") %>%
+allProj <- bind_rows(mstProj, dlcpProj, .id = "method") %>%
   arrange(desc(sampleType), sampleDens)
+
+# add Hardy QGIS results for the same area
+
+allProj <- bind_rows(
+  allProj,
+  tibble(sampleType = "", sampleDens = 0, method = "Hardy QGIS",
+             output = list(here(data_path_drvd, "..", "for_fig", 
+                                "HardyQGISTest.gpkg")))) %>% 
+  mutate(mapTitle =  paste(ifelse(method == 1, "MST",
+                                  ifelse(method == 2, "DLCP", method)),
+                           sampleType,
+                           ifelse(sampleDens == 1e-06 & 
+                                    sampleType != "centroid", "low density", 
+                                  ifelse(sampleDens == 1e-05, 
+                                         "high density", ""))))
+
+# add extra row for legend
+allProj <- bind_rows(allProj, 
+                     slice(allProj, 1))
 
@@
   ~ qtm(tsaCost, raster.style = "cont", raster.palette = "Greys",
-        raster.alpha = 0.5)+
+        raster.alpha = 0.5, raster.title = "Cost")+
     qtm(cutblocks, fill = NULL, borders = "grey40", borders.lwd = 1.5)+
@@
     qtm(exRoads, lines.col = "black", lines.lwd = 2)+
-    tm_layout(legend.show = FALSE,
-              main.title = paste(ifelse(allProj$method[.x]== 1, "MST",
-                                        ifelse(allProj$method[.x]== 2, "DLCP", "LCP")),
-                                 allProj$sampleType[.x],
-                                 ifelse(allProj$sampleType[.x] == "centroid", "",
-                                        ifelse(allProj$sampleDens[.x] == 1e-06, "low density", "high density"))),
+    tm_layout(legend.show = .x == nrow(allProj),
+              legend.only = .x == nrow(allProj),
+              main.title = allProj$mapTitle[[.x]],
               main.title.size = 0.75))
 
-tmap_save(tmap_arrange(allMaps, ncol = 2),
+allMaps[[nrow(allProj)]] <- allMaps[[nrow(allProj)]]+
+  tm_add_legend(type = "line", 
+                labels = c("Existing roads", "Projected roads", "Observed roads"),
+                col = c("black", "yellow", "red"))+
+    tm_add_legend(type = "fill", 
+                labels = "Cutblocks",
+                col = NA, border.col = "grey40")+
+  tm_layout(legend.only = TRUE, legend.position = c("center", "center"))
+
+tmap_save(tmap_arrange(allMaps, ncol = 3),
           here("analysis/figures/projection_methods_figure.png"),
-          dpi = 300, height = 8, width = 3.543)
-```
-
-```{r proj-meth-fig, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and least cost path (LCP) road network projection algorithms.
+          dpi = 300, height = 5, width = fig_widths["two"])
+```
+
+```{r proj-meth-fig, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and the Hardy QGIS plugin road network projection algorithms.
 Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2).
 Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red.
-Cost (1 ha resolution) is indicated by grey shading."}
+Cost (1 ha resolution) is indicated by grey shading, with darker grey indicating higher cost.
+The large four sided cutblock in the top right was added to give a better example of the algorithms behaviour inside cutblocks."}
 knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))

### 2023-04-12 analysis/scripts/calc_any_metric.R
finish calc any metric script

@@ -7,6 +7,7 @@ library(tmap)
 library(purrr)
 library(here)
 library(terra)
+library(ggplot2)
 
 # set paths for data
 data_path_raw <- "analysis/data/raw_data/"
@@ -63,6 +64,9 @@ rast_temp <- rast(here(data_path_drvd, "TSA27", "input_cost.tif"))
 # study area boundary
 bound <- read_sf( here(data_path_raw, "tsa27_boundaries.gpkg"))
 
+# Cutblocks
+cutblocks <- read_sf(here(data_path_drvd, "TSA27", "input_cutblocks.gpkg"))
+
 metric_rasts <- map(files, read_sf) %>%
   map(roadDisturbanceFootprint, r = rast_temp, b = bound, d = 250)
 
@@ -70,6 +74,29 @@ metric_rasts <- map(files, read_sf) %>%
 # Step 3:  ----------------------------------------------------------------
 # Calculate the mean of the metric for each raster
 
-metric_means <- metric_rasts %>%
-  map_dbl(~terra::global(.x, "mean", na.rm = TRUE)[1,1])
+metric_means_overall <- metric_rasts %>%
+  map_dfr(~terra::global(.x, "mean", na.rm = TRUE)[1,1]) %>%
+  mutate(metric = "Footprint 250", Scale = "Overall")
 
+metric_means_in_cuts <- metric_rasts %>%
+  map_dfr(~ terra::mask(.x, cutblocks) %>%
+            terra::global("mean", na.rm = TRUE) %>% .[1,1]) %>%
+  mutate(metric = "Footprint 250", Scale = "Cutblocks")
+
+metric_means_out_cuts <- metric_rasts %>%
+  map_dfr(~ terra::mask(.x, cutblocks, inverse = TRUE) %>%
+            terra::global("mean", na.rm = TRUE) %>% .[1,1]) %>%
+  mutate(metric = "Footprint 250", Scale = "Outside cutblocks")
+
+metric_means <- bind_rows(metric_means_overall, metric_means_in_cuts,
+                          metric_means_out_cuts)
+
+metric_means %>%
+  tidyr::pivot_longer(cols = -c(Scale, observed, metric), names_to = "method",
+                      values_to = "response") %>%
+  mutate(prop_dif = (response - observed)/observed) %>%
+  ggplot(aes(method, prop_dif))+
+  geom_col()+
+  facet_grid(Scale ~ metric)+
+  coord_flip()+
+  labs(x = "Method", y = "(Projected - Observed)/Observed")

### 2023-04-12 analysis/paper/paper.Rmd
In progress: writing

@@
     correspondence: true
+    equal_contributor: "yes"
+    orcid_id: 0000-0001-7875-9015
   - name: Sarah Endicott
@@
     correspondence: false
+    equal_contributor: "yes"
   - name: David Lapins
@@
 Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022].
-In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
+In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
 Despite the importance of roads for wildlife, these models generally do not include road network projections.
@@
 Leston et al.
-(2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire.
-This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool.
-However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
-Published methods intended to support strategic planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales.
-More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
-The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
-These methods have not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
-@barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].
-
-In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package.
-We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife.
-The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose.
-Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
-We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here.
-Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.
+(2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
+This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] and tools exist [@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate; @epstein_combinatorial_2006].
+However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
+Methods intended to support tactical and operational planning also require detailed parameterization with inputs that may not be publicly available, and employ algorithms that are too complex for application at regional, provincial or national scales.
+More promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
+The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
+Hardy's tool is designed for integration with other components of the Landis-II modelling framework [e.g.
+@labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+@barros_empowering_2023; @daniel_state-and-transition_2016].
+
+To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package.
+We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia.
+The adequacy of a road network projection methods depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose.
+Our interest is projecting impacts of forest change on wildlife, so we focused on common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 
@@
 We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)).
-We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].
+We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023].
 
@@
 
-To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott2023].
-The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi2006a; @igraph:2022], `terra` for manipulating spatial raster data [@hijmans2023], and `sf` for manipulating spatial vector data [@pebesma2018; @pebesma2022].
+To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott_roads_2023].
+The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
 The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>).
@@
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
-All road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered.
-Within all cutblocks, the Hardy method most accurately projects road density, and least accurately projects distance to road.
-Low density method variants most severely underproject road density within cutblocks, and most accurately project distance to road within cutblocks.
+Within all cutblocks, the Hardy method most accurately projected road density, and least accurately projected distance to road.
+Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks.
 These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
 Figure \@ref(fig:proj-map-fig)b).
-Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method more accurately projects ...
-All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments.
-All the methods also underproject road presence, and the problem is more severe for the low density method variants.
-In general, performance differences among low and high density method variants are larger than differences among projection algorithms.
-With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).
+Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method most accurately projected road density and distance to road.
+However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain (Figure \@ref(fig:proj-meth-fig)).
+The Hardy method projects more wiggles than the other methods, but none of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+
+Another fundamental limitation of all these methods is that they generate straight road segments within pixels.
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse coarse (Figure A-\@ref(fig:figA1)).
+In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).
 
@@
 Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy.
-Another source of discrepancy between observed and projected networks in this mountainous landscape are switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c).
-None of the simple algorithms considered here are capable of projecting switchbacks.
+Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks.
 Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
+Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered.
 
@@
 <!-- Discussion pts
-- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
-- as expected, not great at projecting exact road locations.
-sources of discrepancy highlighted in this mountainous terrain include switchbacks, stream crossings, inaccurate harvest locations.
+- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models).
+Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows.
+- as expected, none of these algorithms are great at projecting exact road locations.
 we selected revelstoke because it is topographically complex, so a good challenge.
@@
  - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing.
-Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm, and the Hardy algorithm more accurately projects road density than either.
-Hardy algorithm is also less prone to putting roads on block edges.
-If accurate road density within blocks is primary goal, consider using Hardy algorithm.
-Thus far only implemented in Landis II and QGIS extension, but source code is open.
-If there is a need could integrate into roads R package.
-Or consider simpler adjustments to algorithms for selecting landing points such as placing landing points away from cutblock edges.
-- However, before investing in more complexity recognize that adequacy depends on the goal.
+Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm.
+Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this.
+But note that these algorithms can get the right density without replicating the pattern or mechanism.
+Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid.
+Projected wiggles are quite different from observed switchbacks.
+- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do.
+But before investing in more complexity recognize that adequacy depends on the goal.
 For some purposes, projections from relatively crude cost models and simple algorithms may be adequate.

### 2023-04-13 analysis/paper/paper.Rmd
Add fig results for real cutblocks only Only added the table and checked the difference in the figure

--- 2023-04-11 analysis/paper/paper.Rmd
@@
 meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
-
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+                                   "mean_table.csv"))
 matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))

### 2023-04-13 analysis/paper/paper.Rmd
writing in progress - reconciling changes

--- 2023-04-12 analysis/paper/paper.Rmd
@@
 The impacts of roads on wildlife are complex and varied.
-Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].\
+Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].
 In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
-Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021].
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
@@
 We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia.
-The adequacy of a road network projection methods depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose.
-Our interest is projecting impacts of forest change on wildlife, so we focused on common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
+The adequacy of a road network projection methods depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose.
+Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 
@@
 
+Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools designed for this purpose have not been available.
+Hardy's
+
 ```{=html}
@@
 Projected wiggles are quite different from observed switchbacks.
+- Focussed here on forest resource roads.
+Note the algorithms are also likely to work for other situations where the objectives of road building are to access target locations and minimize cost.
+Not a useful algorithm for projecting linear features that were constructed for some other purpose such as seismic testing (e.g.
+Figure \@ref(fig:ftNelson)), or for circumstances where information about target locations is not available.
+When considering transferring these results to other contexts, also note that in this example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
+It would be interesting to test performance in a circumstance where target locations are far from existing roads.
 - If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do.
@@
 In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).
-If our wildlife response models depend on distance to road or buffered disturbance footprint, these somewhat crude road projections are likely good enough.
-- A related point is that it may helpful to recognize that some road metrics are easier to project than others when constructing model to project wildlife responses to disturbance ...
-(Bodner,).
-If interested in predictors that we did not consider, see example script.
-Can evaluate alternative metrics without rerunning the projections.
-
-However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
--->
+For wildlife response models informed by distance to road or buffered disturbance footprint, somewhat crude road projections are likely good enough.
+When constructing new wildlife response models intended for projection, it be helpful to select predictor variables such as distance to road, road presence within grid cells, and disturbance footprint that are more easily projectable than road density (Bodner).
+- There are a bewildering array of possible metrics of connectivity and fragmentation we did not consider in this paper.
+In general, performance will vary among metrics, and it is difficult to tell what is good enough without reference to particular circumstances (objectives, species, landscapes).
+Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making the road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals.
+The best performance metrics will be those most closely linked to the outcome of interest.
+For example, where the outcome of interest is space use or connectivity projected from more (e.g.
+Whittington, Labadie) or less (samc, pither etc) species-specific movement models consider testing sensitivity of these models to the difference between observed and retrospectively projected roads.
+We hope R package and open workflow associated with this paper can help enable retrospective projection and relevant performance assessment in other contexts.
+However, we also recognize there is work associated with acquiring data to construct cost surfaces & retrospectively project roads.
+For those seeking a lower effort option, we also provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.
+
 ```

### 2023-04-13 analysis/paper/paper.Rmd
Update paper to use results for real cutblocks

--- 2023-04-13 analysis/paper/paper.Rmd
@@
 # load csv
-meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
 meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
                                    "mean_table.csv"))
-matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
+
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+                                   "agree_table.csv"))
 
@@
 Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
-The cutblocks only method is not defined for the road density, road presence or distance to road characteristics."}
+Cutblocks that did not intersect the observed road network were excluded"}
 # organize data for ggplot
-meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
+doAspatPlot <- function(tab){
+  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>% 
   pivot_longer(-c(sampleType, sampleDens, areaMean, method),
@@
 
-# save runTimes for later comparison
-dlcp_time <- meanTable %>% filter(method == "dlcp", areaMean == "overall") %>% 
-  pull(runTime) %>% round(2)
-
-mst_time <- meanTable %>% filter(method == "mst", areaMean == "overall",
-                                 sampleType == "regular", sampleDens == "high sample density") %>% 
-  pull(runTime) %>% round(2) %>% `*`(60)
-
 prop_dif %>% 
@@
   coord_flip()
+}
+
+doAspatPlot(meanTableRealCuts)
+
+# save runTimes for later comparison
+dlcp_time <- meanTableRealCuts %>% 
+  filter(method == "dlcp", areaMean == "overall") %>% 
+  pull(runTime) %>% round(2)
+
+mst_time <- meanTableRealCuts %>%
+  filter(method == "mst", areaMean == "overall",
+         sampleType == "regular", sampleDens == "high sample density") %>% 
+  pull(runTime) %>% round(2) %>% `*`(60)
 
@@
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
-Values closer to one indicate better performance."}
-
-matchData_sum <- matchData %>% ungroup() %>%
+Values closer to one indicate better performance.
+Cutblocks that did not intersect the observed road network were excluded"}
+doSpatPerf <- function(tab){
+  matchData_sum <- tab %>% ungroup() %>%
     mutate(agreement = factor(agreement,
@@
     select(metric, sampleDens, sensitivity, precision, F_measure) 
-
+}
+
+perf_tbl <- doSpatPerf(matchDataRealCuts)
 perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
@@
 
-perf_tbl %>% 
+doSpatPlot <- function(tab){
+  tab %>% 
     pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
@@
     labs(y = "Performance", x = NULL)
+}
+
+doSpatPlot(perf_tbl)
 
@@
 
+## All cutblocks results
+
+```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster.
+Metrics were summarised within cutblocks, outside cutblocks, and overall.
+Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
+Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
+meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
+doAspatPlot(meanTable)
+```
+
+```{r spat-perf-fig-all, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
+Values closer to one indicate better performance.
+Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
+matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
+perf_tbl2 <- doSpatPerf(matchData)
+doSpatPlot(perf_tbl2)
+
+```
+
 ## Coarse resolution results
@@
 
-```{r figA1, fig.height=12, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
+```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
 Metrics were summarised within cutblocks, outside cutblocks, and overall.
@@
                                       width = 0.75))+
-  facet_grid(areaMean + metric ~"")+
+  facet_grid(areaMean ~ metric, scales = "free")+
   theme_classic()+
@@
         axis.title.y = element_blank(),
-        strip.background.x = element_blank(),
         legend.position = "right",
@@
 
-```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
+```{r figA2, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
 F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
@@
                                     "Forestry disturbance\nfootprint",
-                                    "Road disturbance footprint"))) %>%
+                                    "Road disturbance\nfootprint"))) %>%
   group_by(metric, sampleDens, agreement, resolution) %>%
@@
   coord_flip()+
-  ylim(0, 1)+
+  scale_y_continuous(limits = c(0, 1), breaks = 0:5/5)+
   theme_classic()+

### 2023-04-13 analysis/paper/paper.Rmd

--- 2023-04-13 analysis/paper/paper.Rmd
@@
 # load csv
-meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
-
-matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+                                   "mean_table.csv"))
+
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+                                   "agree_table.csv"))
 

### 2023-04-13 analysis/paper/paper.Rmd
reconciling.

--- 2023-04-13 analysis/paper/paper.Rmd
@@
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
-Within all cutblocks, the Hardy method most accurately projected road density, and least accurately projected distance to road.
+Within all mapped cutblocks, the Hardy method most accurately projected road density, and least accurately projected distance to road.
 Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks.
@@
 Figure \@ref(fig:proj-map-fig)b).
-Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method most accurately projected road density and distance to road.
+Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road.
 However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain (Figure \@ref(fig:proj-meth-fig)).
-The Hardy method projects more wiggles than the other methods, but none of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+The Hardy method projects more tortuous roads within cutblocks, but none of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
 
@@
 meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
-meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
-                                   "mean_table.csv"))
+#meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+#                                   "mean_table.csv"))
 matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
@@
 Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows.
-- as expected, none of these algorithms are great at projecting exact road locations.
-we selected revelstoke because it is topographically complex, so a good challenge.
-Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for.
-Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface.
+- we selected revelstoke because it is topographically complex, so a good challenge.
+Hardy result in eastern Canadian shield showed better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for.
+Stream crossing behaviour in Revelstoke landscape could be improved with more a detailed high resolution cost surface.
 But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain.
@@
 To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies.
-Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors.
-Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
+Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory result finding that me Better harvest & existing road data would allow us to more distinguish errors in projections from errors in the input data (cite lit in coops paper)
  - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing.

### 2023-04-13 analysis/paper/paper.Rmd

--- 2023-04-13 analysis/paper/paper.Rmd
@@
 # load csv
+meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
 meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",

### 2023-04-13 analysis/paper/paper.Rmd
writing in progress

@@
 All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
-The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy.
+The DLCP projection method had comparable performance to the MST projection method with the same sampling strategy.
 The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
@@
 But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain.
-In particular, none of these simple algorithms will project switchbacks on steep slopes.
+None of these simple algorithms will project switchbacks on steep slopes.
 A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases.
@@
 To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies.
-Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory result finding that me Better harvest & existing road data would allow us to more distinguish errors in projections from errors in the input data (cite lit in coops paper)
- - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing.
+Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results.
+Better harvest & existing road data would allow us to more distinguish errors in projections from errors in the input data (cite lit in coops paper)
+ - Outside cutblocks, we saw very little variation in performance among our method variants - any of them are somewhat better than nothing, but they don't make a big differene, partly because in this example projected new roads over fairly short distances, and new landings were always fairly close to existing roads or other landings.
+Would be interesting to test in a context where targets are further from existing roads, but outside cutblocks our DLCP algorithm is very similar to Hardy's algorithm, so expect similar performance.
 Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm.
 Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this.
-But note that these algorithms can get the right density without replicating the pattern or mechanism.
+But examination of spatial patterns clarifies that algorithms can get the right density without replicating the pattern or mechanism.
 Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid.
@@
 - Focussed here on forest resource roads.
-Note the algorithms are also likely to work for other situations where the objectives of road building are to access target locations and minimize cost.
+Note the algorithms could also work in other situations where the objectives of road building are to access target locations and minimize cost.
 Not a useful algorithm for projecting linear features that were constructed for some other purpose such as seismic testing (e.g.
 Figure \@ref(fig:ftNelson)), or for circumstances where information about target locations is not available.
-When considering transferring these results to other contexts, also note that in this example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong.
-It would be interesting to test performance in a circumstance where target locations are far from existing roads.
 - If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do.
@@
 In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).
-For wildlife response models informed by distance to road or buffered disturbance footprint, somewhat crude road projections are likely good enough.
-When constructing new wildlife response models intended for projection, it be helpful to select predictor variables such as distance to road, road presence within grid cells, and disturbance footprint that are more easily projectable than road density (Bodner).
+For wildlife response models informed by distance to road () or buffered disturbance footprint (), somewhat crude road projections are likely good enough.
+When constructing new wildlife response models intended for projection, it may be helpful to select these more easily projectable (Bodner) predictors.
 - There are a bewildering array of possible metrics of connectivity and fragmentation we did not consider in this paper.
-In general, performance will vary among metrics, and it is difficult to tell what is good enough without reference to particular circumstances (objectives, species, landscapes).
-Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making the road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals.
+In general, performance will vary among metrics, and it is difficult determine what is good enough without reference to particular circumstances (objectives, species, landscapes).
+Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals.
 The best performance metrics will be those most closely linked to the outcome of interest.
@@
 However, we also recognize there is work associated with acquiring data to construct cost surfaces & retrospectively project roads.
-For those seeking a lower effort option, we also provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.
+For those seeking a lower effort option, we provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.
 

### 2023-04-14 analysis/paper/paper.Rmd
writing in progress

@@
 
-Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools designed for this purpose have not been available.
-Hardy's
+Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available.
+Hardy's Landis-II extension [-@hardy_forest_2019; -@hardy_landis-ii_2023] can be integrated with other components of the Landis-II modelling framework [e.g.
+@labadie_global_2023].
+Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provides a comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+@barros_empowering_2023; @daniel_state-and-transition_2016].
+
+We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model.
+Thus, we were not surprised to less accurate projections of road density and presence than Hardy found in less mountainous regions of eastern Canada with a model that more carefully accounted for water crossing costs [-@hardy_landis-ii_2023].
+The stream crossing behaviour of road projection algorithms in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains.
+A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases.
+The performance of these algorithms varies with terrain; in topographically complex landscape high resolution cost surfaces that account for the costs of stream and river crossings will address some but not all of the challenges of forest road projection in mountain regions.
+
+Aspatial metrics of overall projection accuracy can be difficult to interpret.
+To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies.
+Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results.
+The lack of consistent accurate information on the location of resource roads across Canada and elsewhere [@poley_identifying_2022](Poley, ) Better harvest & existing road data would help errors in projections from errors in the input data (cite lit in coops paper)
 
@@
 <!-- Discussion pts
-- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models).
-Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows.
 - we selected revelstoke because it is topographically complex, so a good challenge.
@@
 None of these simple algorithms will project switchbacks on steep slopes.
-A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases.
 - Aspatial overall performance metrics can be difficult to interpret.

### 2023-04-19 analysis/paper/paper.Rmd
Writing in progress - this version contains road network projection details in appendix A for reference. The details will reduced and moved into the main text in future versions.

@@
 In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
-Despite the importance of roads for wildlife, these models generally do not include road network projections.
+Despite the importance of roads for wildlife, very few of these projections have included road networks.
 For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].
@@
 The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
-Hardy's tool is designed for integration with other components of the Landis-II modelling framework [e.g.
-@labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 The adequacy of a road network projection methods depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose.
-Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
+Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023; @stewart_climate-informed_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 
@@
 
-Continue reviewing pdfs in downloads -
-
 -->
 ```
-<!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.).
--->
 
@@
 
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)).
-This allows projection of the forest road network from historic forest harvest data.
-Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g.
-alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observe the responses of road projection algorithms to barriers.
+We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because it is a mountainous areas with approximately 55% of non-forested land where road building is difficult or impossible (e.g.
+alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010].
+Thus, the area provides opportunities to observe the responses of road projection algorithms to steep complex terrain.
+Forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)), which allows projection of the forest road network from historic forest harvest data.
 Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021].
-The area is home to wildlife species that are sensitive to roads including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010].
+The area is home to wildlife of concern including including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010].
 Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020].
@@
 
-Our road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC.
+Road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC.
 To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set.
@@
 All roads from the DRA data set were assumed to be built before 1990.
-Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021].
-Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
+Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
 The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021].
@@
 
-To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1).
+We developed a simple model of road building cost informed by slope and major barriers (Table 1).
 Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)).
@@
 All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].
-These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (https://osf.io/8vmqh/).
+These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository <https://osf.io/8vmqh/>.
 We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.
-
-<!-- TO DO: add link to OSF repository once public-->
+Scripts used for data acquisition and analysis associated with this paper are available on GitHub <https://github.com/LandSciTech/RoadPaper>.
+
+<!-- TO DO: number analysis scripts to clarify workflow -->
 
@@
 
-Simple methods for projecting low-cost routes from an existing road network to target locations (i.e.
-harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)).
-We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g.
-centroid of a cutblock) and the existing road network.
-Least cost paths to each cutblock are determined independently in the simple method (LCP).
-The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths.
-Sequence is determined by a heuristic, such as the distance to the nearest existing road in these examples.
-The minimum spanning tree (MST) method calculates the minimum cost network to connect all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956].
-See Appendix A for a more detailed explanation of these methods.
-We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)).
-We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023].
+We focus on methods that link multiple target locations to an existing road network for strategic purposes, described by Dean (1997) as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network.
+We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail; in this manuscript we focus on comparing the dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+
+The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAP).
+For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
+The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
+Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)).
+The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design.
+The DLCP approach solves each STAP sequentially, starting with target that is nearest to an existing road.
+The existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target.
+The method was first developed by Anderson and Nelson (2004) who noted that the projected network is sensitive to the ordering of the STAP and recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
+
+The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
+Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
+The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network.
+These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
+The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node in graph with a tree that minimizes edge weights and does not include loops.
+This method differs from the minimum spanning tree method used by Clark et al.
+(2000) who used Euclidean distances rather than least costs for edge weights.
 
@@
 
-To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott_roads_2023].
-The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>).
-The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>).
+To help make these road network projection methods more easily and widely useable, we implemented the DLCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
+The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The package is available on CRAN (The Comprehensive R Archive Network).
+The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>), along with tutorials and documentation (<https://landscitech.github.io/roads/>).
 
@@
 
-In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
-The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more].
+To assess performance we calculated five characteristics of the projected and observed road networks: road presence, measured as whether or not roads are present in each pixel; road density, measured as density of roads in meters per hectare (m/ha); distance to nearest road, measured as distance from each raster cell to the nearest road; road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
+The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020] and to quantify impacts of disturbance on Southern Mountain caribou [@lochhead_linking_2022; @polfus_identifying_2011].
+The other metrics are also often used in studies of anthropogenic impacts on wildlife [e.g.
+@labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
 Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
-All road characteristics were stored in raster format.
+
+TO DO: Sarah are these statements still correct? Please revise as necessary.
 
@@
 
-The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values.
-This was done across the timber supply area and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks.
+Aspatial performance of the projections was assessed by comparing projected and observed mean values of each road network characteristic across the timber supply area and within cutblocks.
+We further distinguished mapped cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
 Spatial performance was assessed in two different ways.
@@
 The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
-Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed.
+Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footpint was observed.
 These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
@@
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
-Within all mapped cutblocks, the Hardy method most accurately projected road density, and least accurately projected distance to road.
-Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks.
+Within all mapped cutblocks (Figure \@(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road.
+Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@(fig:aspat-perf-fig-all)).
 These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
@@
 Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road.
-However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain (Figure \@ref(fig:proj-meth-fig)).
-The Hardy method projects more tortuous roads within cutblocks, but none of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
-
-Another fundamental limitation of all these methods is that they generate straight road segments within pixels.
+However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuousity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
+None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+
+Another fundamental limitation of these methods is that they generate straight road segments within pixels.
 As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse coarse (Figure A-\@ref(fig:figA1)).
@@
 Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
-Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered.
+Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.
 
@@
 
+TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.
+
 # Discussion
@@
 Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available.
-Hardy's Landis-II extension [-@hardy_forest_2019; -@hardy_landis-ii_2023] can be integrated with other components of the Landis-II modelling framework [e.g.
+Hardy et al.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g.
 @labadie_global_2023].
-Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provides a comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide a complementary and comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model.
-Thus, we were not surprised to less accurate projections of road density and presence than Hardy found in less mountainous regions of eastern Canada with a model that more carefully accounted for water crossing costs [-@hardy_landis-ii_2023].
-The stream crossing behaviour of road projection algorithms in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains.
-A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases.
-The performance of these algorithms varies with terrain; in topographically complex landscape high resolution cost surfaces that account for the costs of stream and river crossings will address some but not all of the challenges of forest road projection in mountain regions.
+Thus, we were not surprised to see less accurate projections of road density and presence than in less mountainous regions of eastern Canada with more careful accounting for the cost of water crossings [@hardy_landis-ii_2023].
+Water crossing behaviour in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains.
+A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density becomes less accurate as raster cell resolution increases.
 
@@
 Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results.
-The lack of consistent accurate information on the location of resource roads across Canada and elsewhere [@poley_identifying_2022](Poley, ) Better harvest & existing road data would help errors in projections from errors in the input data (cite lit in coops paper)
+These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+Outside cutblocks, we saw very little variation in performance among projection method variants.
+All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads.
+Outside cutblocks our DLCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads.
+Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm.
+The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid.
+One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.
+
+Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are more costly to develop and use.
+Before investing in additional complexity, it is important to recognize that adequacy depends on the goal.
+For some purposes, projections from relatively crude cost models and simple algorithms may be adequate.
+For example, models of caribou movement often include distance to roads as a predictor [e.g.
+@labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020].
+In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
+
+There are many metrics of connectivity and fragmentation [@francis_routledge_2021] we did not consider in this paper.
+In general, performance will vary among metrics, and it is difficult determine what is good enough without reference to particular circumstances (i.e.
+objectives, species, landscapes).
+The most informative performance metrics will be those most closely linked to outcomes of interest.
+For example, where the outcome of interest is space use or connectivity projected from more [e.g.
+@whittington_towns_2022; @labadie_global_2023] or less [e.g.
+@hughes_comparison_2023; @diniz_landscape_2020] species-specific movement models it may be helpful to test sensitivity of these models to the difference between observed and retrospectively projected roads.
+When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
+Rather than adding additional performance metrics to this paper, we have focussed our efforts on development of an open source R package [@endicott_roads_2023] and workflows <https://github.com/LandSciTech/RoadPaper> that we hope will help enable users devise performance tests appropriate to their particular circumstances and goals.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (github link) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
 
 ```{=html}
-<!-- Discussion pts
-- we selected revelstoke because it is topographically complex, so a good challenge.
-Hardy result in eastern Canadian shield showed better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for.
-Stream crossing behaviour in Revelstoke landscape could be improved with more a detailed high resolution cost surface.
-But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain.
-None of these simple algorithms will project switchbacks on steep slopes.
-- Aspatial overall performance metrics can be difficult to interpret.
-To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies.
-Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results.
-Better harvest & existing road data would allow us to more distinguish errors in projections from errors in the input data (cite lit in coops paper)
- - Outside cutblocks, we saw very little variation in performance among our method variants - any of them are somewhat better than nothing, but they don't make a big differene, partly because in this example projected new roads over fairly short distances, and new landings were always fairly close to existing roads or other landings.
-Would be interesting to test in a context where targets are further from existing roads, but outside cutblocks our DLCP algorithm is very similar to Hardy's algorithm, so expect similar performance.
-Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm.
-Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this.
-But examination of spatial patterns clarifies that algorithms can get the right density without replicating the pattern or mechanism.
-Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid.
-Projected wiggles are quite different from observed switchbacks.
-- Focussed here on forest resource roads.
-Note the algorithms could also work in other situations where the objectives of road building are to access target locations and minimize cost.
-Not a useful algorithm for projecting linear features that were constructed for some other purpose such as seismic testing (e.g.
-Figure \@ref(fig:ftNelson)), or for circumstances where information about target locations is not available.
-- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do.
-But before investing in more complexity recognize that adequacy depends on the goal.
-For some purposes, projections from relatively crude cost models and simple algorithms may be adequate.
-For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates ().
-In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing).
-For wildlife response models informed by distance to road () or buffered disturbance footprint (), somewhat crude road projections are likely good enough.
-When constructing new wildlife response models intended for projection, it may be helpful to select these more easily projectable (Bodner) predictors.
-- There are a bewildering array of possible metrics of connectivity and fragmentation we did not consider in this paper.
-In general, performance will vary among metrics, and it is difficult determine what is good enough without reference to particular circumstances (objectives, species, landscapes).
-Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals.
-The best performance metrics will be those most closely linked to the outcome of interest.
-For example, where the outcome of interest is space use or connectivity projected from more (e.g.
-Whittington, Labadie) or less (samc, pither etc) species-specific movement models consider testing sensitivity of these models to the difference between observed and retrospectively projected roads.
-We hope R package and open workflow associated with this paper can help enable retrospective projection and relevant performance assessment in other contexts.
-However, we also recognize there is work associated with acquiring data to construct cost surfaces & retrospectively project roads.
-For those seeking a lower effort option, we provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.
-
-```
+
+Discuss possible other uses of algorithms - non forest roads, attributing age to segments as in lochhead, linking disconnected networks (but see coops et al refs)
+
+Discuss challenges on large landscapes & solutions...
+
 <!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice.
@@
 
-```{=html}
 <!--  What David had
@@
 Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network.
-We used four methods for solving the MTAP in the roads package.
-These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP), dynamic least cost paths (DLCP) approach and finally, the minimum spanning tree approach (MST).
-In this paper, we focus on the LCP, DLCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
-
-The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
-For each STAP, a least cost path from a single source to sink was solved using Dijkstra's shortest path algorithm (1959).
-Dijkstra's algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section "weighted graph").
-Witht the LCP approach each STAP is solved independently but for the DLCP approach each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction.
-This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP.
+We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by independent and dynamic least cost paths (LCP, DLCP) and finally, the minimum spanning tree approach (MST).
+
+The LCP and DLCP approaches solve the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
+For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm (1959) which minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel.
+A mathematical weighted graph consisting of nodes and edges is used to parameterize Dijkstra's algorithm wherein each node corresponds to a raster pixel which may be any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point that is a plausible location for road construction.
+Each node is connected with up to eight neighbours, and the cost of construction along each edge is weighted by the distance to the neighbouring node and the average of the construction cost at the source and destination nodes (or pixels) (Table \@ref(tab:cost-tbl)).
+The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design.
+
+The LCP and DLCP approaches solve each STAP sequentially, starting with target that is nearest to an existing road.
+In the DLCP algorithm, the existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target, whereas the LCP algorithm builds a connection to each target independently.
+The DLCP method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP but the projected network is sensitive to the ordering of the STAP.
 They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
@@
 Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
-Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph.
-The minimum spanning tree was solved using Kruskal's algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop.
-This method was most similar to Clark et al.
-(2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling.
-An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.
-
-## Weighted graph
-
-A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms.
-When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (1 ha resolution).
-Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network.
-Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point.
-The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up.
-The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies.
-The edges connecting the nodes follow an "octagonal" neighbourhood or allow a maximum of eight possible edges for a given node, with diagonal edges having a higher cost.
-Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design.
-Therefore, our road projection methods are for strategic modelling purposes only.
-Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs.
-When solving least cost paths, a cost surface raster was used.
-The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels.
-When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.
+Our MST approach consists of three steps for evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) create a weighted graph where each node is either a landing or a point on the existing road network and edge weights correspond to the total cost of the least cost path between pairs of nodes from the previous step, and 3) solve the minimum spanning tree for this graph.
+The minimum spanning tree is solved using Kruskal's algorithm (1956) where every node in the graph is linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop.
+Clark et al.
+(2000) also used a minimum spanning tree to project simple road networks during harvest scheduling, but they used Euclidean distance rather than least cost paths for edge weights.
+
+TO DO: add references to appendix.
 

### 2023-04-19 analysis/paper/paper.Rmd
writing in progress

@@
 ---
-title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife"
+title: "Implementation and assessment of methods for projecting forest road network development"
 author:
@@
   keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 editor_options: 
@@
 <!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->
+
+<!-- TO DO CJFR guidelines to note:
+* 6000 words, 50 references - need to remove ~30 refs
+* abstract 200 words
+* page #s
+* double spacing
+* affiliation - insitution, city, prov, country
+* CMJE definition of authorship - need CRedit author contribution statement
+* need funding statement, data availability statement, competing interests
+* review figure guidance
+-->
 
@@
 (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
-This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] and tools exist [@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate; @epstein_combinatorial_2006].
+This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012] and tools exist [e.g.
+@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate].
 However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
-Methods intended to support tactical and operational planning also require detailed parameterization with inputs that may not be publicly available, and employ algorithms that are too complex for application at regional, provincial or national scales.
-More promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
+Promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
 The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
@@
 
--->
-```
-
-<!-- Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations.
-Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes.
-Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network.
-In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially.
--->
-
 <!-- value of roadless areas: [@ibisch_a_2016; @talty_conservation_2020; @psaralexi_importance_2017] -->
@@
 
+-->
+```
+
 # Methods
@@
 
-We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because it is a mountainous areas with approximately 55% of non-forested land where road building is difficult or impossible (e.g.
-alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010].
-Thus, the area provides opportunities to observe the responses of road projection algorithms to steep complex terrain.
-Forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)), which allows projection of the forest road network from historic forest harvest data.
+We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land where road building is difficult or impossible [new tsa data package ref].
+Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain.
+Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data.
+The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River.
+At low to mid-elevations, the forests of the interior cedar-hemlock biogeoclimatic zone (bec ref) include remnants of rare old growth inland temperate rainforest (parks canada website).
+At higher elevations, forests are dominated by Englemann spruce and subalpine fir (bec ref).
+The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt (new tsa data package ref).
+Two of six
+
 Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021].
@@
 Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].
+
+<!--	TO DO: consider these comments from Kyle: Need a description about road development.
+In BC, forestry roads are developed by forest licensees under a timber harvesting permit.
+Various considerations go into road development including soils, grade, and economics.
+What type/classification of roads are you looking at (primitive roads, paved roads, etc).
+What’s the maximum grade?
+---can find a lot of this verbiage in the Interior Appraisal Manual.
+How are they maintained – do they last forever -see Auditor General Report
+•	How much of the Revelstoke TSA overlaps with caribou range? Elevation? Climate? Flora/Fauna?
+•	Maybe add a table of descriptive statistics presenting information about roads in the Revelstoke TSA and the other TSAs you mention?
+By Caribou Range?
+– road density, road length, cutblock area, cutblock density, etc
+
+-->
 
@@
 
-Road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC.
-To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set.
+Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia.
+To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads digital road atlas that were within 50 m of forest tenure roads.
 The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990.
-All roads from the DRA data set were assumed to be built before 1990.
+We assumed everything in the digital road atlas was built before 1990.
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021].
+The data does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021].
 Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
 
-We developed a simple model of road building cost informed by slope and major barriers (Table 1).
-Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)).
+Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)).
 Existing roads are assigned a cost of 0.
-The cost raster consists of 1,703,936 pixels with a resolution of 1 ha.
+The road construction cost raster consists of 1,703,936 pixels with a resolution of 1 ha.
 
@@
 These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository <https://osf.io/8vmqh/>.
-We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.
+We were not granted permission to share the digital road atlas so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.
 Scripts used for data acquisition and analysis associated with this paper are available on GitHub <https://github.com/LandSciTech/RoadPaper>.
@@
 
-We focus on methods that link multiple target locations to an existing road network for strategic purposes, described by Dean (1997) as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network.
-We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail; in this manuscript we focus on comparing the dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
-
-The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAP).
-For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
+We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
+Specifically, we compare dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+
+The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems.
+For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
 The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
@@
 The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design.
-The DLCP approach solves each STAP sequentially, starting with target that is nearest to an existing road.
-The existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target.
-The method was first developed by Anderson and Nelson (2004) who noted that the projected network is sensitive to the ordering of the STAP and recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
-
-The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
-Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
+The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.
+The method was first developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.
+
+The MST approach extends the LCP approach by providing a heuristic for linking landings and evaluating road branching opportunities.
+Dean [-@dean_finding_1997] showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
 The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network.
 These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
-The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node in graph with a tree that minimizes edge weights and does not include loops.
-This method differs from the minimum spanning tree method used by Clark et al.
-(2000) who used Euclidean distances rather than least costs for edge weights.
-
-The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
+The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node with a tree that minimizes edge weights and does not include loops.
+This method differs from Clark et al.
+[@clark_three-stage_2000] who used Euclidean distances for edge weights.
+
+The DLCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
 We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)).
@@
 
-TO DO: Sarah are these statements still correct? Please revise as necessary.
+<!-- TO DO: Sarah are these statements still correct? Please revise as necessary.-->
 
@@
 
-TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.
+<!-- TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->
 
@@
 ```{=html}
+<!-- The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries.
+In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).
+-->
 
@@
 
-<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice.
+<!-- TO DO: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice.
 They don't need to run the projections to do this - just summarize the outputs using a method of their choice.
--->
-
 <!--  What David had
-Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters.
-However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model.
-The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc.
-Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.
-
-Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data.
-Often major discrepancies were related to cutblocks that were likely incorrectly labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks.
-These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections.
-In order to get more accurate road projections, one could simply remove the cutblocks derived from LANDSAT as these were often the cutblocks which caused issues, however, these cutblocks were left in as a major goal of this assessment was to create road projections while adjusting input parameters as little as possible.
-Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue.
-*Another potential issue was with roads not being included in the data set that were seen in the satellite imagery.* (this is no longer identified in the results section since it wasn't a major discrepancy after fixing the road data source) This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets.
+
 Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc).
-This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery.
-These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn't have and should be treated with caution.
-It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets.
-Again, this would not present an issue when predicting future impacts on landscapes, only when assessing the model accuracy in this context.
-Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.
-
-Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines).
-Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc (\@ref(fig:ftNelson)).
-Because of this, forestry roads make up a much smaller percentage of the overall road network.
-In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors.
-However, other resource extraction activities could likely be incorporated into this model relatively easily.
-The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas.
-
-It is also worth noting that although the model can project road networks fairly accurately across a large landscape, it may differ from the way in which roads are planned in real life.
+
 The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries.
 In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).
-
-Certain projection methods were more accurate at projected some metrics compared to others.
-Similarly, certain metrics were sensitive to pixel size (road density, road presence) and others were not.
-Depending on what the goal of running the projections is, then the projection method and resolution may or may not play an important role.
-If you would like to predict road density as accurately as possible, then running the projections using as fine a resolution of cost surface as possible is the best option.
-If you're wanting to only predict roughly the location of the roads/save computing time and are more interested in a larger area, then a coarse resolution may suit these requirements perfectly fine.
-The tradeoffs of using a coarser resolution of cost surface may be less extreme when projecting roads across a flatter landscape compared to a mountainous landscape like was used for this paper since there would likely be less variability in cost across larger areas.
-
-With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which "roads" can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios.
-Because caribou are also considered an umbrella species within forests (Drever et al., 2019), projecting forestry impacts on their habitats may also provide valuable insight towards other key species as well.
-With the input of future cutblocks and the projections of forest road networks to these cutblocks, the results from the projection can provide a great outlook at the impacts of forestry scenarios on woodland caribou habitats using the forestry disturbance footprint metric.
-The distance to nearest road metric appears to be a very interesting metric to consider as well when looking specifically at larger areas.
-Using this metric to make projections at larger scales, such as at a TSA level, may provide very beneficial information about forestry impacts especially in a wildlife context when assessing the influence of forestry on various species with different tolerances to disturbance.
-
-Further improvements could be made that would increase the accuracy of the model.
-The cost surface used for this paper was developed at a provincial scale, however, a higher resolution cost surface with more detail could further improve the accuracy of the projection results.
-A finer resolution cost surface with more detail will ultimately create a more realistic road network and would possibly allow for more accuracy within cutblocks (where fine micro-topography, water tables, etc.
-will play a role in road development).
-The downside being that it would also increase computing time and may not always be necessary depending on the application of the model, especially when using the model for a large-scale landscape.
-Adding too much complexity to the cost surface may also take away from the goal of this model to be simple and easy to use, however, the algorithm itself would work on a very fine and detailed cost surface.
-Along with increased computing time, a key limitation of using this model on a finer cost surface would be that eventually cost pixels would need to have directionality associated with them, which is something this model does not take into account.
-The accuracy of comparing the projections to the observed network could also simply be improved by using higher quality/more up to date road and cutblock data.
-
-Assessing the performance of this model also presented some unique challenges with comparing two networks with each other across a landscape.
-We needed to carefully think about what we were trying to achieve.
-The goal wasn't to present a model that matched the observed road network pixel by pixel, but instead was to simulate disturbance across a large-scale landscape to an acceptable level while still being a relatively simple and straightforward tool.
-It's for this reason why metrics were chosen that are commonly used in road impact assessments on wildlife; such as the road density and forestry disturbance footprint metrics.
-Once the vector networks were converted into these raster layers, it also facilitated comparing observed versus projected as we could compare the means across the TSA and study where pixels matched and didn't match.
-Incorporating these metrics into landscape level simulations is also a realistic goal for this tool (ie projecting forest road density across a landscape) so a goal was to therefore see how well the model did at simulating these metrics when compared to the observed landscape.
 
@@
 For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.
-
--The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm.
-
-# Conclusion
-
-The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort.
-This model, however, is not suitable for strategic planning.
-The results from our projections in our study area overall matched up adequately with the observed present day forest road networks, with major discrepancies being attributed to issues in the available road and cutblock data.
-These discrepancies would not be issues in future landscape simulations, only when assessing accuracy.
-In areas where the quality of the data was high, projections matched well with the observed network.
-Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations.
-
-This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape.
-With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale.
-With land use activities transforming a large proportion of the planet's land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats.
-Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of \~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005).
-With only 7% of the planet's roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016).
-The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.
 -->
@@
 # Appendix
-
-## Road network projection method details
-
-Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes.
-Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network.
-We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by independent and dynamic least cost paths (LCP, DLCP) and finally, the minimum spanning tree approach (MST).
-
-The LCP and DLCP approaches solve the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP).
-For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm (1959) which minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel.
-A mathematical weighted graph consisting of nodes and edges is used to parameterize Dijkstra's algorithm wherein each node corresponds to a raster pixel which may be any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along each edge is weighted by the distance to the neighbouring node and the average of the construction cost at the source and destination nodes (or pixels) (Table \@ref(tab:cost-tbl)).
-The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design.
-
-The LCP and DLCP approaches solve each STAP sequentially, starting with target that is nearest to an existing road.
-In the DLCP algorithm, the existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target, whereas the LCP algorithm builds a connection to each target independently.
-The DLCP method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP but the projected network is sensitive to the ordering of the STAP.
-They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
-
-The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities.
-Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
-Our MST approach consists of three steps for evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) create a weighted graph where each node is either a landing or a point on the existing road network and edge weights correspond to the total cost of the least cost path between pairs of nodes from the previous step, and 3) solve the minimum spanning tree for this graph.
-The minimum spanning tree is solved using Kruskal's algorithm (1956) where every node in the graph is linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop.
-Clark et al.
-(2000) also used a minimum spanning tree to project simple road networks during harvest scheduling, but they used Euclidean distance rather than least cost paths for edge weights.
-
-TO DO: add references to appendix.
 

### 2023-04-21 analysis/paper/paper.Rmd
writing in progress

@@
 ---
-title: "Implementation and assessment of methods for projecting forest road network development"
+title: "Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife"
 author:
@@
 bibliography: Roads.bib
-csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
+csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
 abstract: |
-  Text of abstract
+  abstract to do
 keywords: |
@@
 Despite the importance of roads for wildlife, very few of these projections have included road networks.
-For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].
+For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].
 
@@
 (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
-This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012] and tools exist [e.g.
-@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate].
+This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g.
+@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist.
 However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
@@
 
-We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land where road building is difficult or impossible [new tsa data package ref].
+We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is difficult or impossible.
 Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain.
@@
 The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River.
-At low to mid-elevations, the forests of the interior cedar-hemlock biogeoclimatic zone (bec ref) include remnants of rare old growth inland temperate rainforest (parks canada website).
-At higher elevations, forests are dominated by Englemann spruce and subalpine fir (bec ref).
-The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt (new tsa data package ref).
-Two of six
-
-Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021].
-The area is home to wildlife of concern including including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010].
-Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020].
+At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_and_range_biogeoclimatic_nodate] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017].
+At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_and_range_biogeoclimatic_nodate].
+The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022].
+Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020;@british_columbia_ministry_of_forests_revelstoke_2022;@environment_canada_recovery_2014].
 Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022].
@@
 – road density, road length, cutblock area, cutblock density, etc
-
 -->
@@
 Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)).
-The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design.
+The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007].
 The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.
-The method was first developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.
+The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.
 
@@
 These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
-The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node with a tree that minimizes edge weights and does not include loops.
+The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights and does not include loops.
 This method differs from Clark et al.
 [@clark_three-stage_2000] who used Euclidean distances for edge weights.
+
+TO DO: talk about key differences from Hardy algorithm.
+TO DO: talk about implications of algorithms for speed/computational cost.
 
@@
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
-Within all mapped cutblocks (Figure \@(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road.
-Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@(fig:aspat-perf-fig-all)).
+Within all mapped cutblocks (Figure \@ref(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road.
+Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@ref(fig:aspat-perf-fig-all)).
 These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
@@
 Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results.
-These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+
 Outside cutblocks, we saw very little variation in performance among projection method variants.
@@
 
+In this paper our focus is tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife.
+However, we note there are other possible uses for these road network projection methods.
+Cost minimizing methods are not suitable for projecting development of seismic lines (Figure \@ref(fig:ftNelson)), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
+Lochhead et al.
+[-@lochhead_linking_2022] used the least cost path algorithm to reconstruct the likely history of road network development in a case where harvest dates were known but road construction dates were not.
+In principle, these algorithms could also be used to link disconnected road segments into a network, but local route finding algorithms [@roussel_vectorial_2023] that do not require construction of large graphs are better suited to this purpose.
+More obviously, spatial stochastic models are used to project impacts of disturbance on a variety of values other than wildlife [@albrich_simulating_2020; @hof_editorial_2021; @sturtevant_understanding_2021].
+It would also be interesting to account for persistent impacts of resource roads on vegetation, productivity and carbon sequestration [@braham_characterization_2023].
+
 ```{=html}
@@
 
-Discuss possible other uses of algorithms - non forest roads, attributing age to segments as in lochhead, linking disconnected networks (but see coops et al refs)
-
 Discuss challenges on large landscapes & solutions...
@@
 They don't need to run the projections to do this - just summarize the outputs using a method of their choice.
-<!--  What David had
-
-Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc).
-
-The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries.
-In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).
-
-Many spatial stochastic models don't include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010).
-Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these.
-For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.
 -->

### 2023-04-24 analysis/paper/paper.Rmd
SE initial edit

@@
 In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
-Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020[a]], and genetics [@habrich_varying_2021].
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
@@
 Leston et al.
-(2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
+(2020) projected forest [b]road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
 This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g.
@@
 
-To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package.
+To help address the need for adequately accurate forest[c] road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package.
 We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia.
-The adequacy of a road network projection methods depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose.
+The adequacy of a road network projection method depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose.
 Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023; @stewart_climate-informed_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
@@
 Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia.
-To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads digital road atlas that were within 50 m of forest tenure roads.
+To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads.
 The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990.
@@
 The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
-Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footpint was observed.
+Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed.
 These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
@@
 Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road.
-However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuousity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
+However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
 None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
@@
 Another fundamental limitation of these methods is that they generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse coarse (Figure A-\@ref(fig:figA1)).
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure A-\@ref(fig:figA1)).
 In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).
@@
 Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a).
-Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy.
+Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms would likely reduce this discrepancy.
 Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks.
@@
 Water crossing behaviour in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains.
-A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density becomes less accurate as raster cell resolution increases.
+A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density become less accurate as raster cell resolution increases.
 
@@
 There are many metrics of connectivity and fragmentation [@francis_routledge_2021] we did not consider in this paper.
-In general, performance will vary among metrics, and it is difficult determine what is good enough without reference to particular circumstances (i.e.
+In general, performance will vary among metrics, and it is difficult to determine what is good enough without reference to particular circumstances (i.e.
 objectives, species, landscapes).
@@
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
-Rather than adding additional performance metrics to this paper, we have focussed our efforts on development of an open source R package [@endicott_roads_2023] and workflows <https://github.com/LandSciTech/RoadPaper> that we hope will help enable users devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (github link) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
+Rather than adding additional performance metrics to this paper, we have focussed our efforts on development of an open source R package [@endicott_roads_2023] and workflows <https://github.com/LandSciTech/RoadPaper> that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (<https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/calc_any_metric.R>) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
 
@@
 ```
+
+[a]This and a few other spots is a lot of citations for the same fact.
+I have been told before not more that 3.
+It would help reduce overall refs to remove any that are not also used else where.
+If you want to be clear that it is not an exhaustive list I have seen others use (E.g Ref 1, Ref2)
+[b]This feels a bit backwards since road projection for harvesting is it's own thing not really something that needs to be highlighted by a wildlife paper example.
+I would restructure more as "While road projection tools are not typically used in forest landscape models, there are existing tools for operational forest road planning and design.
+These have occasionally been incorporated into forest landscape models (Leston et al) but there are x reasons why this is not ideal.
+" and then carry on about promising options
+[c]I feel like the concept of being "adequately accurate" is fairly key to the whole paper and should be defined and explained in the previous paragraph in relation to why the operational tools are not as useful for landscape modelling.

### 2023-04-24 analysis/paper/paper.Rmd
rename scripts tidy up comments

@@
 In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
-Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020[a]], and genetics [@habrich_varying_2021].
+Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021].
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
@@
 Leston et al.
-(2020) projected forest [b]road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
+(2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
 This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g.
@@
 
-To help address the need for adequately accurate forest[c] road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package.
+To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package.
 We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia.
@@
 Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
-
-<!-- TO DO: Sarah are these statements still correct? Please revise as necessary.-->
 
@@
 Rather than adding additional performance metrics to this paper, we have focussed our efforts on development of an open source R package [@endicott_roads_2023] and workflows <https://github.com/LandSciTech/RoadPaper> that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (<https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/calc_any_metric.R>) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (<https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/4_calc_any_metric.R>) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/1_prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
 
@@
 
-<!-- TO DO: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice.
-They don't need to run the projections to do this - just summarize the outputs using a method of their choice.
--->
 ```
@@
 ```
-
-[a]This and a few other spots is a lot of citations for the same fact.
-I have been told before not more that 3.
-It would help reduce overall refs to remove any that are not also used else where.
-If you want to be clear that it is not an exhaustive list I have seen others use (E.g Ref 1, Ref2)
-[b]This feels a bit backwards since road projection for harvesting is it's own thing not really something that needs to be highlighted by a wildlife paper example.
-I would restructure more as "While road projection tools are not typically used in forest landscape models, there are existing tools for operational forest road planning and design.
-These have occasionally been incorporated into forest landscape models (Leston et al) but there are x reasons why this is not ideal.
-" and then carry on about promising options
-[c]I feel like the concept of being "adequately accurate" is fairly key to the whole paper and should be defined and explained in the previous paragraph in relation to why the operational tools are not as useful for landscape modelling.

### 2023-04-24 analysis/scripts/exploratory/reconstructRoadYear.R

@@ -69,3 +69,7 @@ write_sf(newSim$roads, here(data_path_drvd, "reconstructedRoads.gpkg"))
 
 
 plot(newSim$roads["Year"])
+
+# this didn't work that well becasue the reconstructed roads do not follow the
+# observed roads as well as expected. Maybe need cost of existing roads to be
+# closer to 0? or might just be related to rasterization and resolution

### 2023-04-24 analysis/paper/paper.Rmd
Sarah edits to intro

@@
 Despite the importance of roads for wildlife, very few of these projections have included road networks.
-For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].
+For example, in a recent collection of articles focused on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].
 
-Leston et al.
-(2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire.
-This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g.
-@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist.
-However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
+While road projection tools are not typically used in forest landscape models, they are important for harvest planning, and methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g.
+@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist to accomplish this task in an operational context.
+These have occasionally been incorporated into forest landscape models [@leston_quantifying_2020], however they were not designed for this purpose.
+The algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] and the level of detail required to parameterize them is often difficult to obtain at landscape scales.
+Harvest road planning tools likely produce the most accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.
+
 Promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].

### 2023-04-25 analysis/paper/paper.Rmd
Add descrip of dif btw algoritms and comp and mem costs

@@
 
-<!-- TO DO: number analysis scripts to clarify workflow -->
-
 ## Road Projection Methods and Variants
@@
 
-The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems.
+The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
 For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
@@
 
-TO DO: talk about key differences from Hardy algorithm.
-TO DO: talk about implications of algorithms for speed/computational cost.
+Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our DLCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network.
+One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
+On the other hand, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
+However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network.
+It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network.
+Hardy's method also has a different approach for building roads inside cutblocks, it iterates over each cell in the harvest area and builds a road to it if it is not within the skidding distance of an existing road.
+In our algorithm we create landing points in the harvested area to serve as the starting points for the roads to be built based on an expected landing density.
+
+<!--
+TO DO: should this paragraph about speed and memory go in the results or discussion rather than methods?
+-->
+The MST method is expected to be closer to an optimal solution than either of the LCP methods described above but at the cost of being more computationally intensive and taking longer to solve.
+The Hardy QGIS plugin was much faster to run than our R package and more robust to analyzing large landscapes.
+The `igraph` package that we used to implement the Dijkstra algorithm requires the values of the cost raster to be read into memory in R which can lead to high memory usage on larger landscapes.
+See Appendix X for a more detailed exploration of the speed and memory requirements of the different methods.
 

### 2023-04-26 analysis/paper/paper.Rmd
writing in progress

@@
 @remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist to accomplish this task in an operational context.
-These have occasionally been incorporated into forest landscape models [@leston_quantifying_2020], however they were not designed for this purpose.
+These have occasionally been incorporated into forest landscape models [@leston_quantifying_2020] but they were not designed for that purpose.
 The algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] and the level of detail required to parameterize them is often difficult to obtain at landscape scales.
-Harvest road planning tools likely produce the most accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.
+Harvest road planning tools may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.
 
@@
 
-```{r study-area, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel.", cache=TRUE}
+```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table \@ref(tab:cost-tbl)) in the Revelstoke timber supply area in British Colombia, Canada.", cache=TRUE}
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
@@
 One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
-On the other hand, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
+In contrast our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
 However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network.
 It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network.
-Hardy's method also has a different approach for building roads inside cutblocks, it iterates over each cell in the harvest area and builds a road to it if it is not within the skidding distance of an existing road.
-In our algorithm we create landing points in the harvested area to serve as the starting points for the roads to be built based on an expected landing density.
+Within cutblocks, Hardy's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
 
@@
 -->
-The MST method is expected to be closer to an optimal solution than either of the LCP methods described above but at the cost of being more computationally intensive and taking longer to solve.
-The Hardy QGIS plugin was much faster to run than our R package and more robust to analyzing large landscapes.
+The MST method is expected to be closer to an optimal solution than iterative least cost path methods, and also more computationally costly.
+The Hardy QGIS plugin is much faster than our R package, and better able to handle large landscapes.
 The `igraph` package that we used to implement the Dijkstra algorithm requires the values of the cost raster to be read into memory in R which can lead to high memory usage on larger landscapes.
@@
 
-```{r proj-meth-fig, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and the Hardy QGIS plugin road network projection algorithms.
+```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), dynamic least cost paths (DLCP), and the Hardy QGIS plugin road network projection algorithms.
 Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2).
 Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red.
-Cost (1 ha resolution) is indicated by grey shading, with darker grey indicating higher cost.
-The large four sided cutblock in the top right was added to give a better example of the algorithms behaviour inside cutblocks."}
+Cost (1 ha resolution) is indicated by grey shading.
+The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks."}
 knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
@@
 
-```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster.
-Metrics were summarised within cutblocks, outside cutblocks, and overall.
-Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
-Cutblocks that did not intersect the observed road network were excluded"}
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
+Values greater than zero indicate overprojection and negative values indicate underprojection.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
 # organize data for ggplot
@@
 
-```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network.
-The primary map shows the observed distance to the nearest road minus the projected distance and the insets show close-ups of the road networks with satellite imagery to give more context.
-a) Projected and observed road networks follow similar patterns although projected roads cross valley bottoms more frequently.
-b) An example of a misidentified cutblock.
-c) Projected roads do not match observed switchback roads.
-The satellite images are from Google (2022)."}
+```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network.
+The main map shows observed distance to the nearest road minus the projected distance.
+Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected methods do no mimic observed switchbacks."}
 
@@
 
-```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
-F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster).
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
 Values closer to one indicate better performance.
-Cutblocks that did not intersect the observed road network were excluded"}
+Cutblocks not accessible via the observed road network were excluded from analysis."}
 doSpatPerf <- function(tab){

### 2023-04-26 analysis/paper/paper.Rmd
Writing in progress.

@@
 abstract: |
-  abstract to do
+  Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available.
+Hardy et al.
+[-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension.
+As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
+Comparison of these methods in a mountainous region of British Columbia clarifies that performance is adequate according to some metrics but not others, that aspatial performance metrics obscure important differences in alorithm behaviour within and outside cutblocks, and that none of these algorithms mimic switchbacks in steep terrain.
+Method assessment is complicated by errors in available in cutblock and forest road data.
+Simple iterative least cost path road network simulation methods may be adequate for projecting important impacts on wildlife across large areas, and we hope that more easily useable open source implementations will help integrate resource roads into forest change projections.
 keywords: |
@@
 
-```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table \@ref(tab:cost-tbl)) in the Revelstoke timber supply area in British Colombia, Canada.", cache=TRUE}
+```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada.", cache=TRUE}
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))

### 2024-03-11 analysis/paper/paper.Rmd
revision in progress

@@
     correspondence: false
+  - name: Gregory Paradis
+    email: gregory.paradis@ubc.ca
+    institute: [UBC]
+    correspondence: false
   - name: Kyle Lochhead
@@
     institute: [FLNR]
-    correspondence: false
-  - name: Gregory Paradis
-    email: gregory.paradis@ubc.ca
-    institute: [UBC]
     correspondence: false
@@
 tsa <- tm_shape(tsaCost, name = "Cost")+
-  tm_raster(style = "cont", title = "Cost")+
+  tm_raster(style = "cont", title = "Cost",palette="Greys")+
   tm_shape(st_union(tsaBoundary))+
@@
 library(caribouMetrics)
-library(pfocal)
+#library(pfocal)
 
@@
   1:nrow(allProj),
-  ~ qtm(tsaCost, raster.style = "cont", raster.palette = "Greys",
-        raster.alpha = 0.5, raster.title = "Cost")+
-    qtm(cutblocks, fill = NULL, borders = "grey40", borders.lwd = 1.5)+
-    qtm(roads, lines.col = "red", lines.lwd = 2)+
-    qtm(read_sf(here(allProj$output[[.x]])), lines.col = "yellow", lines.lwd = 2)+
+  ~ qtm(cutblocks, borders = "#92c5de", fill="#92c5de", borders.lwd = 1,)+
+    qtm(tsaCost, raster.style = "cont", raster.palette = "Greys",
+        raster.alpha = 0.25, raster.title = "Cost")+
+    qtm(roads, lines.col = "#0571b0", lines.lwd = 2,lines.lty="solid")+
+    qtm(read_sf(here(allProj$output[[.x]])), lines.col = "#ca0020", lines.lwd = 2)+
     qtm(exRoads, lines.col = "black", lines.lwd = 2)+
@@
                 labels = c("Existing roads", "Projected roads", "Observed roads"),
-                col = c("black", "yellow", "red"))+
+                col = c("black", "#ca0020", "#0571b0"),lty=c("solid","solid","solid"))+
     tm_add_legend(type = "fill", 
                 labels = "Cutblocks",
-                col = NA, border.col = "grey40")+
+                col = "#92c5de", border.col ="#92c5de")+
   tm_layout(legend.only = TRUE, legend.position = c("center", "center"))

### 2024-03-11 R/functionsSource.R

@@ -5,18 +5,41 @@
 
 # functions for roads
 
-
 # rasterize line density
+rasterizeLineDensity <- function(x, r, ptDensity = 1) {
+  if(any(c("POINT", "MULTIPOINT") %in%
+         sf::st_geometry_type(x, by_geometry = TRUE))){
+    lfPt <- sf::st_collection_extract(x, "POINT")
+    x <- sf::st_collection_extract(x, "LINESTRING")
+  } else {
+    lfPt <- slice(x, 0)
+  }
 
-rasterizeLineDensity <- function(x, r) {
-   spst_im <- spatstat.geom::pixellate(x = spatstat.geom::as.psp(sf::st_geometry(x)),
-                                      W = maptools::as.im.RasterLayer(r),
-                                      DivideByPixelArea = F)
-  spst_rast <- raster::raster(spst_im)/(res(r)[1]*res(r)[2]/10000)
-  spst_rast <- round(spst_rast, digits = 1)
-  spst_rast <- raster::`crs<-`(spst_rast, value = raster::crs(r))
+  line_len <- terra::rasterizeGeom(terra::vect(x), r, fun = "length")
 
-  return(terra::rast(spst_rast))
+  cell_area <- terra::cellSize(r)/10000
+
+  r <- round(line_len/cell_area, digits = 1)
+
+
+  if(!is.null(ptDensity)){
+    if(ptDensity > 2+2*2^0.5){
+      warning("ptDensity is greater than the expected max of 4.828.",
+              " see ?rasterizeLineDensity for details",
+              call. = FALSE)
+    }
+
+    if(nrow(lfPt) > 0){
+      lfR <- terra::rasterizeGeom(terra::vect(lfPt), r, fun = "count")
+
+      lfR <- lfR * ptDensity * terra::res(r)[1]
+
+      lfR <- round(lfR / (terra::res(r)[1] * terra::res(r)[2] / 10000), digits = 1)
+      r <- r + lfR
+    }
+  }
+
+  return(r)
 }
 
 roadDisturbanceFootprint <- function(x, r, b) {

### 2024-03-11 analysis/scripts/5_make_fig4.Rmd
Update fig 4

Done in QGIS and PowerPoint to keep it simple

In QGIS open input_cutblocks and input_roads for TSA 27 as well as combined_revelstoke_roads.  Get Google satellite imagery from XYZ tile service.

Assign colors to as follows
Existing roads: black
Projected roads: #ca0020
Observed roads: #0571b0
Cutblocks: #92c5de

Then capture images in the map views desired by clicking Project > Import/Export > Export map as image,
setting the dpi to 300 and saving as png.

To make the raster map we need to manipulate some of the rasters.

```{r}
library(terra)
library(sf)
library(dplyr)
library(tmap)

data_path_raw <- "analysis/data/raw_data/"
data_path_drvd <- "analysis/data/derived_data/TSA27"

dist_obs <- rast(here::here(file.path(data_path_drvd, "distanceToRoadobserved_NA_NA.tif")))

dist_proj <- rast(here::here(file.path(data_path_drvd, "distanceToRoadregular_1e-05_mst.tif")))

dist_dif <- dist_obs - dist_proj

# need to clip/mask everything to the tsa boundary to avoid problems outside tsa

tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg")) %>% st_union()

dist_dif <- mask(dist_dif, vect(tsaBoundary))

e_rds <- st_read(here::here(file.path(data_path_drvd, "input_roads.gpkg")))
p_rds <- st_read(here::here(file.path(data_path_drvd, "regular_1e-05_mst.gpkg")))
o_rds <- st_read(here::here(file.path(data_path_drvd, "../combined_revelstoke_roads.gpkg")))

rds <- lst(e_rds, p_rds, o_rds) %>% 
  purrr::map(st_intersection, tsaBoundary)

mp <- tm_shape(dist_dif)+
  tm_raster(palette = "BrBG", breaks = c(-2000, -1000, 1000, 2000), 
            midpoint = 0, style = "cont", 
            labels = c("<= -2000", "     -1000", "      1000", ">=  2000"),
            title = "Observed - projected\ndistance to road")+
  tm_shape(rds$o_rds)+
  tm_lines(col = "#0571b0")+
  tm_shape(rds$p_rds)+
  tm_lines(col = "#ca0020")+
  tm_shape(rds$e_rds)+
  tm_lines(col = "black")+
  tm_layout(legend.position = c("LEFT", "BOTTOM"), legend.just = "left", frame = FALSE)+
  tm_add_legend(type = "line", col = c("black", "#ca0020", "#0571b0"), 
                labels = c("Existing roads", "Projected roads", "Observed roads"))+
  tm_add_legend(type = "fill", col = "white", border.col = "#92c5de", labels = "Cutblocks")

tmap_save(mp,
          here("analysis/figures/spatial_projections.png"),
          dpi = 300, height = 7, width = fig_widths["one"])

```

Then arrange the images in powerpoint, modifying the contrast slightly to make the colours show on the satellite imagery. Save to png for inclusion in Rmd and pdf for final product.

### 2024-03-12 analysis/paper/paper.Rmd
Update fig 4
change to ILCP throughout

@@
 We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
-Specifically, we compare dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
-
-The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
+Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+
+The ILCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
 For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
@@
 
-Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our DLCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network.
+Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network.
 One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
@@
 
-The DLCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
+The ILCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
 We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)).
@@
 MST|Centroid|N/A|1 ha and 100 ha
-DLCP|Regular|10|1 ha 
+ILCP|Regular|10|1 ha 
 Hardy QGIS|N/A|~9|1 ha
@@
 
-To help make these road network projection methods more easily and widely useable, we implemented the DLCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
+To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
 The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
@@
 
-dlcpProj <- projectAll(tsbs = tsb,
+ilcpProj <- projectAll(tsbs = tsb,
                       paramTable = paramTable %>% 
                         filter(sampleType == "regular") %>% 
-                        mutate(method = "dlcp"),
+                        mutate(method = "ilcp"),
                       costSurface = tsaCost,
@@
 
-allProj <- bind_rows(mstProj, dlcpProj, .id = "method") %>%
+allProj <- bind_rows(mstProj, ilcpProj, .id = "method") %>%
   arrange(desc(sampleType), sampleDens)
@@
   mutate(mapTitle =  paste(ifelse(method == 1, "MST",
-                                  ifelse(method == 2, "DLCP", method)),
+                                  ifelse(method == 2, "ILCP", method)),
                            sampleType,
@@
 
-```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), dynamic least cost paths (DLCP), and the Hardy QGIS plugin road network projection algorithms.
+```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), iterative least cost paths (ILCP), and the Hardy QGIS plugin road network projection algorithms.
 Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2).
@@
                              "mst regular high sample density",
-                             "dlcp regular high sample density",
+                             "ilcp regular high sample density",
                              "NA NA klementQGIS",
@@
                              "MST regular high density",
-                             "DLCP regular high density",
+                             "ILCP regular high density",
                              "Hardy QGIS",
@@
 # save runTimes for later comparison
-dlcp_time <- meanTableRealCuts %>% 
-  filter(method == "dlcp", areaMean == "overall") %>% 
+ilcp_time <- meanTableRealCuts %>% 
+  filter(method == "ilcp", areaMean == "overall") %>% 
   pull(runTime) %>% round(2)
@@
 
-knitr::include_graphics(here("analysis/figures/Figure4_diff_map_update.png"))
+knitr::include_graphics(here("analysis/figures/Figure4_map_update.png"))
 
@@
                                "mst regular high density",
-                               "dlcp regular high density",
+                               "ilcp regular high density",
                                "NA klementQGIS", 
@@
                                "MST regular high density",
-                               "DLCP regular high density",
+                               "ILCP regular high density",
                                "Hardy QGIS", 
@@
 All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
-The DLCP projection method had comparable performance to the MST projection method with the same sampling strategy.
-The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.
-
-<!-- TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->
+The ILCP projection method had comparable performance to the MST projection method with the same sampling strategy.
+The ILCP method was faster to compute however, with a run time of `r ilcp_time` minutes compared to `r mst_time` minutes for the MST method.
+
+<!-- TO DO: Sarah figure out why `r ilcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->
 
@@
 All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads.
-Outside cutblocks our DLCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads.
+Outside cutblocks our ILCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads.
 Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm.
@@
 meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
-  filter((method != "dlcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))
+  filter((method != "ilcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))
 
 matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
-  filter(method != "dlcp", sampleType != "cutOnly")
+  filter(method != "ilcp", sampleType != "cutOnly")
 
@@
                              "mst regular high sample density",
-                             "dlcp regular high sample density",
+                             "ilcp regular high sample density",
                              "NA klementQGIS", 
@@
                              "mst regular high density",
-                             "dlcp regular high density",
+                             "ilcp regular high density",
                              "NA klementQGIS"),
@@
                              "Regular high density",
-                             "DLCP regular high density",
+                             "ILCP regular high density",
                              "Hardy QGIS")),

### 2024-03-12 analysis/scripts/2_runProjections.R

@@ -80,8 +80,8 @@ paramTable <- tibble(sampleType, sampleDens, method = "mst",
                      forestryDisturbance = vector("list", length(sampleDens))) %>%
   distinct()
 
-# add row for dlcp with regular high density sampling
-paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "dlcp"))
+# add row for ilcp with regular high density sampling
+paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "ilcp"))
 
 #
 run_projections(

### 2024-03-12 analysis/scripts/exploratory/testRunProjections.R

@@ -37,9 +37,9 @@ paramTable <- tibble(sampleType, sampleDens, method = "mst",
                      forestryDisturbance = vector("list", length(sampleDens))) %>%
   distinct()
 
-# add row for dlcp with regular high density sampling
+# add row for ilcp with regular high density sampling
 paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>%
-                          mutate(method = "dlcp"))
+                          mutate(method = "ilcp"))
 
 
 run_projections(

### 2024-03-12 analysis/paper/paper.Rmd
more update citations for data/repo

@@
 
+```{=html}
 <!-- TO DO CJFR guidelines to note:
@@
 -->
-
+```
 Keywords: `r rmarkdown::metadata$keywords`
@@
 ```
-
 # Methods
@@
 The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022].
-Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020;@british_columbia_ministry_of_forests_revelstoke_2022;@environment_canada_recovery_2014].
+Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014].
 Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022].
@@
 
+```{=html}
 <!--        TO DO: consider these comments from Kyle: Need a description about road development.
@@
 -->
-
+```
 ```{r setup2}
@@
 All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].
-These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository <https://osf.io/8vmqh/>.
-We were not granted permission to share the digital road atlas so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.
-Scripts used for data acquisition and analysis associated with this paper are available on GitHub <https://github.com/LandSciTech/RoadPaper>.
+These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository [@endicott2024a].
+We were not granted permission to share the digital road atlas so it must be accessed from the B.C.
+Data Catalogue [@governmentofbritishcolumbia].
+Scripts used for data acquisition and analysis associated with this paper are available on GitHub [@endicott2024].
 
@@
 We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
-Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
 
@@
 
-Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network.
+Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network.
 One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
@@
 
+```{=html}
 <!--
@@
 -->
+```
 The MST method is expected to be closer to an optimal solution than iterative least cost path methods, and also more computationally costly.
@@
 The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-The package is available on CRAN (The Comprehensive R Archive Network).
-The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>), along with tutorials and documentation (<https://landscitech.github.io/roads/>).
+The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
 
@@
 ## Aspatial performance
+
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
@@
 Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm.
-The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid.
+The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations \>200 m away on a 100 m grid.
 One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.
@@
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
-Rather than adding additional performance metrics to this paper, we have focussed our efforts on development of an open source R package [@endicott_roads_2023] and workflows <https://github.com/LandSciTech/RoadPaper> that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (<https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/4_calc_any_metric.R>) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/1_prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
+Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott2024a] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
 
@@
 Discuss challenges on large landscapes & solutions...
-
 ```

### 2024-03-12 analysis/scripts/6_benchmark_methods.R
start benchmark script for running in cloud

@@ -1,0 +1,76 @@
+# Compare speeds and memory usage for different methods
+
+# Plan: run on cloud on separate nodes
+# First crop cost layer and dissaggregate so that we can then compare more levels of aggregation
+# in Previous runs ~ 1 hour was max for non aggregated raster
+# only compare one density level
+# Also record memory usage. Using separate nodes will ensure to memory overlap between runs
+
+row_ind <- commandArgs(trailingOnly = TRUE)
+
+#set path for data
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
+
+###### Source scripts ######
+# load in required libraries
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+
+
+# load functions used in script
+devtools::load_all()
+
+# Prepare disaggregated cost surface
+# bc_cost <- rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
+#
+# tsaBoundary <- read_sf(paste0(data_path_raw, "tsa27_boundaries.gpkg"))
+#
+# rev_cost <- crop(bc_cost, vect(tsaBoundary), snap = "out")
+#
+# rev_cost10 <- disagg(rev_cost, 10,
+#                      filename = file.path(data_path_drvd, "revelstoke_cost_10.tif"))
+
+param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
+                         stringsAsFactors = FALSE)
+
+method <- param_tbl$method[row_ind]
+agg <- param_tbl$agg[row_ind]
+sampleDens <- 0.00001
+
+inputs <- prepInputs(
+  cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
+  roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
+  tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
+  costPth = paste0(data_path_drvd, "revelstoke_cost_10.tif"),
+  outPth = paste0(data_path_drvd, "TSA27/"),
+  aggFact = agg, #factor of aggregation of cost surface. 1 = no aggregation.
+  saveInputs = FALSE
+)
+
+
+landings <- getLandingsFromTarget(inputs$cutblocks,
+                                     landingDens = sampleDens,
+                                     sampleType = "regular")
+
+# record memory, time, graph size and total cost
+prof <- peakRAM::peakRAM(
[…]

### 2024-03-12 R/functionsSource.R
move supplement to separate file

@@ -541,3 +541,133 @@ writeRastNotNull <- function(rast, filename){
   terra::writeRaster(rast, filename, overwrite = TRUE)
 
 }
+
+doAspatPlot <- function(tab){
+  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>%
+    pivot_longer(-c(sampleType, sampleDens, areaMean, method),
+                 names_to = "metric", values_to = "response")%>%
+    mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_,
+                               sampleType),
+           sampleDens = paste(method, sampleType, sampleDens) %>%
+             factor(levels = c("mst NA centroid",
+                               "NA NA observed",
+                               "mst random low sample density",
+                               "mst regular low sample density",
+                               "mst random high sample density",
+                               "mst regular high sample density",
+                               "ilcp regular high sample density",
+                               "NA NA klementQGIS",
+                               "NA NA NA"),
+                    labels = c("MST centroid",
+                               "Observed",
+                               "MST random low density",
+                               "MST regular low density",
+                               "MST random high density",
+                               "MST regular high density",
+                               "ILCP regular high density",
+                               "Hardy QGIS",
+                               "Cutblocks only")),
+           metric = factor(metric,
+                           levels = c("roadDensityMean", "roadPresenceMean",
+                                      "distanceToRoadMean",
+                                      "forestryDisturbanceMean",
+                                      "roadDisturbanceMean"),
+                           labels = c("Road\ndensity", "Road\npresence",
+                                      "Distance\nto road",
+                                      "Forestry\ndisturbance\nfootprint",
+                                      "Road\ndisturbance\nfootprint")))
+
+  observed_values <- filter(meanTable_long, sampleDens == "Observed")
+  projected_values <- filter(meanTable_long, sampleDens != "Observed")
+
+  # get proportional difference from observed
+  prop_dif <- projected_values %>%
+    left_join(observed_values %>% select(areaMean, metric, response),
+              by = c("areaMean", "metric"),
+              suffix = c("_proj", "_obs")) %>%
+    mutate(prop_dif = (response_proj-response_obs)/response_obs,
+           areaMean = factor(areaMean, labels = c("Cutblocks", "Outside\ncutblocks",
[…]

### 2024-03-13 analysis/paper/paper.Rmd

@@
 The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
+The package is available on CRAN (The Comprehensive R Archive Network).
+The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>), along with tutorials and documentation (<https://landscitech.github.io/roads/>).
 
@@
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
-Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott2024a] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
+Rather than adding additional performance metrics to this paper, we have focussed our efforts on development of an open source R package [@endicott_roads_2023] and workflows <https://github.com/LandSciTech/RoadPaper> that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps <https://osf.io/8vmqh/> and an example script (<https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/4_calc_any_metric.R>) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script <https://github.com/LandSciTech/RoadPaper/blob/master/analysis/scripts/1_prepareData.R> that could be modified to extract required input data for other areas in British Columbia.
 

### 2024-03-13 analysis/scripts/5_make_fig4.Rmd

@@
 mp <- tm_shape(dist_dif)+
-  tm_raster(palette = "BrBG", breaks = c(-2000, -1000, 1000, 2000), 
+  tm_raster(palette = "PuOr", breaks = c(-2000, -1000, 1000, 2000), 
             midpoint = 0, style = "cont", 
+            contrast = 0.5,
             labels = c("<= -2000", "     -1000", "      1000", ">=  2000"),
@@
   tm_layout(legend.position = c("LEFT", "BOTTOM"), legend.just = "left", frame = FALSE)+
-  tm_add_legend(type = "line", col = c("black", "#ca0020", "#0571b0"), 
+  tm_add_legend(type = "line", col = c("black", "#ca0020", "#0571b0"), lwd = 2,
                 labels = c("Existing roads", "Projected roads", "Observed roads"))+
-  tm_add_legend(type = "fill", col = "white", border.col = "#92c5de", labels = "Cutblocks")
+  tm_add_legend(type = "fill", col = "white", border.col = "#92c5de", labels = "Cutblocks",
+                border.lwd = 2)
 

### 2024-03-13 analysis/cloud/make.R
add cloud stuff

@@ -1,0 +1,40 @@
+#' Road Paper cloud benchmarking
+#'
+#' @description
+#' Benchmarking the roads paper algorithms on the cloud
+#'
+#' @author Sarah Endicott \email{sarah.endicott@canada.ca}
+#'
+#' @date 2024/03/12
+
+# I copied these from the rocker/r-bspm Dockerfile because it sets them in the
+# RProfile but it wasn't working
+if(requireNamespace("bspm", quietly = TRUE)){
+  message("setting up bspm for use on cloud")
+  bspm::enable()
+  options(pkgType="binary", install.packages.check.source = "no")
+} else {
+  r = getOption("repos")
+  r["CRAN"] = "https://cran.rstudio.com/"
+  options(repos = r)
+}
+
+row_ind <- as.numeric(commandArgs(trailingOnly = TRUE)[[1]])
+cat(print(row_ind))
+
+# set PAT to avoid api limit errors on GitHub
+Sys.setenv(GITHUB_PAT = "<pat>")
+
+install.packages("Require")
+## Install Dependencies (listed in DESCRIPTION) ----
+print("install deps")
+out <- try(Require::Require(c("dplyr", "sf", "terra", "tpq/peakRAM", "LandSciTech/roads")))
+cat("installed from binary:\n")
+print(out)
+
+out2 <- try(Require::Require(c("dplyr", "sf", "terra", "tpq/peakRAM", "LandSciTech/roads"),
+                             type = "source", upgrade = "never"))
+cat("installed from source?:\n")
+print(out2)
+
+source("6_benchmark_methods.R")

### 2024-03-13 analysis/scripts/6_benchmark_methods.R

@@ -6,8 +6,6 @@
 # only compare one density level
 # Also record memory usage. Using separate nodes will ensure to memory overlap between runs
 
-row_ind <- commandArgs(trailingOnly = TRUE)
-
 #set path for data
 data_path_raw <- "analysis/data/raw_data/"
 data_path_drvd <- "analysis/data/derived_data/"
@@ -18,7 +16,6 @@ library(dplyr)
 library(sf)
 library(roads)
 library(terra)
-
 
 # load functions used in script
 devtools::load_all()
@@ -39,6 +36,8 @@ param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
 method <- param_tbl$method[row_ind]
 agg <- param_tbl$agg[row_ind]
 sampleDens <- 0.00001
+
+message("running row: ", row_ind, "agg: ", agg, "; method: ", method)
 
 inputs <- prepInputs(
   cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
@@ -63,10 +62,6 @@ prof <- peakRAM::peakRAM(
                          roads = inputs$roadsExist,
                          roadMethod = method)
 )
-
-
-n_verticies <- igraph::vcount(output$g)
-n_edges <- igraph::ecount(output$g)
 
 out <- data.frame(method = method, resolution = res(output$costSurface)[1],
                   n_verticies = igraph::vcount(output$g),

### 2024-03-13 analysis/scripts/exploratory/testAltGraphFns.R
more comparison of octagon method results.

@@ -37,8 +37,8 @@ text(x=5.75,y=4.2,labels='roads',adj=c(0,0.3),xpd=TRUE)
 
 sim = list(costSurface=costRaster)
 
-g1 = getGraph(sim,"rook")
-gNew = getGraph(sim,"rook",method="ours")
+g1 = getGraph(sim,"queen")
+gNew = getGraph(sim,"queen",method="ours")
 igraph::identical_graphs(g1$g,gNew$g,attrs=F)
 
 edge_attr(g1$g,"weight")=round(edge_attr(g1$g,"weight"),5)
@@ -46,17 +46,21 @@ edge_attr(gNew$g,"weight")=round(edge_attr(gNew$g,"weight"),5)
 
 edge_attr(g1$g,"weight")==edge_attr(gNew$g,"weight")
 igraph::identical_graphs(g1$g,gNew$g,attrs=T)
-#rook methods yield identical graphs, up to rounding errors.
+#queen and rook methods yield identical graphs, up to rounding errors.
 
 g1 = getGraph(sim,"octagon")
 gNew = getGraph(sim,"octagon",method="ours")
 igraph::identical_graphs(g1$g,gNew$g,attrs=F)
-#these graphs are not identical because the order of nodes is different.
-# TO DO: figure out how to confirm that they are otherwise identical.
 
+edge_attr(g1$g,"weight")=round(edge_attr(g1$g,"weight"),0)
+edge_attr(gNew$g,"weight")=round(edge_attr(gNew$g,"weight"),0)
 
+edge_attr(g1$g,"weight")==edge_attr(gNew$g,"weight")
+igraph::identical_graphs(g1$g,gNew$g,attrs=T)
+#octagon methods yield similar but not identical graphs because gdistance::geoCorrection
+#method accounts for lengths of diagonals and other geographic distortions on the grid.
 
-#speed/memory testing with big raster
+#speed/memory benchmarking
 data_path_raw <- "C:/Users/HughesJo/Documents/gitprojects/RoadPaper/analysis/data/raw_data/"
 out_path <- "C:/Users/HughesJo/Documents/gitprojects/RoadPaper/analysis/figures/"
 

### 2024-03-13 analysis/paper/paper.Rmd

--- 2024-03-12 analysis/paper/paper.Rmd
@@
 # organize data for ggplot
-doAspatPlot <- function(tab){
-  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>% 
-  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
-               names_to = "metric", values_to = "response")%>%
-  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
-                             sampleType),
-         sampleDens = paste(method, sampleType, sampleDens) %>% 
-           factor(levels = c("mst NA centroid",
-                             "NA NA observed",
-                             "mst random low sample density",
-                             "mst regular low sample density",
-                             "mst random high sample density",
-                             "mst regular high sample density",
-                             "ilcp regular high sample density",
-                             "NA NA klementQGIS",
-                             "NA NA NA"),
-                  labels = c("MST centroid", 
-                             "Observed",
-                             "MST random low density",
-                             "MST regular low density",
-                             "MST random high density",
-                             "MST regular high density",
-                             "ILCP regular high density",
-                             "Hardy QGIS",
-                             "Cutblocks only")),
-         metric = factor(metric,
-                         levels = c("roadDensityMean", "roadPresenceMean",
-                                    "distanceToRoadMean",
-                                    "forestryDisturbanceMean",
-                                    "roadDisturbanceMean"),
-                         labels = c("Road\ndensity", "Road\npresence",
-                                    "Distance\nto road",
-                                    "Forestry\ndisturbance\nfootprint",
-                                    "Road\ndisturbance\nfootprint")))
-
-observed_values <- filter(meanTable_long, sampleDens == "Observed")
-projected_values <- filter(meanTable_long, sampleDens != "Observed")
-
-# get proportional difference from observed
-prop_dif <- projected_values %>% 
-  left_join(observed_values %>% select(areaMean, metric, response),
-            by = c("areaMean", "metric"), 
-            suffix = c("_proj", "_obs")) %>% 
-  mutate(prop_dif = (response_proj-response_obs)/response_obs,
-         areaMean = factor(areaMean, labels = c("Cutblocks", "Outside\ncutblocks",
-                                                "Overall")))
-
-prop_dif %>% 
-  ggplot(aes(x = sampleDens, prop_dif, fill = sampleDens))+
-  geom_hline(aes(yintercept = 0), color = "grey")+
-  geom_col(position = position_dodge2(preserve = "single",
-                                      width = 0.75))+
-  facet_grid(areaMean~ metric, scales = "free")+
-  theme_classic()+
-  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
-                    name = "Sampling\nMethod", guide = "none")+
-  theme(text = element_text(size = 10), axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.position = "right",
-        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
-  coord_flip()
-}
-
 doAspatPlot(meanTableRealCuts)
@@
 Cutblocks not accessible via the observed road network were excluded from analysis."}
-doSpatPerf <- function(tab){
-  matchData_sum <- tab %>% ungroup() %>%
-    mutate(agreement = factor(agreement,
-                              levels = c("False negative", "False positive",
-                                         "Agree roaded","Agree roadless",
-                                         "Pre-existing roads")),
-           sampleDens = paste(method, sampleType, sampleDens) %>%
-             str_replace("1e-06", "low density") %>%
-             str_replace("1e-05", "high density") %>%
-             str_replace("centroid low density", "centroid") %>%
-             str_replace("klementQGIS NA", "klementQGIS") %>%
-             factor(levels = c("mst centroid",
-                               "mst random low density",
-                               "mst regular low density",
-                               "mst random high density",
-                               "mst regular high density",
-                               "ilcp regular high density",
-                               "NA klementQGIS", 
-                               "NA cutOnly NA"),
-                    labels = c("MST centroid", 
-                               "MST random low density",
-                               "MST regular low density",
-                               "MST random high density",
-                               "MST regular high density",
-                               "ILCP regular high density",
-                               "Hardy QGIS", 
-                               "Cutblocks only")),
-           metric = factor(metric,
-                           levels = c("roadPresence",
-                                      "forestryDisturbance",
-                                      "roadDisturbance"),
-                           labels = c("Road presence",
-                                      "Forestry disturbance\nfootprint",
-                                      "Road disturbance\nfootprint"))) %>%
-    group_by(metric, sampleDens, agreement) %>%
-    summarise(count = sum(count), .groups = "drop_last") %>%
-    mutate(perc = count/sum(count)*100)
-
-  # performance table
-  perf_tbl <- matchData_sum %>% 
-    pivot_wider(c(metric, sampleDens), names_from = agreement, 
-                values_from = count, values_fill = 0) %>% 
-    mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
-           precision = `Agree roaded`/(`Agree roaded` + `False positive`),
-           F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
-    select(metric, sampleDens, sensitivity, precision, F_measure) 
-}
-
 perf_tbl <- doSpatPerf(matchDataRealCuts)
@@
   summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
-
-doSpatPlot <- function(tab){
-  tab %>% 
-    pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
-                 values_to = "value") %>% 
-    mutate(perf_meas = factor(perf_meas,
-                              labels = c("F-measure", "Precision", "Sensitivity")),
-           value = value  +0.01) %>%   # adding so 0 shows in figure
-    ggplot(aes(sampleDens, value, fill = sampleDens))+
-    geom_col()+
-    scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
-                      name = "Sampling\nMethod", guide = "none")+
-    facet_grid(metric ~ perf_meas)+
-    coord_flip()+
-    ylim(0, 1.01)+
-    theme_classic()+
-    labs(y = "Performance", x = NULL)
-}
 
@@
 
-# Appendix
-
-## All cutblocks results
-
-```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster.
-Metrics were summarised within cutblocks, outside cutblocks, and overall.
-Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
-Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
-meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
-doAspatPlot(meanTable)
-```
-
-```{r spat-perf-fig-all, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
-F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
-Values closer to one indicate better performance.
-Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
-matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
-perf_tbl2 <- doSpatPerf(matchData)
-doSpatPlot(perf_tbl2)
-
-```
-
-## Coarse resolution results
-
-```{r get-data2}
-data_path_drvd <- "analysis/data/derived_data/TSA27_1000"
-
-# load csv
-meanTable2 <- read.csv(here(data_path_drvd, "mean_table.csv")) %>% 
-  mutate(resolution = "1000")
-
-matchData2 <- read.csv(here(data_path_drvd, "agree_table.csv")) %>% 
-  mutate(resolution = "1000")
-
-meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
-  filter((method != "ilcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))
-
-matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
-  filter(method != "ilcp", sampleType != "cutOnly")
-
-```
-
-```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
-Metrics were summarised within cutblocks, outside cutblocks, and overall.
-Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
-# organize data for ggplot
-meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
-  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
-               names_to = "metric", values_to = "response")%>%
-  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
-                             sampleType),
-         sampleDens = paste(method, sampleType, sampleDens) %>% 
-           factor(levels = c("mst NA centroid",
-                             "NA NA observed",
-                             "mst random low sample density",
-                             "mst regular low sample density",
-                             "mst random high sample density",
-                             "mst regular high sample density",
-                             "ilcp regular high sample density",
-                             "NA klementQGIS", 
-                             "NA NA NA"),
-                  labels = c("MST centroid", 
-                             "Observed",
-                             "MST random low density",
-                             "MST regular low density",
-                             "MST random high density",
-                             "MST regular high density",
-                             "ILCP regular high density",
-                             "Hardy QGIS",
-                             "Cutblocks only")),
-         metric = factor(metric,
-                         levels = c("roadDensityMean", "roadPresenceMean",
-                                    "distanceToRoadMean",
-                                    "forestryDisturbanceMean",
-                                    "roadDisturbanceMean"),
-                         labels = c("Road\ndensity", "Road\npresence",
-                                    "Distance\nto road",
-                                    "Forestry\ndisturbance\nfootprint",
-                                    "Road\ndisturbance\nfootprint")))
-
-observed_values <- filter(meanTable_long, sampleDens == "Observed")
-projected_values <- filter(meanTable_long, sampleDens != "Observed")
-
-# get proportional difference from observed
-prop_dif <- projected_values %>% 
-  left_join(observed_values %>% select(areaMean, metric, resolution, response),
-            by = c("areaMean", "metric", "resolution"), 
-            suffix = c("_proj", "_obs")) %>% 
-  mutate(prop_dif = (response_proj-response_obs)/response_obs,
-         areaMean = factor(areaMean, labels = c("Cutblocks", "Outside\ncutblocks",
-                                                "Overall")))
-
-prop_dif %>% 
-  ggplot(aes(x = sampleDens, prop_dif, fill = resolution))+
-  geom_hline(aes(yintercept = 0), color = "grey")+
-  geom_col(position = position_dodge2(preserve = "single",
-                                      width = 0.75))+
-  facet_grid(areaMean ~ metric, scales = "free")+
-  theme_classic()+
-  scale_fill_manual(values = met.brewer(pal_nm), labels = c("1 ha", "100 ha"))+
-  theme(text = element_text(size = 10), axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.position = "right",
-        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
-  coord_flip()+
-  labs(fill = "Resolution")
-
-```
-
-```{r figA2, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
-F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
-
-matchData_sum <- matchData %>% ungroup() %>%
-  mutate(agreement = factor(agreement,
-                            levels = c("False negative", "False positive",
-                                       "Agree roaded","Agree roadless",
-                                       "Pre-existing roads")),
-         sampleDens = paste(method, sampleType, sampleDens) %>%
-           str_replace("1e-06", "low density") %>%
-           str_replace("1e-05", "high density") %>%
-           str_replace("centroid low density", "centroid") %>%
-           str_replace("klementQGIS NA", "klementQGIS") %>%
-           factor(levels = c("mst centroid",
-                             "mst random low density",
-                             "mst regular low density",
-                             "mst random high density",
-                             "mst regular high density",
-                             "ilcp regular high density",
-                             "NA klementQGIS"),
-                  labels = c("MST centroid", 
-                             "Random low density",
-                             "Regular low density",
-                             "Random high density",
-                             "Regular high density",
-                             "ILCP regular high density",
-                             "Hardy QGIS")),
-         metric = factor(metric,
-                         levels = c("roadPresence",
-                                    "forestryDisturbance",
-                                    "roadDisturbance"),
-                         labels = c("Road presence",
-                                    "Forestry disturbance\nfootprint",
-                                    "Road disturbance\nfootprint"))) %>%
-  group_by(metric, sampleDens, agreement, resolution) %>%
-  summarise(count = sum(count), .groups = "drop_last") %>%
-  mutate(perc = count/sum(count)*100)
-
-# performance table
-perf_tbl <- matchData_sum %>% 
-  pivot_wider(c(metric, sampleDens, resolution), names_from = agreement, 
-              values_from = count) %>% 
-  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
-         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
-         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
-  select(resolution, metric, sampleDens, sensitivity, precision, F_measure)
-
-perf_tbl_rng <- perf_tbl %>% group_by(metric, resolution) %>% 
-  summarise(rng_F = range(F_measure) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
-
-perf_tbl %>% 
-  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
-               values_to = "value") %>% 
-  mutate(perf_meas = factor(perf_meas,
-                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
-  ggplot(aes(sampleDens, value, fill = resolution))+
-  geom_col(position = position_dodge2(preserve = "single",
-                                      width = 0.75))+
-  scale_fill_manual(values = met.brewer(pal_nm), labels = c("1 ha", "100 ha"))+
-  facet_grid(metric ~ perf_meas)+
-  coord_flip()+
-  scale_y_continuous(limits = c(0, 1), breaks = 0:5/5)+
-  theme_classic()+
-  labs(y = "Performance", x = NULL, fill = "Resolution")
-
-```
-
-## Fort Nelson map
-
-```{r ftNelson, cache=TRUE, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson.
-Cost based road projection methods are not expected to be useful in areas where roads are built for purposes other than accessing a target location at minimal cost.
-For example, in the Northeast corner of the Fort Nelson Timber Supply Area seismic lines show a different pattern than foresty roads."}
-ftN_roads_obs <- read_sf(here("analysis/data/derived_data/combined_ft_nelson_roads.gpkg"))
-
-# only show top right corner
-bb <- st_bbox(ftN_roads_obs)
-
-bb[["xmin"]] <- bb[["xmax"]] - (bb[["xmax"]] - bb[["xmin"]])/2
-bb[["ymin"]] <- bb[["ymax"]] - (bb[["ymax"]] - bb[["ymin"]])/2
-
-ftN_roads_obs_NE <- st_crop(ftN_roads_obs, bb)
-
-tm_shape(ftN_roads_obs_NE, is.master = TRUE)+
-  tm_lines()
-```
-
-\newpage
-
 ### Colophon

### 2024-03-13 analysis/paper/supplement.Rmd

---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: yes
bibliography: Roads.bib
csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style

editor_options: 
  chunk_output_type: inline
---

# (APPENDIX) Supplementary Material for Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife {-}

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r setup2}
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)
library(tmap)
library(sf)

devtools::load_all(here())

#set paths
data_path_raw <- "analysis/data/raw_data/"
data_path_drvd <- "analysis/data/derived_data/TSA27"

pal_nm <- "Redon"

fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)

```

# All cutblocks results

```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
doAspatPlot(meanTable)
```

```{r spat-perf-fig-all, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster. F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted. Values closer to one indicate better performance. Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
perf_tbl2 <- doSpatPerf(matchData)
doSpatPlot(perf_tbl2)

```

# Coarse resolution results

```{r get-data2}
data_path_drvd <- "analysis/data/derived_data/TSA27_1000"

# load csv
meanTable2 <- read.csv(here(data_path_drvd, "mean_table.csv")) %>% 
  mutate(resolution = "1000")

matchData2 <- read.csv(here(data_path_drvd, "agree_table.csv")) %>% 
  mutate(resolution = "1000")

meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
  filter((method != "ilcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))

matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
  filter(method != "ilcp", sampleType != "cutOnly")

```

```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "ilcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

observed_values <- filter(meanTable_long, sampleDens == "Observed")
projected_values <- filter(meanTable_long, sampleDens != "Observed")

# get proportional difference from observed
prop_dif <- projected_values %>% 
  left_join(observed_values %>% select(areaMean, metric, resolution, response),
            by = c("areaMean", "metric", "resolution"), 
            suffix = c("_proj", "_obs")) %>% 
  mutate(prop_dif = (response_proj-response_obs)/response_obs,
         areaMean = factor(areaMean, labels = c("Cutblocks", "Outside\ncutblocks",
                                                "Overall")))

prop_dif %>% 
  ggplot(aes(x = sampleDens, prop_dif, fill = resolution))+
  geom_hline(aes(yintercept = 0), color = "grey")+
  geom_col(position = position_dodge2(preserve = "single",
                                      width = 0.75))+
  facet_grid(areaMean ~ metric, scales = "free")+
  theme_classic()+
  scale_fill_manual(values = met.brewer(pal_nm), labels = c("1 ha", "100 ha"))+
  theme(text = element_text(size = 10), axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
  coord_flip()+
  labs(fill = "Resolution")

```

```{r figA2, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters. F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "ilcp regular high density",
                             "NA klementQGIS"),
                  labels = c("MST centroid", 
                             "Random low density",
                             "Regular low density",
                             "Random high density",
                             "Regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance\nfootprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

# performance table
perf_tbl <- matchData_sum %>% 
  pivot_wider(c(metric, sampleDens, resolution), names_from = agreement, 
              values_from = count) %>% 
  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
  select(resolution, metric, sampleDens, sensitivity, precision, F_measure)

perf_tbl_rng <- perf_tbl %>% group_by(metric, resolution) %>% 
  summarise(rng_F = range(F_measure) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)

perf_tbl %>% 
  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
               values_to = "value") %>% 
  mutate(perf_meas = factor(perf_meas,
                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
  ggplot(aes(sampleDens, value, fill = resolution))+
  geom_col(position = position_dodge2(preserve = "single",
                                      width = 0.75))+
  scale_fill_manual(values = met.brewer(pal_nm), labels = c("1 ha", "100 ha"))+
  facet_grid(metric ~ perf_meas)+
  coord_flip()+
  scale_y_continuous(limits = c(0, 1), breaks = 0:5/5)+
  theme_classic()+
  labs(y = "Performance", x = NULL, fill = "Resolution")

```

# Fort Nelson map

```{r ftNelson, cache=TRUE, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson. Cost based road projection methods are not expected to be useful in areas where roads are built for purposes other than accessing a target location at minimal cost. For example, in the Northeast corner of the Fort Nelson Timber Supply Area seismic lines show a different pattern than foresty roads."}
ftN_roads_obs <- read_sf(here("analysis/data/derived_data/combined_ft_nelson_roads.gpkg"))

# only show top right corner
bb <- st_bbox(ftN_roads_obs)

bb[["xmin"]] <- bb[["xmax"]] - (bb[["xmax"]] - bb[["xmin"]])/2
bb[["ymin"]] <- bb[["ymax"]] - (bb[["ymax"]] - bb[["ymin"]])/2

ftN_roads_obs_NE <- st_crop(ftN_roads_obs, bb)

tm_shape(ftN_roads_obs_NE, is.master = TRUE)+
  tm_lines()
```

\newpage

# Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
#if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
#if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2024-03-13 analysis/scripts/exploratory/testAltGraphFns.R
add peak RAM

@@ -61,8 +61,8 @@ igraph::identical_graphs(g1$g,gNew$g,attrs=T)
 #method accounts for lengths of diagonals and other geographic distortions on the grid.
 
 #speed/memory benchmarking
-data_path_raw <- "C:/Users/HughesJo/Documents/gitprojects/RoadPaper/analysis/data/raw_data/"
-out_path <- "C:/Users/HughesJo/Documents/gitprojects/RoadPaper/analysis/figures/"
+data_path_raw <- "C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/data/raw_data/"
+out_path <- "C:/Users/endicotts/Documents/gitprojects/RoadPaper/analysis/figures/"
 
 landscape <- rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
 
@@ -90,6 +90,16 @@ bm <- bench::mark(min_iterations = 1, check = FALSE,
                   )
 
 bm
+
+# this one is supposed to tell us about max RAM used over the whole calculation
+ram_use <- peakRAM::peakRAM(g1_100 = getGraph(landscape_100,"octagon"),
+                            gNew_100 = getGraph(landscape_100,"octagon",method="ours"),
+                            g1_500 = getGraph(landscape_500,"octagon"),
+                            gNew_500 = getGraph(landscape_500,"octagon",method="ours"),
+                            g1_1000 = getGraph(landscape_1000,"octagon"),
+                            gNew_1000 = getGraph(landscape_1000,"octagon",method="ours"),
+                            g1_2000 = getGraph(landscape_2000,"octagon"),
+                            gNew_2000 = getGraph(landscape_2000,"octagon",method="ours"))
 
 pdf(paste0("outputs/figures/getGraph_compareTime.pdf"),width=4,height=4)
 plot(bm,type="boxplot")+

### 2024-03-13 analysis/scripts/exploratory/testAltGraphFns.R
move to roads tests folder

@@ -1,126 +1,0 @@
-library(terra)
-library(dplyr)
-library(sf)
-#library(roads)
-library(igraph)
-library(ggplot2)
-devtools::document();devtools::load_all()
-
-## colours for displaying cost raster
-if(requireNamespace("viridis", quietly = TRUE)){
-  # Use colour blind friendly palette if available
-  rastColours <- c('grey50', viridis::viridis(20))
-} else {
-  rastColours <- c('grey50', terrain.colors(20))
-}
-
-CLUSexample <- prepExData(CLUSexample)
-
-costRaster <- CLUSexample$cost
-roadsLine <- sf::st_sfc(geometry = sf::st_linestring(
-  matrix(c(0.5, 4.5, 4.5, 4.5),
-         ncol = 2, byrow = T)
-)) %>%
-  sf::st_as_sf(crs = sf::st_crs(costRaster))
-
-
-landings <- roads::CLUSexample$landings
-
-## plot example scenario
-plot(costRaster, col = rastColours, main = 'Example Scenario')
-plot(roadsLine, add = TRUE)
-plot(landings, add = TRUE, pch = 19)
-points(x=5.6,y=4.5,pch=19,xpd=TRUE)
-text(x=5.8,y=4.5,labels='landing',adj=c(0,0.4),xpd=TRUE)
-lines(x=c(5.3,5.6),y=c(4.2,4.2),lwd=2,xpd=TRUE)
-text(x=5.75,y=4.2,labels='roads',adj=c(0,0.3),xpd=TRUE)
-
-sim = list(costSurface=costRaster)
-
-g1 = getGraph(sim,"queen")
-gNew = getGraph(sim,"queen",method="ours")
-igraph::identical_graphs(g1$g,gNew$g,attrs=F)
-
-edge_attr(g1$g,"weight")=round(edge_attr(g1$g,"weight"),5)
-edge_attr(gNew$g,"weight")=round(edge_attr(gNew$g,"weight"),5)
-
-edge_attr(g1$g,"weight")==edge_attr(gNew$g,"weight")
-igraph::identical_graphs(g1$g,gNew$g,attrs=T)
-#queen and rook methods yield identical graphs, up to rounding errors.
-
-g1 = getGraph(sim,"octagon")
-gNew = getGraph(sim,"octagon",method="ours")
-igraph::identical_graphs(g1$g,gNew$g,attrs=F)
-
[…]

### 2024-03-14 analysis/paper/paper.Rmd
playing with slope penalty options

@@
 Slope|16178 + 504*slope|Interior appraisal manual
-Major lakes|65000|Major water bodies of BC
+Major lakes|NA|Major water bodies of BC
 Existing roads|0|Digital road atlas and forest tenure roads for BC

### 2024-03-14 R/functionsSource.R

@@ -69,7 +69,7 @@ roadDisturbanceFootprint <- function(x, r, b) {
 
 fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
                            sampleType, costRaster, existingRoads,
-                           projectionMethod, sim = NULL) {
+                           projectionMethod, weightFunction, sim = NULL) {
 
   subLandings <- getLandingsFromTarget(cutblockPolygons,
                                        landingDens = sampleDensity,
@@ -87,12 +87,12 @@ fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
 
     sim$roadMethod <- projectionMethod
 
-    subProjectionResults <- projectRoads(subLandings, sim = sim)
+    subProjectionResults <- projectRoads(subLandings, sim = sim,weightFunction =weightFunction)
   } else{
     subProjectionResults <- projectRoads(subLandings,
                                          cost = costRaster,
                                          roads = existingRoads,
-                                         roadMethod = projectionMethod)
+                                         roadMethod = projectionMethod,weightFunction=weightFunction)
   }
 
 
@@ -106,6 +106,10 @@ fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
 
 projectAll <- function(tsbs,paramTable, costSurface,
                        cutblocks, existingRoads, fileLocation) {
+
+  if(!is.element("weightFunction",names(paramTable))){
+    paramTable$weightFunction = deparse1(formals(projectRoads)$weightFunction)
+  }
 
   for(i in 1:nrow(paramTable)) {
     # i = 1
@@ -122,6 +126,7 @@ projectAll <- function(tsbs,paramTable, costSurface,
                                            costRaster = costSurface,
                                            existingRoads = existingRoads,
                                            projectionMethod = cRow$method,
+                                           weightFunction = eval(str2lang(cRow$weightFunction)),
                                            sim = projectionsList[[1]])
 
 

### 2024-03-14 analysis/scripts/exploratory/runProjectionsSlopePenalties.R
New method for penalizing grade of edges.

@@ -1,0 +1,212 @@
+## Running projections
+
+#set path for data
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data/"
+
+###### Source scripts ######
+# load in required libraries
+library(raster)
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(tmap)
+library(purrr)
+library(caribouMetrics)
+library(pfocal)
+library(ggplot2)
+library(stringr)
+library(ggpubr)
+library(MetBrewer)
+library(here)
+library(tmap)
+
+# load functions used in script
+devtools::load_all()
+
+# set rasterOptions to allow higher memory usage
+#prevOpts <- terraOptions(memfrac = 0.9)
+
+#densities to test
+low <- 0.000001
+high <- 0.00001
+
+# For Revelstoke at fine resolution
+
+#build new cost raster
+
+dem = rast(paste0(data_path_drvd, "TSA27/dem_revelstoke.tif"))
+costOld = rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
+cCost = crop(costOld,dem)
+#get average elevation in each cost cell
+cDem = resample(dem,cCost,method="cubic")
+cDem[is.na(cCost)]=NA
+cDem = cDem/res(cDem)[1] #Put in elevation in units of cell width
+range(cDem)
+terra::writeRaster(cDem,
+                   filename = file.path(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"), overwrite=T)
+
+#This is a simplified version of the grade penalty approach taken by Anderson and Nelson:
+#doesn't distinguish between adverse and favourable grades.
+#construction cost values from interior appraisal manual.
+#ignores (unknown) grade penalties beside roads in order to make do with a single input layer.
+slopePenaltyFn<-function(x1,x2,limit=10,penalty=504){
+  grade = abs(x1-x2)*(pmin(x1,x2)!=0) #percent slope. if one of the locations is a road, don't apply grade penalty
+  slp = 16178+grade*penalty
+  slp[grade>limit]=NA
+  slp[pmax(x1,x2)==0]=0 # if both 0 then this is an existing road link
+
+  return(slp)
+}
+
+see = data.frame(demDif = seq(-30,30,length.out=1000))
+see$cost = slopePenaltyFn(100,100+see$demDif)
+plot(see$cost~see$demDif)
+
+
[…]

### 2024-03-15 analysis/paper/paper.Rmd
Ongoing grade penalty work

@@
                       existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"))
+                      fileLocation = here(data_path_drvd, "..", "for_fig"),roadsInCost = F)
 

### 2024-03-15 R/functionsSource.R

@@ -69,7 +69,7 @@ roadDisturbanceFootprint <- function(x, r, b) {
 
 fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
                            sampleType, costRaster, existingRoads,
-                           projectionMethod, weightFunction, sim = NULL) {
+                           projectionMethod, weightFunction, sim = NULL,roadsInCost=T,...) {
 
   subLandings <- getLandingsFromTarget(cutblockPolygons,
                                        landingDens = sampleDensity,
@@ -87,12 +87,12 @@ fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
 
     sim$roadMethod <- projectionMethod
 
-    subProjectionResults <- projectRoads(subLandings, sim = sim,weightFunction =weightFunction)
+    subProjectionResults <- projectRoads(subLandings, sim = sim,weightFunction =weightFunction,...)
   } else{
     subProjectionResults <- projectRoads(subLandings,
                                          cost = costRaster,
                                          roads = existingRoads,
-                                         roadMethod = projectionMethod,weightFunction=weightFunction)
+                                         roadMethod = projectionMethod,weightFunction=weightFunction,roadsInCost=roadsInCost,...)
   }
 
 
@@ -105,10 +105,10 @@ fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
 # function to generate all projections and raster layers for metrics
 
 projectAll <- function(tsbs,paramTable, costSurface,
-                       cutblocks, existingRoads, fileLocation) {
+                       cutblocks, existingRoads, fileLocation,roadsInCost=T,...) {
 
   if(!is.element("weightFunction",names(paramTable))){
-    paramTable$weightFunction = deparse1(formals(projectRoads)$weightFunction)
+    paramTable$weightFunction = deparse1(formals(projectRoads)$weightFunction,collapse="\n")
   }
 
   for(i in 1:nrow(paramTable)) {
@@ -127,7 +127,8 @@ projectAll <- function(tsbs,paramTable, costSurface,
                                            existingRoads = existingRoads,
                                            projectionMethod = cRow$method,
                                            weightFunction = eval(str2lang(cRow$weightFunction)),
-                                           sim = projectionsList[[1]])
+                                           sim = projectionsList[[1]],
+                                           roadsInCost=roadsInCost,...)
 
 
     projections <- projectionsList[[i]]$roads
[…]

### 2024-03-15 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -40,30 +40,12 @@ dem = rast(paste0(data_path_drvd, "TSA27/dem_revelstoke.tif"))
 costOld = rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
 cCost = crop(costOld,dem)
 #get average elevation in each cost cell
-cDem = resample(dem,cCost,method="cubic")
+cDem = resample(dem,cCost,method="average")
 cDem[is.na(cCost)]=NA
 cDem = cDem/res(cDem)[1] #Put in elevation in units of cell width
 range(cDem)
 terra::writeRaster(cDem,
                    filename = file.path(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"), overwrite=T)
-
-#This is a simplified version of the grade penalty approach taken by Anderson and Nelson:
-#doesn't distinguish between adverse and favourable grades.
-#construction cost values from interior appraisal manual.
-#ignores (unknown) grade penalties beside roads in order to make do with a single input layer.
-slopePenaltyFn<-function(x1,x2,limit=10,penalty=504){
-  grade = abs(x1-x2)*(pmin(x1,x2)!=0) #percent slope. if one of the locations is a road, don't apply grade penalty
-  slp = 16178+grade*penalty
-  slp[grade>limit]=NA
-  slp[pmax(x1,x2)==0]=0 # if both 0 then this is an existing road link
-
-  return(slp)
-}
-
-see = data.frame(demDif = seq(-30,30,length.out=1000))
-see$cost = slopePenaltyFn(100,100+see$demDif)
-plot(see$cost~see$demDif)
-
 
 pal_nm <- "Redon"
 
@@ -80,7 +62,11 @@ roads <- read_sf(here(data_path_drvd,  "testing_obs_roads.gpkg"))
 demCost = crop(cDem,tsaCost)
 demCost[tsaCost==0]=0
 plot(demCost)
-plot(tsaCost)
+range(demCost)
+
+see = data.frame(demDif = seq(-0.3,0.3,length.out=1000))
+see$cost = slopePenaltyFn(1,1+see$demDif)
+plot(see$cost~see$demDif)
 
 # densities
 high <- 0.00001
@@ -100,6 +86,7 @@ paramTable <- tibble(sampleType, sampleDens,
   distinct()
 
 paramTable$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+paramTable$weightFunction = gsub("limitCost = NA","limitCost = 65000",paramTable$weightFunction,fixed=T)
 
[…]

### 2024-03-26 analysis/paper/paper.Rmd
change name to weightRaster, remove no longer needed finResProject fun, make sure runTimes are not lost when re-running metric calcs

--- 2024-03-14 analysis/paper/paper.Rmd
@@
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
-                      costSurface = tsaCost,
+                      weightRaster = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"))
+                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                      roadsInWeight = F)
 
@@
                         mutate(method = "ilcp"),
-                      costSurface = tsaCost,
+                      weightRaster = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"))
+                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                      roadsInWeight = F)
 

### 2024-03-26 R/functionsSource.R

@@ -67,45 +67,10 @@ roadDisturbanceFootprint <- function(x, r, b) {
 }
 
 
-fineResProject <- function(TSAsubset, cutblockPolygons, sampleDensity,
-                           sampleType, costRaster, existingRoads,
-                           projectionMethod, weightFunction, sim = NULL,roadsInCost=T,...) {
-
-  subLandings <- getLandingsFromTarget(cutblockPolygons,
-                                       landingDens = sampleDensity,
-                                       sampleType = sampleType)
-  if(nrow(cutblockPolygons) == 0){
-    return(NULL)
-  }
-
-  if(!is.null(sim)){
-    # if sim list supplied use it so don't redo getGraph but change all the
-    # other parts to the current inputs
-    sim$costSurface <- costRaster
-
-    sim$roads <- existingRoads
-
-    sim$roadMethod <- projectionMethod
-
-    subProjectionResults <- projectRoads(subLandings, sim = sim,weightFunction =weightFunction,...)
-  } else{
-    subProjectionResults <- projectRoads(subLandings,
-                                         cost = costRaster,
-                                         roads = existingRoads,
-                                         roadMethod = projectionMethod,weightFunction=weightFunction,roadsInCost=roadsInCost,...)
-  }
-
-
-  gc(verbose = TRUE)
-
-  return(subProjectionResults)
-
-}
-
 # function to generate all projections and raster layers for metrics
 
-projectAll <- function(tsbs,paramTable, costSurface,
-                       cutblocks, existingRoads, fileLocation,roadsInCost=T,...) {
+projectAll <- function(tsbs,paramTable, weightRaster,
+                       cutblocks, existingRoads, fileLocation, roadsInWeight=T, ...) {
 
   if(!is.element("weightFunction",names(paramTable))){
     paramTable$weightFunction = deparse1(formals(projectRoads)$weightFunction,collapse="\n")
@@ -116,22 +81,18 @@ projectAll <- function(tsbs,paramTable, costSurface,
     cRow = paramTable[i,]
     start <- Sys.time()
 
-    projectionsList <- vector("list", nrow(paramTable))
+    subLandings <- getLandingsFromTarget(cutblocks,
+                                         landingDens = cRow$sampleDens,
+                                         sampleType = cRow$sampleType)
 
[…]

### 2024-03-26 analysis/scripts/6_benchmark_methods.R

@@ -58,12 +58,12 @@ landings <- getLandingsFromTarget(inputs$cutblocks,
 prof <- peakRAM::peakRAM(
   #Running projections
   output <- projectRoads(landings,
-                         cost = inputs$tsaCost_st,
+                         weightRaster = inputs$tsaCost_st,
                          roads = inputs$roadsExist,
                          roadMethod = method)
 )
 
-out <- data.frame(method = method, resolution = res(output$costSurface)[1],
+out <- data.frame(method = method, resolution = res(output$weightRaster)[1],
                   n_verticies = igraph::vcount(output$g),
                   n_edges = igraph::ecount(output$g)) %>%
   bind_cols(prof[,-1])

### 2024-03-26 analysis/scripts/exploratory/closerLookAtPatterns.R

@@ -96,7 +96,7 @@ devtools::load_all(here())
 
 #TO DO: troubleshooting - why isn't projectAll working here?
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
-                      costSurface = tsaCost,
+                      weightRaster = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
                       fileLocation = here(data_path_drvd, "for_fig"))
@@ -105,7 +105,7 @@ ilcpProj <- projectAll(tsbs = tsb,
                        paramTable = paramTable %>%
                          filter(sampleType == "regular") %>%
                          mutate(method = "ilcp"),
-                       costSurface = tsaCost,
+                       weightRaster = tsaCost,
                        cutblocks = cutblocks,
                        existingRoads = exRoads,
                        fileLocation = here(data_path_drvd,"for_fig"))
@@ -114,7 +114,7 @@ lcpProj <- projectAll(tsbs = tsb,
                       paramTable = paramTable %>%
                         filter(sampleType == "centroid") %>%
                         mutate(method = "lcp"),
-                      costSurface = tsaCost,
+                      weightRaster = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
                       fileLocation = here(data_path_drvd,  "for_fig"))

### 2024-03-26 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -86,14 +86,14 @@ paramTable <- tibble(sampleType, sampleDens,
   distinct()
 
 paramTable$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
-paramTable$weightFunction = gsub("limitCost = NA","limitCost = 65000",paramTable$weightFunction,fixed=T)
+paramTable$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTable$weightFunction,fixed=T)
 
 
 f2 <- eval(str2lang(paramTable$weightFunction[1]))
 
 # re-run projections for just the small area with different parameters
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
-                      costSurface = demCost,
+                      weightRaster = demCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
                       fileLocation = here(data_path_drvd, "for_fig"),roadsInCost=F)
@@ -102,7 +102,7 @@ ilcpProj <- projectAll(tsbs = tsb,
                        paramTable = paramTable %>%
                          filter(sampleType == "regular") %>%
                          mutate(method = "ilcp"),
-                       costSurface = demCost,
+                       weightRaster = demCost,
                        cutblocks = cutblocks,
                        existingRoads = exRoads,
                        fileLocation = here(data_path_drvd, "for_fig"),roadsInCost=F)
@@ -163,7 +163,7 @@ f2 <- eval(str2lang(paramTable$weightFunction[1]))
 
 # re-run projections for just the small area with different parameters
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
-                      costSurface = demCost,
+                      weightRaster = demCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
                       fileLocation = here(data_path_drvd, "for_fig"),roadsInCost=F)
@@ -172,7 +172,7 @@ ilcpProj <- projectAll(tsbs = tsb,
                        paramTable = paramTable %>%
                          filter(sampleType == "regular") %>%
                          mutate(method = "ilcp"),
-                       costSurface = demCost,
+                       weightRaster = demCost,
                        cutblocks = cutblocks,
                        existingRoads = exRoads,
                        fileLocation = here(data_path_drvd, "for_fig"),roadsInCost=F)
@@ -247,9 +247,13 @@ if(0){
 
[…]

### 2024-03-26 analysis/paper/paper.Rmd
Writing in progress.

--- 2024-03-15 analysis/paper/paper.Rmd
@@
 tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
-MST|Regular|1|1 ha and 100 ha
-MST|Regular|10|1 ha and 100 ha
-MST|Random|1|1 ha and 100 ha
-MST|Random|10|1 ha and 100 ha
-MST|Centroid|N/A|1 ha and 100 ha
-ILCP|Regular|10|1 ha 
+MST grade penalty & simple cost|Regular|1|1 & 100 ha
+MST grade penalty & simple cost|Regular|10|1 & 100 ha
+MST grade penalty & simple cost|Random|1|1 & 100 ha
+MST grade penalty & simple cost|Random|10|1 & 100 ha
+MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
+ILCP grade penalty & simple cost|Regular|10|1 ha 
 Hardy QGIS|N/A|~9|1 ha

### 2024-03-26 analysis/scripts/1_prepareData.R
switching to dem cost

@@ -126,7 +126,7 @@ if(down_meth == "bcdata"){
   # This is the map with the tile names:
   # https://www2.gov.bc.ca/assets/gov/data/geographic/topography/250kgrid.pdf
 
-  # map tiles for revelstoke 82N, 82M, 82L, 82K
+  # map tiles for revelstoke 82N, 82M, 82L, 82K, 83D
   list.dem <- c("82n", "82m", "82l", "82k", "83d")
   dem_path <- file.path(data_path_raw, "bc_dem_tiles")
   dir.create(dem_path)
@@ -166,7 +166,7 @@ if(down_meth == "bcdata"){
   # project to WGS84
   rev_dem2 <- terra::project(rev_dem, "epsg:3005")
   terra::writeRaster(rev_dem2,
-                     filename = file.path(data_path_drvd, "dem_revelstoke.tif"))
+                     filename = file.path(data_path_drvd, "TSA27/dem_revelstoke.tif"))
 
 
 }
@@ -233,6 +233,19 @@ osf_proj <- osf_retrieve_node("https://osf.io/8vmqh/")
 
 osf_ls_files(osf_proj) %>% filter(name == "cost_surface_bc_ha.tif") %>%
   osf_download(path = "data/raw_data")
+
+# Use NAs from cost surface to get water in DEM
+dem <- rast(paste0(data_path_drvd, "TSA27/dem_revelstoke.tif"))
+costOld <- rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
+cCost <- crop(costOld, dem)
+# get average elevation in each cost cell
+cDem <- resample(dem, cCost, method = "average")
+cDem[is.na(cCost)] <- NA
+terra::writeRaster(cDem,
+                   filename = file.path(data_path_drvd,
+                                        "TSA27/dem_revelstokeCoarse.tif"),
+                   overwrite = T)
+
 
 # Repeat for Fort Nelson --------------------------------------------------
 if(down_meth == "bcdata"){

### 2024-03-26 analysis/scripts/2_runProjections.R

@@ -28,7 +28,8 @@ high <- 0.00001
 
 # For Revelstoke at fine resolution
 
-# First run prepInputs to create filtered inputs for QGIS plugin
+# First run prepInputs to create filtered inputs for QGIS plugin using the
+# original cost surface
 prepInputs(
   cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
   roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
@@ -83,14 +84,21 @@ paramTable <- tibble(sampleType, sampleDens, method = "mst",
 # add row for ilcp with regular high density sampling
 paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "ilcp"))
 
+# add weightFunction
+paramTable$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+
+# this means that slopes that are steeper than limit will be 65000 not NA
+paramTable$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTable$weightFunction,fixed=T)
+
+
 #
 run_projections(
   paramTable,
   cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
   roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
   tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
-  costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif"),
-  outPth = paste0(data_path_drvd, "TSA27/"),
+  costPth = paste0(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"),
+  outPth = paste0(data_path_drvd, "TSA27/dem/"),
 
   #Klement QGIS projection results shapefile
   klementProj = paste0(data_path_drvd, "TSA27/", "klementProjection.shp"),
@@ -106,8 +114,8 @@ run_projections(
   cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
   roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
   tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
-  costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif"),
[…]

### 2024-03-26 analysis/scripts/6_benchmark_methods.R

@@ -21,14 +21,13 @@ library(terra)
 devtools::load_all()
 
 # Prepare disaggregated cost surface
-# bc_cost <- rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
+# dem <- rast(paste0(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"))
 #
 # tsaBoundary <- read_sf(paste0(data_path_raw, "tsa27_boundaries.gpkg"))
 #
-# rev_cost <- crop(bc_cost, vect(tsaBoundary), snap = "out")
-#
-# rev_cost10 <- disagg(rev_cost, 10,
-#                      filename = file.path(data_path_drvd, "revelstoke_cost_10.tif"))
+# # not using original 20m dem because need NAs from old cost
+# rev_cost10 <- disagg(dem, 10,
+#                      filename = file.path(data_path_drvd, "TSA27/dem_revelstoke_10.tif"))
 
 param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
                          stringsAsFactors = FALSE)
@@ -37,14 +36,14 @@ method <- param_tbl$method[row_ind]
 agg <- param_tbl$agg[row_ind]
 sampleDens <- 0.00001
 
-message("running row: ", row_ind, "agg: ", agg, "; method: ", method)
+message("running row: ", row_ind, "; agg: ", agg, "; method: ", method)
 
 inputs <- prepInputs(
   cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg"),
   roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
   tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
-  costPth = paste0(data_path_drvd, "revelstoke_cost_10.tif"),
-  outPth = paste0(data_path_drvd, "TSA27/"),
+  costPth = paste0(data_path_drvd, "TSA27/dem_revelstoke_10.tif"),
+  outPth = paste0(data_path_drvd, "TSA27/bench/"),
   aggFact = agg, #factor of aggregation of cost surface. 1 = no aggregation.
   saveInputs = FALSE
 )

### 2024-04-02 R/functionsSource.R
making it work for cloud results

@@ -115,23 +115,20 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
                         cutblocksPth, existingRoads,
                         weightRaster){
   # row for observed
-  observedRow <- c("observed", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)
-
-  paramTable <- rbind(paramTable, observedRow)
+  paramTable <- dplyr::add_row(paramTable, sampleType = "observed")
 
   paramTable$output[[nrow(paramTable)]] <- observedRoads
 
   # row for QGIS Plugin
   if(!is.null(klementProj)){
-    klementRow <- c("klementQGIS", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)
 
-    paramTable <- rbind(paramTable, klementRow)
+    paramTable <- dplyr::add_row(paramTable, sampleType = "klementQGIS")
 
     paramTable$output[[nrow(paramTable)]] <- klementProj
   }
 
   # row for if only consider roads in disturbance
-  paramTable <- rbind(paramTable, c("cutOnly", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA))
+  paramTable <- dplyr::add_row(paramTable,  sampleType = "cutOnly")
 
   cutblocksRaster <- terra::rasterize(terra::vect(cutblocks), nonAggregatedweightRaster)
 
@@ -143,7 +140,11 @@ calcMetrics <- function(paramTable, klementProj, cutblocks,
     if(cRow$sampleType == "cutOnly"){
       out <- existingRoads
     } else {
-      out <- readRDS(cRow$output)
+      if(stringr::str_detect(cRow$output, "gpkg|shp")){
+        out <- sf::read_sf(cRow$output)
+      } else {
+        out <- readRDS(cRow$output)
+      }
     }
 
     roadDensityResults <- rasterizeLineDensity(out, r = weightRaster)
@@ -187,7 +188,7 @@ getMetricMeans <- function(paramTable, cutblocks){
                          method = paramTable$method,
                          areaMean = "overall")
 
-  metricsAll <- paramTable %>% select(roadDisturbance:forestryDisturbance) %>%
+  metricsAll <- paramTable %>% select(matches("road|Disturbance")) %>%
     unlist() %>%
     map_dbl(~terra::global(.x, "mean", na.rm = TRUE)[1,1]) %>%
     as_tibble(rownames = "name") %>%
[…]

### 2024-04-02 analysis/cloud/setup_azure.sh
compiling cloud results

@@ -61,20 +61,24 @@ az batch pool create --json-file analysis/cloud/pool_to_use.json
 az batch job create --pool-id $poolName --id $jobName
 
 # 1 2 3 4 5 6
-for rowi in {1..24}
+for rowi in 7 8 9 10 11 13 14 15 16 17 19 20 21 22
 do
   az batch task create --json-file analysis/cloud/task_jsons/task_roads_$rowi.json --job-id $jobName
 done
 
 # az batch task delete --task-id connectivity-combine --job-id $jobName --yes
 #
-# for rowi in {1..24}
+# for rowi in 7 8 9 10 11 13 14 15 16 17 19 20 21 22
 # do
-#   az batch task delete --job-id $jobName --task-id roads-benchmark-$rowi --yes
+#   az batch task reactivate --job-id $jobName --task-id roads-benchmark-$rowi
 # done
 
+# set target dedicated nodes
+az batch pool resize --pool-id $poolName --target-dedicated-nodes 14
+
 # prompt auto scaleing of pool by changing time interval
-az batch pool autoscale enable --pool-id $poolName --auto-scale-evaluation-interval "PT6M"\
+# enabling this as soon as the tasks are created seems to make it think there are no tasks
+az batch pool autoscale enable --pool-id $poolName --auto-scale-evaluation-interval "PT5M"\
  --auto-scale-formula 'percentage = 70;
  span = TimeInterval_Second * 15;
  $samples = $ActiveTasks.GetSamplePercent(span);

### 2024-04-02 analysis/scripts/7_compile_results.R

@@ -6,7 +6,7 @@ library(tmap)
 library(purrr)
 library(caribouMetrics)
 
-
+devtools::load_all()
 
 #set path for data
 data_path_raw <- "analysis/data/raw_data/"
@@ -17,10 +17,46 @@ data_path_drvd <- "analysis/data/derived_data/"
 bench_res <- list.files("analysis/data/derived_data/bench_results", pattern = "bench_.*.rds",
                         full.names = TRUE) %>%
   purrr::map(readRDS) %>%
-  bind_rows()
+  bind_rows() %>%
+  tidyr::separate(id, into = c("method", "sampleType", "sampleDens", "agg", "cutblocks_real"),
+                  sep = "_",
+                  extra = "merge", convert = TRUE)
 
+#densities to test
+low <- 0.000001
+high <- 0.00001
 
-param_tbl
+param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
+                         stringsAsFactors = FALSE) %>%
+  mutate(sampleDens = high,
+         sampleType = "regular",
+         cutblocks_real = "revelstoke_real")
+
+# parameter table creation for running projections
+sampleDens <- c(low,high,low,high,low)
+sampleType <- c("regular","regular","random","random","centroid")
+paramTable <- tibble(method = "mst", sampleType, sampleDens, agg = 10,
+                     cutblocks_real = "revelstoke_real") %>%
+  distinct()
+
+# add row for ilcp with regular high density sampling
+paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "ilcp"))
+
+# add coarse resolution
+paramTable2 <- paramTable %>% mutate(agg = 100) %>% bind_rows(paramTable)
+
+paramTable3 <- bind_rows(paramTable2,
+                         paramTable %>% mutate(cutblocks_real = "revelstoke")) %>%
+  bind_rows(param_tbl) %>%
+  distinct()
+
+param_tbl <- paramTable3
+
+param_tbl %>% left_join(bench_res, by = c("method", "sampleType", "sampleDens",
+                                          "agg", "cutblocks_real")) %>%
+  tibble::rowid_to_column() %>%
+  filter(is.na(resolution)) %>%
+  pull(rowid)
 
 run_projections(param_tbl %>% filter(agg == 100) %>%
                   rowwise() %>%
@@ -38,7 +74,7 @@ run_projections(param_tbl %>% filter(agg == 100) %>%
[…]

### 2024-04-02 analysis/paper/paper.Rmd
working on paper fig 2, syncing first

@@
 tsaCost <- raster::raster(here(data_path_drvd, "..", "testing_cost.tif"))
+dem <- raster::raster(here(data_path_drvd, "..", "testing_dem.tif"))
 roads <- read_sf(here(data_path_drvd, "..", "testing_obs_roads.gpkg"))
@@
 
+simpleCostFn = function(x1,x2,...) (x1+x2)/2
+paramTable$weightFunction = deparse1(simpleCostFn,collapse="\n")
+paramTable$weightMethod = "simple cost"
+
+paramTableDem = paramTable
+paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
+
 # re-run projections for just the small area with different parameters
@@
 
-allProj <- bind_rows(mstProj, ilcpProj, .id = "method") %>%
+# re-run projections for just the small area with different parameters
+mstProjDem <- projectAll(tsbs = tsb, paramTable = paramTableDem,
+                      weightRaster = dem,
+                      cutblocks = cutblocks,
+                      existingRoads = exRoads,
+                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                      roadsInWeight = F)
+
+ilcpProjDem <- projectAll(tsbs = tsb,
+                      paramTable = paramTableDem %>% 
+                        filter(sampleType == "regular") %>% 
+                        mutate(method = "ilcp"),
+                      weightRaster = dem,
+                      cutblocks = cutblocks,
+                      existingRoads = exRoads,
+                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                      roadsInWeight = F)
+
+allProj <- bind_rows(mstProj, ilcpProj, mstProjDem, ilcpDem, .id = "method") %>%
   arrange(desc(sampleType), sampleDens)
@@
                                   ifelse(method == 2, "ILCP", method)),
+                           weightMethod,
                            sampleType,

### 2024-04-02 analysis/scripts/3_method_example_fig.R

@@ -20,6 +20,7 @@ cutblocksPth = paste0(data_path_raw, "cutblocks_revelstoke.gpkg")
 roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg")
 tsaBoundaryPth = paste0(data_path_raw, "tsa27_58_testing.gpkg")
 costPth = paste0(data_path_raw, "cost_surface_bc_ha.tif")
+demPth = paste0(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif")
 
 # densities
 high <- 0.00001
@@ -29,6 +30,9 @@ low <-  0.000001
 cutblocks <- read_sf(cutblocksPth)
 roads <- read_sf(roadsPth)
 tsaCost <- rast(costPth)
+dem <- rast(demPth)
+
+#get average elevation in each cost cell
 
 projRoads <- read_sf(paste0(data_path_drvd, "TSA27/", "random_1e-05_mst.gpkg"))
 
@@ -53,6 +57,7 @@ cutblocks <- vect(cutblocks) %>% crop(bound) %>% st_as_sf()
 projRoads <- vect(projRoads) %>% crop(bound) %>% st_as_sf()
 exRoads <- vect(exRoads) %>% crop(bound) %>% st_as_sf()
 tsaCost <- crop(tsaCost, bound)
+dem <- crop(dem, bound)
 
 # add a large cutblock
 plot(roads %>% st_geometry())
@@ -70,4 +75,5 @@ write_sf(exRoads, paste0(data_path_drvd, "testing_ex_roads.gpkg"))
 write_sf(tsb[[1]], paste0(data_path_drvd, "testing_tsb.gpkg"))
 write_sf(cutblocks, paste0(data_path_drvd, "testing_cutblocks.gpkg"))
 writeRaster(tsaCost, paste0(data_path_drvd, "testing_cost.tif"),overwrite=T)
+writeRaster(dem, paste0(data_path_drvd, "testing_dem.tif"),overwrite=T)
 write_sf(roads, paste0(data_path_drvd, "testing_obs_roads.gpkg"))

### 2024-04-02 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -34,19 +34,6 @@ high <- 0.00001
 
 # For Revelstoke at fine resolution
 
-#build new cost raster
-
-dem = rast(paste0(data_path_drvd, "TSA27/dem_revelstoke.tif"))
-costOld = rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
-cCost = crop(costOld,dem)
-#get average elevation in each cost cell
-cDem = resample(dem,cCost,method="average")
-cDem[is.na(cCost)]=NA
-cDem = cDem
-range(cDem)
-terra::writeRaster(cDem,
-                   filename = file.path(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"), overwrite=T)
-
 pal_nm <- "Redon"
 
 fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
@@ -58,11 +45,8 @@ exRoads <- read_sf(here(data_path_drvd, "testing_ex_roads.gpkg"))
 tsb <- read_sf(here(data_path_drvd,  "testing_tsb.gpkg"))
 cutblocks <- read_sf(here(data_path_drvd,  "testing_cutblocks.gpkg"))
 tsaCost <- rast(here(data_path_drvd,  "testing_cost.tif"))
+dem <- rast(here(data_path_drvd,  "testing_dem.tif"))
 roads <- read_sf(here(data_path_drvd,  "testing_obs_roads.gpkg"))
-demCost = crop(cDem,tsaCost)
-demCost[tsaCost==0]=0
-plot(demCost)
-range(demCost)
 
 see = data.frame(demDif = seq(-0.3,0.3,length.out=1000))
 see$cost = slopePenaltyFn(1,1+see$demDif,resolution=100)
@@ -93,19 +77,19 @@ f2 <- eval(str2lang(paramTable$weightFunction[1]))
 
 # re-run projections for just the small area with different parameters
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
-                      weightRaster = demCost,
+                      weightRaster = dem,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "for_fig"),roadsInCost=F)
+                      fileLocation = here(data_path_drvd, "for_fig"),roadsInWeight=F)
 
 ilcpProj <- projectAll(tsbs = tsb,
                        paramTable = paramTable %>%
                          filter(sampleType == "regular") %>%
[…]

### 2024-04-02 analysis/paper/paper.Rmd
working on fig 2

@@
 paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
+paramTableDem$weightMethod = "grade penalty"
+
+paramTableDem<- subset(paramTableDem,sampleType!="centroid")
 
@@
                       existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                      fileLocation = here(data_path_drvd,"..", "for_fig"),
                       roadsInWeight = F)
@@
                        existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                       fileLocation = here(data_path_drvd,"..", "for_fig"),
                        roadsInWeight = F)
@@
                          existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                         fileLocation = here(data_path_drvd,"..", "for_fig"),
                          roadsInWeight = F)
@@
                           existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "..", "for_fig"),
+                          fileLocation = here(data_path_drvd,"..", "for_fig"),
                           roadsInWeight = F)
 
-allProj <- bind_rows(mstProj, ilcpProj, mstProjDem, ilcpDem, .id = "method") %>%
-  arrange(desc(sampleType), sampleDens)
+allProj <- bind_rows(mstProj, ilcpProj, mstProjDem, ilcpProjDem) %>%
+  arrange(weightMethod,desc(sampleType), sampleDens)
 
@@
   tibble(sampleType = "", sampleDens = 0, method = "Hardy QGIS",
-             output = list(here(data_path_drvd, "..", "for_fig", 
+         output = list(here(data_path_drvd,"..", "for_fig",
                             "HardyQGISTest.gpkg")))) %>%
-  mutate(mapTitle =  paste(ifelse(method == 1, "MST",
-                                  ifelse(method == 2, "ILCP", method)),
-                           weightMethod,
+  mutate(mapTitle =  paste(weightMethod,
+                           ifelse(method == "mst", "MST",
+                                  ifelse(method == "ilcp", "ILCP", method)),
                            sampleType,
@@
           here("analysis/figures/projection_methods_figure.png"),
-          dpi = 300, height = 5, width = fig_widths["two"])
+          dpi = 300, height = 8, width = fig_widths["two"])
 ```

### 2024-04-02 R/functionsSource.R

@@ -96,12 +96,22 @@ projectAll <- function(tsbs,paramTable, weightRaster,
 
     end <- Sys.time()
 
-    paramTable$output[[i]] <- file.path(fileLocation, paste0(cRow$sampleType, "_",
-                                     cRow$sampleDens, "_", cRow$method, ".gpkg"))
+    if(is.element("weightMethod",names(paramTable))){
+      paramTable$output[[i]] <- file.path(fileLocation, paste0(cRow$weightMethod, "_",cRow$sampleType, "_",
+                                                               cRow$sampleDens, "_", cRow$method, ".gpkg"))
+    }else{
+      paramTable$output[[i]] <- file.path(fileLocation, paste0(cRow$sampleType, "_",
+                                                               cRow$sampleDens, "_", cRow$method, ".gpkg"))
+    }
     paramTable$runTime[[i]] <- as.numeric(end - start)
 
     # save the projected roads to a file
     sf::write_sf(projections, paramTable$output[[i]])
+
+    if(is.element("weightMethod",names(paramTable))){
+
+      terra::writeRaster(weightRaster, gsub(".gpkg","weight.tif",paramTable$output[[i]],fixed=T),overwrite=T)
+    }
 
     rm(projections)
     gc(verbose = TRUE)

### 2024-04-02 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -25,6 +25,8 @@ library(tmap)
 # load functions used in script
 devtools::load_all()
 
+
+
 # set rasterOptions to allow higher memory usage
 #prevOpts <- terraOptions(memfrac = 0.9)
 
@@ -69,30 +71,63 @@ paramTable <- tibble(sampleType, sampleDens,
                      forestryDisturbance = vector("list", length(sampleDens))) %>%
   distinct()
 
-paramTable$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
-paramTable$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTable$weightFunction,fixed=T)
+simpleCostFn = function(x1,x2,...) (x1+x2)/2
+paramTable$weightFunction = deparse1(simpleCostFn,collapse="\n")
+paramTable$weightMethod = "simple cost"
 
+paramTableDem = paramTable
+paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
+paramTableDem$weightMethod = "grade penalty"
 
-f2 <- eval(str2lang(paramTable$weightFunction[1]))
+paramTableDem<- subset(paramTableDem,sampleType!="centroid")
 
 # re-run projections for just the small area with different parameters
+set.seed(4)
+
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
-                      weightRaster = dem,
+                      weightRaster = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
-                      fileLocation = here(data_path_drvd, "for_fig"),roadsInWeight=F)
+                      fileLocation = here(data_path_drvd, "for_fig"),
+                      roadsInWeight = F)
+set.seed(4)
 
 ilcpProj <- projectAll(tsbs = tsb,
                        paramTable = paramTable %>%
                          filter(sampleType == "regular") %>%
                          mutate(method = "ilcp"),
-                       weightRaster = dem,
+                       weightRaster = tsaCost,
                        cutblocks = cutblocks,
                        existingRoads = exRoads,
-                       fileLocation = here(data_path_drvd, "for_fig"),roadsInWeight=F)
+                       fileLocation = here(data_path_drvd, "for_fig"),
+                       roadsInWeight = F)
 
[…]

### 2024-04-02 analysis/scripts/exploratory/runProjectionsSlopePenalties.R
update maps for fig 2

@@ -63,22 +63,17 @@ sampleType <- c("regular","random","centroid")
 paramTable <- tibble(sampleType, sampleDens,
                      method = "mst",
                      runTime = vector("list", length(sampleDens)),
-                     output = vector("list", length(sampleDens)),
-                     roadDisturbance = vector("list", length(sampleDens)),
-                     roadDensity = vector("list", length(sampleDens)),
-                     roadPresence = vector("list", length(sampleDens)),
-                     distanceToRoad = vector("list", length(sampleDens)),
-                     forestryDisturbance = vector("list", length(sampleDens))) %>%
+                     output = vector("list", length(sampleDens))) %>%
   distinct()
 
 simpleCostFn = function(x1,x2,...) (x1+x2)/2
 paramTable$weightFunction = deparse1(simpleCostFn,collapse="\n")
-paramTable$weightMethod = "simple cost"
+paramTable$weightMethod = "Simple cost"
 
 paramTableDem = paramTable
 paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
 paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
-paramTableDem$weightMethod = "grade penalty"
+paramTableDem$weightMethod = "Grade penalty"
 
 paramTableDem<- subset(paramTableDem,sampleType!="centroid")
 
@@ -125,6 +120,7 @@ ilcpProjDem <- projectAll(tsbs = tsb,
                           fileLocation = here(data_path_drvd, "for_fig"),
                           roadsInWeight = F)
 
+terra::writeRaster(tsaCost, here(data_path_drvd, "for_fig","HardyQGISTestweight.tif"),overwrite=T)
 
 allProj <- bind_rows(mstProj, ilcpProj, mstProjDem, ilcpProjDem) %>%
   arrange(weightMethod,desc(sampleType), sampleDens)
@@ -136,17 +132,11 @@ allProj <- bind_rows(
   tibble(sampleType = "", sampleDens = 0, method = "Hardy QGIS",
          output = list(here(data_path_drvd, "for_fig",
                             "HardyQGISTest.gpkg")))) %>%
-  mutate(mapTitle =  paste(weightMethod,
+  mutate(mapTitle =  paste(ifelse(is.na(weightMethod), "", weightMethod),
                            ifelse(method == "mst", "MST",
[…]

### 2024-04-03 analysis/paper/paper.Rmd
work on fig 1.

@@
 data_path_raw <- "analysis/data/raw_data/"
-data_path_drvd <- "analysis/data/derived_data/TSA27"
+data_path_drvd <- "analysis/data/derived_data/TSA27_real_cuts"
 
@@
 
-```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada.", cache=TRUE}
+<!---BLOCK_MULTICOL_START--->
+
+```{r prepStudyArea, cache=TRUE}
+
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
 #cost surface raster layer
-tsaCost <- terra::rast(here(data_path_drvd, "input_cost.tif"))
+tsaCost <- terra::rast(here(data_path_drvd, "old cost/input_cost.tif"))
+dem <- terra::rast(here(data_path_drvd, "dem/input_cost.tif"))
+
+#tsaCost[tsaCost==65000]=NA
+dem[dem==-65000]=NA
 
 tsaCost <- terra::mask(tsaCost, terra::vect(tsaBoundary))
+dem <- terra::mask(dem, terra::vect(tsaBoundary))
 
@@
   dpi = 600, insets_tm = canada_overlay, insets_vp = vp,
-  height = 6, width = fig_widths["single"], units = "in")
+  height = 6*0.9, width = fig_widths["single"]*0.9, units = "in")
+
+tsaDEM <- tm_shape(dem, name = "Elevation")+
+  tm_raster(style = "cont", title = "Elevation",palette="Greys")+
+  tm_shape(st_union(tsaBoundary))+
+  tm_borders(col = "black")+
+  tm_compass(position = c(0.3, 0.11), size = 1)+
+  tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
+  tm_layout(legend.position = c("left", "bottom"),
+            legend.bg.color = "white")
+
+tmap_save(tsaDEM, filename = here("analysis/figures/Figure1_StudyAreaDEM.png"),
+  dpi = 600, insets_vp = vp,
+  height = 6*0.9, width = fig_widths["single"]*0.9, units = "in")
+
+```
+
+```{r study-area1, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada."}
 
@@
 
+```
+```{r study-area2}
+
+knitr::include_graphics(here("analysis/figures/Figure1_StudyAreaDEM.png"))
+```
+<!---BLOCK_MULTICOL_STOP{widths: [3.6,3.6]}--->
+
+```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada."}
+
+png::writePNG(array(0, dim = c(1,1,4)), here("analysis/figures/empty.png"))
+knitr::include_graphics(here("analysis/figures/empty.png"))
 ```
@@
 # load csv
-meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
-matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
+meanTable <- read.csv(here(data_path_drvd,"../TSA27", "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd,"../TSA27", "agree_table.csv"))
 meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",

### 2024-04-04 R/functionsSource.R
fix segfault when spatrasts are in memory (I think that was the problem)

@@ -297,8 +297,8 @@ calcAgree <- function(obs_rast, proj_rast, prex_rast, return_res = FALSE){
 #' @param boundary polygon boundary to mask rasters
 
 
-agreeMetricsAll <- function(paramTable, prex_rast, prex_vect, boundary, cutblocks, nonAggregatedweightRaster){
-  obs_tbl <- paramTable %>% filter(sampleType == "observed") %>%
+agreeMetricsAll <- function(paramTable2, prex_rast, prex_vect, boundary, cutblocks, nonAggregatedweightRaster){
+  obs_tbl <- paramTable2 %>% filter(sampleType == "observed") %>%
     dplyr::select(roadDisturbance, roadPresence, forestryDisturbance) %>%
     tidyr::pivot_longer(everything(), names_to = "metric", values_to = "obs_rast") %>%
     mutate(obs_rast = obs_rast %>%
@@ -320,7 +320,7 @@ agreeMetricsAll <- function(paramTable, prex_rast, prex_vect, boundary, cutblock
     mutate(prex_rast = prex_rast %>%
              map(~terra::subst(.x, from = 1, to = 10)) )
 
-  paramTable %>% filter(sampleType != "observed") %>%
+  paramTable2 %>% filter(sampleType != "observed") %>%
     dplyr::select(sampleType, sampleDens, method, roadDisturbance, roadPresence, forestryDisturbance) %>%
     tidyr::pivot_longer(-c(contains("Sample"), method), names_to = "metric", values_to = "rast") %>%
     left_join(obs_tbl, by = "metric") %>%
@@ -337,7 +337,7 @@ prepInputs <- function(cutblocksPth, roadsPth, tsaBoundaryPth, costPth,
                        outPth, aggFact, replaceNA = NULL,
                        saveInputs = FALSE){
   if(!dir.exists(outPth)){
-    dir.create(outPth)
+    dir.create(outPth, recursive = TRUE)
   }
   ######### load in data for projections ########################################
 
@@ -406,7 +406,7 @@ prepInputs <- function(cutblocksPth, roadsPth, tsaBoundaryPth, costPth,
 
 # run all projections and summarise results for one tsa
 run_projections <- function(paramTable,cutblocksPth, roadsPth, tsaBoundaryPth, costPth,
-                            outPth, klementProj, aggFact, method = "mst",
+                            outPth, klementProj = NULL, aggFact, method = "mst",
[…]

### 2024-04-04 analysis/cloud/make_combine.R
cloud stuff for combining results

@@ -1,0 +1,49 @@
+#' Road Paper cloud benchmarking
+#'
+#' @description
+#' Benchmarking the roads paper algorithms on the cloud
+#'
+#' @author Sarah Endicott \email{sarah.endicott@canada.ca}
+#'
+#' @date 2024/03/12
+
+# I copied these from the rocker/r-bspm Dockerfile because it sets them in the
+# RProfile but it wasn't working
+if(requireNamespace("bspm", quietly = TRUE)){
+  message("setting up bspm for use on cloud")
+  bspm::enable()
+  options(pkgType="binary", install.packages.check.source = "no")
+} else {
+  r = getOption("repos")
+  r["CRAN"] = "https://cran.rstudio.com/"
+  options(repos = r)
+}
+
+real <- as.character(commandArgs(trailingOnly = TRUE)[[1]])
+cat(print(real))
+
+# set PAT to avoid api limit errors on GitHub
+Sys.setenv(GITHUB_PAT = "<pat>")
+
+install.packages("remotes")
+## Install Dependencies (listed in DESCRIPTION) ----
+print("install deps")
+install.packages("dplyr")
+install.packages("sf")
+install.packages("terra")
+install.packages("purrr")
+install.packages("tidyr")
+install.packages("tibble")
+remotes::install_github("LandSciTech/roads")
+remotes::install_github("LandSciTech/caribouMetrics")
+remotes::install_github("LandSciTech/pfocal")
+# out <- try(Require::Require(c("dplyr", "sf", "terra", "tpq/peakRAM", "LandSciTech/roads")))
+# cat("installed from binary:\n")
+# print(out)
+#
+# out2 <- try(Require::Require(c("dplyr", "sf", "terra", "tpq/peakRAM", "LandSciTech/roads"),
+#                              type = "source", upgrade = "never"))
+# cat("installed from source?:\n")
+# print(out2)
+flnm <- paste0("7_compile_results_cloud_", real, ".R")
+source(flnm)

### 2024-04-04 analysis/cloud/setup_azure.sh

@@ -39,6 +39,17 @@ done
 
 sed 's,<pat>,'$ghpat',g' analysis/cloud/make.R > analysis/cloud/make_to_use.R
 
+# files for combine
+sed 's,<subnetId>,'${subnetid//&/\\&}',g' analysis/cloud/pool_combine.json\
+| sed 's,<id>,'${poolName}',g'> analysis/cloud/pool_to_use.json
+
+for rowi in real notreal real1000
+do
+  sed 's,<SASURL>,'${sasurl//&/\\&}',g' analysis/cloud/task_combine.json\
+  | sed 's,<row>,'$rowi',g' > analysis/cloud/task_jsons/task_combine_$rowi.json
+done
+
+sed 's,<pat>,'$ghpat',g' analysis/cloud/make_combine.R > analysis/cloud/make_to_use.R
 
 #### Move files to container ##############
 # First check container is empty. All files in container will be copied to nodes
@@ -56,25 +67,42 @@ az storage copy -d $sasurl -s analysis/data/derived_data/TSA27/dem_revelstoke_10
 az storage copy -d $sasurl -s analysis/scripts/6_benchmark_methods.R
 az storage copy -d $sasurl -s R/functionsSource.R
 
+# files for combine task
+az storage copy -d $sasurl -s analysis/data/derived_data/TSA27/dem_revelstokeCoarse.tif
+
+az storage copy -d $sasurl -s analysis/scripts/7_compile_results_cloud_notreal.R
+az storage copy -d $sasurl -s analysis/scripts/7_compile_results_cloud_real.R
+az storage copy -d $sasurl -s analysis/scripts/7_compile_results_cloud_real1000.R
+
+az storage copy -d $sasurl -s analysis/data/derived_data/TSA27_real_cuts/klementProjection.shp
+az storage copy -d $sasurl -s analysis/data/derived_data/TSA27_real_cuts/klementProjection.shx
+az storage copy -d $sasurl -s analysis/data/derived_data/TSA27_real_cuts/klementProjection.dbf
+az storage copy -d $sasurl -s analysis/data/derived_data/TSA27_real_cuts/klementProjection.prj
 #### Create pool, job, tasks ##########################
 az batch pool create --json-file analysis/cloud/pool_to_use.json
 az batch job create --pool-id $poolName --id $jobName
 
-# 1 2 3 4 5 6
-for rowi in 7 8 9 10 11 13 14 15 16 17 19 20 21 22
+
[…]

### 2024-04-04 analysis/scripts/7_compile_results.R

@@ -69,7 +69,7 @@ run_projections(param_tbl %>% filter(agg == 100) %>%
                 roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
                 tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
                 costPth = paste0(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"),
-                outPth = paste0(data_path_drvd, "TSA27_real_cuts/dem/"),
+                outPth = paste0(data_path_drvd, "TSA27_real_cuts/dem_1000/"),
 
                 #Klement QGIS projection results shapefile
                 klementProj = paste0(data_path_drvd, "TSA27_real_cuts/", "klementProjection.shp"),
@@ -77,4 +77,46 @@ run_projections(param_tbl %>% filter(agg == 100) %>%
                 aggFact = 10, #factor of aggregation of cost surface. 1 = no aggregation.
                 load_file = "results",
                 saveInputs = FALSE,
+                replaceNA = -65000)
+
+run_projections(param_tbl %>% filter(agg == 10, cutblocks_real == "revelstoke_real") %>%
+                  rowwise() %>%
+                  mutate(across(everything(), as.character)) %>%
+                  mutate(output = file.path(data_path_drvd, "bench_results",
+                                            paste0("result_",
+                                                   paste0(c_across(everything()), collapse = "_"),
+                                                   ".rds"))),
+                cutblocksPth = paste0(data_path_drvd, "cutblocks_revelstoke_real.gpkg"),
+                roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
+                tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
+                costPth = paste0(data_path_drvd, "TSA27/dem_revelstokeCoarse.tif"),
+                outPth = paste0(data_path_drvd, "TSA27_real_cuts/dem/"),
+
+                #Klement QGIS projection results shapefile
+                klementProj = paste0(data_path_drvd, "TSA27_real_cuts/", "klementProjection.shp"),
+
+                aggFact = 1, #factor of aggregation of cost surface. 1 = no aggregation.
+                load_file = "results",
+                saveInputs = TRUE,
+                replaceNA = -65000)
+
+run_projections(param_tbl %>% filter(agg == 10, cutblocks_real == "revelstoke") %>%
+                  rowwise() %>%
[…]

### 2024-04-04 analysis/scripts/7_compile_results_cloud_notreal.R

@@ -1,0 +1,76 @@
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(purrr)
+library(caribouMetrics)
+
+# load functions used in script
+source("functionsSource.R")
+
+#set path for data
+data_path_raw <- ""
+data_path_drvd <- ""
+
+
+# compile results after running on cloud
+bench_res <- list.files(".", pattern = "bench_.*.rds",
+                        full.names = TRUE) %>%
+  purrr::map(readRDS) %>%
+  bind_rows() %>%
+  tidyr::separate(id, into = c("method", "sampleType", "sampleDens", "agg", "cutblocks_real"),
+                  sep = "_",
+                  extra = "merge", convert = TRUE)
+
+#densities to test
+low <- 0.000001
+high <- 0.00001
+
+param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
+                         stringsAsFactors = FALSE) %>%
+  mutate(sampleDens = high,
+         sampleType = "regular",
+         cutblocks_real = "revelstoke_real")
+
+# parameter table creation for running projections
+sampleDens <- c(low,high,low,high,low)
+sampleType <- c("regular","regular","random","random","centroid")
+paramTable <- tibble(method = "mst", sampleType, sampleDens, agg = 10,
+                     cutblocks_real = "revelstoke_real") %>%
+  distinct()
+
+# add row for ilcp with regular high density sampling
+paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "ilcp"))
+
+# add coarse resolution
+paramTable2 <- paramTable %>% mutate(agg = 100) %>% bind_rows(paramTable)
+
+paramTable3 <- bind_rows(paramTable2,
+                         paramTable %>% mutate(cutblocks_real = "revelstoke")) %>%
+  bind_rows(param_tbl) %>%
+  distinct()
+
+param_tbl <- paramTable3
+
+param_tbl %>% left_join(bench_res, by = c("method", "sampleType", "sampleDens",
+                                          "agg", "cutblocks_real")) %>%
+  tibble::rowid_to_column() %>%
+  filter(is.na(resolution)) %>%
+  pull(rowid)
+
+run_projections(param_tbl %>% filter(agg == 10, cutblocks_real == "revelstoke") %>%
+                  rowwise() %>%
+                  mutate(across(everything(), as.character)) %>%
+                  mutate(output = file.path(paste0("result_",
[…]

### 2024-04-04 analysis/scripts/7_compile_results_cloud_real.R

@@ -1,0 +1,80 @@
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(purrr)
+library(caribouMetrics)
+
+# load functions used in script
+source("functionsSource.R")
+
+#set path for data
+data_path_raw <- ""
+data_path_drvd <- ""
+
+
+# compile results after running on cloud
+bench_res <- list.files(".", pattern = "bench_.*.rds",
+                        full.names = TRUE) %>%
+  purrr::map(readRDS) %>%
+  bind_rows() %>%
+  tidyr::separate(id, into = c("method", "sampleType", "sampleDens", "agg", "cutblocks_real"),
+                  sep = "_",
+                  extra = "merge", convert = TRUE)
+
+#densities to test
+low <- 0.000001
+high <- 0.00001
+
+param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
+                         stringsAsFactors = FALSE) %>%
+  mutate(sampleDens = high,
+         sampleType = "regular",
+         cutblocks_real = "revelstoke_real")
+
+# parameter table creation for running projections
+sampleDens <- c(low,high,low,high,low)
+sampleType <- c("regular","regular","random","random","centroid")
+paramTable <- tibble(method = "mst", sampleType, sampleDens, agg = 10,
+                     cutblocks_real = "revelstoke_real") %>%
+  distinct()
+
+# add row for ilcp with regular high density sampling
+paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "ilcp"))
+
+# add coarse resolution
+paramTable2 <- paramTable %>% mutate(agg = 100) %>% bind_rows(paramTable)
+
+paramTable3 <- bind_rows(paramTable2,
+                         paramTable %>% mutate(cutblocks_real = "revelstoke")) %>%
+  bind_rows(param_tbl) %>%
+  distinct()
+
+param_tbl <- paramTable3
+
+param_tbl %>% left_join(bench_res, by = c("method", "sampleType", "sampleDens",
+                                          "agg", "cutblocks_real")) %>%
+  tibble::rowid_to_column() %>%
+  filter(is.na(resolution)) %>%
+  pull(rowid)
+
+run_projections(param_tbl %>% filter(agg == 10, cutblocks_real == "revelstoke_real") %>%
+                  rowwise() %>%
+                  mutate(across(everything(), as.character)) %>%
+                  mutate(output = file.path(paste0("result_",
[…]

### 2024-04-04 analysis/scripts/7_compile_results_cloud_real1000.R

@@ -1,0 +1,62 @@
+library(dplyr)
+library(sf)
+library(roads)
+library(terra)
+library(purrr)
+library(caribouMetrics)
+
+# load functions used in script
+source("functionsSource.R")
+
+#set path for data
+data_path_raw <- ""
+data_path_drvd <- ""
+
+#densities to test
+low <- 0.000001
+high <- 0.00001
+
+param_tbl <- expand.grid(method = c("ilcp", "mst"), agg = c(1, 3, 10, 50, 100),
+                         stringsAsFactors = FALSE) %>%
+  mutate(sampleDens = high,
+         sampleType = "regular",
+         cutblocks_real = "revelstoke_real")
+
+# parameter table creation for running projections
+sampleDens <- c(low,high,low,high,low)
+sampleType <- c("regular","regular","random","random","centroid")
+paramTable <- tibble(method = "mst", sampleType, sampleDens, agg = 10,
+                     cutblocks_real = "revelstoke_real") %>%
+  distinct()
+
+# add row for ilcp with regular high density sampling
+paramTable <- bind_rows(paramTable, paramTable %>% slice(2) %>% mutate(method = "ilcp"))
+
+# add coarse resolution
+paramTable2 <- paramTable %>% mutate(agg = 100) %>% bind_rows(paramTable)
+
+paramTable3 <- bind_rows(paramTable2,
+                         paramTable %>% mutate(cutblocks_real = "revelstoke")) %>%
+  bind_rows(param_tbl) %>%
+  distinct()
+
+param_tbl <- paramTable3
+
+
+run_projections(param_tbl %>% filter(agg == 100) %>%
+                  rowwise() %>%
+                  mutate(across(everything(), as.character)) %>%
+                  mutate(output = file.path(paste0("result_",
+                                                   paste0(c_across(everything()), collapse = "_"),
+                                                   ".rds"))),
+                cutblocksPth = paste0(data_path_drvd, "cutblocks_revelstoke_real.gpkg"),
+                roadsPth = paste0(data_path_drvd, "combined_revelstoke_roads.gpkg"),
+                tsaBoundaryPth = paste0(data_path_raw, "tsa27_boundaries.gpkg"),
+                costPth = paste0(data_path_drvd, "dem_revelstokeCoarse.tif"),
+                outPth = paste0(data_path_drvd, "TSA27_real_cuts/dem_1000/"),
+
+                aggFact = 10, #factor of aggregation of cost surface. 1 = no aggregation.
[…]

### 2024-04-04 analysis/paper/paper.Rmd

@@
 # re-run projections for just the small area with different parameters
+set.seed(4)
 mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
@@
                       roadsInWeight = F)
-
+set.seed(4)
 ilcpProj <- projectAll(tsbs = tsb,
@@
 # re-run projections for just the small area with different parameters
+set.seed(4)
 mstProjDem <- projectAll(tsbs = tsb, paramTable = paramTableDem,
@@
                          roadsInWeight = F)
-
+set.seed(4)
 ilcpProjDem <- projectAll(tsbs = tsb,
@@
                           roadsInWeight = F)
+
+terra::writeRaster(tsaCost, here(data_path_drvd,"..",  "for_fig","HardyQGISTestweight.tif"),overwrite=T)
 
@@
                             "HardyQGISTest.gpkg")))) %>%
-  mutate(mapTitle =  paste(weightMethod,
+  mutate(mapTitle =  paste(ifelse(is.na(weightMethod), "", weightMethod),
                            ifelse(method == "mst", "MST",
                                   ifelse(method == "ilcp", "ILCP", method)),
-                           sampleType,
-                           ifelse(sampleDens == 1e-06 &
-                                    sampleType != "centroid", "low density",
-                                  ifelse(sampleDens == 1e-05,
-                                         "high density", ""))))
+                           sampleType) %>%
+           trimws())
 
@@
   ~ qtm(cutblocks, borders = "#92c5de", fill="#92c5de", borders.lwd = 1,)+
-    qtm(tsaCost, raster.style = "cont", raster.palette = "Greys",
-        raster.alpha = 0.25, raster.title = "Cost")+
+    qtm(rast(here(gsub(".gpkg","weight.tif",allProj$output[[.x]],fixed=T))),
+        raster.style = "cont", raster.palette = "Greys",
+        raster.alpha = 0.25, raster.title = NA, raster.legend.show = FALSE)+
     qtm(roads, lines.col = "#0571b0", lines.lwd = 2,lines.lty="solid")+
@@
 allMaps[[nrow(allProj)]] <- allMaps[[nrow(allProj)]]+
+  tm_add_legend(type = "fill",
+                labels = c("low", "med", "high"),
+                col = c("grey99", "grey87", "grey75"),
+                border.col= c("grey99", "grey87", "grey75"),
+                title = "weightRaster")+
   tm_add_legend(type = "line",

### 2024-04-17 analysis/paper/paper.Rmd
integrating new results - in progress

@@
 # load csv
-meanTable <- read.csv(here(data_path_drvd,"../TSA27", "mean_table.csv"))
-matchData <- read.csv(here(data_path_drvd,"../TSA27", "agree_table.csv"))
-meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
                                    "mean_table.csv"))
 
-matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
                                    "agree_table.csv"))

### 2024-04-17 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -69,6 +69,7 @@ paramTable$weightFunction = deparse1(simpleCostFn,collapse="\n")
 paramTable$weightMethod = "Simple cost"
 
 paramTableDem = paramTable
+
 paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
 paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
 paramTableDem$weightMethod = "Grade penalty"

### 2024-04-17 analysis/paper/paper.Rmd
minor changes, found issue with doubled roads, investigating

@@
 #cost surface raster layer
-tsaCost <- terra::rast(here(data_path_drvd, "old cost/input_cost.tif"))
+tsaCost <- terra::rast(here(data_path_drvd, "input_cost.tif"))
 dem <- terra::rast(here(data_path_drvd, "dem/input_cost.tif"))

### 2024-04-17 analysis/scripts/5_make_fig4.Rmd
minor changes, found issue with doubled roads, investigating

@@
 data_path_raw <- "analysis/data/raw_data/"
-data_path_drvd <- "analysis/data/derived_data/TSA27"
+data_path_drvd <- "analysis/data/derived_data/TSA27_real_cuts/dem"
 
@@
 
-e_rds <- st_read(here::here(file.path(data_path_drvd, "input_roads.gpkg")))
-p_rds <- st_read(here::here(file.path(data_path_drvd, "regular_1e-05_mst.gpkg")))
-o_rds <- st_read(here::here(file.path(data_path_drvd, "../combined_revelstoke_roads.gpkg")))
+e_rds <- st_read(here::here(file.path(data_path_drvd, "../input_roads.gpkg")))
+p_rds <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke_real.rds"))
+o_rds <- st_read(here::here(file.path(data_path_drvd, "../../combined_revelstoke_roads.gpkg")))
 
@@
 
+# p_rds_ilcp <- readRDS(here::here("analysis/data/derived_data/bench_results/result_ilcp_regular_1e-05_10_revelstoke_real.rds"))
+# 
+# write_sf(p_rds_ilcp, here::here("analysis/data/derived_data/bench_results/ilcp_regular_1e-05_10_revelstoke_real.gpkg"))
+#  p_rds_mst <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke_real.rds"))
+# 
+# write_sf(p_rds_mst, here::here("analysis/data/derived_data/bench_results/mst_regular_1e-05_10_revelstoke_real.gpkg"))
 ```

### 2024-04-17 analysis/cloud/setup_azure.sh

@@ -102,7 +102,7 @@ done
 # done
 
 # set target dedicated nodes
-az batch pool resize --pool-id $poolName --target-dedicated-nodes 2
+az batch pool resize --pool-id $poolName --target-dedicated-nodes 3
 
 # prompt auto scaleing of pool by changing time interval
 # enabling this as soon as the tasks are created seems to make it think there are no tasks
@@ -156,11 +156,13 @@ az storage blob list -c sendicott --account-name ecdcwls --sas-token $sastoken \
 az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/*?$sastoken \
 -d "analysis/data/derived_data/bench_results" --include-pattern "*.rds"
 
-az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/\$AZ_BATCH_TASK_ID/TSA27_real_cuts?$sastoken \
+az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/\$AZ_BATCH_TASK_ID?$sastoken \
 -d "analysis/data/derived_data/cloud_combine_results" --recursive
 
+az storage remove -c sendicott -n taskId --account-name ecdcwls --sas-token $sastoken --recursive
+
 # Remove just files matching pattern
-az storage remove -c sendicott --include-pattern "*.rds" --account-name ecdcwls --sas-token $sastoken --recursive
+az storage remove -c sendicott --include-pattern "*.txt" --account-name ecdcwls --sas-token $sastoken --recursive
 
 # NOTE removes ***everything*** from the storage container
 az storage remove -c sendicott --exclude-pattern "*.gpkg" --account-name ecdcwls --sas-token $sastoken --recursive

### 2024-04-18 analysis/paper/paper.Rmd
add benchmarking results to paper, fixed missing times

@@
 doAspatPlot(meanTableRealCuts)
-
-# save runTimes for later comparison
-ilcp_time <- meanTableRealCuts %>% 
-  filter(method == "ilcp", areaMean == "overall") %>% 
-  pull(runTime) %>% round(2)
-
-mst_time <- meanTableRealCuts %>%
-  filter(method == "mst", areaMean == "overall",
-         sampleType == "regular", sampleDens == "high sample density") %>% 
-  pull(runTime) %>% round(2) %>% `*`(60)
-
 ```
@@
 The ILCP projection method had comparable performance to the MST projection method with the same sampling strategy.
-The ILCP method was faster to compute however, with a run time of `r ilcp_time` minutes compared to `r mst_time` minutes for the MST method.
+The ILCP method was faster to compute however, with a run time of `r ilcp_time` hours compared to `r mst_time` hours for the MST method.
 
-<!-- TO DO: Sarah figure out why `r ilcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->
+```{r speed-fig, fig.height=5, fig.width=fig_widths["one"], fig.cap=""}
+# Get run times from bench marking results
+# compile results after running on cloud
+bench_res <- list.files(here::here("analysis/data/derived_data/bench_results"),
+                        pattern = "bench_.*.rds",
+                        full.names = TRUE) %>%
+  purrr::map(readRDS) %>%
+  bind_rows() %>%
+  tidyr::separate(id, into = c("method", "sampleType", "sampleDens", "agg", "cutblocks_real"),
+                  sep = "_",
+                  extra = "merge", convert = TRUE)
+
+bench_res %>% filter(cutblocks_real == "revelstoke_real", sampleDens == 1e-05,
+                     sampleType == "regular") %>% 
+  pivot_longer(c(Elapsed_Time_sec, Peak_RAM_Used_MiB), names_to = "variable", 
+               values_to = "values") %>% 
+  ggplot(aes(y = values, x = toupper(method), fill = as.factor(resolution)))+
+  geom_col(position = position_dodge())+
+  scale_fill_discrete(type = met.brewer(pal_nm, 3))+
+  facet_wrap(~variable, nrow = 2, scales = "free", 
+             labeller = labeller(variable = c(Elapsed_Time_sec = "Time (s)", 
+                                              Peak_RAM_Used_MiB = "Peak RAM used (MiB)"))) +
+  labs(y = NULL, x = "Projection method", fill = "Resolution")
+
+ilcp_time <- bench_res %>% 
+  filter(method == "ilcp", cutblocks_real == "revelstoke_real",
+         resolution == 100) %>% 
+  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
+  units::set_units("hours") %>% round(2) %>% units::drop_units()
+
+mst_time <- bench_res %>% 
+  filter(method == "mst", cutblocks_real == "revelstoke_real",
+         resolution == 100, sampleDens == 1e-05, sampleType == "regular") %>% 
+  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
+  units::set_units("hours") %>% round(2) %>% units::drop_units()
+
+```
 

### 2024-04-18 analysis/paper/paper.Rmd
update paths are required packages, paper and supplement now run on my machine

@@
 
-As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
-the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
-Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
-The ILCP projection method had comparable performance to the MST projection method with the same sampling strategy.
-The ILCP method was faster to compute however, with a run time of `r ilcp_time` hours compared to `r mst_time` hours for the MST method.
-
-```{r speed-fig, fig.height=5, fig.width=fig_widths["one"], fig.cap=""}
+```{r speed-fig, fig.height=5, fig.width=fig_widths["single"], fig.cap=""}
 # Get run times from bench marking results
@@
 ```
+
+As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
+the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
+Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
+The ILCP projection method had comparable performance to the MST projection method with the same sampling strategy.
+The ILCP method was faster to compute however, with a run time of `r ilcp_time` hours compared to `r mst_time` hours for the MST method.
 

### 2024-04-18 analysis/paper/supplement.Rmd
update paths are required packages, paper and supplement now run on my machine

@@
 Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
-meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))
+meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
 doAspatPlot(meanTable)
@@
 Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
-matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))
+matchData <- read.csv(here(data_path_drvd, "dem", "agree_table.csv"))
 perf_tbl2 <- doSpatPerf(matchData)
@@
 ```{r get-data2}
-data_path_drvd <- "analysis/data/derived_data/TSA27_1000"
+data_path_drvd <- "analysis/data/derived_data/TSA27_real_cuts/"
 
 # load csv
-meanTable2 <- read.csv(here(data_path_drvd, "mean_table.csv")) %>% 
+meanTable2 <- read.csv(here(data_path_drvd, "dem_1000", "mean_table.csv")) %>% 
   mutate(resolution = "1000")
 
-matchData2 <- read.csv(here(data_path_drvd, "agree_table.csv")) %>% 
+matchData2 <- read.csv(here(data_path_drvd, "dem_1000", "agree_table.csv")) %>% 
   mutate(resolution = "1000")
 
+# compare to real cuts
+meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
 meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
@@
 
+matchData <- read.csv(here(data_path_drvd, "dem", "agree_table.csv"))
 matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 

### 2024-04-18 R/functionsSource.R

@@ -657,7 +657,7 @@ doSpatPerf <- function(tab){
 
   # performance table
   perf_tbl <- matchData_sum %>%
-    pivot_wider(c(metric, sampleDens), names_from = agreement,
+    pivot_wider(id_cols = c(metric, sampleDens), names_from = agreement,
                 values_from = count, values_fill = 0) %>%
     mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
            precision = `Agree roaded`/(`Agree roaded` + `False positive`),

### 2024-04-18 analysis/cloud/setup_azure.sh

@@ -4,7 +4,9 @@
 
 az login
 az batch account login -g EcDc-WLS-rg -n ecdcwlsbatch
-az batch pool list
+az batch pool list  \
+--query "[].{id:id, curNodes: currentDedicatedNodes,tarNodes: targetDedicatedNodes, allocState:allocationState}" \
+--output yaml
 
 #### Set ids and names ##############
 subnetid=$(az network vnet subnet list --resource-group EcDc-WLS-rg --vnet-name EcDc-WLS-vnet \
@@ -102,7 +104,7 @@ done
 # done
 
 # set target dedicated nodes
-az batch pool resize --pool-id $poolName --target-dedicated-nodes 3
+az batch pool resize --pool-id $poolName --target-dedicated-nodes 20
 
 # prompt auto scaleing of pool by changing time interval
 # enabling this as soon as the tasks are created seems to make it think there are no tasks
@@ -159,10 +161,10 @@ az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/*?$sastoken \
 az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/\$AZ_BATCH_TASK_ID?$sastoken \
 -d "analysis/data/derived_data/cloud_combine_results" --recursive
 
-az storage remove -c sendicott -n taskId --account-name ecdcwls --sas-token $sastoken --recursive
+az storage remove -c sendicott -n \$AZ_BATCH_TASK_ID --account-name ecdcwls --sas-token $sastoken --recursive
 
 # Remove just files matching pattern
-az storage remove -c sendicott --include-pattern "*.txt" --account-name ecdcwls --sas-token $sastoken --recursive
+az storage remove -c sendicott --include-pattern "*.rds" --account-name ecdcwls --sas-token $sastoken --recursive
 
 # NOTE removes ***everything*** from the storage container
 az storage remove -c sendicott --exclude-pattern "*.gpkg" --account-name ecdcwls --sas-token $sastoken --recursive

### 2024-04-18 analysis/cloud/setup_azure.sh
commit csv tables of results

@@ -104,7 +104,7 @@ done
 # done
 
 # set target dedicated nodes
-az batch pool resize --pool-id $poolName --target-dedicated-nodes 20
+az batch pool resize --pool-id $poolName --target-dedicated-nodes 3
 
 # prompt auto scaleing of pool by changing time interval
 # enabling this as soon as the tasks are created seems to make it think there are no tasks
@@ -131,14 +131,14 @@ az batch node list --pool-id $poolName --query "{nodes: [].[id, state][]}" --out
 #### Monitor tasks ############################
 
 # details for a single task filtered by query
-az batch task show --job-id $jobName --task-id roads-benchmark-1 \
+az batch task show --job-id $jobName --task-id roads-benchmark-14 \
 --query "{state: state, executionInfo: executionInfo}" --output yaml
 
 # az batch task delete --job-id $jobName --task-id roads-benchmark-1 --yes
 # az batch task reactivate --job-id $jobName --task-id roads-benchmark-1
 
 # download output file for a task
-az batch task file download --task-id roads-benchmark-1 --job-id $jobName --file-path "stdout.txt" --destination "analysis/cloud/stdout.txt"
+az batch task file download --task-id roads-benchmark-14 --job-id $jobName --file-path "stdout.txt" --destination "analysis/cloud/stdout.txt"
 
 # List of all tasks and their state
 # See here for making fancy queries https://jmespath.org/tutorial.html

### 2024-04-18 analysis/scripts/5_make_fig4.Rmd
update Fig 4

@@
 
-In QGIS open input_cutblocks and input_roads for TSA 27 as well as combined_revelstoke_roads.  Get Google satellite imagery from XYZ tile service.
+```{r}
+library(terra)
+library(sf)
+library(dplyr)
+library(tmap)
+
+# run to save projected roads for use in QGIS
+p_rds_mst <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke.rds"))
+
+write_sf(p_rds_mst, here::here("analysis/data/derived_data/bench_results/mst_regular_1e-05_10_revelstoke.gpkg"))
+```
+
+In QGIS open input_cutblocks and input_roads for TSA 27 as well as combined_revelstoke_roads and projected roads for all cutblocks.  Get Google satellite imagery from XYZ tile service.
 
@@
 ```{r}
-library(terra)
-library(sf)
-library(dplyr)
-library(tmap)
-
 data_path_raw <- "analysis/data/raw_data/"
-data_path_drvd <- "analysis/data/derived_data/TSA27_real_cuts/dem"
+data_path_drvd <- "analysis/data/derived_data/TSA27/dem"
 
@@
 e_rds <- st_read(here::here(file.path(data_path_drvd, "../input_roads.gpkg")))
-p_rds <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke_real.rds"))
+p_rds <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke.rds"))
 o_rds <- st_read(here::here(file.path(data_path_drvd, "../../combined_revelstoke_roads.gpkg")))

### 2024-04-29 analysis/paper/paper.Rmd
writing in progress

@@
 ```
+
 ```{r setup2}
+library(sf)
 library(ggplot2)
@@
 library(tmap)
-library(sf)
 
+#devtools::install_deps(here()) #try this if the next line is giving trouble.
 devtools::load_all(here())
@@
 #cost surface raster layer
-tsaCost <- terra::rast(here(data_path_drvd, "input_cost.tif"))
+tsaCost <- terra::rast(here(data_path_drvd, "old cost/input_cost.tif"))
 dem <- terra::rast(here(data_path_drvd, "dem/input_cost.tif"))
@@
 
-## Aspatial performance
+According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and supplemental figure with old results).
+We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
+The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density.
+All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projects of road presence and distance to road.
 
-For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)).
-Within all mapped cutblocks (Figure \@ref(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road.
-Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@ref(fig:aspat-perf-fig-all)).
-These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
-Figure \@ref(fig:proj-map-fig)b).
-Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road.
-However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
-None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads.
+Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
+The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
+Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms.
+Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g.
+\@ref(fig:proj-map-fig)b).
+We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement B).
 
-Another fundamental limitation of these methods is that they generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure A-\@ref(fig:figA1)).
-In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).
-
-Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)).
-Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a).
-Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms would likely reduce this discrepancy.
-Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks.
-Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
-Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.
+Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
+In other words, these road network projection methods project fewer roads than we observe.
+Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
+One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure C.1).
+In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure C.1).
 
@@
 The main map shows observed distance to the nearest road minus the projected distance.
-Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected methods do no mimic observed switchbacks."}
+Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks."}
 
@@
 
-```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster).
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Variation in spatial performance among projection method variants (1 ha cost raster).
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
@@
 
-As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
+Although these methods project fewer roads than we observe, they represent a substantial improvement over the "cutblocks only" scenario, in which no new roads are added (Figure \@ref(fig:aspat-perf-fig)).
+That said, these methods are not designed to project the exact locations of new roads, so they (unsurprisingly) do not yield accurate spatial projections of road presence (i.e.
 the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
@@
 All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
-The ILCP projection method had comparable performance to the MST projection method with the same sampling strategy.
-The ILCP method was faster to compute however, with a run time of `r ilcp_time` hours compared to `r mst_time` hours for the MST method.
+
+Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement D).
+All of these methods require construction of a large graph and therefore have substantial memory requirements; graph size depends on the number of pixels in the cost surface where road construction is possible (i.e.
+not NA), so options such as increasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements.
+Memory requirements and computation time can also be reduced by setting a smaller number of landing targets (e.g.
+only target the centroid of each cutblock).
+Hardy comparison....
 

### 2024-04-30 analysis/paper/supplement.Rmd
Add simple cost and benchmark results to Supp

@@
 ```
+
+# Simple cost results
+
+For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig-simp)).
+Within all mapped cutblocks (Figure \@ref(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road.
+Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@ref(fig:aspat-perf-fig-all)).
+These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
+Figure \@ref(fig:proj-map-fig)b).
+Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig-simp)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road.
+However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
+None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+
+Another fundamental limitation of these methods is that they generate straight road segments within pixels.
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure A-\@ref(fig:figA1)).
+In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).
+
+Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig-simp)).
+Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a).
+Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms would likely reduce this discrepancy.
+Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks.
+Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig-simp)), suggesting that these measures are easier to project.
+Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.
+
+```{r get-data}
+# load csv
+meanTable <- read.csv(here(data_path_drvd,"../TSA27", "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd,"../TSA27", "agree_table.csv"))
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+                                   "mean_table.csv"))
+
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
+                                   "agree_table.csv"))
+
+```
+
+```{r aspat-perf-fig-simp, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
+Values greater than zero indicate overprojection and negative values indicate underprojection.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
+# organize data for ggplot
+doAspatPlot(meanTableRealCuts)
+```
+
+<!-- I have not remade Figure 4 with the old results but we could pull it out of GitHub if we want it.-->
+
+```{r spat-perf-fig-simp, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster).
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
+Values closer to one indicate better performance.
+Cutblocks not accessible via the observed road network were excluded from analysis."}
+perf_tbl <- doSpatPerf(matchDataRealCuts)
+perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
+  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
+
+doSpatPlot(perf_tbl)
+
+```
+
+As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
+the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
+Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig-simp)).
 
@@
 
+# Benchmarking road projection methods
+
+```{r speed-fig, fig.width=fig_widths["two"], fig.cap=""}
+# Get run times from bench marking results
+# compile results after running on cloud
+bench_res <- list.files(here::here("analysis/data/derived_data/bench_results"),
+                        pattern = "bench_.*.rds",
+                        full.names = TRUE) %>%
+  purrr::map(readRDS) %>%
+  bind_rows() %>%
+  tidyr::separate(id, into = c("method", "sampleType", "sampleDens", "agg", "cutblocks_real"),
+                  sep = "_",
+                  extra = "merge", convert = TRUE)
+
+# Hardy timing added below
+# To run Hardy alg: 2mins 10s
+# Plus to Union with existing roads 2 mins 34s
+
+# HArdy with 10m resolution 14097.66 seconds
+
+bench_res %>%
+  add_row(cutblocks_real = "revelstoke_real", sampleType = "regular", sampleDens = 1.5e-05, 
+          method = "Hardy QGIS", Elapsed_Time_sec = 2*60+34, n_verticies = 1815825) %>% 
+  filter(cutblocks_real == "revelstoke_real", sampleType %in% c("regular", "centroid"), 
+         sampleDens > 1e-06 | sampleType == "centroid") %>% 
+  mutate(Peak_RAM_Used_MiB = Peak_RAM_Used_MiB * 0.001048576, 
+         sampleDens = ifelse(sampleType == "centroid", 0.1, sampleDens * 1000^2)) %>% 
+  pivot_longer(c(Elapsed_Time_sec, Peak_RAM_Used_MiB), names_to = "variable", 
+               values_to = "values") %>% 
+  ggplot(aes(y = values, col = interaction(method, sampleDens), x = n_verticies,
+             group = interaction(method, sampleDens), 
+             shape = interaction(method, sampleDens)))+
+  geom_line(linewidth = 1.05)+
+  geom_point(size = 2)+
+  scale_colour_manual(labels = c("ILCP centroid", "MST centroid",
+                                 "ILCP high density", "MST high density", "Hardy QGIS"),
+                      values = c("#ab84a5","#732f30", "#df8d71", "#1e5a46", "#af4f2f"))+
+  scale_shape_discrete(solid = FALSE, 
+                       labels = c("ILCP centroid", "MST centroid", 
+                                  "ILCP high density", "MST high density", "Hardy QGIS"))+
+  scale_y_log10()+
+  scale_x_log10()+
+  facet_wrap(~variable, nrow = 2, scales = "free", 
+             labeller = labeller(variable = c(Elapsed_Time_sec = "Time (s)", 
+                                              Peak_RAM_Used_MiB = "Peak RAM used (GB)"))) +
+  labs(y = NULL, x = "# Nodes", col = "Method", shape = "Method")+
+  theme_bw()
+
+ilcp_time <- bench_res %>% 
+  filter(method == "ilcp", cutblocks_real == "revelstoke_real",
+         resolution == 100) %>% 
+  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
+  units::set_units("hours") %>% round(2) %>% units::drop_units()
+
+mst_time <- bench_res %>% 
+  filter(method == "mst", cutblocks_real == "revelstoke_real",
+         resolution == 100, sampleDens == 1e-05, sampleType == "regular") %>% 
+  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
+  units::set_units("hours") %>% round(2) %>% units::drop_units()
+```
+
 # Fort Nelson map

### 2024-04-30 R/functionsSource.R

@@ -560,6 +560,7 @@ doAspatPlot <- function(tab){
     mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_,
                                sampleType),
            sampleDens = paste(method, sampleType, sampleDens) %>%
+             str_replace("dlcp", "ilcp") %>%
              factor(levels = c("mst NA centroid",
                                "NA NA observed",
                                "mst random low sample density",
@@ -627,6 +628,7 @@ doSpatPerf <- function(tab){
              str_replace("1e-05", "high density") %>%
              str_replace("centroid low density", "centroid") %>%
              str_replace("klementQGIS NA", "klementQGIS") %>%
+             str_replace("dlcp", "ilcp") %>%
              factor(levels = c("mst centroid",
                                "mst random low density",
                                "mst regular low density",

### 2024-04-30 analysis/cloud/setup_azure.sh

@@ -33,7 +33,7 @@ sed 's,<subnetId>,'${subnetid//&/\\&}',g' analysis/cloud/pool_roads.json\
 # make a different json for each task
 mkdir -p analysis/cloud/task_jsons
 
-for rowi in {1..24}
+for rowi in {25..27}
 do
   sed 's,<SASURL>,'${sasurl//&/\\&}',g' analysis/cloud/task_roads.json\
   | sed 's,<row>,'$rowi',g' > analysis/cloud/task_jsons/task_roads_$rowi.json
@@ -85,7 +85,7 @@ az batch pool create --json-file analysis/cloud/pool_to_use.json
 az batch job create --pool-id $poolName --id $jobName
 
 
-for rowi in {1..24}
+for rowi in {25..27}
 do
   az batch task create --json-file analysis/cloud/task_jsons/task_roads_$rowi.json --job-id $jobName
 done
@@ -164,7 +164,7 @@ az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/\$AZ_BATCH_TA
 az storage remove -c sendicott -n \$AZ_BATCH_TASK_ID --account-name ecdcwls --sas-token $sastoken --recursive
 
 # Remove just files matching pattern
-az storage remove -c sendicott --include-pattern "*.rds" --account-name ecdcwls --sas-token $sastoken --recursive
+az storage remove -c sendicott --include-pattern "*.txt" --account-name ecdcwls --sas-token $sastoken --recursive
 
 # NOTE removes ***everything*** from the storage container
 az storage remove -c sendicott --exclude-pattern "*.gpkg" --account-name ecdcwls --sas-token $sastoken --recursive

### 2024-04-30 analysis/paper/paper.Rmd
Writing in progress.

@@
 Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].
-In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
+In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012;@st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
 Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021].
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
+Impacts on wildlife persist after roads are no longer driveable as they to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021].
+Measures of road impacts are important for informing wildlife management decisions, and there is a need to include roads in ecological forecasting models.
 
@@
 For example, in a recent collection of articles focused on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].
+In the absence of road projection methods, researchers are left to conservatively infer the amount and randomly assign the location of roaded areas [@johnson_witnessing_2015].
 
-While road projection tools are not typically used in forest landscape models, they are important for harvest planning, and methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g.
-@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist to accomplish this task in an operational context.
-These have occasionally been incorporated into forest landscape models [@leston_quantifying_2020] but they were not designed for that purpose.
-The algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] and the level of detail required to parameterize them is often difficult to obtain at landscape scales.
-Harvest road planning tools may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.
+While road projection tools are not typically used in forest landscape models, methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g.
+@remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include road network projection tools.
+These have occasionally been incorporated into forest landscape models [e.g.
+@leston_quantifying_2020] but they were not designed for that purpose.
+The algorithms contained in proprietary closed-source software are difficult to  fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] used for research because costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for research.
+The level of detail required to parameterize forest management planning tools is also often difficult to obtain at landscape scales.
+Harvest road planning methods and tools [e.g.
+@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.
+There is a need for simple open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
 
-Promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
+Geographic route finding algorithms often draw from a larger body of work on methods for finding paths among points that minimize some measure of cost [@deo_graph_2016].
+General-purpose open source tools for graph analysis [e.g.
+@igraph_core_team_igraph_nodate] include methods for building graphs from adjacency matrices, finding “least cost paths” between pairs of points, and finding a least cost network of routes (i.e.
+a minimum spanning tree) that connects a larger set of points [@deo_graph_2016].
+More domain-specific tools [e.g.
+@van_etten_r_2017] can help make these general-purpose tools easier to use and apply in specific contexts.
+Promising simple graph-based options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
 The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
@@
 
-To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package.
+We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006;@anderson_projecting_2004;@stuckelberger_improved_2007;@clark_three-stage_2000].
+Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
+However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.
+We focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs on a regular isotropic graph from a single input raster.
+If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context.
+Both Hardy et al.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al.
+[-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water.
+We also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.
+
+To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] and a simplified version of the grade penalty approach suggested by Anderson and Nelson [-@anderson_projecting_2004] in an open source R package.
+Our intention is to produce outputs comparable to Hardy et al.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] with a tool that can be used in modelling contexts outside of QGIS/Landis, and can serve as a more easily modifiable starting point for further refinement and development.
 We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia.
@@
 Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023; @stewart_climate-informed_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
+
+We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our simplified method for penalizing steep grades to yield more realistic road networks in steep terrain.
+We do not have a similarly strong hypothesis about whether minimimum spanning tree  projections are likely to be more or less realistic than interative least cost path projections.
+If designers of real roads take a larger view to accessing multiple sites at the lowest cost we might expect minimum spanning tree algorithms to yield more realistic networks.
+However, real road networks are built over time, and it is not necessarily true that they designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.
 
@@
 The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022].
-Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014].
+Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by the Committee on the Status of Endangered Wildlife in Canada and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014].
 Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022].
@@
 
-```{r study-area1, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada."}
+```{r study-area1, fig.cap= "see next fig"}
 
@@
 
-```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada."}
+```{r study-area, fig.cap="Road construction cost surface and an elevation surface used to calculate grade penalties.
+Costs are calculated using a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada.
+In both layers, pixels containing existing roads are assigned a value of 0."}
 
@@
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-The data does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021].
 Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
@@
 Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)).
-Existing roads are assigned a cost of 0.
-The road construction cost raster consists of 1,703,936 pixels with a resolution of 1 ha.
+Existing roads are assigned a construction cost of 0.
+We compare two different methods for integrating these costs into the cost of connections between adjacent raster cells.
+In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells.
+In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure @\ref(fig:study-area)b).
+This second method allows us to penalize edge grade as suggested by Anderson and Nelson (2004), and we refer to it as the “grade penalty” method.
+The road construction cost and elevation rasters consist of 1.7 million    pixels with a resolution of 1 ha.
+We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha cost rasters.
+We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.
 
@@
 tbl <- "Cost Type|Value ($ per km)|Source
-Slope|16178 + 504*slope|Interior appraisal manual
+Slope (%)|16178 + 504*slope|Interior appraisal manual
 Major lakes|NA|Major water bodies of BC
@@
 row.names(df)<-NULL
-knitr::kable(df, caption = "A simple model of road construction costs.")
+knitr::kable(df, caption = "A simple model of road construction costs. In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000.")
 ```
@@
 All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021].
-These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository [@endicott2024a].
+These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository [@endicott_assessment_2024].
 We were not granted permission to share the digital road atlas so it must be accessed from the B.C.
-Data Catalogue [@governmentofbritishcolumbia].
-Scripts used for data acquisition and analysis associated with this paper are available on GitHub [@endicott2024].
+Data Catalogue [@data_bc_digital_2021].
+Scripts used for data acquisition and analysis associated with this paper are available on GitHub [@endicott_roadpaper_2024].
 
-## Road Projection Methods and Variants
+## Road Projection Methods
 
-We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
-Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+We focus on methods that link multiple target locations to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
+Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy personal communication).
 
-The ILCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
+The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
 For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
-The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)).
-The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007].
-The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.
-The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.
+There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006].
+The iterative least cost path algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
+Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl, Supplement X).
+Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004].
+We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.
 
-The MST approach extends the LCP approach by providing a heuristic for linking landings and evaluating road branching opportunities.
-Dean [-@dean_finding_1997] showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area.
-The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network.
-These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
-The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights and does not include loops.
-This method differs from Clark et al.
-[@clark_three-stage_2000] who used Euclidean distances for edge weights.
-
-Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network.
-One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
-In contrast our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
+Hardy’s method [-@hardy_forest_2019; -@hardy_landis-ii_2023]  is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network.
+A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
+In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
 However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network.
 It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network.
-Within cutblocks, Hardy's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
+Within cutblocks, Hardy’s method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
+
+Minimum spanning trees (MST) can also be used to link landings and evaluate road branching opportunities [@picard_finding_2006; @clark_three-stage_2000].
+MST algorithms do not require users to select the order in which landings are processed, but it remains necessary to select a set of points on the existing road network to include in the tree.
+The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network.
+These results are used to create a weighted graph wherein nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
+The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights.
+This method differs from Clark et al.
+[-@clark_three-stage_2000] who used Euclidean distances for edge weights, and thus did not incorporate spatial variation in road construction costs.
 
@@
 ```
-The MST method is expected to be closer to an optimal solution than iterative least cost path methods, and also more computationally costly.
-The Hardy QGIS plugin is much faster than our R package, and better able to handle large landscapes.
-The `igraph` package that we used to implement the Dijkstra algorithm requires the values of the cost raster to be read into memory in R which can lead to high memory usage on larger landscapes.
-See Appendix X for a more detailed exploration of the speed and memory requirements of the different methods.
+
+Note that the evolution of a road network over time can be modeled by iteratively applying any of the methods, but in this manuscript we focus on comparing outcomes of a single application of each of the methods to a network of forest roads developed since 1990.
+
+To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
+The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
+
+## Comparison of Method Variants
 
@@
 row.names(df2)<-NULL
-knitr::kable(df2, caption = "Road projection method variants.")
+knitr::kable(df2, caption = "Road projection method variants.
+We examined sensitivity of minimimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
 ```
 
-To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
+## Performance Metrics
+
+To assess performance we calculated four characteristics of the projected and observed road networks that are relevant for wildlife: road density, measured as density of roads in meters per hectare (m/ha); distance to nearest road, measured as distance from each raster cell to the nearest road; road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
+The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020] and to quantify impacts of disturbance on Southern Mountain caribou [@lochhead_linking_2022; @polfus_identifying_2011].
+The other metrics are also often used in studies of anthropogenic impacts on wildlife [e.g.
+@labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
+We also calculated road presence, measured as whether or not roads are present in each pixel, but we note that accurate projection of the exact locations of forest roads is not generally a requirement in stochastic ecological forecasting models wherein the exact locations of future disturbances are often not known.
+Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
+Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
+
+<!-- TO DO: Discuss why we do not calculate the 50m buffer metric used by Kyle for SMC.
+50 m buffer is basically built in to the 100 m raster resolution.
+Road presence would be our closest equivalent measure - and we don't project that measure very well.
+-->
+
+Aspatial performance of the projections was assessed by comparing projected and observed mean values of each road network characteristic across the timber supply area and within cutblocks.
+As an example, distance to roads within cutblocks is the average distance from non-road pixels within cutblocks to the nearest roads.
+We further distinguished mapped cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
+Spatial performance was assessed in two different ways.
+First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks.
+The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
+Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed.
+These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
+Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true.
+Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed.
+F-measure attempts to balance sensitivity and precision to give an overall score.
+
+# Results
+
+According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and supplemental figure with old results).
+We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
+The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density.
+All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projects of road presence and distance to road.
+
+Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads.
+Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
+The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
+Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms.
+Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g.
+\@ref(fig:proj-map-fig)b).
+We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement B).
+
+Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
+In other words, these road network projection methods project fewer roads than we observe.
+Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
+One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure C.1).
+In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure C.1).
+
+```{r get-data}
+# load csv
+meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
+                                   "mean_table.csv"))
+
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
+                                   "agree_table.csv"))
+
+```
+
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
+Values greater than zero indicate overprojection and negative values indicate underprojection.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
+# organize data for ggplot
+doAspatPlot(meanTableRealCuts)
+```
 
@@
 
-```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), iterative least cost paths (ILCP), and the Hardy QGIS plugin road network projection algorithms.
-Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2).
-Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red.
-Cost (1 ha resolution) is indicated by grey shading.
-The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks."}
+```{r proj-meth-fig, fig.cap="Comparison of road projection method variants (Table 2).
+The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
+Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks.
+In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row).
+Landing density is high for all except the centroid and Hardy variants."}
 knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
-```
-
-## Performance Metrics
-
-To assess performance we calculated five characteristics of the projected and observed road networks: road presence, measured as whether or not roads are present in each pixel; road density, measured as density of roads in meters per hectare (m/ha); distance to nearest road, measured as distance from each raster cell to the nearest road; road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks.
-The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020] and to quantify impacts of disturbance on Southern Mountain caribou [@lochhead_linking_2022; @polfus_identifying_2011].
-The other metrics are also often used in studies of anthropogenic impacts on wildlife [e.g.
-@labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
-Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
-Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
-
-<!-- TO DO: Discuss why we do not calculate the 50m buffer metric used by Kyle for SMC.
-50 m buffer is basically built in to the 100 m raster resolution.
-Road presence would be our closest equivalent measure - and we don't project that measure very well.
--->
-
-Aspatial performance of the projections was assessed by comparing projected and observed mean values of each road network characteristic across the timber supply area and within cutblocks.
-We further distinguished mapped cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
-Spatial performance was assessed in two different ways.
-First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks.
-The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
-Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed.
-These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
-Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true.
-Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed.
-F-measure attempts to balance sensitivity and precision to give an overall score.
-
-# Results
-
-According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and supplemental figure with old results).
-We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
-The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density.
-All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projects of road presence and distance to road.
-
-Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads.
-Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
-The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
-Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms.
-Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g.
-\@ref(fig:proj-map-fig)b).
-We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement B).
-
-Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
-In other words, these road network projection methods project fewer roads than we observe.
-Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
-One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure C.1).
-In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure C.1).
-
-```{r get-data}
-# load csv
-meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
-matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
-meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
-                                   "mean_table.csv"))
-
-matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
-                                   "agree_table.csv"))
-
-```
-
-```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
-Values greater than zero indicate overprojection and negative values indicate underprojection.
-Cutblocks not accessible via the observed road network were excluded from analysis.
-Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
-# organize data for ggplot
-doAspatPlot(meanTableRealCuts)
 ```
@@
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
-Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott2024a] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
+Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_2024_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
 

### 2024-04-30 analysis/paper/supplement.Rmd

@@
 
-# (APPENDIX) Supplementary Material for Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife {-}
+# (APPENDIX) Supplementary Material for Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife {.unnumbered}
 
@@
 
+# Grade penalty function
+
+The grade penalty method for calculating edge weights is a simplified version of the approach taken by @anderson2004a that does not distinguish between adverse and favourable grades.
+In addition, we use a modified Digital Elevation Model (DEM) as the input to the edge weight calculation to allow barriers other than slope to be included.
+The DEM is modified such that barriers where roads cannot be built are assigned NA values, barriers with high cost (e.g.
+stream crossings) can be represented with negative values, and existing roads are assigned a value of 0.
+All other values in the DEM are assumed to be elevation and the difference between the two elevations is used to determine the grade.
+Default construction cost values are taken from the BC Interior Appraisal Manual such that the base cost of building a road is \$16 178 with an additional cost of \$504 for every 1% grade and a limit of 20% grade.
+Edges with greater than 20% grade are assigned either NA or an arbitrarily high value.
+A value of 65 000 is used in our projections to avoid errors when a landing cannot be accessed without crossing an NA cell.
+If any of the cells to be connected by an edge has a negative value the edge weight is set to the larger value after taking the absolute value.
+
+```{r echo=TRUE}
+
+slopePenaltyFn <- function(x1, x2, resolution = 1, baseCost = 16178, limit = 20,
+                           penalty = 504, limitWeight = NA){
+  # If one of the nodes is a road or barrier ignore grade penalty
+  cond <- pmin(x1, x2) >= 0
+  cond[is.na(cond)] <- F
+
+  # percent slope, if both nodes have elevation values.
+  grade <- 100 * abs(x1 - x2) * cond / resolution
+
+  slp <- baseCost + grade * penalty * (pmin(x1, x2) > 0)
+  slp[grade > limit] <- limitWeight
+
+  # if both 0 this is an existing road link. Otherwise it is a barrier.
+  slp[!cond] <- abs(pmin(x1, x2))[!cond]
+
+  return(slp)
+}
+
+```
+
 # Simple cost results
 
+The simple cost method for setting edge weights generally performed worse than the grade penalty method.
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig-simp)).
@@
 However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
-None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+None of the simple cost methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
+
+```{=html}
+<!-- To remove? This is all covered in the main text and I don't think differs much here.
 
@@
 Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.
-
+-->
+```
 ```{r get-data}
@@
 
-```{r speed-fig, fig.width=fig_widths["two"], fig.cap=""}
+We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of three different resolutions 1000 m, 500 m and 100 m, which corresponded to 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
+Each cell in the raster becomes a node in the graph used to determine the shortest path which increases the time and memory requirements of the projection.
+Benchmarking was done using Azure Standard A4 v2 nodes which have 4 cores and 8 GB of RAM available, which was chosen to match a typical laptop.
+Each projection was run on a separate node and the `peakRAM` package was used to measure the total time to run the projection and the maximum memory used at any point in the analysis [@quinn2024].
+In addition, we timed the execution of the Hardy QGIS plugin with the 100 m resolution weights raster.
+
+```{r speed-fig, fig.width=fig_widths["two"], fig.cap="Speed and peak memory usage of the ILCP and MST road projection methods with landings regularly sampled at 10 per km^2^ (high density) or using cutblock centroids as landings with weight rasters of three different resolutions: 1000 m, 500 m and 100 m, which for our study area is 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
+In addition, the time to run the Hardy QGIS plugin with the 100 m raster is also shown.
+Both number of nodes and time are displayed on log transformed axes."}
 # Get run times from bench marking results
@@
 
-# Hardy timing added below
+# Hardy timing added below for 100m
 # To run Hardy alg: 2mins 10s
 # Plus to Union with existing roads 2 mins 34s
+# Hardy with 10m resolution 14097.66 seconds
+hardy_time <- 2*60+34
 
-# HArdy with 10m resolution 14097.66 seconds
+ilcp_time <- bench_res %>% 
+  filter(method == "ilcp", cutblocks_real == "revelstoke_real",
+         resolution == 100) %>% 
+  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
+  units::set_units("hours") %>% round(2) %>% units::drop_units()
 
+mst_time <- bench_res %>% 
+  filter(method == "mst", cutblocks_real == "revelstoke_real",
+         resolution == 100, sampleDens == 1e-05, sampleType == "regular") %>% 
+  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
+  units::set_units("hours") %>% round(2) %>% units::drop_units()
+
+minor_breaks <- rep(1:4*2, 21)*(10^rep(-10:10, each=9))
 bench_res %>%
   add_row(cutblocks_real = "revelstoke_real", sampleType = "regular", sampleDens = 1.5e-05, 
-          method = "Hardy QGIS", Elapsed_Time_sec = 2*60+34, n_verticies = 1815825) %>% 
+          method = "Hardy QGIS", Elapsed_Time_sec = hardy_time, n_verticies = 1815825) %>% 
   filter(cutblocks_real == "revelstoke_real", sampleType %in% c("regular", "centroid"), 
@@
                                   "ILCP high density", "MST high density", "Hardy QGIS"))+
-  scale_y_log10()+
-  scale_x_log10()+
+  scale_y_log10(minor_breaks = minor_breaks)+
+  scale_x_log10(minor_breaks = minor_breaks, limits = c(1e04, NA))+
   facet_wrap(~variable, nrow = 2, scales = "free", 
@@
   theme_bw()
+```
 
-ilcp_time <- bench_res %>% 
-  filter(method == "ilcp", cutblocks_real == "revelstoke_real",
-         resolution == 100) %>% 
-  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
-  units::set_units("hours") %>% round(2) %>% units::drop_units()
-
-mst_time <- bench_res %>% 
-  filter(method == "mst", cutblocks_real == "revelstoke_real",
-         resolution == 100, sampleDens == 1e-05, sampleType == "regular") %>% 
-  pull(Elapsed_Time_sec) %>% units::set_units("sec") %>%
-  units::set_units("hours") %>% round(2) %>% units::drop_units()
-```
+Our benchmarking showed a log-linear relationship between the number of nodes in the graph and the total execution time and memory requirements with the MST method, while the ILCP method had a shallower slope that increased slightly with larger numbers of nodes (Figure \@ref(fig:speed-fig)).
+Other than for the smallest graph the ILCP method was always faster than the MST method and projections were always faster when the centroid was used rather than high density landings (Figure \@ref(fig:speed-fig)).
+The Hardy method was significantly faster than any of the R package methods.
 
@@
 
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage
+
 # Colophon
@@
 # what commit is this file at? 
-#if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())
 ```

### 2024-04-30 analysis/paper/paper.Rmd
minor edits to intro, methods, results

@@
 Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].
-Impacts on wildlife persist after roads are no longer driveable as they to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021].
+Impacts on wildlife persist after roads are no longer driveable as they continue to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021].
 Measures of road impacts are important for informing wildlife management decisions, and there is a need to include roads in ecological forecasting models.
@@
 [-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water.
-We also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.
+Here we assess this approach and also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.
 
@@
 Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
+We further filtered cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
+Results including all cutblocks are also presented.
 
@@
 We compare two different methods for integrating these costs into the cost of connections between adjacent raster cells.
-In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells.
-In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure @\ref(fig:study-area)b).
+In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells, and we refer to it as the “simple cost” method.
+In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b).
 This second method allows us to penalize edge grade as suggested by Anderson and Nelson (2004), and we refer to it as the “grade penalty” method.
@@
 The iterative least cost path algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl, Supplement X).
-Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004].
+Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A).
+Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004].
 We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.
@@
 As an example, distance to roads within cutblocks is the average distance from non-road pixels within cutblocks to the nearest roads.
-We further distinguished mapped cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
 Spatial performance was assessed in two different ways.
@@
 
-According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and supplemental figure with old results).
+According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure B.1).
 We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
 The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density.
-All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projects of road presence and distance to road.
+All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.
 
@@
 \@ref(fig:proj-map-fig)b).
-We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement B).
+We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement C).
 
@@
 One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure C.1).
-In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure C.1).
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure D.1).
+In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure D.1).
 
@@
 
-bench_res %>% filter(cutblocks_real == "revelstoke_real", sampleDens == 1e-05,
-                     sampleType == "regular") %>% 
-  pivot_longer(c(Elapsed_Time_sec, Peak_RAM_Used_MiB), names_to = "variable", 
-               values_to = "values") %>% 
-  ggplot(aes(y = values, x = toupper(method), fill = as.factor(resolution)))+
-  geom_col(position = position_dodge())+
-  scale_fill_discrete(type = met.brewer(pal_nm, 3))+
-  facet_wrap(~variable, nrow = 2, scales = "free", 
-             labeller = labeller(variable = c(Elapsed_Time_sec = "Time (s)", 
-                                              Peak_RAM_Used_MiB = "Peak RAM used (MiB)"))) +
-  labs(y = NULL, x = "Projection method", fill = "Resolution")
+# Hardy time measured in QGIS with 100m raster
+hardy_time <- 2*60+34
 
@@
 
-Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement D).
+Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement E).
 All of these methods require construction of a large graph and therefore have substantial memory requirements; graph size depends on the number of pixels in the cost surface where road construction is possible (i.e.
-not NA), so options such as increasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements.
+not NA), so options such as decreasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements.
 Memory requirements and computation time can also be reduced by setting a smaller number of landing targets (e.g.
 only target the centroid of each cutblock).
-Hardy comparison....
+The Hardy method, run in QGIS, was an order of magnitude faster than any of our methods run in R.
 

### 2024-05-01 analysis/paper/supplement.Rmd
add profiling to Supp

@@
 
+# Profiling
+
+Profiling with the `Rprof` function can reveal which parts of a function take the longest.
+On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods.
+Constructing the graph took ~ 4 s for both methods.
+For the ILCP method, the iteratively determining the shortest path is the only other step that takes more that a second to run.
+Each iteration is relatively fast but must be repeated for each landing.
+For the MST method, the longest running step is the building of the second graph which is used to determine the MST.
+Determining the shortest path to connect the parts of the MST also takes about 2 seconds.
+
+```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings.
+Functions higher in the call stack are called by functions lower down.", fig.height=2}
+if(!file.exists(here::here("analysis/data/derived_data/ilcp_Rprof.out"))){
+  source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
+}
+
+profstat(here::here("analysis/data/derived_data/ilcp_Rprof.out"))
+
+# interactive version
+# profvis::profvis(prof_input = here::here("analysis/data/derived_data/ilcp_Rprof.out"))
+```
+
+```{r profiling-mst, fig.cap="Profiling results for the MST method showing functions that take > 1 s to run on a cost raster with 1 million cells.
+Functions higher in the call stack are called by functions lower down.", fig.height=5}
+profstat(here::here("analysis/data/derived_data/mst_Rprof.out"))
+```
+
 # Fort Nelson map

### 2024-05-01 R/functionsSource.R

@@ -684,3 +684,37 @@ doSpatPlot <- function(tab){
     theme_classic()+
     labs(y = "Performance", x = NULL)
 }
+
+# static version of profvis results filtered of only long running parts
+profstat <- function(file_nm, min_depth = 6, max_depth = 11, min_time = 1){
+  pr <- profvis::parse_rprof(file_nm)
+
+  pr[["prof"]] %>% filter(between(depth, min_depth, max_depth)) %>%
+    select(-c(filenum, filename, linenum, memalloc, meminc)) %>%
+    pivot_wider(names_from = depth, values_from = label, values_fill = "blank") %>%
+    mutate(across(c(everything(), -time), ~ifelse(lag(.x) != .x, 1, 0) %>%
+                    coalesce(0), .names = "{.col}_fun_id" )
+    ) %>%
+    pivot_longer(-time, names_to = c("depth"), values_to = c("value"),
+                 values_transform = as.character, values_drop_na = TRUE) %>%
+    separate(depth, into = c("depth", "fun_id"), sep = "_", extra = "merge",
+             fill = "right") %>%
+    mutate(fun_id = replace_na(fun_id, "label")) %>%
+    pivot_wider(names_from = fun_id, values_from = value) %>%
+    filter(!is.na(label)) %>%
+    group_by(depth) %>%
+    mutate(fun_id2 = cumsum(fun_id) + 1,
+           depth = as.numeric(depth)) %>%
+    group_by(depth, label, fun_id2) %>%
+    # convert from intervals to seconds
+    summarise(st_time = min(time)*pr$interval/1000,
+              end_time = (max(time)*pr$interval/1000) -0.01, .groups = "drop") %>%
+    mutate(tot_time = end_time - st_time) %>%
+    filter(tot_time >= min_time, label != "blank") %>%
+    ggplot(aes(y = depth, label = label))+
+    geom_linerange(aes(xmin = st_time, xmax = end_time))+
+    geom_text(aes(y = depth+0.2, x = st_time), hjust = 0)+
+    scale_x_continuous(expand = expansion(mult = c(0, 0.15)))+
+    labs(y = "Call stack", x = "Time (s)")+
+    theme_classic()
+}

### 2024-05-08 analysis/paper/paper.Rmd
writing in progress

@@
 @labadie_global_2023].
-Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide a complementary and comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that better predicts road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
+Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in this context.
 
-We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model.
-Thus, we were not surprised to see less accurate projections of road density and presence than in less mountainous regions of eastern Canada with more careful accounting for the cost of water crossings [@hardy_landis-ii_2023].
-Water crossing behaviour in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains.
-A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density become less accurate as raster cell resolution increases.
+We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region.
+The simplified grade penalty method does not increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A).
+We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it.
+However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have also allowed the possibility for users of the `roads` R package to define their own method for determining edge weights from adjacent node values.
+
+Among the method variants we considered, the minimum spanning tree (MST) regular high density method yielded better projections of average road density and the forestry disturbance footprint (Figure 3).
+Aspatial projections of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3), so overall the MST regular high density projections were more accurate.
+However, the MST algorithm is also more computationally demanding than the other methods we considered (Supplement E), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
 
 Aspatial metrics of overall projection accuracy can be difficult to interpret.
-To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies.
-Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results.
-These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns.
+This examination highlighted some errors that could be fixed by better accounting for the cost of water crossing in our weight raster, and other more fundamental limitations and differences among methods.
+Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while the Hardy method produces curves (Figure 4).
+We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.
 
-Outside cutblocks, we saw very little variation in performance among projection method variants.
-All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads.
-Outside cutblocks our ILCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads.
-Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm.
-The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations \>200 m away on a 100 m grid.
-One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.
-
-Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are more costly to develop and use.
+Although the grade penalty method improves performance by aligning roads with elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in real road networks.
+We chose not to implement Anderson and Nelson's methods (2004) for penalizing sharp curves and large junction angles because these complications would increase algorithmic complexity and computational costs.
+Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are generally more costly to develop and use.
 Before investing in additional complexity, it is important to recognize that adequacy depends on the goal.
@@
 @labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020].
-In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
+In our example landscape, projections of disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
+
+Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data.
+We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+
+These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible.
+The associated memory and computational requirements may be prohibitive for large landscapes (Supplement E).
+Options for reducing those requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s less costly implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023].
+The first three options are available to users of the `roads` R package.
+However, those contemplating reducing computational cost by increasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D).
+The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
+More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
 

### 2024-05-08 analysis/paper/supplement.Rmd
writing in progress

@@
                            penalty = 504, limitWeight = NA){
-  # If one of the nodes is a road or barrier ignore grade penalty
+  # Don't calculate grade penalty cost if one of the nodes is a barrier.
   cond <- pmin(x1, x2) >= 0
@@
 
-  # percent slope, if both nodes have elevation values.
+  # Apply grade penalty if both nodes have elevation values. 
+  # If one node is a road use base cost.
   grade <- 100 * abs(x1 - x2) * cond / resolution
@@
 
-  # if both 0 this is an existing road link. Otherwise it is a barrier.
+  # If both 0 this is an existing road link. Otherwise it is a barrier.
   slp[!cond] <- abs(pmin(x1, x2))[!cond]
@@
 Functions higher in the call stack are called by functions lower down.", fig.height=2}
-if(!file.exists(here::here("analysis/data/derived_data/ilcp_Rprof.out"))){
+if(!file.exists(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))){
   source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
@@
 
-profstat(here::here("analysis/data/derived_data/ilcp_Rprof.out"))
+profstat(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))
 
@@
 Functions higher in the call stack are called by functions lower down.", fig.height=5}
-profstat(here::here("analysis/data/derived_data/mst_Rprof.out"))
+profstat(here::here("analysis/data/derived_data/mst_100_Rprof.out"))
 ```

### 2024-05-08 analysis/scripts/test_profile.R

@@ -9,13 +9,13 @@ library(igraph)
 library(ggplot2)
 
 #speed/memory benchmarking
-data_path_raw <- "~/gitprojects/RoadPaper/analysis/data/raw_data/"
-out_path <- "~/gitprojects/RoadPaper/analysis/figures/"
+data_path_raw <- "analysis/data/raw_data/"
+out_path <- "analysis/figures/"
 
 size <- 1000
 lnds <- 10
 
-landscape <- rast(paste0(data_path_raw, "cost_surface_bc_ha.tif"))
+landscape <- rast(here::here(paste0(data_path_raw, "cost_surface_bc_ha.tif")))
 
 base_point <- c((1881188-159588)/2, (1748188-173788)/2)
 

### 2024-05-09 analysis/paper/paper.Rmd
Writing in progress

@@
   - ECCC: Environment and Climate Change Canada 
+  - UBC: University of British Columbia 
   - FLNR: British Columbia Ministry of Forests 
-  - UBC: University of British Columbia 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 Impacts on wildlife persist after roads are no longer driveable as they continue to be used by wildlife and soil compaction impedes vegetation recovery [@st-pierre_drivers_2021; @lacerte_determinants_2021].
-Measures of road impacts are important for informing wildlife management decisions, and there is a need to include roads in ecological forecasting models.
+Measures of road impacts are important for informing wildlife management decisions, and there is therefore a need to include roads in ecological forecasting models.
 
@@
 
-While road projection tools are not typically used in forest landscape models, methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g.
-@remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include road network projection tools.
+While road projections are not typically integrated with forest landscape models, projection methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g.
+@remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include projection tools.
 These have occasionally been incorporated into forest landscape models [e.g.
 @leston_quantifying_2020] but they were not designed for that purpose.
-The algorithms contained in proprietary closed-source software are difficult to  fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] used for research because costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for research.
+The algorithms contained in proprietary closed-source software are difficult to  fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] used for research because costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for use in research.
 The level of detail required to parameterize forest management planning tools is also often difficult to obtain at landscape scales.
@@
 [-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water.
-Here we assess this approach and also propose and assess a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.
+We compare this approach to a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.
 
@@
 ```{r study-area, fig.cap="Road construction cost surface and an elevation surface used to calculate grade penalties.
-Costs are calculated using a simple model informed by slope and major barriers (Table 1) in the Revelstoke timber supply area in British Colombia, Canada.
+Costs are calculated using a simple model informed by slope and major barriers (Table 1, Supplement A) in the Revelstoke timber supply area in British Colombia, Canada.
 In both layers, pixels containing existing roads are assigned a value of 0."}
@@
 We further filtered cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
-Results including all cutblocks are also presented.
 
@@
 Existing roads are assigned a construction cost of 0.
-We compare two different methods for integrating these costs into the cost of connections between adjacent raster cells.
-In the first method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells, and we refer to it as the “simple cost” method.
-In the second method, we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b).
-This second method allows us to penalize edge grade as suggested by Anderson and Nelson (2004), and we refer to it as the “grade penalty” method.
+We compare two methods for integrating these costs into the cost of connections between adjacent raster cells.
+In the first "simple cost" method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells.
+In the second "grade penalty" method (Supplement A), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), which penalization of edge grades as suggested by Anderson and Nelson (2004).
 The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha.
@@
 Slope (%)|16178 + 504*slope|Interior appraisal manual
-Major lakes|NA|Major water bodies of BC
+Major lakes|65000|Major water bodies of BC
 Existing roads|0|Digital road atlas and forest tenure roads for BC
@@
 row.names(df)<-NULL
-knitr::kable(df, caption = "A simple model of road construction costs. In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000.")
+knitr::kable(df, caption = "A simple model of road construction costs.
+In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
+These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.")
 ```
@@
 There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006].
-The iterative least cost path algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A).
+The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
+Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes.
 Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004].
@@
 It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network.
-Within cutblocks, Hardy’s method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
+Within cutblocks, Hardy et al's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
 
@@
 To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads.
-Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
-The grade penalty method improves realism by making roads follow elevation contours, but all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
+Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks, in that the grade penalty method encourages roads to follow slope contours, while the Hardy method creates more curves (Figure \@ref(fig:proj-meth-fig)).
+Although the grade penalty method improves realism by following elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
 Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms.
@@
 \@ref(fig:proj-map-fig)b).
-We omitted false cutblocks with no associated access roads from our analysis; failure to do this would distort results (Supplement C).
+We omitted false cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Supplement C).
 
@@
 As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure D.1).
-In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure D.1).
 
@@
 
-Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and more time to compute on large landscapes (Supplement E).
-All of these methods require construction of a large graph and therefore have substantial memory requirements; graph size depends on the number of pixels in the cost surface where road construction is possible (i.e.
-not NA), so options such as decreasing raster resolution or constraining the portion of the landscape where roads are plausible can help reduce memory requirements.
-Memory requirements and computation time can also be reduced by setting a smaller number of landing targets (e.g.
-only target the centroid of each cutblock).
-The Hardy method, run in QGIS, was an order of magnitude faster than any of our methods run in R.
+Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and time to compute on large landscapes (Supplement E).
+All of these methods require construction of a large graph with a node for each pixel in the cost surface where road construction is possible (i.e.
+not NA), and memory requirements increase with the number of nodes in the graph (Supplement E).
+Compute time also increases with the number of landing targets (Supplement E).
+The Hardy QGIS plugin was an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).
 
@@
 @barros_empowering_2023; @daniel_state-and-transition_2016].
-Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in this context.
+Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in that context.
 
 We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region.
-The simplified grade penalty method does not increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A).
+The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A).
 We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it.
-However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have also allowed the possibility for users of the `roads` R package to define their own method for determining edge weights from adjacent node values.
+However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have allowed the possibility for users of the `roads` R package to specify their own method for determining edge weights from adjacent node values in a weight function.
 
@@
 This examination highlighted some errors that could be fixed by better accounting for the cost of water crossing in our weight raster, and other more fundamental limitations and differences among methods.
-Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while the Hardy method produces curves (Figure 4).
+Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure 4).
 We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.
@@
 Although the grade penalty method improves performance by aligning roads with elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in real road networks.
-We chose not to implement Anderson and Nelson's methods (2004) for penalizing sharp curves and large junction angles because these complications would increase algorithmic complexity and computational costs.
+We chose not to implement Anderson and Nelson's methods (2004) for penalizing sharp curves, large junction angles and switchback spacing because these complications would increase algorithmic complexity and computational costs.
 Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are generally more costly to develop and use.
@@
 These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible.
-The associated memory and computational requirements may be prohibitive for large landscapes (Supplement E).
-Options for reducing those requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s less costly implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023].
+The associated memory and computational requirements may present challenges on large landscapes (Supplement E).
+Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s more efficient implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023].
 The first three options are available to users of the `roads` R package.
@@
 Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_2024_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
 
@@
 However, we note there are other possible uses for these road network projection methods.
-Cost minimizing methods are not suitable for projecting development of seismic lines (Figure \@ref(fig:ftNelson)), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
+Cost minimizing methods are not suitable for projecting development of seismic lines (e.g.
+Supplement G), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
 Lochhead et al.

### 2024-05-13 analysis/paper/paper.Rmd
writing in progress

@@
 [-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension.
-As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
-Comparison of these methods in a mountainous region of British Columbia clarifies that performance is adequate according to some metrics but not others, that aspatial performance metrics obscure important differences in alorithm behaviour within and outside cutblocks, and that none of these algorithms mimic switchbacks in steep terrain.
-Method assessment is complicated by errors in available in cutblock and forest road data.
-Simple iterative least cost path road network simulation methods may be adequate for projecting important impacts on wildlife across large areas, and we hope that more easily useable open source implementations will help integrate resource roads into forest change projections.
+As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
+Grade penalties improve performance in a mountainous region of British Columbia, as does the minimum spanning tree method.
+However, the minimum spanning tree method requires more time and memory to compute, so users must weigh the benefits of improved performance against computational costs.
+These simple road network simulation methods can improve projections of important impacts of development on wildlife across large areas, and we hope that more easily useable open source implementations will encourage integration of resource roads into forest change projections.
 keywords: |
@@
 tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
-#cost surface raster layer
+#weight raster layers
 tsaCost <- terra::rast(here(data_path_drvd, "old cost/input_cost.tif"))
@@
 The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha.
-We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha cost rasters.
+We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha weight rasters.
 We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.
@@
 Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004].
-We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing.
+We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
 
@@
 A skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
-We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution.
+We compared two resolutions of the weight raster by aggregating the 1 ha resolution weight raster to a coarser 100 ha (1000 by 1000 m) resolution.
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
@@
 One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure D.1).
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the weight raster resolution is coarse (Figure D.1).
 
@@
 
-```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
 Values greater than zero indicate overprojection and negative values indicate underprojection.
@@
 
-```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Variation in spatial performance among projection method variants (1 ha cost raster).
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Variation in spatial performance among projection method variants (1 ha weight raster).
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
@@
 Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and time to compute on large landscapes (Supplement E).
-All of these methods require construction of a large graph with a node for each pixel in the cost surface where road construction is possible (i.e.
+All of these methods require construction of a large graph with a node for each pixel in the weight raster where road construction is possible (i.e.
 not NA), and memory requirements increase with the number of nodes in the graph (Supplement E).

### 2024-05-13 analysis/paper/supplement.Rmd
writing in progress

@@
 The grade penalty method for calculating edge weights is a simplified version of the approach taken by @anderson2004a that does not distinguish between adverse and favourable grades.
-In addition, we use a modified Digital Elevation Model (DEM) as the input to the edge weight calculation to allow barriers other than slope to be included.
-The DEM is modified such that barriers where roads cannot be built are assigned NA values, barriers with high cost (e.g.
-stream crossings) can be represented with negative values, and existing roads are assigned a value of 0.
-All other values in the DEM are assumed to be elevation and the difference between the two elevations is used to determine the grade.
-Default construction cost values are taken from the BC Interior Appraisal Manual such that the base cost of building a road is \$16 178 with an additional cost of \$504 for every 1% grade and a limit of 20% grade.
+We use a modified Digital Elevation Model (DEM) as the input to the edge weight calculation that allows barriers other than slope to be included without incurring additional computation and memory costs.
+Pixels where roads cannot be built are assigned a value of NA, and negative values indicate costs associated with other features (e.g.
+stream crossings).
+Existing roads are assigned a value of 0.
+All other values in the weight raster are assumed to be elevation and grade is calculated from differences between adjacent locations.
+Default construction cost values are taken from the BC Interior Appraisal Manual: the base cost of building a road is \$16 178 with an additional cost of \$504 for every 1% grade and a limit of 20% grade.
 Edges with greater than 20% grade are assigned either NA or an arbitrarily high value.
 A value of 65 000 is used in our projections to avoid errors when a landing cannot be accessed without crossing an NA cell.
-If any of the cells to be connected by an edge has a negative value the edge weight is set to the larger value after taking the absolute value.
+Note that grade penalties are not applied adjacent to existing roads, nor do they apply when connecting to a node with a negative value.
+In the latter case, the edge weight is set to the highest absolute cost.
 
@@
 
-The simple cost method for setting edge weights generally performed worse than the grade penalty method.
+The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 3 and \@ref(fig:aspat-perf-fig-simp)).
+
+```{=html}
+<!-- To remove? This is all covered in the main text and I don't think differs much here.
 For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig-simp)).
@@
 None of the simple cost methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
-
-```{=html}
-<!-- To remove? This is all covered in the main text and I don't think differs much here.
 
@@
 
-```{r aspat-perf-fig-simp, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
+```{r aspat-perf-fig-simp, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different simple cost projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster).
 Values greater than zero indicate overprojection and negative values indicate underprojection.
@@
 
-```{r spat-perf-fig-simp, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster).
+```{r spat-perf-fig-simp, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Spatial variation of in performance among simple cost projection method variants (1 ha cost raster).
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
@@
 
-As expected, none of the projection methods yield accurate spatial projections of road presence (i.e.
-the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
-Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig-simp)).
-
 # All cutblocks results
 
-```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster.
+```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution weight raster in analysis that includes false cutblocks with no observed access roads.
 Metrics were summarised within cutblocks, outside cutblocks, and overall.
 Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
-Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
+"}
 meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
@@
 
-```{r spat-perf-fig-all, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster.
+```{r spat-perf-fig-all, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution weight raster, including false cutblocks with no observed access roads.
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted.
-Values closer to one indicate better performance.
-Cutblocks included all cutblocks present in the harvest dataset including falsly detected cutblocks with no observed roads accessing them."}
+Values closer to one indicate better performance."}
 matchData <- read.csv(here(data_path_drvd, "dem", "agree_table.csv"))
@@
 
-```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
+```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) weight rasters.
 Metrics were summarised within cutblocks, outside cutblocks, and overall.
@@
 
-```{r figA2, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters.
+```{r figA2, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (1 ha) and coarse resolution (100 ha) weight rasters.
 F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
@@
 
-We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of three different resolutions 1000 m, 500 m and 100 m, which corresponded to 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
+We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of three different resolutions (1000 m, 500 m and 100 m), giving 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
 Each cell in the raster becomes a node in the graph used to determine the shortest path which increases the time and memory requirements of the projection.
-Benchmarking was done using Azure Standard A4 v2 nodes which have 4 cores and 8 GB of RAM available, which was chosen to match a typical laptop.
-Each projection was run on a separate node and the `peakRAM` package was used to measure the total time to run the projection and the maximum memory used at any point in the analysis [@quinn2024].
-In addition, we timed the execution of the Hardy QGIS plugin with the 100 m resolution weights raster.
+Benchmarking was done using Azure Standard A4 v2 nodes which have 4 cores and 8 GB of RAM, chosen to match a typical laptop.
+Each projection was run on a separate node and the `peakRAM` package was used to measure the total time to run the projection and the maximum memory used at any point in the analysis [@quinn_peakRAM_2024].
+We also timed the execution of the Hardy QGIS plugin with the 100 m resolution weights raster.
 
-```{r speed-fig, fig.width=fig_widths["two"], fig.cap="Speed and peak memory usage of the ILCP and MST road projection methods with landings regularly sampled at 10 per km^2^ (high density) or using cutblock centroids as landings with weight rasters of three different resolutions: 1000 m, 500 m and 100 m, which for our study area is 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
-In addition, the time to run the Hardy QGIS plugin with the 100 m raster is also shown.
+Execution time and memory requirements increase with the number of nodes in the graph, and the slope of that increase is highest for the MST method (Figure \@ref(fig:speed-fig)).
+Except on the very small landscape the ILCP method is faster than the MST method.
+Reducing the number of landings by targeting centroids also reduces execution time (Figure \@ref(fig:speed-fig)).
+The Hardy method is significantly faster than any of the R package methods.
+
+```{r speed-fig, fig.width=fig_widths["two"], fig.cap="Speed and peak memory usage of the ILCP and MST road projection methods with landings regularly sampled at 10 per km^2^ (high density) or using cutblock centroids as landings, with weight rasters of three different resolutions (1000 m, 500 m and 100 m) giving 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
+The time to run the Hardy QGIS plugin with the 100 m raster is also shown.
 Both number of nodes and time are displayed on log transformed axes."}
@@
 
-Our benchmarking showed a log-linear relationship between the number of nodes in the graph and the total execution time and memory requirements with the MST method, while the ILCP method had a shallower slope that increased slightly with larger numbers of nodes (Figure \@ref(fig:speed-fig)).
-Other than for the smallest graph the ILCP method was always faster than the MST method and projections were always faster when the centroid was used rather than high density landings (Figure \@ref(fig:speed-fig)).
-The Hardy method was significantly faster than any of the R package methods.
-
 # Profiling
 
-Profiling with the `Rprof` function can reveal which parts of a function take the longest.
+Profiling with the `Rprof` function shows which parts of a function take the longest.
 On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods.
-Constructing the graph took ~ 4 s for both methods.
-For the ILCP method, the iteratively determining the shortest path is the only other step that takes more that a second to run.
-Each iteration is relatively fast but must be repeated for each landing.
-For the MST method, the longest running step is the building of the second graph which is used to determine the MST.
-Determining the shortest path to connect the parts of the MST also takes about 2 seconds.
+Both methods begin by constructing a full graph, which took ~ 4 s.
+For the ILCP method, iteratively determining the shortest paths is the only other step that takes more that a second to run, and this run time will increase with the number of landings.
+For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road distance takes the most time.
+Determining the locations of the paths in the minimum spanning tree also takes about 2 seconds.
 

### 2024-05-13 analysis/paper/supplement.Rmd
more writing in progress

@@
 
-The grade penalty method for calculating edge weights is a simplified version of the approach taken by @anderson2004a that does not distinguish between adverse and favourable grades.
+The grade penalty method for calculating edge weights is a simplified version of the approach taken by Anderson and Nelson [-@anderson_projecting_2004] that does not distinguish between adverse and favourable grades.
 We use a modified Digital Elevation Model (DEM) as the input to the edge weight calculation that allows barriers other than slope to be included without incurring additional computation and memory costs.
@@
 Benchmarking was done using Azure Standard A4 v2 nodes which have 4 cores and 8 GB of RAM, chosen to match a typical laptop.
-Each projection was run on a separate node and the `peakRAM` package was used to measure the total time to run the projection and the maximum memory used at any point in the analysis [@quinn_peakRAM_2024].
+Each projection was run on a separate node and the `peakRAM` package was used to measure the total time to run the projection and the maximum memory used at any point in the analysis [@quinn_peakram_2024].
 We also timed the execution of the Hardy QGIS plugin with the 100 m resolution weights raster.

### 2024-05-14 analysis/paper/paper.Rmd
edits for journal format

@@
 institute:
-  - ECCC: Environment and Climate Change Canada 
-  - UBC: University of British Columbia 
-  - FLNR: British Columbia Ministry of Forests 
+  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, Ottawa, ON, Canada  
+  - UBC:  Faculty of Forestry, Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada 
+  - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
-abstract: |
-  Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available.
-Hardy et al.
-[-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension.
-As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
-Grade penalties improve performance in a mountainous region of British Columbia, as does the minimum spanning tree method.
-However, the minimum spanning tree method requires more time and memory to compute, so users must weigh the benefits of improved performance against computational costs.
-These simple road network simulation methods can improve projections of important impacts of development on wildlife across large areas, and we hope that more easily useable open source implementations will encourage integration of resource roads into forest change projections.
 keywords: |
-  keyword 1; keyword 2; keyword 3
+   forest roads, road networks, forestry, spatially explicit modeling, projections, least cost path
 editor_options: 
@@
 
-Highlights: `r rmarkdown::metadata$highlights`
+[To do list](https://docs.google.com/document/d/1msHZuJGv0C67Avbl67YXm_P6fBSAvGw-m9qIzNDic64/edit?usp=sharing)
 
@@
   fig.path = "../figures/",
+  dev = "tiff",
+  dev.args=list(compression="lzw"),
   dpi = 300
@@
 ```
+
+\newpage
+
+# Abstract
+
+Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available.
+Hardy et al.
+[-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension.
+As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
+Grade penalties improve performance in a mountainous region of British Columbia, as does the minimum spanning tree method.
+However, the minimum spanning tree method requires more time and memory to compute, so users must weigh the benefits of improved performance against computational costs.
+These simple road network simulation methods can improve projections of important impacts of development on wildlife across large areas, and we hope that more easily useable open source implementations will encourage integration of resource roads into forest change projections.
 
@@
 Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].
-In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012;@st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
+In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012; @st-pierre_stairway_2022]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020].
 Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021].
@@
 General-purpose open source tools for graph analysis [e.g.
-@igraph_core_team_igraph_nodate] include methods for building graphs from adjacency matrices, finding “least cost paths” between pairs of points, and finding a least cost network of routes (i.e.
+@csardi_igraph_2006] include methods for building graphs from adjacency matrices, finding “least cost paths” between pairs of points, and finding a least cost network of routes (i.e.
 a minimum spanning tree) that connects a larger set of points [@deo_graph_2016].
@@
 
-We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006;@anderson_projecting_2004;@stuckelberger_improved_2007;@clark_three-stage_2000].
+We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006; @anderson_projecting_2004; @stuckelberger_improved_2007; @clark_three-stage_2000].
 Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
@@
 The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River.
-At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_and_range_biogeoclimatic_nodate] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017].
-At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_and_range_biogeoclimatic_nodate].
+At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017].
+At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022].
 The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022].
@@
 ```
-
 ```{r setup2}
@@
 library(tmap)
+library(cowplot)
 
@@
   tm_borders(col = "black")+
-  tm_compass(position = c(0.3, 0.11), size = 1)+
-  tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
+  # tm_compass(position = c(0.3, 0.11), size = 1)+
+  # tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
   tm_layout(legend.position = c("left", "bottom"),
             legend.bg.color = "white")
+
+vp2 <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
+                      just=c("right", "top"))
+
+both_maps <- tmap_arrange(tsa, tsaDEM, outer.margins = 0)
+
+tsa_grob <- tmap_grob(tsa)
+tsaDEM_grob <- tmap_grob(tsaDEM)
+canada_overlay_grob <- tmap_grob(canada_overlay)
+
+out_map <- plot_grid(tsa_grob, tsaDEM_grob)+
+    draw_grob(canada_overlay_grob, x=0.965, y=0.945,
+            width = 0.35, height=0.2,
+            hjust= 1, vjust = 1)
+
+tmap_save(both_maps, filename = here("analysis/figures/Figure1_StudyArea_both.tiff"),
+  dpi = 600, insets_tm = canada_overlay, insets_vp = vp2,
+  height = 6*0.9, width = fig_widths["two"], units = "in")
 
@@
 
+# try with ggspatial
+
+library(ggspatial)
+library(ggplot2)
+
+ggplot() + 
+  layer_spatial(dem) +
+  geom_sf(data = st_union(tsaBoundary), fill = NA) +
+  annotation_scale() +
+  scale_fill_continuous(na.value = NA)+
+  theme_void()
 ```
@@
 ```
+
 ```{r study-area2}
@@
 ```
+
 <!---BLOCK_MULTICOL_STOP{widths: [3.6,3.6]}--->
@@
 We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha weight rasters.
-We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions >= 1 ha.
+We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions \>= 1 ha.
 
@@
 For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
-There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004;@hardy_forest_2019; @hardy_landis-ii_2023;@picard_finding_2006].
+There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
 The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
 Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes.
-Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007;@anderson_projecting_2004].
+Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007; @anderson_projecting_2004].
 We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
@@
 ```
-
 Note that the evolution of a road network over time can be modeled by iteratively applying any of the methods, but in this manuscript we focus on comparing outcomes of a single application of each of the methods to a network of forest roads developed since 1990.
@@
 To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 
+# Competing interests
+
+# Author Contributions
+
+Conceptualization: JH, SE, DL, GP, KL Data curation: SE, DL Formal analysis: JH, SE, DL Funding acquisition: JH Investigation: JH, SE Methodology: JH, SE, DL, KL Project administration: JH Resources: JH Software: JH, SE Supervision: JH Visualization: SE Writing – original draft: JH, SE, DL Writing – review & editing: JH, SE, DL, GP, KL
+
+# Funding
+
+The authors declare no specific funding for this work.
+
+# Data availability
+
+Data generated or analyzed during this study are available in the Open Science Framework repository, <https://osf.io/8vmqh/?view_only=d168967961914f22af89e44c19aec019>.
+The roads package can be installed from CRAN and the code and a user guide are available on GitHub at <https://github.com/LandSciTech/roads>.
+
 <!-- The following line inserts a page break  -->

### 2024-05-14 analysis/paper/paper.Rmd
delete old figures, reformat study area

@@
 library(purrr)
-library(ggpubr)
 library(MetBrewer)
@@
 library(cowplot)
+library(ggspatial)
 
@@
 ```
-
-<!---BLOCK_MULTICOL_START--->
 
@@
 
-canada_overlay <- tm_shape(canada, projection = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")+
-  tm_borders(col="black")+
-  tm_shape(st_buffer(tsaBoundary, 10000))+
-  tm_polygons(border.col = met.brewer(pal_nm, 1), col = met.brewer(pal_nm, 1))+
-  tm_layout(frame=TRUE)
+# using ggspatial because tmap doesn't work with both multiplots and insets
+tsaBound <- st_union(tsaBoundary)
 
-tsa <- tm_shape(tsaCost, name = "Cost")+
-  tm_raster(style = "cont", title = "Cost",palette="Greys")+
-  tm_shape(st_union(tsaBoundary))+
-  tm_borders(col = "black")+
-  tm_compass(position = c(0.3, 0.11), size = 1)+
-  tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
-  tm_layout(legend.position = c("left", "bottom"),
-            legend.bg.color = "white")
+dem_map <- ggplot() + 
+  layer_spatial(dem) +
+  geom_sf(data = tsaBound, fill = NA) +
+  annotation_scale(width_hint = 0.4) +
+  scale_fill_continuous(low = "white", high = "black",
+                        na.value = NA)+
+  theme_void()+
+  theme(plot.background = element_rect(colour = NA, fill = "white"))+
+  labs(fill = "Elevation")
 
-vp <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
-               just=c("right", "top"))
+cost_map <- ggplot() + 
+  layer_spatial(tsaCost) +
+  geom_sf(data = tsaBound, fill = NA) +
+  annotation_north_arrow(style = north_arrow_orienteering(fill = c("black", "black")), height = unit(1, "cm"), width = unit(0.75, "cm")) +
+  scale_fill_continuous(low = "white", high = "black",
+                        na.value = NA)+
+  theme_void()+
+  theme(plot.background = element_rect(colour = NA, fill = "white"))+
+  labs(fill = "Cost")
 
-tmap_save(tsa, filename = here("analysis/figures/Figure1_StudyArea.png"),
-  dpi = 600, insets_tm = canada_overlay, insets_vp = vp,
-  height = 6*0.9, width = fig_widths["single"]*0.9, units = "in")
+can_map <- ggplot() + 
+  geom_sf(data = canada %>% st_transform(crs = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"), fill = NA) +
+  geom_sf(data = st_buffer(tsaBound, 10000) %>% st_transform(crs = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"), 
+          fill = met.brewer(pal_nm, 1)) +
+  theme_void()+
+  theme(panel.border = element_rect(colour = "black", fill = NA))
 
-tsaDEM <- tm_shape(dem, name = "Elevation")+
-  tm_raster(style = "cont", title = "Elevation",palette="Greys")+
-  tm_shape(st_union(tsaBoundary))+
-  tm_borders(col = "black")+
-  # tm_compass(position = c(0.3, 0.11), size = 1)+
-  # tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
-  tm_layout(legend.position = c("left", "bottom"),
-            legend.bg.color = "white")
-
-vp2 <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
-                      just=c("right", "top"))
-
-both_maps <- tmap_arrange(tsa, tsaDEM, outer.margins = 0)
-
-tsa_grob <- tmap_grob(tsa)
-tsaDEM_grob <- tmap_grob(tsaDEM)
-canada_overlay_grob <- tmap_grob(canada_overlay)
-
-out_map <- plot_grid(tsa_grob, tsaDEM_grob)+
-    draw_grob(canada_overlay_grob, x=0.965, y=0.945,
+out_map <- plot_grid(cost_map, dem_map)+
+  draw_plot(can_map, x=1.09, y=0.945,
             width = 0.35, height=0.2,
@@
 
-tmap_save(both_maps, filename = here("analysis/figures/Figure1_StudyArea_both.tiff"),
-  dpi = 600, insets_tm = canada_overlay, insets_vp = vp2,
-  height = 6*0.9, width = fig_widths["two"], units = "in")
-
-tmap_save(tsaDEM, filename = here("analysis/figures/Figure1_StudyAreaDEM.png"),
-  dpi = 600, insets_vp = vp,
-  height = 6*0.9, width = fig_widths["single"]*0.9, units = "in")
-
-# try with ggspatial
-
-library(ggspatial)
-library(ggplot2)
-
-ggplot() + 
-  layer_spatial(dem) +
-  geom_sf(data = st_union(tsaBoundary), fill = NA) +
-  annotation_scale() +
-  scale_fill_continuous(na.value = NA)+
-  theme_void()
+save_plot(here("analysis/figures/Figure1_StudyArea_both.tiff"), 
+          out_map,
+          dpi = knitr::opts_chunk$get("dpi"),
+          base_width = fig_widths["two"], 
+          base_height = 6*0.9)
 ```
-
-```{r study-area1, fig.cap= "see next fig"}
-
-knitr::include_graphics(here("analysis/figures/Figure1_StudyArea.png"))
-
-```
-
-```{r study-area2}
-
-knitr::include_graphics(here("analysis/figures/Figure1_StudyAreaDEM.png"))
-```
-
-<!---BLOCK_MULTICOL_STOP{widths: [3.6,3.6]}--->
 
@@
 
-png::writePNG(array(0, dim = c(1,1,4)), here("analysis/figures/empty.png"))
-knitr::include_graphics(here("analysis/figures/empty.png"))
+knitr::include_graphics(here("analysis/figures/Figure1_StudyArea_both.tiff"))
 ```
@@
 To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csárdi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 tmap_save(tmap_arrange(allMaps, ncol = 3),
-          here("analysis/figures/projection_methods_figure.png"),
+          here("analysis/figures/projection_methods_figure.tiff"),
           dpi = 300, height = 8, width = fig_widths["two"])
@@
 Landing density is high for all except the centroid and Hardy variants."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"))
 ```

### 2024-05-14 analysis/paper/paper.Rmd
writing in progress

--- 2024-05-14 analysis/paper/paper.Rmd
@@
 The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use our more computationally efficient methods (Supplement E) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 not NA), and memory requirements increase with the number of nodes in the graph (Supplement E).
+Our method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017], in part because we avoid the additional cost of constructing a transition matrix, and in part because we ignore variation in the size of raster cells that is accounted for in the 'gdistance' _geoCorrection_ function.
 Compute time also increases with the number of landing targets (Supplement E).
-The Hardy QGIS plugin was an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).
+The Hardy QGIS plugin is an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).
 
@@
 We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it.
-However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have allowed the possibility for users of the `roads` R package to specify their own method for determining edge weights from adjacent node values in a weight function.
+However, we also acknowledge that we have used a very simple cost model, and that the important determinants of road construction costs may vary among landscapes.
+We have therefore allowed the possibility for users of the `roads` R package to specify their own method for determining edge weights from adjacent node values in a weight function.
+A more complex cost model could include variation among substrate types (e.g.
+peat, clay, gravel) and more nuanced consideration of the costs of various types of water crossings (e.g.
+culverts and bridges).
+Real construction costs and impacts also vary substantially among types of roads (e.g.
+long-term vs temporary haul roads).
+If distinguishing among types of roads is important, an effective approach might be to first project the paths of lengthy long-term haul roads using a cost model and set of targets that is appropriate for this type of road, and then project the paths of shorter temporary haul roads from this long-term network using a different cost model and set of landing locations.
 
@@
 @labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020].
-In our example landscape, projections of disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
+In our example landscape, projections of the disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
+These methods do not predict the exact spatial locations of roads, so outcomes and impacts should not be reported or interpreted at the level of individual pixels; for example, these projections should not guide selection of sample locations for studies of the local (pixel level) impacts of anticipated roads.
 
@@
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
+Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete, though it may also be adequate for projecting the combined footprint of roads and cutblocks.
 More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
@@
 @whittington_towns_2022; @labadie_global_2023] or less [e.g.
-@hughes_comparison_2023; @diniz_landscape_2020] species-specific movement models it may be helpful to test sensitivity of these models to the difference between observed and retrospectively projected roads.
+@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity of these models to the difference between observed and retrospectively projected roads.
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
@@
 
-Conceptualization: JH, SE, DL, GP, KL Data curation: SE, DL Formal analysis: JH, SE, DL Funding acquisition: JH Investigation: JH, SE Methodology: JH, SE, DL, KL Project administration: JH Resources: JH Software: JH, SE Supervision: JH Visualization: SE Writing – original draft: JH, SE, DL Writing – review & editing: JH, SE, DL, GP, KL
+Conceptualization: JH, SE, DL, GP, KL Data curation: SE, DL Formal analysis: JH, SE, DL Funding acquisition: JH Investigation: JH, SE Methodology: JH, SE, DL, KL Project administration: JH Resources: JH Software: JH, SE, KL Supervision: JH Visualization: SE Writing – original draft: JH, SE, DL Writing – review & editing: JH, SE, DL, GP, KL
 

### 2024-05-14 analysis/paper/paper.Rmd
update fig 4

--- 2024-05-14 analysis/paper/paper.Rmd
@@
 
-fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
+fig_widths <- c(min = 1.18, single = 3.48, mid = 5.51, two = 7.16)
 ```
@@
           out_map,
+          compression = "lzw",
           dpi = knitr::opts_chunk$get("dpi"),
@@
           here("analysis/figures/projection_methods_figure.tiff"),
-          dpi = 300, height = 8, width = fig_widths["two"])
+          dpi = 300, compression="lzw", 
+          height = 8, width = fig_widths["two"])
 ```
 
-```{r proj-meth-fig, fig.cap="Comparison of road projection method variants (Table 2).
+```{r proj-meth-fig, fig.height=8, fig.width=fig_widths["two"], fig.cap="Comparison of road projection method variants (Table 2).
 The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
@@
 
-knitr::include_graphics(here("analysis/figures/Figure4_map_update.png"))
+knitr::include_graphics(here("analysis/figures/Figure4_map_update.tif"))
 

### 2024-05-14 analysis/paper/paper.Rmd

@@
 To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csárdi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use our more computationally efficient methods (Supplement E) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

### 2024-05-14 analysis/paper/paper.Rmd
resolving conflicts

--- 2024-05-14 analysis/paper/paper.Rmd
@@
 
-Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia.
+Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
 To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads.
@@
 Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data.
-We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
 

### 2024-05-14 analysis/paper/paper.Rmd
writing in progress

@@
     correspondence: false
+  - name: Kyle Lochhead
+    email: Kyle.Lochhead@gov.bc.ca
+    institute: [FLNR]
+    correspondence: false
   - name: Gregory Paradis
@@
     correspondence: false
-  - name: Kyle Lochhead
-    email: Kyle.Lochhead@gov.bc.ca
-    institute: [FLNR]
-    correspondence: false
 
@@
   - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, Ottawa, ON, Canada  
+  - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada
   - UBC:  Faculty of Forestry, Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada
-  - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada 
+
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 keywords: |
-   forest roads, road networks, forestry, spatially explicit modeling, projections, least cost path
+   forest roads, road networks, forest dynamics, spatially explicit modeling, predictive ecology, least cost paths
 editor_options: 
@@
 In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
-Despite the importance of roads for wildlife, very few of these projections have included road networks.
+Despite the importance of roads for wildlife, very few of these projections have included road networks [@labadie_global_2023; @leston_quantifying_2020].
 For example, in a recent collection of articles focused on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].
@@
 Harvest road planning methods and tools [e.g.
-@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.
+@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of the road network may suffice.
 There is a need for simple open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
-However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.
-We focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs on a regular isotropic graph from a single input raster.
+However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding, and we focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs (represented as edge weights on a regular isotropic graph) from values of adjacent locations on a single user-supplied input raster.
+This constraint permits penalization of steep grades as suggested by Anderson and Nelson [-@anderson_projecting_2004], but does not permit other aspects of their method.
+We compare a simple grade penalty approach to the method of calculating edge weights from a cost surface raster used by both Hardy et al.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al.
+[-@lochhead_linking_2022].
 If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context.
-Both Hardy et al.
-[-@hardy_forest_2019; -@hardy_landis-ii_2023] and Lochhead et al.
-[-@lochhead_linking_2022] calculate the cost of constructing a road from one raster pixel to another (represented as graph edge weights) from a single user-supplied cost surface raster that can vary with attributes that alter construction costs such as slope and the presence of water.
-We compare this approach to a simplified method for penalizing steep grades [as suggested by @anderson_projecting_2004] that does not incur the additional computational costs associated with other aspects of their more complex method.
 
@@
 In the first "simple cost" method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells.
-In the second "grade penalty" method (Supplement A), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), which penalization of edge grades as suggested by Anderson and Nelson (2004).
+In the second "grade penalty" method (Supplement A), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
 The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha.
@@
 Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes.
-Both the grade penalty and simple cost methods require construction of a simple regular isotropic graph from a single input grid, which is less computationally costly than more complex proposed methods for consideration of curvature and switchbacks [@stuckelberger_improved_2007; @anderson_projecting_2004].
+Both the grade penalty and simple cost methods involve setting edge weights on a simple regular isotropic graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation.
 We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
@@
 However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network.
-It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network.
 Within cutblocks, Hardy et al's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
@@
 The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use our more computationally efficient methods (Supplement E) by default.
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement E) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 not NA), and memory requirements increase with the number of nodes in the graph (Supplement E).
-Our method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017], in part because we avoid the additional cost of constructing a transition matrix, and in part because we ignore variation in the size of raster cells that is accounted for in the 'gdistance' _geoCorrection_ function.
 Compute time also increases with the number of landing targets (Supplement E).
+Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Supplement F), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the 'gdistance' _geoCorrection_ function.
 The Hardy QGIS plugin is an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).
@@
 @barros_empowering_2023; @daniel_state-and-transition_2016].
-Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in that context.
+Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can help guide method selection in that context.
 
@@
 However, we also acknowledge that we have used a very simple cost model, and that the important determinants of road construction costs may vary among landscapes.
-We have therefore allowed the possibility for users of the `roads` R package to specify their own method for determining edge weights from adjacent node values in a weight function.
+We have therefore also allowed the possibility for users of the `roads` R package to specify their own method for determining edge weights from adjacent node values in a weight function.
 A more complex cost model could include variation among substrate types (e.g.
@@
 In our example landscape, projections of the disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
-These methods do not predict the exact spatial locations of roads, so outcomes and impacts should not be reported or interpreted at the level of individual pixels; for example, these projections should not guide selection of sample locations for studies of the local (pixel level) impacts of anticipated roads.
+On the other hand, these methods do not predict the exact spatial locations of roads, so outcomes and impacts should not be reported or interpreted at the level of individual pixels; for example, these projections should not guide selection of sample locations for studies of the local (pixel level) impacts of anticipated roads.
 
@@
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
-Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete, though it may also be adequate for projecting the combined footprint of roads and cutblocks.
+Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete even if it is also adequate for projecting the combined footprint of roads and cutblocks.
 More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
@@
 
+This work was motivated by discussions with Colin Daniel (Apex Resource Management Solutions) and needs of the Western Boreal Initiative, an effort to project cumulative effects of disturbances on forests and wildlife led by Eliot McIntire (Canadian Forest Service, Natural Resources Canada) and Samuel Haché (Canadian Wildlife Service, Environment and Climate Change Canada).
+Questions from Steve Cumming (Université Laval) and Lisa Venier (Canadian Forest Service, Natural Resources Canada) helped refine our focus and approach.
+We thank Clément Hardy for answering our questions about his method.
+
 # Competing interests
+
+The authors declare there are no competing interests.
 
@@
 
-The authors declare no specific funding for this work.
+Funding for this work was provided by Environment and Climate Change Canada.
 

### 2024-05-14 analysis/paper/paper.Rmd
added ilcp gdistance to supp, trying to fix fig 3 size WIP

--- 2024-05-14 analysis/paper/paper.Rmd
@@
 
-Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
+Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia.
 To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads.
@@
 
-```{r proj-meth-fig, fig.height=8, fig.width=fig_widths["two"], fig.cap="Comparison of road projection method variants (Table 2).
+```{r proj-meth-fig, fig.height=9.33, fig.width=fig_widths["two"], out.width="100%", out.height="100%", fig.cap="Comparison of road projection method variants (Table 2).
 The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
@@
 Landing density is high for all except the centroid and Hardy variants."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"))
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"), dpi=NA)
 ```
@@
 Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data.
-We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
 

### 2024-05-14 analysis/paper/supplement.Rmd
added ilcp gdistance to supp, trying to fix fig 3 size WIP

@@
 
+```{r profiling-ilcp_gdistance, fig.cap="Profiling results for the ILCP method using the gdistance package internally showing functions that take > 1 s to run on a cost raster with 1 million cells.
+Functions higher in the call stack are called by functions lower down.", fig.height=5}
+profstat(here::here("analysis/data/derived_data/ilcp_gdistance_100_Rprof.out"))
+```
+
 # Fort Nelson map

### 2024-05-15 analysis/paper/paper.Rmd
fixed figure size problem

--- 2024-05-14 analysis/paper/paper.Rmd
@@
 )
+# Setting to avoid re-running projections
+rerun <- FALSE
 ```
@@
 
-```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
 Values greater than zero indicate overprojection and negative values indicate underprojection.
@@
 
+if(rerun){
   # re-run projections for just the small area with different parameters
@@
 
+saveRDS(allProj, here("analysis/data/derived_data/test_prog_fig3.rds"))
+
+} else {
+  allProj <- readRDS(here("analysis/data/derived_data/test_prog_fig3.rds"))
+}
+
 allMaps <- purrr::map(
@@
           here("analysis/figures/projection_methods_figure.tiff"),
-          dpi = 300, compression="lzw", 
+          dpi = knitr::opts_chunk$get("dpi"), compression="lzw", 
           height = 8, width = fig_widths["two"])
@@
 
-```{r proj-meth-fig, fig.height=8, fig.width=fig_widths["two"], fig.cap="Comparison of road projection method variants (Table 2).
+```{r proj-meth-fig, out.width=paste0(fig_widths["two"], "in"), out.height="8in", fig.cap="Comparison of road projection method variants (Table 2).
 The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
@@
 Landing density is high for all except the centroid and Hardy variants."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"))
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"),
+                        dpi = knitr::opts_chunk$get("dpi"))
 ```
@@
 
-```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Variation in spatial performance among projection method variants (1 ha weight raster).
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4.5in", fig.cap="Variation in spatial performance among projection method variants (1 ha weight raster).
 F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
@@
 
-```{r speed-fig, fig.height=5, fig.width=fig_widths["single"], fig.cap=""}
+```{r speed-fig}
 # Get run times from bench marking results
@@
 Data generated or analyzed during this study are available in the Open Science Framework repository, <https://osf.io/8vmqh/?view_only=d168967961914f22af89e44c19aec019>.
+The code to reproduce the analysis and figures in this paper is available on GitHub at <https://github.com/LandSciTech/RoadPaper>.
 The roads package can be installed from CRAN and the code and a user guide are available on GitHub at <https://github.com/LandSciTech/roads>.

### 2024-05-15 R/functionsSource.R
Fix so colours stay the same when Hardy not included

@@ -601,6 +601,9 @@ doAspatPlot <- function(tab){
            areaMean = factor(areaMean, labels = c("Cutblocks", "Outside\ncutblocks",
                                                   "Overall")))
 
+  pal <- met.brewer(pal_nm, 8) %>% .[c(8, 1:7)]
+  names(pal) <- levels(prop_dif$sampleDens) %>% setdiff("Observed")
+
   prop_dif %>%
     ggplot(aes(x = sampleDens, prop_dif, fill = sampleDens))+
     geom_hline(aes(yintercept = 0), color = "grey")+
@@ -608,7 +611,7 @@ doAspatPlot <- function(tab){
                                         width = 0.75))+
     facet_grid(areaMean~ metric, scales = "free")+
     theme_classic()+
-    scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
+    scale_fill_manual(values = pal,
                       name = "Sampling\nMethod", guide = "none")+
     theme(text = element_text(size = 10), axis.title.x = element_blank(),
           axis.title.y = element_blank(),
@@ -668,6 +671,10 @@ doSpatPerf <- function(tab){
 }
 
 doSpatPlot <- function(tab){
+
+  pal <- met.brewer(pal_nm, 8) %>% .[c(8, 1:7)]
+  names(pal) <- levels(tab$sampleDens) %>% setdiff("Observed")
+
   tab %>%
     pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
                  values_to = "value") %>%
@@ -676,7 +683,7 @@ doSpatPlot <- function(tab){
            value = value  +0.01) %>%   # adding so 0 shows in figure
     ggplot(aes(sampleDens, value, fill = sampleDens))+
     geom_col()+
-    scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:7)],
+    scale_fill_manual(values = pal,
                       name = "Sampling\nMethod", guide = "none")+
     facet_grid(metric ~ perf_meas)+
     coord_flip()+

### 2024-05-15 analysis/paper/supplement.Rmd
fixing gdistance profiling and updating text to include it

--- 2024-05-13 analysis/paper/supplement.Rmd
@@
 
-The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 3 and \@ref(fig:aspat-perf-fig-simp)).
+The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 2 and \@ref(fig:aspat-perf-fig-simp), and Figures 5 and \@ref(fig:spat-perf-fig-simp)).
 
@@
 On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods.
-Both methods begin by constructing a full graph, which took ~ 4 s.
-For the ILCP method, iteratively determining the shortest paths is the only other step that takes more that a second to run, and this run time will increase with the number of landings.
-For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road distance takes the most time.
-Determining the locations of the paths in the minimum spanning tree also takes about 2 seconds.
+Both methods begin by constructing a full graph, which took 3-5 s.
+For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling-ilcp)).
+For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road distance takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling-mst)).
+Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds.
+We also considered using the `gDistance` package [@van_etten_r_2017] to build the graph but it took longer than our method (Figure \@ref(fig:profiling-ilcp-gdistance)).
 
@@
 # interactive version
-# profvis::profvis(prof_input = here::here("analysis/data/derived_data/ilcp_Rprof.out"))
+# profvis::profvis(prof_input = here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))
 ```
@@
 profstat(here::here("analysis/data/derived_data/mst_100_Rprof.out"))
+```
+
+```{r profiling-ilcp-gdistance, fig.cap="Profiling results for the ILCP method using the gdistance package internally showing functions that take > 1 s to run on a cost raster with 1 million cells.
+Functions higher in the call stack are called by functions lower down.", fig.height=5}
+profstat(here::here("analysis/data/derived_data/ilcp_gdistance_100_Rprof.out"))
 ```

### 2024-05-15 analysis/paper/paper.Rmd
change to gradePenaltyFn

@@
 paramTableDem = paramTable
-paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+paramTableDem$weightFunction = deparse1(gradePenaltyFn,collapse="\n")
 paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
@@
 Compute time also increases with the number of landing targets (Supplement E).
-Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Supplement F), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the 'gdistance' _geoCorrection_ function.
+Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Supplement F), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function.
 The Hardy QGIS plugin is an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).
@@
 The code to reproduce the analysis and figures in this paper is available on GitHub at <https://github.com/LandSciTech/RoadPaper>.
-The roads package can be installed from CRAN and the code and a user guide are available on GitHub at <https://github.com/LandSciTech/roads>.
+The `roads` package can be installed from CRAN and the code and a user guide are available on GitHub at <https://github.com/LandSciTech/roads>.
 

### 2024-05-15 analysis/paper/supplement.Rmd
change to gradePenaltyFn

@@
 
-slopePenaltyFn <- function(x1, x2, resolution = 1, baseCost = 16178, limit = 20,
+gradePenaltyFn <- function(x1, x2, resolution = 1, baseCost = 16178, limit = 20,
                            penalty = 504, limitWeight = NA){

### 2024-05-15 analysis/scripts/6_benchmark_methods.R

@@ -89,7 +89,7 @@ prof <- peakRAM::peakRAM(
                          weightRaster = inputs$tsaCost_st,
                          roads = inputs$roadsExist,
                          roadMethod = cRow$method,
-                         weightFunction = slopePenaltyFn,
+                         weightFunction = gradePenaltyFn,
                          limitWeight = 65000)
 )
 

### 2024-05-15 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -49,7 +49,7 @@ dem <- rast(here(data_path_drvd,  "testing_dem.tif"))
 roads <- read_sf(here(data_path_drvd,  "testing_obs_roads.gpkg"))
 
 see = data.frame(demDif = seq(-0.3,0.3,length.out=1000))
-see$cost = slopePenaltyFn(1,1+see$demDif,resolution=100)
+see$cost = gradePenaltyFn(1,1+see$demDif,resolution=100)
 plot(see$cost~see$demDif)
 
 # densities
@@ -70,7 +70,7 @@ paramTable$weightMethod = "Simple cost"
 
 paramTableDem = paramTable
 
-paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+paramTableDem$weightFunction = deparse1(gradePenaltyFn,collapse="\n")
 paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)
 paramTableDem$weightMethod = "Grade penalty"
 
@@ -197,7 +197,7 @@ if(0){
 
   paramTable<-paramTable[nrow(paramTable)-1,]
 
-  paramTable$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
+  paramTable$weightFunction = deparse1(gradePenaltyFn,collapse="\n")
 
   # this means that slopes that are steeper than limit will be 65000 not NA
   paramTable$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTable$weightFunction,fixed=T)

### 2024-05-15 analysis/paper/paper.Rmd

@@
 paramTableDem = paramTable
-paramTableDem$weightFunction = deparse1(gradePenaltyFn,collapse="\n")
+paramTableDem$weightFunction = deparse1(slopePenaltyFn,collapse="\n")
 paramTableDem$weightFunction = gsub("limitWeight = NA","limitWeight = 65000",paramTableDem$weightFunction,fixed=T)

### 2024-05-15 analysis/paper/paper.Rmd
fis abbreviations

--- 2024-05-15 analysis/paper/paper.Rmd
@@
 Hardy et al.
-[-@hardy_landis-ii_2023] partially addressed this need with a Landis-II extension.
+[-@hardy_landis-ii_2023] partially addressed this need with a LANDIS-II extension.
 As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
@@
 Promising simple graph-based options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
-The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
-Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+The former method is available as QGIS and LANDIS-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
+Hardy's tool is designed for integration with other components of the LANDIS-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 Our intention is to produce outputs comparable to Hardy et al.
-[-@hardy_forest_2019; -@hardy_landis-ii_2023] with a tool that can be used in modelling contexts outside of QGIS/Landis, and can serve as a more easily modifiable starting point for further refinement and development.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] with a tool that can be used in modelling contexts outside of QGIS/LANDIS, and can serve as a more easily modifiable starting point for further refinement and development.
 We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia.
@@
 Cutblocks not accessible via the observed road network were excluded from analysis.
-Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
+Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks.
+Projection method variants are defined in Table 2."}
 # organize data for ggplot
@@
 Values closer to one indicate better performance.
-Cutblocks not accessible via the observed road network were excluded from analysis."}
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Projection method variants are defined in Table 2."}
 perf_tbl <- doSpatPerf(matchDataRealCuts)
@@
 Hardy et al.
-[-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the LANDIS-II modelling framework [e.g.
 @labadie_global_2023].

### 2024-05-17 analysis/paper/paper.Rmd
Sent to co-authors for review & comment.

@@
 # Setting to avoid re-running projections
-rerun <- FALSE
+rerun <- TRUE
 ```
@@
 Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
-Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.
 

### 2024-05-17 analysis/paper/supplement.Rmd
Sent to co-authors for review & comment.

@@
 Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds.
-We also considered using the `gDistance` package [@van_etten_r_2017] to build the graph but it took longer than our method (Figure \@ref(fig:profiling-ilcp-gdistance)).
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (compare Figures \@ref(fig:profiling-ilcp) and \@ref(fig:profiling-ilcp-gdistance)).
 
 ```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings.
-Functions higher in the call stack are called by functions lower down.", fig.height=2}
+Functions higher in the call stack are called by functions lower down.", fig.height=3}
 if(!file.exists(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))){

### 2024-05-27 analysis/paper/paper.Rmd
minor edits based on Greg's comments, more to come on bigger things

--- 2024-05-15 analysis/paper/paper.Rmd
@@
   - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada
-  - UBC:  Faculty of Forestry, Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada
+  - UBC:  Faculty of Forestry, Department of Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada
 
@@
 We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our simplified method for penalizing steep grades to yield more realistic road networks in steep terrain.
-We do not have a similarly strong hypothesis about whether minimimum spanning tree projections are likely to be more or less realistic than interative least cost path projections.
+We do not have a similarly strong hypothesis about whether minimum spanning tree projections are likely to be more or less realistic than iterative least cost path projections.
 If designers of real roads take a larger view to accessing multiple sites at the lowest cost we might expect minimum spanning tree algorithms to yield more realistic networks.
-However, real road networks are built over time, and it is not necessarily true that they designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.
+However, real road networks are built over time, and it is not necessarily true that they are designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.
 
@@
 At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017].
-At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022].
+At higher elevations, forests are dominated by Engelmann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022].
 The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022].
@@
 Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022].
-Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].
+Roads also alter grizzly bear habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].
 
@@
 Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
-We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy personal communication).
+We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy, personal communication).
 
 The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
-For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
+For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [@dijkstra_note_1959].
 There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
@@
 
-Hardy’s method [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network.
+The method described in Hardy et al.
+[-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network.
 A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
-In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
+In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve the least cost path between the two.
 However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network.
-Within cutblocks, Hardy et al's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
+Within cutblocks, the Hardy et al.
+method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
 
@@
 The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network.
-These results are used to create a weighted graph wherein nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
-The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights.
+These results are used to create a weighted graph where nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
+The minimum spanning tree for this graph is solved using Kruskal's algorithm [@kruskal_shortest_1956], which links all nodes with a tree that minimizes the sum of edge weights.
 This method differs from Clark et al.
@@
 The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement E) by default.
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement F) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 knitr::kable(df2, caption = "Road projection method variants.
-We examined sensitivity of minimimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
+We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
 ```
@@
 One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the weight raster resolution is coarse (Figure D.1).
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).
 
@@
 The associated memory and computational requirements may present challenges on large landscapes (Supplement E).
-Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s more efficient implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023].
+Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations.
 The first three options are available to users of the `roads` R package.
-However, those contemplating reducing computational cost by increasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D).
+However, those contemplating reducing computational cost by decreasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D).
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
@@
 Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].
 

### 2024-05-27 analysis/paper/supplement.Rmd
Update profiling graph

--- 2024-05-15 analysis/paper/supplement.Rmd
@@
 
+# Formula for extrapolating RAM needed for road projection
+# bench_res %>% 
+#   filter(method == "mst", sampleType == "regular", sampleDens == 1e-05,
+#          cutblocks_real== "revelstoke_real") %>% 
+#   arrange(desc(resolution)) %>% 
+#   {lm(log10(Peak_RAM_Used_MiB)~log10(n_verticies), data = .)}
+# 10^(-1.6+0.769*log10(21791850))
+
 minor_breaks <- rep(1:4*2, 21)*(10^rep(-10:10, each=9))
@@
 Both methods begin by constructing a full graph, which took 3-5 s.
-For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling-ilcp)).
-For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road distance takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling-mst)).
+For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling)).
+For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling)).
 Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds.
-We also considered using the `gDistance` package [@van_etten_r_2017] to build the graph but it took longer than our method (Figure \@ref(fig:profiling-ilcp-gdistance)).
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (Figure \@ref(fig:profiling)).
 
-```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings.
-Functions higher in the call stack are called by functions lower down.", fig.height=2}
-if(!file.exists(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))){
-  source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
-}
+```{r profiling, fig.cap="Profiling results for the ILCP, MST, and ILCP using the gdistance package methods, showing functions that take more than one second to run on a cost raster with 1 million cells and 10 landings.
+Functions higher in the call stack are called by functions lower down.", fig.height=6, fig.width=fig_widths["two"]}
 
-profstat(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))
+ilcp_100 <- profstat(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))
+mst_100 <- profstat(here::here("analysis/data/derived_data/mst_100_Rprof.out"))
+ilcp_gd_100 <- profstat(here::here("analysis/data/derived_data/ilcp_gdistance_100_Rprof.out"))
 
+bind_rows(ILCP = ilcp_100, MST = mst_100, `ILCP gdistance` = ilcp_gd_100, .id = "method") %>% 
+  mutate(label = case_when(label == ".External2" ~ "", 
+                           label == "lazy_eval" ~ "", 
+                           label == "gdistance::geoCorrection" & depth == 8 ~ "", 
+                           label == "graph.adjacency.sparse" ~ "", 
+                           TRUE ~ label)) %>% 
+  mutate(method = factor(method, levels = c("ILCP", "MST", "ILCP gdistance"))) %>% 
+  ggplot(aes(y = depth, label = label))+
+  geom_linerange(aes(xmin = st_time, xmax = end_time-0.1))+
+  geom_text(aes(y = depth+0.35, x = st_time), hjust = 0, 
+            size.unit = "pt", size = 10)+
+  scale_x_continuous(expand = expansion(mult = c(0, 0.15)))+
+  labs(y = "Call stack", x = "Time (s)")+
+  theme_bw()+
+  theme(panel.grid = element_blank())+
+  facet_grid(method~.)
 # interactive version
 # profvis::profvis(prof_input = here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))
-```
-
-```{r profiling-mst, fig.cap="Profiling results for the MST method showing functions that take > 1 s to run on a cost raster with 1 million cells.
-Functions higher in the call stack are called by functions lower down.", fig.height=5}
-profstat(here::here("analysis/data/derived_data/mst_100_Rprof.out"))
-```
-
-```{r profiling-ilcp-gdistance, fig.cap="Profiling results for the ILCP method using the gdistance package internally showing functions that take > 1 s to run on a cost raster with 1 million cells.
-Functions higher in the call stack are called by functions lower down.", fig.height=5}
-profstat(here::here("analysis/data/derived_data/ilcp_gdistance_100_Rprof.out"))
 ```

### 2024-05-28 analysis/paper/paper.Rmd
Updates for Forest Science

@@
 bibliography: Roads.bib
-csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
+csl: "../templates/chicago-author-date.csl" # Insert path for the bib-style
 keywords: |
@@
 
-```{r study-area, fig.cap="Road construction cost surface and an elevation surface used to calculate grade penalties.
-Costs are calculated using a simple model informed by slope and major barriers (Table 1, Supplement A) in the Revelstoke timber supply area in British Colombia, Canada.
-In both layers, pixels containing existing roads are assigned a value of 0."}
-
-knitr::include_graphics(here("analysis/figures/Figure1_StudyArea_both.tiff"))
-```
-
 Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
@@
 In the first "simple cost" method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells.
-In the second "grade penalty" method (Supplement A), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
+In the second "grade penalty" method (Supplement 1-1), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
 The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha.
@@
 We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions \>= 1 ha.
-
-```{r cost-tbl}
-tbl <- "Cost Type|Value ($ per km)|Source
-Slope (%)|16178 + 504*slope|Interior appraisal manual
-Major lakes|65000|Major water bodies of BC
-Existing roads|0|Digital road atlas and forest tenure roads for BC
-"
-df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
-names(df) <- unname(as.list(df[1,]))
-df <- df[-1,] # remove first row
-row.names(df)<-NULL
-knitr::kable(df, caption = "A simple model of road construction costs.
-In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
-These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.")
-```
 
@@
 The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes.
+Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement 1-1), weighted the distance to neighbouring nodes.
 Both the grade penalty and simple cost methods involve setting edge weights on a simple regular isotropic graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation.
@@
 The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement F) by default.
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement 1-7) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
-
-```{r method-tbl}
-tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
-MST grade penalty & simple cost|Regular|1|1 & 100 ha
-MST grade penalty & simple cost|Regular|10|1 & 100 ha
-MST grade penalty & simple cost|Random|1|1 & 100 ha
-MST grade penalty & simple cost|Random|10|1 & 100 ha
-MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
-ILCP grade penalty & simple cost|Regular|10|1 ha 
-Hardy QGIS|N/A|~9|1 ha
-Cutblocks only|N/A|N/A|1 ha
-
-"
-df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
-names(df2) <- unname(as.list(df2[1,]))
-df2 <- df2[-1,] # remove first row
-row.names(df2)<-NULL
-knitr::kable(df2, caption = "Road projection method variants.
-We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
-```
 
@@
 \@ref(fig:proj-map-fig)b).
-We omitted false cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Supplement C).
+We omitted false cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Figures S3 and S4).
 
@@
 
-```
+perf_tbl <- doSpatPerf(matchDataRealCuts)
+perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
+  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
 
-```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
-Values greater than zero indicate overprojection and negative values indicate underprojection.
-Cutblocks not accessible via the observed road network were excluded from analysis.
-Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks.
-Projection method variants are defined in Table 2."}
-# organize data for ggplot
-doAspatPlot(meanTableRealCuts)
 ```
@@
 
-```{r proj-meth-fig, out.width=paste0(fig_widths["two"], "in"), out.height="8in", fig.cap="Comparison of road projection method variants (Table 2).
-The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
-Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks.
-In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row).
-Landing density is high for all except the centroid and Hardy variants."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"),
-                        dpi = knitr::opts_chunk$get("dpi"))
-```
-
-```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network.
-The main map shows observed distance to the nearest road minus the projected distance.
-Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks."}
-
-knitr::include_graphics(here("analysis/figures/Figure4_map_update.tif"))
-
-```
-
-```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4.5in", fig.cap="Variation in spatial performance among projection method variants (1 ha weight raster).
-F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
-Values closer to one indicate better performance.
-Cutblocks not accessible via the observed road network were excluded from analysis.
-Projection method variants are defined in Table 2."}
-perf_tbl <- doSpatPerf(matchDataRealCuts)
-perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
-  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
-
-doSpatPlot(perf_tbl)
-
-```
-
 ```{r speed-fig}
@@
 
-Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and time to compute on large landscapes (Supplement E).
+Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and time to compute on large landscapes (Figure S7).
 All of these methods require construction of a large graph with a node for each pixel in the weight raster where road construction is possible (i.e.
-not NA), and memory requirements increase with the number of nodes in the graph (Supplement E).
-Compute time also increases with the number of landing targets (Supplement E).
-Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Supplement F), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function.
-The Hardy QGIS plugin is an order of magnitude faster than the MST and ILCP methods implemented in R (Supplement E).
+not NA), and memory requirements increase with the number of nodes in the graph (Figure S7).
+Compute time also increases with the number of landing targets (Figure S7).
+Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function.
+The Hardy QGIS plugin is an order of magnitude faster than the MST and ILCP methods implemented in R (Figure S7).
 
@@
 We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region.
-The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A).
+The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement 1-1).
 We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it.
@@
 Aspatial projections of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3), so overall the MST regular high density projections were more accurate.
-However, the MST algorithm is also more computationally demanding than the other methods we considered (Supplement E), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
+However, the MST algorithm is also more computationally demanding than the other methods we considered (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
 
@@
 These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible.
-The associated memory and computational requirements may present challenges on large landscapes (Supplement E).
+The associated memory and computational requirements may present challenges on large landscapes (Figure S7).
 Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations.
 The first three options are available to users of the `roads` R package.
-However, those contemplating reducing computational cost by decreasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D).
+However, those contemplating reducing computational cost by decreasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Figures S5 and S6).
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
 Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete even if it is also adequate for projecting the combined footprint of roads and cutblocks.
-More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
+More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Figure S8) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
 
@@
 Cost minimizing methods are not suitable for projecting development of seismic lines (e.g.
-Supplement G), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
+Figure S9), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
 Lochhead et al.
@@
 
-# References
+# Literature Cited
 
@@
 \newpage
+# Supplementary Material
 
+*Supplement 1.* Detailed methods and additional results and figures, including: detailed methods for the grade penalty function; results using simple cost, all cutblocks, and a coarse resolution raster; projection method benchmarking and profiling; and a map of roads in Fort Nelson timber supply area.
+
+\newpage
+# Figures
+
+Figure 1: Road construction cost surface and an elevation surface used to calculate grade penalties.
+Costs are calculated using a simple model informed by slope and major barriers (Table 1, Supplement 1-1) in the Revelstoke timber supply area in British Colombia, Canada.
+In both layers, pixels containing existing roads are assigned a value of 0.
+
+```{r study-area, fig.cap="."}
+
+knitr::include_graphics(here("analysis/figures/Figure1_StudyArea_both.tiff"))
+```
+
+\newpage
+Figure 2: Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
+Values greater than zero indicate overprojection and negative values indicate underprojection.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks.
+Projection method variants are defined in Table 2.
+
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.cap="."}
+# organize data for ggplot
+doAspatPlot(meanTableRealCuts)
+```
+\newpage
+
+Figure 3: Comparison of road projection method variants (Table 2).
+The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
+Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks.
+In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row).
+Landing density is high for all except the centroid and Hardy variants.
+
+```{r proj-meth-fig, out.width=paste0(fig_widths["two"], "in"), out.height="8in", fig.cap="."}
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"),
+                        dpi = knitr::opts_chunk$get("dpi"))
+```
+
+\newpage
+
+Figure 4: Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network.
+The main map shows observed distance to the nearest road minus the projected distance.
+Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks.
+```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="."}
+
+knitr::include_graphics(here("analysis/figures/Figure4_map_update.tif"))
+
+```
+\newpage
+
+Figure 5: Variation in spatial performance among projection method variants (1 ha weight raster).
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
+Values closer to one indicate better performance.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Projection method variants are defined in Table 2.
+
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4.5in", fig.cap="."}
+
+doSpatPlot(perf_tbl)
+
+```
+
+\newpage
+
+# Tables
+
+Table 1: A simple model of road construction costs.
+In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
+These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.
+```{r cost-tbl}
+tbl <- "Cost Type|Value ($ per km)|Source
+Slope (%)|16178 + 504*slope|Interior appraisal manual
+Major lakes|65000|Major water bodies of BC
+Existing roads|0|Digital road atlas and forest tenure roads for BC
+"
+df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df) <- unname(as.list(df[1,]))
+df <- df[-1,] # remove first row
+row.names(df)<-NULL
+knitr::kable(df, caption = ".")
+```
+
+\newpage
+Table 2: Road projection method variants.
+We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.
+```{r method-tbl}
+tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
+MST grade penalty & simple cost|Regular|1|1 & 100 ha
+MST grade penalty & simple cost|Regular|10|1 & 100 ha
+MST grade penalty & simple cost|Random|1|1 & 100 ha
+MST grade penalty & simple cost|Random|10|1 & 100 ha
+MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
+ILCP grade penalty & simple cost|Regular|10|1 ha 
+Hardy QGIS|N/A|~9|1 ha
+Cutblocks only|N/A|N/A|1 ha
+
+"
+df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df2) <- unname(as.list(df2[1,]))
+df2 <- df2[-1,] # remove first row
+row.names(df2)<-NULL
+knitr::kable(df2, caption = ".")
+```
+
+\newpage
 ### Colophon

### 2024-05-28 analysis/paper/supplement/supplement.Rmd
Updates for Forest Science

--- 2024-05-27 analysis/paper/supplement.Rmd
@@
       number_sections: yes
-bibliography: Roads.bib
-csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
+bibliography: ../Roads.bib
+csl: "../../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
 
@@
 
-# Grade penalty function
+## Grade penalty function
 
@@
 
-# Simple cost results
+## Simple cost results
 
-The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 2 and \@ref(fig:aspat-perf-fig-simp), and Figures 5 and \@ref(fig:spat-perf-fig-simp)).
+The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 2 and S\@ref(fig:aspat-perf-fig-simp), and Figures 5 and S\@ref(fig:spat-perf-fig-simp)).
 
-```{=html}
-<!-- To remove? This is all covered in the main text and I don't think differs much here.
-For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig-simp)).
-Within all mapped cutblocks (Figure \@ref(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road.
-Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@ref(fig:aspat-perf-fig-all)).
-These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g.
-Figure \@ref(fig:proj-map-fig)b).
-Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig-simp)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road.
-However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)).
-None of the simple cost methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).
-
-Another fundamental limitation of these methods is that they generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure A-\@ref(fig:figA1)).
-In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).
-
-Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig-simp)).
-Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a).
-Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms would likely reduce this discrepancy.
-Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks.
-Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig-simp)), suggesting that these measures are easier to project.
-Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.
--->
-```
 ```{r get-data}
@@
 
-# All cutblocks results
+## All cutblocks results
 
@@
 
-# Coarse resolution results
+## Coarse resolution results
 
@@
 
-# Benchmarking road projection methods
+## Benchmarking road projection methods
 
@@
 
-Execution time and memory requirements increase with the number of nodes in the graph, and the slope of that increase is highest for the MST method (Figure \@ref(fig:speed-fig)).
+Execution time and memory requirements increase with the number of nodes in the graph, and the slope of that increase is highest for the MST method (Figure S\@ref(fig:speed-fig)).
 Except on the very small landscape the ILCP method is faster than the MST method.
-Reducing the number of landings by targeting centroids also reduces execution time (Figure \@ref(fig:speed-fig)).
+Reducing the number of landings by targeting centroids also reduces execution time (Figure S\@ref(fig:speed-fig)).
 The Hardy method is significantly faster than any of the R package methods.
@@
                   extra = "merge", convert = TRUE)
+
+ch_time <- structure(list(id = 1, resolution = 50, n_verticies = 21791850, 
+    n_edges = 87138977, Elapsed_Time_sec = 10001.606, Total_RAM_Used_MiB = 4018.7, 
+    Peak_RAM_Used_MiB = 10752.7), class = "data.frame", row.names = c(NA, 
+-1L))
 
@@
           method = "Hardy QGIS", Elapsed_Time_sec = hardy_time, n_verticies = 1815825) %>% 
+  # add_row(cutblocks_real = "revelstoke_real", sampleType = "centroid", sampleDens = 1.5e-05, 
+  #         method = "ilcp", Elapsed_Time_sec = ch_time$Elapsed_Time_sec, 
+  #         n_verticies = ch_time$n_verticies, Peak_RAM_Used_MiB = ch_time$Peak_RAM_Used_MiB) %>% 
   filter(cutblocks_real == "revelstoke_real", sampleType %in% c("regular", "centroid"), 
@@
 
-# Profiling
+## Profiling
 
@@
 Both methods begin by constructing a full graph, which took 3-5 s.
-For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling)).
-For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling)).
+For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure S\@ref(fig:profiling)).
+For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road takes the most time, which also increases with the number of landings (Figure S\@ref(fig:profiling)).
 Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds.
-We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (Figure \@ref(fig:profiling)).
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (Figure S\@ref(fig:profiling)).
 
@@
 
-# Fort Nelson map
+## Fort Nelson map
 
@@
 
-# References
+# Literature Cited {.unnumbered}
 
@@
 
-# Colophon
+# Colophon {.unnumbered}
 

### 2024-05-28 analysis/paper/paper.Rmd
weight function adjustments

--- 2024-05-27 analysis/paper/paper.Rmd
@@
   - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada
-  - UBC:  Faculty of Forestry, Department of Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada
+  - UBC:  Faculty of Forestry, Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada
 
@@
 We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our simplified method for penalizing steep grades to yield more realistic road networks in steep terrain.
-We do not have a similarly strong hypothesis about whether minimum spanning tree projections are likely to be more or less realistic than iterative least cost path projections.
+We do not have a similarly strong hypothesis about whether minimimum spanning tree projections are likely to be more or less realistic than interative least cost path projections.
 If designers of real roads take a larger view to accessing multiple sites at the lowest cost we might expect minimum spanning tree algorithms to yield more realistic networks.
-However, real road networks are built over time, and it is not necessarily true that they are designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.
+However, real road networks are built over time, and it is not necessarily true that they designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.
 
@@
 At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017].
-At higher elevations, forests are dominated by Engelmann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022].
+At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022].
 The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022].
@@
 Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022].
-Roads also alter grizzly bear habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].
+Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].
 
@@
 Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
-We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy, personal communication).
+We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy personal communication).
 
 The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
-For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [@dijkstra_note_1959].
+For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959].
 There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
@@
 
-The method described in Hardy et al.
-[-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network.
+Hardy’s method [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network.
 A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
-In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve the least cost path between the two.
+In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two.
 However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network.
-Within cutblocks, the Hardy et al.
-method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
+Within cutblocks, Hardy et al's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.
 
@@
 The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network.
-These results are used to create a weighted graph where nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
-The minimum spanning tree for this graph is solved using Kruskal's algorithm [@kruskal_shortest_1956], which links all nodes with a tree that minimizes the sum of edge weights.
+These results are used to create a weighted graph wherein nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
+The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights.
 This method differs from Clark et al.
@@
 The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement F) by default.
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement E) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 knitr::kable(df2, caption = "Road projection method variants.
-We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
+We examined sensitivity of minimimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
 ```
@@
 One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the weight raster resolution is coarse (Figure D.1).
 
@@
 
-simpleCostFn = function(x1,x2,...) (x1+x2)/2
+simpleCostFn = function(x1,x2,hdistance,...) hdistance*(x1+x2)/2
 paramTable$weightFunction = deparse1(simpleCostFn,collapse="\n")
@@
 The associated memory and computational requirements may present challenges on large landscapes (Supplement E).
-Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations.
+Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s more efficient implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023].
 The first three options are available to users of the `roads` R package.
-However, those contemplating reducing computational cost by decreasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D).
+However, those contemplating reducing computational cost by increasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D).
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
@@
 Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
-We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].
+We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.
 

### 2024-05-28 analysis/scripts/exploratory/runProjectionsSlopePenalties.R

@@ -49,7 +49,7 @@ dem <- rast(here(data_path_drvd,  "testing_dem.tif"))
 roads <- read_sf(here(data_path_drvd,  "testing_obs_roads.gpkg"))
 
 see = data.frame(demDif = seq(-0.3,0.3,length.out=1000))
-see$cost = gradePenaltyFn(1,1+see$demDif,resolution=100)
+see$cost = gradePenaltyFn(1,1+see$demDif,hdistance=100)
 plot(see$cost~see$demDif)
 
 # densities
@@ -64,7 +64,7 @@ paramTable <- tibble(sampleType, sampleDens,
                      output = vector("list", length(sampleDens))) %>%
   distinct()
 
-simpleCostFn = function(x1,x2,...) (x1+x2)/2
+simpleCostFn = function(x1,x2,hdistance,...) hdistance*(x1+x2)/2
 paramTable$weightFunction = deparse1(simpleCostFn,collapse="\n")
 paramTable$weightMethod = "Simple cost"
 

### 2024-05-28 analysis/paper/supplement/supplement.Rmd
More explanation of grade penalty and simple cost methods

@@
 
-## Grade penalty function
+## Grade penalty and simple cost functions
 
@@
 In the latter case, the edge weight is set to the highest absolute cost.
+Weights are adjusted for the horizontal distance between nodes to penalize longer diagonal road segments.
 
@@
 
-gradePenaltyFn <- function(x1, x2, resolution = 1, baseCost = 16178, limit = 20,
+gradePenaltyFn <- function(x1, x2, hdistance, baseCost = 16178, limit = 20,
                            penalty = 504, limitWeight = NA){
+  #For grade calculation, x1, x2 and hdistance (i.e. horizontal distance between nodes) 
+  #must have the same units.
+
   # Don't calculate grade penalty cost if one of the nodes is a barrier.
@@
   # If one node is a road use base cost.
-  grade <- 100 * abs(x1 - x2) * cond / resolution
+  grade <- 100 * abs(x1 - x2) * cond / hdistance #This is % (dimensionless)
 
-  slp <- baseCost + grade * penalty * (pmin(x1, x2) > 0)
+  slp <- baseCost + grade * penalty * (pmin(x1, x2) > 0) #This is cost per km.
   slp[grade > limit] <- limitWeight
@@
 
+  # Multiply by hdistance to penalize diagonals.
+  # Note that multiplying all edge weights by a constant does not change algorithm behaviour,
+  # so we ignore the unit of hdistance here.
+  slp = slp*hdistance
+
   return(slp)
@@
 
+```
+
+In the simple cost method, we first calculate a slope raster $S$, then calculate an input cost raster $X$ that depends on slope $S$:
+\begin{equation}
+   X = 
+   \begin{cases}
+     0 \text{ if existing road} \\
+     65000 \text{ if lake}\\
+     16178+504S \text{ otherwise}
+   \end{cases}
+\end{equation}
+The weight of an edge between two nodes is the mean value of the input cost raster at each of those nodes, adjusted for the euclidean distance between the nodes: 
+```{r echo=TRUE}
+simpleCostFn <- function(x1,x2,hdistance){
+  # Multiply by hdistance to penalize diagonals.
+  # Note that multiplying all edge weights by a constant does not change algorithm behaviour,
+  # so we ignore the unit of hdistance here.
+  return(hdistance*(x1+x2)/2)
+}
 ```

### 2024-05-30 analysis/paper/supplement/supplement.Rmd
More explanation of grade penalty and simple cost methods
re-run results

@@
 
-## Grade penalty and simple cost functions
+## Grade penalty and simple cost methods for setting edge weights
 
@@
 
-We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of three different resolutions (1000 m, 500 m and 100 m), giving 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
+We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of 3 different resolutions (1000 m, 500 m, 100 m, 30 m and 10m), giving 18 297, 72 633, and 1 815 825nodes in the cost graph, respectively.
+We also attempted to include resolutions of 30 m and 10 m (23 787 140 and 214 072 800 nodes, respectively) but these crashed due to insufficient memory.
 Each cell in the raster becomes a node in the graph used to determine the shortest path which increases the time and memory requirements of the projection.

### 2024-05-30 analysis/cloud/setup_azure.sh

@@ -40,7 +40,7 @@ sed 's,<subnetId>,'${subnetid//&/\\&}',g' analysis/cloud/pool_roads.json\
 # make a different json for each task
 mkdir -p analysis/cloud/task_jsons
 
-for rowi in {25..27}
+for rowi in {1..27}
 do
   sed 's,<SASURL>,'${sasurl//&/\\&}',g' analysis/cloud/task_roads.json\
   | sed 's,<row>,'$rowi',g' > analysis/cloud/task_jsons/task_roads_$rowi.json
@@ -92,7 +92,7 @@ az batch pool create --json-file analysis/cloud/pool_to_use.json
 az batch job create --pool-id $poolName --id $jobName
 
 
-for rowi in {25..27}
+for rowi in {1..27}
 do
   az batch task create --json-file analysis/cloud/task_jsons/task_roads_$rowi.json --job-id $jobName
 done
@@ -105,9 +105,9 @@ done
 
 # az batch task delete --task-id connectivity-combine --job-id $jobName --yes
 #
-# for rowi in real1000 real notreal
+# for rowi in {1..27}
 # do
-#   az batch task delete --job-id $jobName --task-id roads-benchmark-$rowi --yes
+#   az batch task reactivate --job-id $jobName --task-id roads-benchmark-$rowi
 # done
 
 # set target dedicated nodes
@@ -171,7 +171,7 @@ az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/\$AZ_BATCH_TA
 az storage remove -c sendicott -n \$AZ_BATCH_TASK_ID --account-name ecdcwls --sas-token $sastoken --recursive
 
 # Remove just files matching pattern
-az storage remove -c sendicott --include-pattern "*.txt" --account-name ecdcwls --sas-token $sastoken --recursive
+az storage remove -c sendicott --include-pattern "*.rds" --account-name ecdcwls --sas-token $sastoken --recursive
 
 # NOTE removes ***everything*** from the storage container
 az storage remove -c sendicott --exclude-pattern "*.gpkg" --account-name ecdcwls --sas-token $sastoken --recursive

### 2024-05-30 analysis/cloud/setup_azure.sh
updated results with fixed diagonals

@@ -105,7 +105,7 @@ done
 
 # az batch task delete --task-id connectivity-combine --job-id $jobName --yes
 #
-# for rowi in {1..27}
+# for rowi in real1000 real notreal
 # do
 #   az batch task reactivate --job-id $jobName --task-id roads-benchmark-$rowi
 # done
@@ -165,7 +165,7 @@ az storage blob list -c sendicott --account-name ecdcwls --sas-token $sastoken \
 az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/*?$sastoken \
 -d "analysis/data/derived_data/bench_results" --include-pattern "*.rds"
 
-az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/\$AZ_BATCH_TASK_ID?$sastoken \
+az storage copy -s https://ecdcwls.blob.core.windows.net/sendicott/results_notreal?$sastoken \
 -d "analysis/data/derived_data/cloud_combine_results" --recursive
 
 az storage remove -c sendicott -n \$AZ_BATCH_TASK_ID --account-name ecdcwls --sas-token $sastoken --recursive
@@ -184,3 +184,6 @@ az batch pool delete --pool-id $poolName
 rm analysis/cloud/make_to_use.R
 rm analysis/cloud/pool_to_use.json
 rm analysis/cloud/task_jsons --recursive
+
+
+

### 2024-05-30 analysis/scripts/7_compile_results_cloud_notreal.R

@@ -49,7 +49,7 @@ param_tbl <- paramTable3
 run_projections(param_tbl %>% filter(agg == 10, cutblocks_real == "revelstoke") %>%
                   rowwise() %>%
                   mutate(across(everything(), as.character)) %>%
-                  mutate(output = file.path(paste0("result_",
+                  mutate(output = file.path(paste0("bench_results/result_",
                                                    paste0(c_across(everything()), collapse = "_"),
                                                    ".rds"))),
                 cutblocksPth = paste0(data_path_drvd, "cutblocks_revelstoke.gpkg"),

### 2024-05-30 analysis/scripts/7_compile_results_cloud_real.R

@@ -51,7 +51,7 @@ param_tbl <- paramTable3
 run_projections(param_tbl %>% filter(agg == 10, cutblocks_real == "revelstoke_real") %>%
                   rowwise() %>%
                   mutate(across(everything(), as.character)) %>%
-                  mutate(output = file.path(paste0("result_",
+                  mutate(output = file.path(paste0("bench_results/result_",
                                                    paste0(c_across(everything()), collapse = "_"),
                                                    ".rds"))),
                 cutblocksPth = paste0(data_path_drvd, "cutblocks_revelstoke_real.gpkg"),

### 2024-05-30 analysis/scripts/7_compile_results_cloud_real1000.R

@@ -49,7 +49,7 @@ param_tbl <- paramTable3
 run_projections(param_tbl %>% filter(agg == 100) %>%
                   rowwise() %>%
                   mutate(across(everything(), as.character)) %>%
-                  mutate(output = file.path(paste0("result_",
+                  mutate(output = file.path(paste0("bench_results/result_",
                                                    paste0(c_across(everything()), collapse = "_"),
                                                    ".rds"))),
                 cutblocksPth = paste0(data_path_drvd, "cutblocks_revelstoke_real.gpkg"),

### 2024-06-10 analysis/paper/paper.Rmd
updated results with fixed diagonals
Implement Kyle's comments

@@
 ---
-title: "Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife"
+title: "Development and assessment of automated forest road projection methods using performance metrics relevant for wildlife"
 author:
@@
 
-Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available.
+Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other forest values, and the increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included automated road network projections, in part because tools suited for this purpose have not been available.
 Hardy et al.
@@
 As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
-Grade penalties improve performance in a mountainous region of British Columbia, as does the minimum spanning tree method.
-However, the minimum spanning tree method requires more time and memory to compute, so users must weigh the benefits of improved performance against computational costs.
-These simple road network simulation methods can improve projections of important impacts of development on wildlife across large areas, and we hope that more easily useable open source implementations will encourage integration of resource roads into forest change projections.
+Grade penalties improve performance in a mountainous region of British Columbia, as does our minimum spanning tree method.
+However, the minimum spanning tree method requires more computing time and memory, so users must weigh the benefits of improved performance against computational costs.
+Our road network simulation methods can improve projections of anticipated resource development impacts on wildlife across large areas, and we hope that more easily useable open source implementations will encourage integration of resource roads into forest change projections.
+
+# Study Implications
+
+This study offers open source tools for automated forest road network projection that are suited for integration into modelling frameworks used to project the implications of forest change for wildlife.
+We assess road network projection methods using ecologically relevant metrics, and show that grade penalties can improve performance in a topographically complex landscape.
+These tools can help improve projections of the cumulative effects of natural and anthropogenic disturbances on wildlife in an era of rapid change.
 
@@
 Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022].
-In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
+In particular, forest landscape models that can account for harvest, fire, and other disturbances are increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
 Despite the importance of roads for wildlife, very few of these projections have included road networks [@labadie_global_2023; @leston_quantifying_2020].
@@
 
-While road projections are not typically integrated with forest landscape models, projection methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g.
-@remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include projection tools.
-These have occasionally been incorporated into forest landscape models [e.g.
+While road projections are not typically integrated with forest landscape models, methods have been developed to support harvest planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007], and some forest management planning software suites [e.g.
+@remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] include road projection tools.
+These methods have occasionally been incorporated into forest landscape models [e.g.
 @leston_quantifying_2020] but they were not designed for that purpose.
-The algorithms contained in proprietary closed-source software are difficult to fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] used for research because costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for use in research.
-The level of detail required to parameterize forest management planning tools is also often difficult to obtain at landscape scales.
-Harvest road planning methods and tools [e.g.
-@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of the road network may suffice.
-There is a need for simple open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
+Existing road projection algorithms contained in proprietary closed-source software are difficult to fully integrate into stochastic models, which often require numerous replicates [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017].
+Further, costs of licensing are generally prohibitive, deploying large commercial software suites designed for desktops to cloud or high performance computing resources is difficult or impossible, and closed-source software is often not sufficiently transparent for use in research.
+At the operational scale, road planning methods and tools [e.g.
+@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of scaling their inputs to larger spatial extents can be prohibitive in contexts where a coarser approximation of the road network may suffice.
+Clearly, there is a need for open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
 
-Geographic route finding algorithms often draw from a larger body of work on methods for finding paths among points that minimize some measure of cost [@deo_graph_2016].
+Automated road projection methods often rely on geographic route finding algorithms that find a path between two points that minimizes some measure of cost defined by a mathematical graph [@deo_graph_2016].
 General-purpose open source tools for graph analysis [e.g.
@@
 @van_etten_r_2017] can help make these general-purpose tools easier to use and apply in specific contexts.
-Promising simple graph-based options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
+Promising simple graph-based options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to reconstruct road development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022].
 The former method is available as QGIS and LANDIS-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate].
-Hardy's tool is designed for integration with other components of the LANDIS-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+The Hardy et al.
+[-@hardy_landis-ii_2023] tool is designed for integration with other components of the LANDIS-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006; @anderson_projecting_2004; @stuckelberger_improved_2007; @clark_three-stage_2000].
-Anderson and Nelson’s (2004) more complex iterative least cost path method allows for consideration of horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
-However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding, and we focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs (represented as edge weights on a regular isotropic graph) from values of adjacent locations on a single user-supplied input raster.
+Anderson and Nelson’s (2004) iterative least cost path method was applied to a 7500 ha area, which allowed them to to increase the complexity of the cost surface by considering horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
+However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding, and we focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs (represented as edge weights on a simple undirected graph) from values of adjacent locations on a single user-supplied input raster.
 This constraint permits penalization of steep grades as suggested by Anderson and Nelson [-@anderson_projecting_2004], but does not permit other aspects of their method.
@@
 [-@lochhead_linking_2022].
-If these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context.
+We argue, if these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context.
 
-To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] and a simplified version of the grade penalty approach suggested by Anderson and Nelson [-@anderson_projecting_2004] in an open source R package.
-Our intention is to produce outputs comparable to Hardy et al.
+To help address the need for forest road network projection tools that can be easily integrated into large spatial scale stochastic models of forest change we implemented two alternative approaches (described in further detail in the next section): 1) iterative least cost paths and 2) a minimum spanning tree  [@lochhead_demo_2018; @lochhead_linking_2022], with two different methods for determining the cost of road building: 1) a simple elevation driven cost surface and 2) a simplified version of the grade penalty approach suggested by Anderson and Nelson [-@anderson_projecting_2004] in an open source R package.
+Our intention was to produce outputs comparable to Hardy et al.
 [-@hardy_forest_2019; -@hardy_landis-ii_2023] with a tool that can be used in modelling contexts outside of QGIS/LANDIS, and can serve as a more easily modifiable starting point for further refinement and development.
@@
 The adequacy of a road network projection method depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose.
-Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023; @stewart_climate-informed_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
+Given our interest in projecting impacts on wildlife we focused on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023; @stewart_climate-informed_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
 
-We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our simplified method for penalizing steep grades to yield more realistic road networks in steep terrain.
+We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our method for penalizing steep grades to yield more realistic road networks in steep terrain relative to our simple cost surface.
 We do not have a similarly strong hypothesis about whether minimum spanning tree projections are likely to be more or less realistic than iterative least cost path projections.
-If designers of real roads take a larger view to accessing multiple sites at the lowest cost we might expect minimum spanning tree algorithms to yield more realistic networks.
+We expect minimum spanning tree projections to be more accurate when resource roads are designed to simultaneously access multiple sites at the lowest cost.
 However, real road networks are built over time, and it is not necessarily true that they are designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.
@@
 We compare two methods for integrating these costs into the cost of connections between adjacent raster cells.
-In the first "simple cost" method, the cost of the connection between two adjacent cells is simply the mean of a cost raster value at each of those cells.
-In the second "grade penalty" method (Supplement 1-1), we calculate the cost of a connection between neighbouring cells as a function of the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
+In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Supplement A).
+The cost of the connection between two adjacent cells is simply the mean of the cost raster value at each of those cells, adjusted for euclidean distance between the cells to penalize longer diagonal road segments.
+In the "grade penalty" method (Supplement 1-1), we use the same construction costs (Table 1), but we calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
+The most important difference between these two approaches is that grade is a property of the edge between two nodes in the grade penalty method, while slope is a property of each node in the simple cost method.
+When slope is a node property, the same slope penalty applies regardless of the alignment of the road segment.
+Calculating grade from the difference in elevation between pairs of nodes allows us to penalize roads that follow elevation contours differently from roads that go up or down.
 The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha.
-We examined sensitivity to our choice of resolution by comparing results from 1 ha and 100 ha weight rasters.
+We examined sensitivity to pixel resolution by comparing results from 1 ha and 100 ha weight rasters.
 We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions \>= 1 ha.
@@
 
-We focus on methods that link multiple target locations to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
-Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
-We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy, personal communication).
+Our road projection methods focused on linking multiple target locations (also known as landings within harvest cutblocks (i.e., log decks)) to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
+Specifically, we compared iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+We used the QGIS version of Hardy’s method (2019), which was easier for us to apply and does not differ substantially from the LANDIS extension (Hardy, personal communication).
 
 The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs).
-For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [@dijkstra_note_1959].
-There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
-The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement 1-1), weighted the distance to neighbouring nodes.
-Both the grade penalty and simple cost methods involve setting edge weights on a simple regular isotropic graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation.
-We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
+For each landing (target location), the least cost path to the existing road network was solved using a mathematical graph and  Dijkstra's shortest path algorithm [@dijkstra_note_1959].
+There are a variety of possible methods for building graphs, choosing branch locations, and ordering target locations, and; these different methods will produce different projected road networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
+The ILCP algorithm we used [@lochhead_demo_2018; @lochhead_linking_2022] was parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
+Each node was connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement 1-1), which is then weighted by the distance to neighbouring nodes.
+Both the grade penalty and simple cost methods involve setting edge weights on a graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation.
+We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest (by Euclidean distance) point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
 
@@
 [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network.
-A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network.
+A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the landing cell and one of multiple potential connection points on the road network.
 In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve the least cost path between the two.
@@
 
-Minimum spanning trees (MST) can also be used to link landings and evaluate road branching opportunities [@picard_finding_2006; @clark_three-stage_2000].
+One of the limitations with ILCP methods is the sensitivity of the road projection to the ordering of the landings (Anderson and Nelson 2004).
+Minimum spanning trees (MST) offer an alternative approach to link landings and evaluate road branching opportunities [@picard_finding_2006; @clark_three-stage_2000].
 MST algorithms do not require users to select the order in which landings are processed, but it remains necessary to select a set of points on the existing road network to include in the tree.
-The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network.
-These results are used to create a weighted graph where nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
+As with the ILCP approach, the first step is to construct a graph in which each node is connected with up to eight neighbours, and set the cost of construction along edges using either the grade penalty or simple cost methods.
+This graph is used to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network.
+These results are used to create a second weighted graph where nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
 The minimum spanning tree for this graph is solved using Kruskal's algorithm [@kruskal_shortest_1956], which links all nodes with a tree that minimizes the sum of edge weights.
@@
 [-@clark_three-stage_2000] who used Euclidean distances for edge weights, and thus did not incorporate spatial variation in road construction costs.
+This extra step increases the number of computational steps, which may be important to the user when conducting large scale anlayses that are compute heavy.
 
@@
 ```
-Note that the evolution of a road network over time can be modeled by iteratively applying any of the methods, but in this manuscript we focus on comparing outcomes of a single application of each of the methods to a network of forest roads developed since 1990.
+Note that the evolution of a road network over time can be modeled by iteratively applying any of the methods, but in this study we focus on comparing a single application of each of the methods to a network of forest roads developed since 1990.
 
-To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
+To help make these road network projection methods more easily accessible, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
 The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
@@
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
+The 12 projections were all run on separate compute nodes and the compute time and peak RAM usage were recorded in order to compare the speed and memory requirements of each method (Supplement E).
+We also ran projections with the ILCP and MST methods with regular, high density sampling and centroids at resolutions of 50 ha, 0.3 ha and 0.1 ha to more fully explore how speed and memory requirements change with number of landings and number of nodes in the graph, but we did not assess their performance (Supplement E)
 
@@
 First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks.
-The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS.
+The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were visaully investigated using Google satellite imagery in QGIS.
 Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed.
@@
 
-Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from real roads.
-Even in cases where the Hardy method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks, in that the grade penalty method encourages roads to follow slope contours, while the Hardy method creates more curves (Figure \@ref(fig:proj-meth-fig)).
-Although the grade penalty method improves realism by following elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in the engineering of real roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
-Our projected roads also cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by better accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms.
-Finally, it is important to note that some discrepancies between observed and projected road networks may be caused by errors in the observed data, such as cutblocks that do not exist (e.g.
+Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from observed roads.
+While, the Hardy et al.(2023) method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks.
+In particular, the grade penalty method incentivized roads to follow slope contours, while the Hardy method created more road curvatures (Figure \@ref(fig:proj-meth-fig)).
+Although the grade penalty method improved realism by following elevation contours, all these methods created dead ends, intersections, and abrupt turns that would be avoided in the engineering of operational roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
+All of the projected road methods also crossed valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by improved costs of crossing streams, rivers and other wet areas in valley bottoms.
+Finally, it is important to note that some discrepancies between observed and projected road networks were caused by errors in the observed data, such as cutblocks that were incorrectly classified (e.g.
 \@ref(fig:proj-map-fig)b).
-We omitted false cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Figures S3 and S4).
+We omitted cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Figures S3 and S4).
 
-Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
+All of the road projection methods underestimate road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
 In other words, these road network projection methods project fewer roads than we observe.
-Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.
-One reason for underprojection of road density is that all these methods generate straight road segments within pixels.
+Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures were more robust to our coarse assumptions about road projection.
+One reason for underestimation of road density is that all these methods generate straight road segments within a pixel.
 As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).
@@
 
-Although these methods project fewer roads than we observe, they represent a substantial improvement over the "cutblocks only" scenario, in which no new roads are added (Figure \@ref(fig:aspat-perf-fig)).
-That said, these methods are not designed to project the exact locations of new roads, so they (unsurprisingly) do not yield accurate spatial projections of road presence (i.e.
-the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`).
-Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
+Although these methods underestimate road presence, they represent a substantial improvement over the "cutblocks only" scenario, which assumes no new roads are built (Figure \@ref(fig:aspat-perf-fig)).
+These relatively coarse methods were not designed to project the exact spatial location of new roads (within 100 m), so it is not surprising that they did not yield accurate spatial projections of road presence (F-Measure: `r perf_tbl_rng["Road presence"]`).
+On the other hand, none of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+This is likely because many of the projected roads are within 500 m of either the existing road network or a cutblock so the false negatives created by failing to project roads are balanced by the lack of false positives.
+All methods other than "cutblocks only" provided comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yielded fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
 
-Although the MST method performed better than alternatives according to several metrics (Figure \@ref(fig:aspat-perf-fig)), it also requires more memory and time to compute on large landscapes (Figure S7).
-All of these methods require construction of a large graph with a node for each pixel in the weight raster where road construction is possible (i.e.
-not NA), and memory requirements increase with the number of nodes in the graph (Figure S7).
+Although the MST method performed better than the alternatives we tested in terms of several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it requires greater computation resources for large landscapes (Figure S7).
+All of these methods require construction of a large mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e.
+not NA), and as such, memory requirements will increase with the number of nodes in the graph (Figure S7).
 Compute time also increases with the number of landing targets (Figure S7).
 Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function.
-The Hardy QGIS plugin is an order of magnitude faster than the MST and ILCP methods implemented in R (Figure S7).
+The Hardy QGIS plugin is more than an order of magnitude faster than the MST and ILCP methods implemented in R (Figure S7).
 
@@
 @labadie_global_2023].
-Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that better predicts road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that improved predicted road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
-Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can help guide method selection in that context.
+Although projecting roads within a stochastic ecological forecasting model was beyond the scope of this study, our analysis of performance based on metrics relevant to wildlife habitat can help guide method selection in that context.
 
@@
 The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement 1-1).
-We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it.
+We found several advantages to this approach in regions where slope is an important determinant of road construction cost, and we recommend its use.
 However, we also acknowledge that we have used a very simple cost model, and that the important determinants of road construction costs may vary among landscapes.
@@
 culverts and bridges).
-Real construction costs and impacts also vary substantially among types of roads (e.g.
+Road construction costs and wildlife impacts also vary substantially among types of roads (e.g.
 long-term vs temporary haul roads).
-If distinguishing among types of roads is important, an effective approach might be to first project the paths of lengthy long-term haul roads using a cost model and set of targets that is appropriate for this type of road, and then project the paths of shorter temporary haul roads from this long-term network using a different cost model and set of landing locations.
+If distinguishing among types of roads is important, an effective approach might be to first project the paths of lengthy long-term haul roads using a cost model and a set of targets that are appropriate for this type of road, and then project the paths of shorter temporary haul roads or spur roads from this long-term network using a different cost model and set of landing locations.
 
-Among the method variants we considered, the minimum spanning tree (MST) regular high density method yielded better projections of average road density and the forestry disturbance footprint (Figure 3).
-Aspatial projections of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3), so overall the MST regular high density projections were more accurate.
-However, the MST algorithm is also more computationally demanding than the other methods we considered (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
+Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Figure 3).
+Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3).
+However, the MST algorithm was also more computationally demanding than the other methods we tested (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
 
@@
 To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns.
-This examination highlighted some errors that could be fixed by better accounting for the cost of water crossing in our weight raster, and other more fundamental limitations and differences among methods.
+This examination highlighted some errors that could be fixed by better accounting for the cost of water crossings in our weight raster, and other more fundamental limitations and differences among methods.
 Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure 4).
@@
 These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible.
-The associated memory and computational requirements may present challenges on large landscapes (Figure S7).
+While we were able to run the models in our example landscape up to 100 m resolution in a reasonable amount of time (< 2.4 hours) using modest amounts of memory (<3 GB RAM), projections at 30 m and 10 m resolutions caused R to crash due to a lack of available memory (Supplement E).
+Although this challenge can be fairly easily overcome by running the projections on a machine with more RAM, the memory and computational requirements may present a challenge on larger landscapes or when incorporated into more complex simulation models.
+For example, for a project studying the effects of different forest management and climate change scenarios on wildlife in the Churchill caribou range in Ontario, we projected road development across the landscape, which at 50 m resolution has a graph with greater than 21 million nodes.
+Projecting roads using the ILCP method on this landscape took 2.7 hours to run and required 11 GB of RAM for one realization.
+Overall we hope to complete 600 different realizations which will require significant time and resources.
+
 Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations.
 The first three options are available to users of the `roads` R package.
-However, those contemplating reducing computational cost by decreasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Figures S5 and S6).
+However, those contemplating reducing computational cost by decreasing raster resolution should be cautious of the results given that all of these methods become less accurate as raster cell size increases (Figures S5 and S6).
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
-Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete even if it is also adequate for projecting the combined footprint of roads and cutblocks.
+Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length or road density will be underestimated, while the combined footprint of roads and cutblocks should be reasonably accurate.
 More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Figure S8) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
@@
 objectives, species, landscapes).
-The most informative performance metrics will be those most closely linked to outcomes of interest.
 For example, where the outcome of interest is space use or connectivity projected from more [e.g.
 @whittington_towns_2022; @labadie_global_2023] or less [e.g.
-@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity of these models to the difference between observed and retrospectively projected roads.
+@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it is recommended to test their sensitivity to the difference between observed and retrospectively projected roads.
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
-Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals.
+In lieu of testing additional performance metrics, we have focused our efforts on the development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that enable users to devise performance tests appropriate to their particular circumstances and goals.
 We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].
 
-In this paper our focus is tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife.
-However, we note there are other possible uses for these road network projection methods.
+In this study our focus was developing tools capable of integrating forest road development into projections of the cumulative effects of disturbance on wildlife.
+However, we note there are other possible uses for our road network projection methods.
 Cost minimizing methods are not suitable for projecting development of seismic lines (e.g.
@@
 [-@lochhead_linking_2022] used the least cost path algorithm to reconstruct the likely history of road network development in a case where harvest dates were known but road construction dates were not.
-In principle, these algorithms could also be used to link disconnected road segments into a network, but local route finding algorithms [@roussel_vectorial_2023] that do not require construction of large graphs are better suited to this purpose.
-More obviously, spatial stochastic models are used to project impacts of disturbance on a variety of values other than wildlife [@albrich_simulating_2020; @hof_editorial_2021; @sturtevant_understanding_2021].
+In principle, these algorithms could also be used to link disconnected road segments into a network.
 It would also be interesting to account for persistent impacts of resource roads on vegetation, productivity and carbon sequestration [@braham_characterization_2023].
+
+# Conclusion
+
+TODO
+In conclusion we found coarse road projection methods can accurately project some disturbance metrics.
+We recommend a high density of landings, a grade penalty in the cost surface and our MST algorithm for ordering landings with iterative least cost paths approaches.
+……
 
@@
 ```
+
 \newpage
@@
 
-Table 1: A simple model of road construction costs.
+Table 1: Road construction costs.
 In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
@@
 \newpage
+
 Table 2: Road projection method variants.

### 2024-06-10 analysis/paper/paper.Rmd
WIP updating results based on most recent run.

@@
 We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
-The MST regular high density method yields the best projections of road density within cutblocks, and all MST variants yield comparably good projections outside cutblocks, so overall the MST regular high density variant most accurately projects road density.
+The MST regular high density method yields the best projections of road density overall, while the MST random high density method had the best projections within cutblocks, and all MST variants yield comparably good projections outside cutblocks.
 All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.
@@
 
-All of the road projection methods underestimate road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
+All of the road projection methods underestimate, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
 In other words, these road network projection methods project fewer roads than we observe.
@@
 Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function.
-The Hardy QGIS plugin is more than an order of magnitude faster than the MST and ILCP methods implemented in R (Figure S7).
+The Hardy QGIS plugin is 65 times faster than the MST and ILCP methods implemented in R (Figure S7).
 

### 2024-06-11 analysis/paper/paper.Rmd
make spat perf plot easier to compare and return table

@@
 
+```{r get-data}
+# load csv
+meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
+                                   "mean_table.csv"))
+
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
+                                   "agree_table.csv"))
+
+aspat_perf <- doAspatPlot(meanTableRealCuts)
+
+# need to make plots more comparable
+# One option is removing extreme value in distance to road
+aspat_perf$plt$data <- aspat_perf$plt$data %>% mutate(prop_dif = ifelse(metric == "Distance\nto road" & sampleDens == "Cutblocks only", NA, prop_dif))
+
+# Other option is setting limits but want to avoid excess white space
+ann_plus <- aspat_perf$prop_dif %>% 
+  filter(abs(prop_dif) > 0.5) %>% 
+  mutate(lab = round(prop_dif, 2), 
+         prop_dif = 0.37*sign(prop_dif))
+
+aspat_perf$plt+coord_flip(ylim = c(-0.35, 0.35))+
+  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.2,
+            size = 10, size.unit = "pt")
+
+perf_tbl <- doSpatPerf(matchDataRealCuts)
+perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
+  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
+```
+
 According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure B.1).
@@
 As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).
-
-```{r get-data}
-# load csv
-meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
-matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
-meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
-                                   "mean_table.csv"))
-
-matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
-                                   "agree_table.csv"))
-
-perf_tbl <- doSpatPerf(matchDataRealCuts)
-perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
-  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
-
-```
 
@@
 # organize data for ggplot
-doAspatPlot(meanTableRealCuts)
+aspat_perf$plt
 ```

### 2024-06-11 analysis/paper/supplement/supplement.Rmd
make spat perf plot easier to compare and return table

@@
 # organize data for ggplot
-doAspatPlot(meanTableRealCuts)
+aspat_simp <- doAspatPlot(meanTableRealCuts)
+aspat_simp$plt
 ```
@@
 meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
-doAspatPlot(meanTable)
+aspat_all <- doAspatPlot(meanTable)
+aspat_all$plt
 ```

### 2024-06-11 R/functionsSource.R

@@ -604,7 +604,7 @@ doAspatPlot <- function(tab){
   pal <- met.brewer(pal_nm, 8) %>% .[c(8, 1:7)]
   names(pal) <- levels(prop_dif$sampleDens) %>% setdiff("Observed")
 
-  prop_dif %>%
+  plt <- prop_dif %>%
     ggplot(aes(x = sampleDens, prop_dif, fill = sampleDens))+
     geom_hline(aes(yintercept = 0), color = "grey")+
     geom_col(position = position_dodge2(preserve = "single",
@@ -618,6 +618,7 @@ doAspatPlot <- function(tab){
           legend.position = "right",
           axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
     coord_flip()
+  return(lst(prop_dif, plt))
 }
 
 doSpatPerf <- function(tab){

### 2024-06-11 analysis/paper/paper.Rmd
finish updating for new results and incorporate JH edits

--- 2024-06-10 analysis/paper/paper.Rmd
@@
 
-We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is difficult or impossible.
+We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is challenging.
 Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain.
@@
 
-According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure B.1).
+```{r get-data}
+# load csv
+meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
+matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
+meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
+                                   "mean_table.csv"))
+
+matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
+                                   "agree_table.csv"))
+
+aspat_perf <- doAspatPlot(meanTableRealCuts)
+
+perf_tbl <- doSpatPerf(matchDataRealCuts)
+perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
+  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
+
+mst_reg_dens <- aspat_perf$prop_dif %>% 
+  filter(metric == "Road\ndensity", sampleDens == "MST regular high density", 
+         areaMean == "Overall")
+
+mst_ran_dens <- aspat_perf$prop_dif %>% 
+  filter(metric == "Road\ndensity", sampleDens == "MST random high density", 
+         areaMean == "Cutblocks")
+
+mst_out_dens <- aspat_perf$prop_dif %>% 
+  filter(metric == "Road\ndensity", method == "mst", areaMean == "Outside\ncutblocks")
+
+all_over_fordens <- aspat_perf$prop_dif %>% 
+  filter(metric == "Forestry\ndisturbance\nfootprint", areaMean == "Overall")
+```
+
+According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure S1).
 We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
-The MST regular high density method yields the best projections of road density overall, while the MST random high density method had the best projections within cutblocks, and all MST variants yield comparably good projections outside cutblocks.
-All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.
+The MST regular high density method yields the best projections of road density overall (projected density: `r mst_reg_dens$response_proj %>% round(2)` m roads/ha, observed density: `r mst_reg_dens$response_obs %>% round(2)` m roads/ha), while the MST random high density method had the best projections within cutblocks (projected density: `r mst_ran_dens$response_proj %>% round(2)` m roads/ha, observed density: `r mst_ran_dens$response_obs %>% round(2)` m roads/ha), and all MST variants yield comparably good projections outside cutblocks (projected density: `r mst_out_dens$response_proj %>% range() %>% round(2) %>% paste0(collapse = " - ")` m roads/ha, observed density: `r mst_out_dens$response_obs %>% unique() %>% round(2)` m roads/ha).
+All MST variants yield better projections of the forestry disturbance footprint (projected: `r all_over_fordens %>% filter(method == "mst") %>% pull(response_proj) %>% range() %>% round(2) %>% unique()`, observed: `r all_over_fordens %>% filter(method == "mst") %>% pull(response_obs) %>% unique() %>% round(2)`) than the Hardy or ILCP methods (`r all_over_fordens %>% filter(sampleDens %in% c("Hardy QGIS", "ILCP regular high density")) %>% pull(response_proj) %>% round(2) %>% unique()`), and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.
 
@@
 In other words, these road network projection methods project fewer roads than we observe.
-Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures were more robust to our coarse assumptions about road projection.
+Projections of the forestry disturbance footprint, the road disturbance footprint, and, to a lesser extent, distance to road were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures were more robust to our coarse assumptions about road projection.
 One reason for underestimation of road density is that all these methods generate straight road segments within a pixel.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).
-
-```{r get-data}
-# load csv
-meanTable <- read.csv(here(data_path_drvd,"../TSA27/dem", "mean_table.csv"))
-matchData <- read.csv(here(data_path_drvd,"../TSA27/dem", "agree_table.csv"))
-meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
-                                   "mean_table.csv"))
-
-matchDataRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts/dem",
-                                   "agree_table.csv"))
-
-perf_tbl <- doSpatPerf(matchDataRealCuts)
-perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
-  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
-
-```
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure S6).
 
@@
 Compute time also increases with the number of landing targets (Figure S7).
-Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function.
+Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` *geoCorrection* function.
 The Hardy QGIS plugin is 65 times faster than the MST and ILCP methods implemented in R (Figure S7).
@@
 
-Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Figure 3).
-Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3).
+Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
+Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure \@ref(fig:aspat-perf-fig)).
 However, the MST algorithm was also more computationally demanding than the other methods we tested (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
@@
 This examination highlighted some errors that could be fixed by better accounting for the cost of water crossings in our weight raster, and other more fundamental limitations and differences among methods.
-Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure 4).
+Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure \@ref(fig:proj-meth-fig)).
 We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.
@@
 These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible.
-While we were able to run the models in our example landscape up to 100 m resolution in a reasonable amount of time (< 2.4 hours) using modest amounts of memory (<3 GB RAM), projections at 30 m and 10 m resolutions caused R to crash due to a lack of available memory (Supplement E).
+While we were able to run the models in our example landscape up to 100 m resolution in a reasonable amount of time (\< 2.4 hours) using modest amounts of memory (\<3 GB RAM), projections at 30 m and 10 m resolutions caused R to crash due to a lack of available memory (Supplement E).
 Although this challenge can be fairly easily overcome by running the projections on a machine with more RAM, the memory and computational requirements may present a challenge on larger landscapes or when incorporated into more complex simulation models.
@@
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
-Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length or road density will be underestimated, while the combined footprint of roads and cutblocks should be reasonably accurate.
-More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Figure S8) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
+Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length and road density will be underestimated.
+More involved possibilities for reducing computational costs include adopting faster alternative routing algorithms [@larmet_cppRouting_2022], implementing slow components of the R package (Figure S8) in a faster compiled language, parallel processing, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
+Several of these options have been implemented in the cppRouting R package [@larmet_cppRouting_2022], and future versions of the roads R package could borrow and build on cppRouting functionality.
 
@@
 objectives, species, landscapes).
-For example, where the outcome of interest is space use or connectivity projected from more [e.g.
+For example, if the outcome of interest is space use or connectivity projected from more [e.g.
 @whittington_towns_2022; @labadie_global_2023] or less [e.g.
-@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it is recommended to test their sensitivity to the difference between observed and retrospectively projected roads.
+@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity to the difference between observed and retrospectively projected roads, in order to confirm that projections are adequate for the intended purpose.
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
@@
 
-TODO
-In conclusion we found coarse road projection methods can accurately project some disturbance metrics.
+TODO In conclusion we found coarse road projection methods can accurately project some disturbance metrics.
 We recommend a high density of landings, a grade penalty in the cost surface and our MST algorithm for ordering landings with iterative least cost paths approaches.
@@
 \newpage
+
 # Supplementary Material
@@
 \newpage
+
 # Figures
@@
 \newpage
+
 Figure 2: Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
@@
 Projection method variants are defined in Table 2.
+Note that extreme values for the cutblocks only method within cutblocks are cut off to ensure patterns in the other methods to remain visible.
+Bars that are cutoff are labeled with the total value.
 
@@
 # organize data for ggplot
-doAspatPlot(meanTableRealCuts)
+# need to make plots more comparable
+# Other option is setting limits but want to avoid excess white space
+ann_plus <- aspat_perf$prop_dif %>% 
+  filter(abs(prop_dif) > 0.5) %>% 
+  mutate(lab = round(prop_dif, 2), 
+         prop_dif = 0.37*sign(prop_dif))
+
+aspat_perf$plt+coord_flip(ylim = c(-0.35, 0.35))+
+  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
+            size = 8, size.unit = "pt")
 ```
@@
 Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks.
+
 ```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="."}
@@
 Table 1: Road construction costs.
-In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
+In the “grade penalty” method, we assign grades \>20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
 These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.
+
 ```{r cost-tbl}
@@
 We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.
+
 ```{r method-tbl}
@@
 \newpage
+
 ### Colophon

### 2024-06-11 analysis/paper/supplement/supplement.Rmd
finish updating for new results and incorporate JH edits

--- 2024-05-30 analysis/paper/supplement/supplement.Rmd
@@
 # organize data for ggplot
-doAspatPlot(meanTableRealCuts)
+aspat_simp <- doAspatPlot(meanTableRealCuts)
+
+ann_plus <- aspat_simp$prop_dif %>% 
+  filter(abs(prop_dif) > 0.5) %>% 
+  mutate(lab = round(prop_dif, 2), 
+         prop_dif = 0.37*sign(prop_dif))
+
+aspat_simp$plt+
+  coord_flip(ylim = c(-0.35, 0.35))+
+  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
+            size = 10, size.unit = "pt")
+
 ```
@@
 Cutblocks not accessible via the observed road network were excluded from analysis."}
-perf_tbl <- doSpatPerf(matchDataRealCuts)
-perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
+perf_tbl_simp <- doSpatPerf(matchDataRealCuts)
+perf_tbl_rng_simp <- perf_tbl_simp %>% group_by(metric) %>% 
   summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)
 
-doSpatPlot(perf_tbl)
+doSpatPlot(perf_tbl_simp)
 
@@
 Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection.
-"}
+Note that extreme values for the cutblocks only method within cutblocks are cut off to ensure patterns in the other methods to remain visible.
+Bars that are cutoff are labeled with the total value."}
 meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
-doAspatPlot(meanTable)
+aspat_all <- doAspatPlot(meanTable)
+
+ann_plus <- aspat_all$prop_dif %>% 
+  filter(abs(prop_dif) > 0.5) %>% 
+  mutate(lab = round(prop_dif, 2), 
+         prop_dif = 0.37*sign(prop_dif))
+
+aspat_all$plt+
+  coord_flip(ylim = c(-0.35, 0.35))+
+  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
+            size = 10, size.unit = "pt")
 ```
@@
 Metrics were summarised within cutblocks, outside cutblocks, and overall.
-Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
+Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation.
+Note that each panel has a different scale so bars cannot be compared across metrics."}
 # organize data for ggplot

### 2024-06-11 analysis/paper/paper.Rmd
missed space

@@
 Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from observed roads.
-While, the Hardy et al.(2023) method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks.
+While, the Hardy et al.
+(2023) method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks.
 In particular, the grade penalty method incentivized roads to follow slope contours, while the Hardy method created more road curvatures (Figure \@ref(fig:proj-meth-fig)).

### 2024-06-12 analysis/paper/paper.Rmd
Add alt text

@@
 
+Alt text: Two raster maps of Revelstoke timber supply area in grey scale, with an inset map showing the location of the study area within Canada.
+
 ```{r study-area, fig.cap="."}
@@
 Bars that are cutoff are labeled with the total value.
+Alt text: A faceted bar chart with one bar for each projection method variant and a panel for each combination of performance metrics and within cutblock, outside cutblock and overall.
 
@@
 
+Alt text: A series of 8 maps showing roads projected with different method variants.
+
 ```{r proj-meth-fig, out.width=paste0(fig_widths["two"], "in"), out.height="8in", fig.cap="."}
@@
 
+Alt text: A raster map of the Revelstoke TSA where the colour indicates the difference between obeserve and projected distance to road.
+Three inset maps zoom in of specific areas with Google satelite imagery in the background.
 ```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="."}
@@
 
+Alt text: A faceted bar chart with one bar for each projection method variant and a panel for each combination of binary performance metrics and within cutblock, outside cutblock and overall.
 ```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4.5in", fig.cap="."}

### 2024-06-12 analysis/paper/paper.Rmd
reconciling

--- 2024-06-11 analysis/paper/paper.Rmd
@@
 For example, in a recent collection of articles focused on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].
-In the absence of road projection methods, researchers are left to conservatively infer the amount and randomly assign the location of roaded areas [@johnson_witnessing_2015].
+In the absence of road projection methods, researchers are left to conservatively infer the amount and randomly assign the location of roaded areas [e.g.
+@johnson_witnessing_2015].
 
@@
 @anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of scaling their inputs to larger spatial extents can be prohibitive in contexts where a coarser approximation of the road network may suffice.
-Clearly, there is a need for open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
+There is a need for open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
 
-Automated road projection methods often rely on geographic route finding algorithms that find a path between two points that minimizes some measure of cost defined by a mathematical graph [@deo_graph_2016].
+Automated road projection methods often rely on more general purpose route finding algorithms that identify paths among points in a mathematical graph that minimize some measure of cost [@deo_graph_2016].
 General-purpose open source tools for graph analysis [e.g.
@@
 
+#TO DO: resume here.
+
 We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006; @anderson_projecting_2004; @stuckelberger_improved_2007; @clark_three-stage_2000].
-Anderson and Nelson’s (2004) iterative least cost path method was applied to a 7500 ha area, which allowed them to to increase the complexity of the cost surface by considering horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
+Anderson and Nelson (2004) appliedtheir method to a relatively small 7500 ha area,  surface by considering horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
 However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding, and we focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs (represented as edge weights on a simple undirected graph) from values of adjacent locations on a single user-supplied input raster.

### 2024-06-12 analysis/paper/supplement/supplement.Rmd
reconciling

--- 2024-05-30 analysis/paper/supplement/supplement.Rmd
@@
 
-# (APPENDIX) Supplementary Material for Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife {.unnumbered}
+# (APPENDIX) Supplementary Material for Development and assessment of automated forest road projection methods using performance metrics relevant for wildlife {.unnumbered}
 
@@
 
-We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of 3 different resolutions (1000 m, 500 m, 100 m, 30 m and 10m), giving 18 297, 72 633, and 1 815 825nodes in the cost graph, respectively.
-We also attempted to include resolutions of 30 m and 10 m (23 787 140 and 214 072 800 nodes, respectively) but these crashed due to insufficient memory.
+We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of three different resolutions (1000 m, 500 m and 100 m), giving 18 297, 72 633 and 1 815 825 nodes in the cost graph, respectively.
 Each cell in the raster becomes a node in the graph used to determine the shortest path which increases the time and memory requirements of the projection.

### 2024-06-12 analysis/paper/paper.Rmd
Preparing to submit

@@
 
-#TO DO: resume here.
-
 We acknowledge that using graph analysis methods to simulate roads is not new, and that other variants of iterative least cost path (or source-to-closest-target) and minimum spanning tree algorithms have been proposed and explored [@picard_finding_2006; @anderson_projecting_2004; @stuckelberger_improved_2007; @clark_three-stage_2000].
-Anderson and Nelson (2004) appliedtheir method to a relatively small 7500 ha area,  surface by considering horizontal and vertical alignment, turn radius and junction angles, and switchback spacing.
-However, constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding, and we focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs (represented as edge weights on a simple undirected graph) from values of adjacent locations on a single user-supplied input raster.
+Anderson and Nelson (2004) considered horizontal and vertical alignment, turn radius and junction angles, and switchback spacing for a relatively small 7500 ha area.
+However, for wide-ranging wildlife such as boreal caribou projections at much larger spatial scales are needed, and constructing and analyzing graphs for large landscapes is memory intensive and computationally demanding.
+We therefore focus here on assessing the adequacy of methods that minimize memory and computational requirements by setting construction costs (represented as edge weights on a simple undirected graph) from values of adjacent locations on a single user-supplied input raster.
 This constraint permits penalization of steep grades as suggested by Anderson and Nelson [-@anderson_projecting_2004], but does not permit other aspects of their method.
@@
 [-@lochhead_linking_2022].
-We argue, if these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context.
+We argue that if these simple algorithms can produce adequate outcomes for landscape-level stochastic ecological forecasting then the additional burden of more complex realistic algorithms is not warranted in that context.
 
-To help address the need for forest road network projection tools that can be easily integrated into large spatial scale stochastic models of forest change we implemented two alternative approaches (described in further detail in the next section): 1) iterative least cost paths and 2) a minimum spanning tree [@lochhead_demo_2018; @lochhead_linking_2022], with two different methods for determining the cost of road building: 1) a simple elevation driven cost surface and 2) a simplified version of the grade penalty approach suggested by Anderson and Nelson [-@anderson_projecting_2004] in an open source R package.
+To help address the need for forest road network projection tools that can be easily integrated into large spatial scale stochastic models of forest change we implemented two alternative approaches (described in further detail in the next section) in an open source R package: 1) iterative least cost paths and 2) a minimum spanning tree [@lochhead_demo_2018; @lochhead_linking_2022].
+We also consider two different methods for determining the cost of road building: 1) a simple elevation driven cost surface and 2) a simplified version of the grade penalty approach suggested by Anderson and Nelson [-@anderson_projecting_2004].
 Our intention was to produce outputs comparable to Hardy et al.
@@
 
-We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our method for penalizing steep grades to yield more realistic road networks in steep terrain relative to our simple cost surface.
+We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our method for penalizing steep grades to yield more realistic road networks in steep terrain relative to our simple cost method.
 We do not have a similarly strong hypothesis about whether minimum spanning tree projections are likely to be more or less realistic than iterative least cost path projections.
@@
 We compare two methods for integrating these costs into the cost of connections between adjacent raster cells.
-In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Supplement A).
+In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Supplement 1-1).
 The cost of the connection between two adjacent cells is simply the mean of the cost raster value at each of those cells, adjusted for euclidean distance between the cells to penalize longer diagonal road segments.
@@
 For each landing (target location), the least cost path to the existing road network was solved using a mathematical graph and Dijkstra's shortest path algorithm [@dijkstra_note_1959].
-There are a variety of possible methods for building graphs, choosing branch locations, and ordering target locations, and; these different methods will produce different projected road networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
+There are a variety of possible methods for building graphs, choosing branch locations, and ordering target locations, and these different methods will produce different projected road networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
 The ILCP algorithm we used [@lochhead_demo_2018; @lochhead_linking_2022] was parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
@@
 [-@clark_three-stage_2000] who used Euclidean distances for edge weights, and thus did not incorporate spatial variation in road construction costs.
-This extra step increases the number of computational steps, which may be important to the user when conducting large scale anlayses that are compute heavy.
 
@@
 To help make these road network projection methods more easily accessible, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+The `roads` package depends on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
 We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement 1-7) by default.
@@
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
-The 12 projections were all run on separate compute nodes and the compute time and peak RAM usage were recorded in order to compare the speed and memory requirements of each method (Supplement E).
-We also ran projections with the ILCP and MST methods with regular, high density sampling and centroids at resolutions of 50 ha, 0.3 ha and 0.1 ha to more fully explore how speed and memory requirements change with number of landings and number of nodes in the graph, but we did not assess their performance (Supplement E)
+We recorded compute time and peak RAM usage for each run, and ran some projections at 25 ha to show how speed and memory requirements vary with number of landings and number of nodes in the graph (Supplement 1-5), but did not assess other aspects of performance at that resolution.
 
@@
 Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from observed roads.
-While, the Hardy et al.
+While the Hardy et al.
 (2023) method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks.
@@
 
-Although the MST method performed better than the alternatives we tested in terms of several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it requires greater computation resources for large landscapes (Figure S7).
-All of these methods require construction of a large mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e.
-not NA), and as such, memory requirements will increase with the number of nodes in the graph (Figure S7).
+Although the MST method performed better than the alternatives we tested according to several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it requires greater computation resources for large landscapes (Figure S7).
+All of these methods require construction of a mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e.
+not NA), and memory requirements increase with the number of nodes in the graph (Figure S7).
 Compute time also increases with the number of landing targets (Figure S7).
 Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` *geoCorrection* function.
-The Hardy QGIS plugin is 65 times faster than the MST and ILCP methods implemented in R (Figure S7).
+For our example landscape the Hardy QGIS plugin was 65 times faster than the MST and ILCP methods implemented in R (Figure S7).
 
@@
 @barros_empowering_2023; @daniel_state-and-transition_2016].
-Although projecting roads within a stochastic ecological forecasting model was beyond the scope of this study, our analysis of performance based on metrics relevant to wildlife habitat can help guide method selection in that context.
+Although projecting roads within a stochastic ecological forecasting model was beyond the scope of this study, our analysis of performance using metrics relevant to wildlife can help guide method selection in that context.
 
@@
 The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement 1-1).
-We found several advantages to this approach in regions where slope is an important determinant of road construction cost, and we recommend its use.
+We found advantages to this approach in regions where slope is an important determinant of road construction cost, and we recommend its use.
 However, we also acknowledge that we have used a very simple cost model, and that the important determinants of road construction costs may vary among landscapes.
@@
 These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible.
-While we were able to run the models in our example landscape up to 100 m resolution in a reasonable amount of time (\< 2.4 hours) using modest amounts of memory (\<3 GB RAM), projections at 30 m and 10 m resolutions caused R to crash due to a lack of available memory (Supplement E).
-Although this challenge can be fairly easily overcome by running the projections on a machine with more RAM, the memory and computational requirements may present a challenge on larger landscapes or when incorporated into more complex simulation models.
-For example, for a project studying the effects of different forest management and climate change scenarios on wildlife in the Churchill caribou range in Ontario, we projected road development across the landscape, which at 50 m resolution has a graph with greater than 21 million nodes.
-Projecting roads using the ILCP method on this landscape took 2.7 hours to run and required 11 GB of RAM for one realization.
-Overall we hope to complete 600 different realizations which will require significant time and resources.
+In our example landscape we were able to project at 100 m resolution in a reasonable amount of time (\< 2.4 hours) using modest amounts of memory (\<3 GB RAM).
+However, our example landscape is fairly small, and memory requirements increase rapidly with landscape size (Supplement 1-5).
+Although it is possible to use machines with more RAM, the computational cost of adding road projections to spatial stochastic simulations of forest change relevant for wide-ranging wildlife is not trivial.
+For example, in a study of the effects of forest management and climate change scenarios on wildlife in a single boreal caribou range (100 stochastic realizations of 6 scenarios), we projected roads at 50 m resolution using graph with ~21 million nodes.
+Each ILCP road projection required 2.7 hours and 11 GB of RAM, for a total of 1620 compute hours.
 
@@
 Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length and road density will be underestimated.
-More involved possibilities for reducing computational costs include adopting faster alternative routing algorithms [@larmet_cppRouting_2022], implementing slow components of the R package (Figure S8) in a faster compiled language, parallel processing, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
-Several of these options have been implemented in the cppRouting R package [@larmet_cppRouting_2022], and future versions of the roads R package could borrow and build on cppRouting functionality.
+More involved possibilities for reducing computational costs include adopting faster routing algorithms [@larmet_cppRouting_2022], implementing slow components of the R package (Figure S8) in a faster compiled language, parallel processing, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
+Several of these options have been implemented in the `cppRouting` R package [@larmet_cppRouting_2022], and future versions of the `roads` R package could borrow and build on `cppRouting` functionality.
 
@@
 @whittington_towns_2022; @labadie_global_2023] or less [e.g.
-@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity to the difference between observed and retrospectively projected roads, in order to confirm that projections are adequate for the intended purpose.
+@diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity to the difference between observed and retrospectively projected roads to confirm that projections are adequate for the intended purpose.
 When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021].
-In lieu of testing additional performance metrics, we have focused our efforts on the development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that enable users to devise performance tests appropriate to their particular circumstances and goals.
+Rather than testing additional performance metrics, we have focused our efforts on the development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that enable users to devise performance tests appropriate to their particular circumstances and goals.
 We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].
@@
 [-@lochhead_linking_2022] used the least cost path algorithm to reconstruct the likely history of road network development in a case where harvest dates were known but road construction dates were not.
-In principle, these algorithms could also be used to link disconnected road segments into a network.
 It would also be interesting to account for persistent impacts of resource roads on vegetation, productivity and carbon sequestration [@braham_characterization_2023].

### 2024-06-12 analysis/paper/supplement/supplement.Rmd
Preparing to submit

@@
 
-We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of 3 different resolutions (1000 m, 500 m, 100 m, 30 m and 10m), giving 18 297, 72 633, and 1 815 825nodes in the cost graph, respectively.
-We also attempted to include resolutions of 30 m and 10 m (23 787 140 and 214 072 800 nodes, respectively) but these crashed due to insufficient memory.
+We compared the speed and peak RAM usage of the MST and ILCP methods with landings regularly sampled at 10 per km^2^ (high density) or using only the cutblock centroids as landings on weight rasters of 3 different resolutions (1000 m, 500 m, 100 m), giving 18 297, 72 633, and 1 815 825 nodes in the cost graph, respectively.
 Each cell in the raster becomes a node in the graph used to determine the shortest path which increases the time and memory requirements of the projection.

### 2024-06-12 analysis/paper/paper.Rmd
Preparing to submit.

@@
 At the operational scale, road planning methods and tools [e.g.
-@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the cost of scaling their inputs to larger spatial extents can be prohibitive in contexts where a coarser approximation of the road network may suffice.
+@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007; @remsoft_remsoft_nodate; @spatial_planning_systems_patchworks_nodate] may produce more accurate road projections but the costs of applying more complex methods across large spatial extents can be prohibitive in contexts where a coarser approximation of the road network may suffice.
 There is a need for open-source road network projection tools that can be integrated into the open modular platforms that are used for large-scale stochastic landscape simulations [@scheller_design_2007; @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 
-To help address the need for forest road network projection tools that can be easily integrated into large spatial scale stochastic models of forest change we implemented two alternative approaches (described in further detail in the next section) in an open source R package: 1) iterative least cost paths and 2) a minimum spanning tree [@lochhead_demo_2018; @lochhead_linking_2022].
+To help address the need for forest road network projection tools that can be easily integrated into large spatial scale stochastic models of forest change we implemented two alternative approaches in an open source R package: 1) iterative least cost paths and 2) a minimum spanning tree [@lochhead_demo_2018; @lochhead_linking_2022].
 We also consider two different methods for determining the cost of road building: 1) a simple elevation driven cost surface and 2) a simplified version of the grade penalty approach suggested by Anderson and Nelson [-@anderson_projecting_2004].
@@
 Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain.
-Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data.
+Most of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data.
 The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River.
@@
 Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
-To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the digital road atlas that were within 50 m of forest tenure roads.
-The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990.
-We assumed everything in the digital road atlas was built before 1990.
+We removed all roads from the digital road atlas that were within 50 m of forest tenure roads to avoid duplication.
+To obtain a starting point for projections we filtered out forest tenure roads with an award date greater than 1990, and assumed everything in the digital road atlas was built before 1990.
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.
-We further filtered cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks.
+We used cutblocks harvested after 1990 as as targets for the projections, and removed cutblocks that that do not contain observed roads to reduce errors associated with open areas that have been misidentified as cutblocks.
 
@@
 In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Supplement 1-1).
-The cost of the connection between two adjacent cells is simply the mean of the cost raster value at each of those cells, adjusted for euclidean distance between the cells to penalize longer diagonal road segments.
-In the "grade penalty" method (Supplement 1-1), we use the same construction costs (Table 1), but we calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
+The cost of the connection between two adjacent cells is simply the mean of the cost raster (Figure \@ref(fig:study-area)a) value at each of those cells, adjusted for euclidean distance to penalize longer diagonal road segments.
+In the "grade penalty" method (Supplement 1-1) we use the same construction costs (Table 1), but calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
 The most important difference between these two approaches is that grade is a property of the edge between two nodes in the grade penalty method, while slope is a property of each node in the simple cost method.
@@
 Calculating grade from the difference in elevation between pairs of nodes allows us to penalize roads that follow elevation contours differently from roads that go up or down.
-The road construction cost and elevation rasters consist of 1.7 million pixels with a resolution of 1 ha.
-We examined sensitivity to pixel resolution by comparing results from 1 ha and 100 ha weight rasters.
+We examined sensitivity to pixel resolution by comparing results from 1 ha (1.7 million pixels) and 100 ha (17 thousand pixels) weight rasters.
 We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions \>= 1 ha.
@@
 Data Catalogue [@data_bc_digital_2021].
-Scripts used for data acquisition and analysis associated with this paper are available on GitHub [@endicott_roadpaper_2024].
+Scripts used for data acquisition and analysis are available on GitHub [@endicott_roadpaper_2024].
 
@@
 
-Our road projection methods focused on linking multiple target locations (also known as landings within harvest cutblocks (i.e., log decks)) to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
+Our road projection methods link multiple target locations (also known as landings within harvest cutblocks (i.e., log decks)) to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
 Specifically, we compared iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
@@
 For each landing (target location), the least cost path to the existing road network was solved using a mathematical graph and Dijkstra's shortest path algorithm [@dijkstra_note_1959].
-There are a variety of possible methods for building graphs, choosing branch locations, and ordering target locations, and these different methods will produce different projected road networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
-The ILCP algorithm we used [@lochhead_demo_2018; @lochhead_linking_2022] was parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node was connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement 1-1), which is then weighted by the distance to neighbouring nodes.
-Both the grade penalty and simple cost methods involve setting edge weights on a graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation.
+There are a variety of possible methods for building graphs, choosing branch locations, and ordering target locations [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006].
+The ILCP algorithm we used [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
+Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement 1-1), which is then weighted by the distance to neighbouring nodes.
+Both the grade penalty and simple cost methods involve setting edge weights on a graph from adjacent locations on a single input grid.
 We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest (by Euclidean distance) point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
@@
 This graph is used to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network.
-These results are used to create a second weighted graph where nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
+These results are used to create a second weighted graph wherein nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes.
 The minimum spanning tree for this graph is solved using Kruskal's algorithm [@kruskal_shortest_1956], which links all nodes with a tree that minimizes the sum of edge weights.
@@
 
-To help make these road network projection methods more easily accessible, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023].
-The `roads` package depends on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
+To help make these methods accessible, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023] which depends on several other R packages: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
 We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement 1-7) by default.
@@
 We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)).
-We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200m.
-A skidding distance of 200 m corresponds to \~ 9 pts/km^2^.
-We compared two resolutions of the weight raster by aggregating the 1 ha resolution weight raster to a coarser 100 ha (1000 by 1000 m) resolution.
+We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200 m which gives \~ 9 pts/km^2^.
+We compared two resolutions of the weight raster (1 ha and 100 ha).
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
@@
 @labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].
-We also calculated road presence, measured as whether or not roads are present in each pixel, but we note that accurate projection of the exact locations of forest roads is not generally a requirement in stochastic ecological forecasting models wherein the exact locations of future disturbances are often not known.
+We also calculated road presence (i.e.
+whether or not roads are present in each pixel), but note that accurate projection of the exact locations of roads is not generally a requirement in stochastic ecological forecasts wherein the exact locations of future disturbances are often not known.
 Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].
@@
 First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks.
-The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were visaully investigated using Google satellite imagery in QGIS.
-Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed.
-These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
+The observed and projected road networks were then overlaid on the difference maps and regions of major discrepancies were investigated visually.
+Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed assigning pixels to 5 categories: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footprint was observed.
+We used these categories to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021].
 Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true.
@@
 
-All of the road projection methods underestimate, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
-In other words, these road network projection methods project fewer roads than we observe.
-Projections of the forestry disturbance footprint, the road disturbance footprint, and, to a lesser extent, distance to road were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures were more robust to our coarse assumptions about road projection.
+Projections of the forestry disturbance footprint, the road disturbance footprint, and, to a lesser extent, distance to road were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)).
+However, all of the road projection methods underestimate road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)), projecting fewer roads than we observe.
 One reason for underestimation of road density is that all these methods generate straight road segments within a pixel.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure S6).
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, making coarse resolution projections less accurate (Figure S6).
 
@@
 Although these methods underestimate road presence, they represent a substantial improvement over the "cutblocks only" scenario, which assumes no new roads are built (Figure \@ref(fig:aspat-perf-fig)).
-These relatively coarse methods were not designed to project the exact spatial location of new roads (within 100 m), so it is not surprising that they did not yield accurate spatial projections of road presence (F-Measure: `r perf_tbl_rng["Road presence"]`).
-On the other hand, none of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+The methods were not designed to project the exact spatial location of new roads (within 100 m), so it is not surprising that they did not yield accurate spatial projections of road presence (F-Measure: `r perf_tbl_rng["Road presence"]`).
+It is somewhat more surprising that none of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
 This is likely because many of the projected roads are within 500 m of either the existing road network or a cutblock so the false negatives created by failing to project roads are balanced by the lack of false positives.
@@
 
-Although the MST method performed better than the alternatives we tested according to several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it requires greater computation resources for large landscapes (Figure S7).
+Although the MST method performed better than the alternatives we tested according to several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it is also more costly to compute (Figure S7).
 All of these methods require construction of a mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e.
@@
 @labadie_global_2023].
-Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that improved predicted road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g.
+Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that improved predicted road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change [e.g.
 @barros_empowering_2023; @daniel_state-and-transition_2016].
@@
 We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region.
-The simplified grade penalty method does not substantially increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement 1-1).
-We found advantages to this approach in regions where slope is an important determinant of road construction cost, and we recommend its use.
+The simplified grade penalty method does not substantially increase memory or computational requirements, and we recommend its use in regions where slope is an important determinant of road construction cost.
 However, we also acknowledge that we have used a very simple cost model, and that the important determinants of road construction costs may vary among landscapes.
@@
 
-Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data.
-We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data that are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
 
@@
 In this study our focus was developing tools capable of integrating forest road development into projections of the cumulative effects of disturbance on wildlife.
-However, we note there are other possible uses for our road network projection methods.
+However, we note there are other possible uses for these road network projection methods.
 Cost minimizing methods are not suitable for projecting development of seismic lines (e.g.
@@
 
-TODO In conclusion we found coarse road projection methods can accurately project some disturbance metrics.
-We recommend a high density of landings, a grade penalty in the cost surface and our MST algorithm for ordering landings with iterative least cost paths approaches.
-……
+Our open source automated forest road projection tools can be integrated into spatial stochastic models of the implications of forest change for wide-ranging wildlife such as woodland caribou.
+In a mountainous test landscape, a high density of landings, a method for penalizing road construction on steep grades, and a minimum spanning tree algorithm all improved projection accuracy, and we recommend these options.
+We hope that these tools will reduce barriers to integrating resource roads into spatial stochastic models of forest dynamics, and thereby help improve projections of anticipated resource development impacts on wildlife across large areas.
 

### 2024-06-12 analysis/paper/paper.Rmd
very minor edits

@@
 Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes.
-We used cutblocks harvested after 1990 as as targets for the projections, and removed cutblocks that that do not contain observed roads to reduce errors associated with open areas that have been misidentified as cutblocks.
+We used cutblocks harvested after 1990 as targets for the projections, and removed cutblocks that do not contain observed roads to reduce errors associated with open areas that have been misidentified as cutblocks.
 
@@
 
-Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data that are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.
+Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data that are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_resource_nodate] that users should keep in mind when interpreting discrepancies between observed and projected road networks.
 

### 2024-06-13 analysis/paper/paper.Rmd
minor edit to fit word limit

@@
 keywords: |
-   forest roads, road networks, forest dynamics, spatially explicit modeling, predictive ecology, least cost paths
+   forest roads, forest dynamics, spatially explicit modeling, predictive ecology, least cost paths
 editor_options: 
@@
 
-Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other forest values, and the increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included automated road network projections, in part because tools suited for this purpose have not been available.
+Despite rapidly growing evidence of complex and varied impacts of road networks on wildlife and other forest values, and the increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included automated road network projections, in part because tools suited for this purpose have not been available.
 Hardy et al.
@@
 However, the minimum spanning tree method requires more computing time and memory, so users must weigh the benefits of improved performance against computational costs.
-Our road network simulation methods can improve projections of anticipated resource development impacts on wildlife across large areas, and we hope that more easily useable open source implementations will encourage integration of resource roads into forest change projections.
+Our road network simulation methods can improve projections of anticipated resource development impacts on wildlife across large areas, and we hope that more accessible open source implementations will encourage integration of resource roads into forest change projections.
 

### 2025-04-07 analysis/paper/paper.Rmd
reformat for landscape ecology

@@
 bibliography: Roads.bib
-csl: "../templates/chicago-author-date.csl" # Insert path for the bib-style
+csl: "../templates/landscape-ecology.csl" # Insert path for the bib-style
+citation-abbreviations: "../templates/ncbi-abbreviations.json"
 keywords: |
@@
 ```
+
 Keywords: `r rmarkdown::metadata$keywords`
@@
 Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022].
-In particular, forest landscape models that can account for harvest, fire, and other disturbances are increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
+In particular, forest landscape models that can account for harvest, fire, and other disturbances are increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2023; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021].
 Despite the importance of roads for wildlife, very few of these projections have included road networks [@labadie_global_2023; @leston_quantifying_2020].
@@
 ```
+
 ```{=html}
@@
 ```
+
 # Methods
@@
 ```
+
 ```{r setup2}
@@
 
+```{r study-area, fig.alt="Two raster maps of Revelstoke timber supply area in grey scale, with an inset map showing the location of the study area within Canada.", fig.cap="Road construction cost surface and an elevation surface used to calculate grade penalties.
+Costs are calculated using a simple model informed by slope and major barriers (Table 1, Online Resource 1-1) in the Revelstoke timber supply area in British Colombia, Canada.
+In both layers, pixels containing existing roads are assigned a value of 0"}
+knitr::include_graphics(here("analysis/figures/Figure1_StudyArea_both.tiff"))
+```
+
 Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
@@
 We compare two methods for integrating these costs into the cost of connections between adjacent raster cells.
-In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Supplement 1-1).
+In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Online Resource 1-1).
 The cost of the connection between two adjacent cells is simply the mean of the cost raster (Figure \@ref(fig:study-area)a) value at each of those cells, adjusted for euclidean distance to penalize longer diagonal road segments.
-In the "grade penalty" method (Supplement 1-1) we use the same construction costs (Table 1), but calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
+In the "grade penalty" method (Online Resource 1-1) we use the same construction costs (Table 1), but calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
 The most important difference between these two approaches is that grade is a property of the edge between two nodes in the grade penalty method, while slope is a property of each node in the simple cost method.
@@
 We examined sensitivity to pixel resolution by comparing results from 1 ha (1.7 million pixels) and 100 ha (17 thousand pixels) weight rasters.
-We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions \>= 1 ha.
+We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at reso
 
@@
 The ILCP algorithm we used [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction.
-Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement 1-1), which is then weighted by the distance to neighbouring nodes.
+Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Online Resource 1-1), which is then weighted by the distance to neighbouring nodes.
 Both the grade penalty and simple cost methods involve setting edge weights on a graph from adjacent locations on a single input grid.
@@
 ```
+
 Note that the evolution of a road network over time can be modeled by iteratively applying any of the methods, but in this study we focus on comparing a single application of each of the methods to a network of forest roads developed since 1990.
@@
 To help make these methods accessible, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023] which depends on several other R packages: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022].
-We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement 1-7) by default.
+We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Online Resource 1-7) by default.
 The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].
@@
 In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).
-We recorded compute time and peak RAM usage for each run, and ran some projections at 25 ha to show how speed and memory requirements vary with number of landings and number of nodes in the graph (Supplement 1-5), but did not assess other aspects of performance at that resolution.
+We recorded compute time and peak RAM usage for each run, and ran some projections at 25 ha to show how speed and memory requirements vary with number of landings and number of nodes in the graph (Online Resource 1-5), but did not assess other aspects of performance at that resolution.
 
@@
 
+```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.alt="A faceted bar chart with one bar for each projection method variant and a panel for each combination of performance metrics and within cutblock, outside cutblock and overall.", fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
+Values greater than zero indicate overprojection and negative values indicate underprojection.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks.
+Projection method variants are defined in Table 2.
+Note that extreme values for the cutblocks only method within cutblocks are cut off to ensure patterns in the other methods to remain visible.
+Bars that are cutoff are labeled with the total value"}
+# organize data for ggplot
+# need to make plots more comparable
+# Other option is setting limits but want to avoid excess white space
+ann_plus <- aspat_perf$prop_dif %>% 
+  filter(abs(prop_dif) > 0.5) %>% 
+  mutate(lab = round(prop_dif, 2), 
+         prop_dif = 0.37*sign(prop_dif))
+
+aspat_perf$plt+coord_flip(ylim = c(-0.35, 0.35))+
+  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
+            size = 8, size.unit = "pt")
+```
+
 Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from observed roads.
@@
 We omitted cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Figures S3 and S4).
+
+```{r proj-meth-fig, out.width=paste0(fig_widths["two"], "in"), out.height="8in", fig.alt="A series of 8 maps showing roads projected with different method variants.", fig.cap="Comparison of road projection method variants (Table 2).
+The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
+Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks.
+In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row).
+Landing density is high for all except the centroid and Hardy variants"}
+knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"),
+                        dpi = knitr::opts_chunk$get("dpi"))
+```
+
+```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.alt="A raster map of the Revelstoke TSA where the colour indicates the difference between obeserve and projected distance to road.
+Three inset maps zoom in of specific areas with Google satelite imagery in the background.", fig.cap="Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network.
+The main map shows observed distance to the nearest road minus the projected distance.
+Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks"}
+
+knitr::include_graphics(here("analysis/figures/Figure4_map_update.tif"))
+
+```
 
@@
 
+```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4.5in", fig.alt="A faceted bar chart with one bar for each projection method variant and a panel for each combination of binary performance metrics and within cutblock, outside cutblock and overall.", fig.cap="Variation in spatial performance among projection method variants (1 ha weight raster).
+F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
+Values closer to one indicate better performance.
+Cutblocks not accessible via the observed road network were excluded from analysis.
+Projection method variants are defined in Table 2"}
+
+doSpatPlot(perf_tbl)
+
+```
+
 Although the MST method performed better than the alternatives we tested according to several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it is also more costly to compute (Figure S7).
@@
 In our example landscape we were able to project at 100 m resolution in a reasonable amount of time (\< 2.4 hours) using modest amounts of memory (\<3 GB RAM).
-However, our example landscape is fairly small, and memory requirements increase rapidly with landscape size (Supplement 1-5).
+However, our example landscape is fairly small, and memory requirements increase rapidly with landscape size (Online Resource 1-5).
 Although it is possible to use machines with more RAM, the computational cost of adding road projections to spatial stochastic simulations of forest change relevant for wide-ranging wildlife is not trivial.
-For example, in a study of the effects of forest management and climate change scenarios on wildlife in a single boreal caribou range (100 stochastic realizations of 6 scenarios), we projected roads at 50 m resolution using graph with ~21 million nodes.
+For example, in a study of the effects of forest management and climate change scenarios on wildlife in a single boreal caribou range (100 stochastic realizations of 6 scenarios), we projected roads at 50 m resolution using graph with \~21 million nodes.
 Each ILCP road projection required 2.7 hours and 11 GB of RAM, for a total of 1620 compute hours.
@@
 ```
+
 # Acknowledgements
@@
 
-# Competing interests
+# Statements & Declarations
+
+## Funding
+
+Funding for this work was provided by Environment and Climate Change Canada.
+
+## Competing interests
 
@@
 
-# Author Contributions
+## Author Contributions
 
 Conceptualization: JH, SE, DL, GP, KL Data curation: SE, DL Formal analysis: JH, SE, DL Funding acquisition: JH Investigation: JH, SE Methodology: JH, SE, DL, KL Project administration: JH Resources: JH Software: JH, SE, KL Supervision: JH Visualization: SE Writing – original draft: JH, SE, DL Writing – review & editing: JH, SE, DL, GP, KL
-
-# Funding
-
-Funding for this work was provided by Environment and Climate Change Canada.
 
@@
 
-*Supplement 1.* Detailed methods and additional results and figures, including: detailed methods for the grade penalty function; results using simple cost, all cutblocks, and a coarse resolution raster; projection method benchmarking and profiling; and a map of roads in Fort Nelson timber supply area.
-
-\newpage
-
-# Figures
-
-Figure 1: Road construction cost surface and an elevation surface used to calculate grade penalties.
-Costs are calculated using a simple model informed by slope and major barriers (Table 1, Supplement 1-1) in the Revelstoke timber supply area in British Colombia, Canada.
-In both layers, pixels containing existing roads are assigned a value of 0.
-
-Alt text: Two raster maps of Revelstoke timber supply area in grey scale, with an inset map showing the location of the study area within Canada.
-
-```{r study-area, fig.cap="."}
-
-knitr::include_graphics(here("analysis/figures/Figure1_StudyArea_both.tiff"))
-```
-
-\newpage
-
-Figure 2: Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster).
-Values greater than zero indicate overprojection and negative values indicate underprojection.
-Cutblocks not accessible via the observed road network were excluded from analysis.
-Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks.
-Projection method variants are defined in Table 2.
-Note that extreme values for the cutblocks only method within cutblocks are cut off to ensure patterns in the other methods to remain visible.
-Bars that are cutoff are labeled with the total value.
-Alt text: A faceted bar chart with one bar for each projection method variant and a panel for each combination of performance metrics and within cutblock, outside cutblock and overall.
-
-```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.cap="."}
-# organize data for ggplot
-# need to make plots more comparable
-# Other option is setting limits but want to avoid excess white space
-ann_plus <- aspat_perf$prop_dif %>% 
-  filter(abs(prop_dif) > 0.5) %>% 
-  mutate(lab = round(prop_dif, 2), 
-         prop_dif = 0.37*sign(prop_dif))
-
-aspat_perf$plt+coord_flip(ylim = c(-0.35, 0.35))+
-  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
-            size = 8, size.unit = "pt")
-```
-
-\newpage
-
-Figure 3: Comparison of road projection method variants (Table 2).
-The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row).
-Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks.
-In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row).
-Landing density is high for all except the centroid and Hardy variants.
-
-Alt text: A series of 8 maps showing roads projected with different method variants.
-
-```{r proj-meth-fig, out.width=paste0(fig_widths["two"], "in"), out.height="8in", fig.cap="."}
-knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"),
-                        dpi = knitr::opts_chunk$get("dpi"))
-```
-
-\newpage
-
-Figure 4: Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network.
-The main map shows observed distance to the nearest road minus the projected distance.
-Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks.
-
-Alt text: A raster map of the Revelstoke TSA where the colour indicates the difference between obeserve and projected distance to road.
-Three inset maps zoom in of specific areas with Google satelite imagery in the background.
-```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="."}
-
-knitr::include_graphics(here("analysis/figures/Figure4_map_update.tif"))
-
-```
-
-\newpage
-
-Figure 5: Variation in spatial performance among projection method variants (1 ha weight raster).
-F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted.
-Values closer to one indicate better performance.
-Cutblocks not accessible via the observed road network were excluded from analysis.
-Projection method variants are defined in Table 2.
-
-Alt text: A faceted bar chart with one bar for each projection method variant and a panel for each combination of binary performance metrics and within cutblock, outside cutblock and overall.
-```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4.5in", fig.cap="."}
-
-doSpatPlot(perf_tbl)
-
-```
+*Online Resource 1.* Detailed methods and additional results and figures, including: detailed methods for the grade penalty function; results using simple cost, all cutblocks, and a coarse resolution raster; projection method benchmarking and profiling; and a map of roads in Fort Nelson timber supply area.
 

### 2025-04-07 analysis/paper/supplement/supplement.Rmd
reformat for landscape ecology

@@
 bibliography: ../Roads.bib
-csl: "../../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
+csl: "../../templates/landscape-ecology.csl" # Insert path for the bib-style
 
@@
 ```
+Journal: Landscape Ecology
+
+Authors: Josie Hughes, Sarah Endicott, David Lapins, Kyle Lochhead, and Gregory Paradis
+
+Corresponding author: Josie Hughes; josie.hughes@ec.gc.ca; Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, Ottawa, ON, Canada
 

### 2025-04-08 analysis/paper/paper.Rmd
update abstract and citations

@@
 
-Despite rapidly growing evidence of complex and varied impacts of road networks on wildlife and other forest values, and the increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included automated road network projections, in part because tools suited for this purpose have not been available.
+## Context
+
+Road networks have complex and varied impacts on wildlife and other forest values, yet spatial stochastic models forecasting impacts of forest disturbance rarely include automated road network projections.
 Hardy et al.
-[-@hardy_landis-ii_2023] partially addressed this need with a LANDIS-II extension.
-As a complementary and comparable alternative, we provide implementations of simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere.
-Grade penalties improve performance in a mountainous region of British Columbia, as does our minimum spanning tree method.
+[-@hardy_landis-ii_2023] partially addressed this need with a LANDIS-II extension, but there remains a need for tools that can be integrated into other modelling frameworks.
+
+## Objectives
+
+Our goal is an open source forest road network projection tool that can be easily applied and integrated into modelling frameworks used for projecting the implications of forest change for wildlife.
+We assess the performance of our road network projection methods using ecologically relevant metrics.
+
+## Methods
+
+We implemented simple iterative least cost path and minimum spanning tree methods with grade penalties in an open source R package.
+We assessed performance by comparing projections to observed resource road development since 1990 in a mountainous region of British Columbia.
+
+## Results
+
+Grade penalties improve performance, as does our minimum spanning tree method.
 However, the minimum spanning tree method requires more computing time and memory, so users must weigh the benefits of improved performance against computational costs.
+
+## Conclusions
+
 Our road network simulation methods can improve projections of anticipated resource development impacts on wildlife across large areas, and we hope that more accessible open source implementations will encourage integration of resource roads into forest change projections.
-
-# Study Implications
-
-This study offers open source tools for automated forest road network projection that are suited for integration into modelling frameworks used to project the implications of forest change for wildlife.
-We assess road network projection methods using ecologically relevant metrics, and show that grade penalties can improve performance in a topographically complex landscape.
 These tools can help improve projections of the cumulative effects of natural and anthropogenic disturbances on wildlife in an era of rapid change.
@@
 
-# Literature Cited
+# Reference list
 

### 2025-04-09 analysis/paper/paper.Rmd
Submitted! last edits here

@@
   - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, Ottawa, ON, Canada  
-  - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada
+  - FLNR: Ministry of Forests, Victoria, BC, Canada
   - UBC:  Faculty of Forestry, Department of Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada
@@
 
-[To do list](https://docs.google.com/document/d/1msHZuJGv0C67Avbl67YXm_P6fBSAvGw-m9qIzNDic64/edit?usp=sharing)
-
 <!-- The following code chunk defines some general settings how code chunks should behave. -->
@@
 
-Road networks have complex and varied impacts on wildlife and other forest values, yet spatial stochastic models forecasting impacts of forest disturbance rarely include automated road network projections.
+Resource road networks have complex and varied impacts on wildlife and other forest values, yet spatial stochastic models forecasting impacts of forest disturbance rarely include automated road network projections.
 Hardy et al.
-[-@hardy_landis-ii_2023] partially addressed this need with a LANDIS-II extension, but there remains a need for tools that can be integrated into other modelling frameworks.
+[-@hardy_landis-ii_2023] partially addressed this need with a LANDIS-II extension, but there remains a need for tools that can be integrated into other modelling frameworks while identifying a pragmatic balance between achieving ecological relevancy and computational cost.
 
@@
 
-Our goal is an open source forest road network projection tool that can be easily applied and integrated into modelling frameworks used for projecting the implications of forest change for wildlife.
-We assess the performance of our road network projection methods using ecologically relevant metrics.
+Our goal is an open source resource road network projection tool that can be easily incorporated into modelling frameworks that assess the implications of forest change for wildlife.
+We compared the performance of several resource road network projection methods using ecologically relevant metrics.
 
@@
 
-Grade penalties improve performance, as does our minimum spanning tree method.
-However, the minimum spanning tree method requires more computing time and memory, so users must weigh the benefits of improved performance against computational costs.
+All resource road projection methods that we tested performed relatively well.
+Grade penalties improved performance, as did our minimum spanning tree method.
+However, the minimum spanning tree method required more computing time and memory, so users must weigh the benefits of improved performance against computational costs.
 
@@
 
-Our road network simulation methods can improve projections of anticipated resource development impacts on wildlife across large areas, and we hope that more accessible open source implementations will encourage integration of resource roads into forest change projections.
-These tools can help improve projections of the cumulative effects of natural and anthropogenic disturbances on wildlife in an era of rapid change.
+Our resource road network simulation methods can improve projections of anticipated resource development impacts on wildlife across large areas.
+Our open source implementation will be useful for improving projections of the cumulative effects of natural and anthropogenic disturbances on wildlife in an era of rapid change.
 
@@
 
-We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is challenging.
+We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Fig.
+\@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is challenging.
 Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain.
@@
 
-Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
+Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021; a static dataset in our study area] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development_forest_2021] data sets for British Columbia.
+We treated the digital road atlas as a static dataset in our study area, while the forest tenure roads are a dynamic dataset, with new road segments added yearly as road permits are issued to licensees.
 We removed all roads from the digital road atlas that were within 50 m of forest tenure roads to avoid duplication.
@@
 In the "simple cost" method, we first calculate a slope raster from a digital elevation model, then calculate an input cost raster that depends on slope (Table 1, Online Resource 1-1).
-The cost of the connection between two adjacent cells is simply the mean of the cost raster (Figure \@ref(fig:study-area)a) value at each of those cells, adjusted for euclidean distance to penalize longer diagonal road segments.
-In the "grade penalty" method (Online Resource 1-1) we use the same construction costs (Table 1), but calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Figure \@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
+The cost of the connection between two adjacent cells is simply the mean of the cost raster (Fig.
+\@ref(fig:study-area)a) value at each of those cells, adjusted for euclidean distance to penalize longer diagonal road segments.
+In the "grade penalty" method (Online Resource 1-1) we use the same construction costs (Table 1), but calculate the cost of a connection between neighbouring cells from the elevation difference between the cells, using an elevation raster as input (Fig.
+\@ref(fig:study-area)b), and penalizing edge grades as suggested by Anderson and Nelson (2004).
 The most important difference between these two approaches is that grade is a property of the edge between two nodes in the grade penalty method, while slope is a property of each node in the simple cost method.
@@
 We examined sensitivity to pixel resolution by comparing results from 1 ha (1.7 million pixels) and 100 ha (17 thousand pixels) weight rasters.
-We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at reso
+We did not consider finer resolutions because stochastic forest landscape models typically simulate dynamics at resolutions \>= 1 ha.
+
+```{r cost-tbl}
+tbl <- "Cost Type|Value ($ per km)|Source
+Slope (%)|16178 + 504*slope|Interior appraisal manual
+Major lakes|65000|Major water bodies of BC
+Existing roads|0|Digital road atlas and forest tenure roads for BC
+"
+df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df) <- unname(as.list(df[1,]))
+df <- df[-1,] # remove first row
+row.names(df)<-NULL
+knitr::kable(df, caption = "Road construction costs.
+In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
+These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.")
+```
 
@@
 
-Our road projection methods link multiple target locations (also known as landings within harvest cutblocks (i.e., log decks)) to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
-Specifically, we compared iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).
+Our road projection methods link multiple target locations (also known as landings within harvest cutblocks (i.e., log decks)) to an existing road network by finding minimum-cost locations for new roadways, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP).
+Specifically, we compared iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Fig.
+\@ref(fig:proj-meth-fig)).
 We used the QGIS version of Hardy’s method (2019), which was easier for us to apply and does not differ substantially from the LANDIS extension (Hardy, personal communication).
@@
 Both the grade penalty and simple cost methods involve setting edge weights on a graph from adjacent locations on a single input grid.
-We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest (by Euclidean distance) point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
+We used a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest (by Euclidean distance) point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.
 
@@
 The ILCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block.
-We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)).
+We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Fig.
+\@ref(fig:proj-meth-fig)).
 We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200 m which gives \~ 9 pts/km^2^.
@@
 We recorded compute time and peak RAM usage for each run, and ran some projections at 25 ha to show how speed and memory requirements vary with number of landings and number of nodes in the graph (Online Resource 1-5), but did not assess other aspects of performance at that resolution.
+
+```{r method-tbl}
+tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
+MST grade penalty & simple cost|Regular|1|1 & 100 ha
+MST grade penalty & simple cost|Regular|10|1 & 100 ha
+MST grade penalty & simple cost|Random|1|1 & 100 ha
+MST grade penalty & simple cost|Random|10|1 & 100 ha
+MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
+ILCP grade penalty & simple cost|Regular|10|1 ha 
+Hardy QGIS|N/A|~9|1 ha
+Cutblocks only|N/A|N/A|1 ha
+
+"
+df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
+names(df2) <- unname(as.list(df2[1,]))
+df2 <- df2[-1,] # remove first row
+row.names(df2)<-NULL
+knitr::kable(df2, caption = "Road projection method variants.
+We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
+```
 
@@
 
-According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure S1).
-We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)).
+According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Fig.
+\@ref(fig:aspat-perf-fig) and Fig.
+S1).
+We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Fig.
+\@ref(fig:aspat-perf-fig)).
 The MST regular high density method yields the best projections of road density overall (projected density: `r mst_reg_dens$response_proj %>% round(2)` m roads/ha, observed density: `r mst_reg_dens$response_obs %>% round(2)` m roads/ha), while the MST random high density method had the best projections within cutblocks (projected density: `r mst_ran_dens$response_proj %>% round(2)` m roads/ha, observed density: `r mst_ran_dens$response_obs %>% round(2)` m roads/ha), and all MST variants yield comparably good projections outside cutblocks (projected density: `r mst_out_dens$response_proj %>% range() %>% round(2) %>% paste0(collapse = " - ")` m roads/ha, observed density: `r mst_out_dens$response_obs %>% unique() %>% round(2)` m roads/ha).
@@
 
-Maps (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differ from one another, and from observed roads.
+Maps (Figs.
+\@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)) help clarify exactly how the projection methods differed from one another, and from observed roads.
 While the Hardy et al.
-(2023) method and the grade penalty MST and ILCP methods project similar mean road densities within cutblocks, they create distinctly different road patterns within cutblocks.
-In particular, the grade penalty method incentivized roads to follow slope contours, while the Hardy method created more road curvatures (Figure \@ref(fig:proj-meth-fig)).
-Although the grade penalty method improved realism by following elevation contours, all these methods created dead ends, intersections, and abrupt turns that would be avoided in the engineering of operational roads in steep terrain (Figures \@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
-All of the projected road methods also crossed valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by improved costs of crossing streams, rivers and other wet areas in valley bottoms.
+(2023) method and the grade penalty MST and ILCP methods projected similar mean road densities within cutblocks, they created distinctly different road patterns within cutblocks.
+In particular, the grade penalty method incentivized roads to follow slope contours, while the Hardy method created more road curvatures (Fig.
+\@ref(fig:proj-meth-fig)).
+Although the grade penalty method improved realism by following elevation contours, all these methods created dead ends, intersections, and abrupt turns that would be avoided in the engineering of operational roads in steep terrain (Figs.
+\@ref(fig:proj-meth-fig) and \@ref(fig:proj-map-fig)c).
+All of the projected road methods also crossed valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Fig.
+\@ref(fig:proj-map-fig)a), but this discrepancy could likely be remedied by improved costs of crossing streams, rivers and other wet areas in valley bottoms.
 Finally, it is important to note that some discrepancies between observed and projected road networks were caused by errors in the observed data, such as cutblocks that were incorrectly classified (e.g.
+Fig.
 \@ref(fig:proj-map-fig)b).
-We omitted cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Figures S3 and S4).
+We omitted cutblocks with no associated access roads from our analysis, and note that failure to do this would distort results (Figs.
+S3 and S4).
 
@@
 
-Projections of the forestry disturbance footprint, the road disturbance footprint, and, to a lesser extent, distance to road were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)).
-However, all of the road projection methods underestimate road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)), projecting fewer roads than we observe.
-One reason for underestimation of road density is that all these methods generate straight road segments within a pixel.
-As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, making coarse resolution projections less accurate (Figure S6).
+Projections of the forestry disturbance footprint, the road disturbance footprint, and, to a lesser extent, distance to road were more accurate than projections of road density and road presence (Fig.
+\@ref(fig:aspat-perf-fig)).
+However, all of the road projection methods underestimate road presence and forestry disturbance footprint (Fig.
+\@ref(fig:aspat-perf-fig)), projecting fewer roads than we observe.
+One reason for underestimation of road density was that all these methods generate straight road segments within a pixel.
+As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increased, making coarse resolution projections less accurate (Fig.
+S6).
 
@@
 
-Although these methods underestimate road presence, they represent a substantial improvement over the "cutblocks only" scenario, which assumes no new roads are built (Figure \@ref(fig:aspat-perf-fig)).
-The methods were not designed to project the exact spatial location of new roads (within 100 m), so it is not surprising that they did not yield accurate spatial projections of road presence (F-Measure: `r perf_tbl_rng["Road presence"]`).
-It is somewhat more surprising that none of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
-This is likely because many of the projected roads are within 500 m of either the existing road network or a cutblock so the false negatives created by failing to project roads are balanced by the lack of false positives.
-All methods other than "cutblocks only" provided comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yielded fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)).
+Although these methods underestimate road presence, they represented a substantial improvement over the "cutblocks only" scenario, which assumed no new roads were built (Fig.
+\@ref(fig:aspat-perf-fig)).
+The methods were not designed to project the exact spatial location of new roads (within 100 m), so it was not surprising that they did not yield accurate spatial projections of road presence (F-Measure: `r perf_tbl_rng["Road presence"]`).
+It was somewhat more surprising that none of the methods provided better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`).
+This was likely due to many of the projected roads being within 500 m of either the existing road network or a cutblock so the false negatives created by failing to project roads are balanced by the lack of false positives.
+All methods other than "cutblocks only" provided comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yielded fewer false negatives and more false positives (Fig.
+\@ref(fig:spat-perf-fig)).
 
@@
 
-Although the MST method performed better than the alternatives we tested according to several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it is also more costly to compute (Figure S7).
-All of these methods require construction of a mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e.
-not NA), and memory requirements increase with the number of nodes in the graph (Figure S7).
-Compute time also increases with the number of landing targets (Figure S7).
-Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` *geoCorrection* function.
-For our example landscape the Hardy QGIS plugin was 65 times faster than the MST and ILCP methods implemented in R (Figure S7).
+Although the MST method performed better than the alternatives we tested according to several important wildlife habitat metrics (Fig.
+\@ref(fig:aspat-perf-fig)), it was also more costly to compute (Fig.
+S7).
+All of these methods required construction of a mathematical graph with a node for each pixel in the weight raster where road construction was possible (i.e.
+not NA), and memory requirements increase with the number of nodes in the graph (Fig.
+S7).
+Compute time also increased with the number of landing targets (Fig.
+S7).
+Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction was more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Fig.
+S8), in part because it avoided the additional cost of constructing a transition matrix, and in part because it ignored variation in the size of raster cells that is accounted for in the `gdistance` *geoCorrection* function.
+For our example landscape the Hardy QGIS plugin was 65 times faster than the MST and ILCP methods implemented in R (Fig.
+S7).
 
@@
 
-Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)).
-Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure \@ref(fig:aspat-perf-fig)).
-However, the MST algorithm was also more computationally demanding than the other methods we tested (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
+Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Fig.
+\@ref(fig:aspat-perf-fig)).
+Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Fig.
+\@ref(fig:aspat-perf-fig)).
+However, the MST algorithm was also more computationally demanding than the other methods we tested (Fig.
+S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.
 
@@
 This examination highlighted some errors that could be fixed by better accounting for the cost of water crossings in our weight raster, and other more fundamental limitations and differences among methods.
-Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure \@ref(fig:proj-meth-fig)).
+Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Fig.
+\@ref(fig:proj-meth-fig)).
 We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.
@@
 The first three options are available to users of the `roads` R package.
-However, those contemplating reducing computational cost by decreasing raster resolution should be cautious of the results given that all of these methods become less accurate as raster cell size increases (Figures S5 and S6).
+However, those contemplating reducing computational cost by decreasing raster resolution should be cautious of the results given that all of these methods become less accurate as raster cell size increases (Figs.
+S5 and S6).
 The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000].
 Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length and road density will be underestimated.
-More involved possibilities for reducing computational costs include adopting faster routing algorithms [@larmet_cppRouting_2022], implementing slow components of the R package (Figure S8) in a faster compiled language, parallel processing, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
+More involved possibilities for reducing computational costs include adopting faster routing algorithms [@larmet_cppRouting_2022], implementing slow components of the R package (Fig.
+S8) in a faster compiled language, parallel processing, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.
 Several of these options have been implemented in the `cppRouting` R package [@larmet_cppRouting_2022], and future versions of the `roads` R package could borrow and build on `cppRouting` functionality.
@@
 Cost minimizing methods are not suitable for projecting development of seismic lines (e.g.
-Figure S9), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
+Fig.
+S9), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable.
 Lochhead et al.
@@
 
-# Tables
-
-Table 1: Road construction costs.
-In the “grade penalty” method, we assign grades \>20% a more prohibitive cost of 65000, which is also assigned to major water bodies.
-These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.
-
-```{r cost-tbl}
-tbl <- "Cost Type|Value ($ per km)|Source
-Slope (%)|16178 + 504*slope|Interior appraisal manual
-Major lakes|65000|Major water bodies of BC
-Existing roads|0|Digital road atlas and forest tenure roads for BC
-"
-df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
-names(df) <- unname(as.list(df[1,]))
-df <- df[-1,] # remove first row
-row.names(df)<-NULL
-knitr::kable(df, caption = ".")
-```
-
-\newpage
-
-Table 2: Road projection method variants.
-We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.
-
-```{r method-tbl}
-tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
-MST grade penalty & simple cost|Regular|1|1 & 100 ha
-MST grade penalty & simple cost|Regular|10|1 & 100 ha
-MST grade penalty & simple cost|Random|1|1 & 100 ha
-MST grade penalty & simple cost|Random|10|1 & 100 ha
-MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
-ILCP grade penalty & simple cost|Regular|10|1 ha 
-Hardy QGIS|N/A|~9|1 ha
-Cutblocks only|N/A|N/A|1 ha
-
-"
-df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
-names(df2) <- unname(as.list(df2[1,]))
-df2 <- df2[-1,] # remove first row
-row.names(df2)<-NULL
-knitr::kable(df2, caption = ".")
-```
-
-\newpage
-
 ### Colophon
